### 2020-12-18 qs_for_jess/demography.Rmd
adding prelim fig/table code + q's for Jess

---
title: "Demographic estimates from age data"
author: "Malavika Rajeev"
date: "12/9/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# pkgs
library(dplyr)
library(readr)
library(ggplot2)
library(tidyr)
library(patchwork)
library(cowplot)

```

## Age data
I have this age data from campaigns from 2018 & 2019 for doggoes in Moramanga District:

```{r}
# Age pyramid
animals_age <- read_csv("animals_age.csv")

animals_age %>%
  mutate(Sex = toupper(Sex), age = floor(age_mos/12)) %>%
  group_by(Sex, age) %>%
  summarize(N  = n()) %>%
  mutate(N = ifelse(Sex == "M", -N, N)) -> age_pyramid

ggplot(age_pyramid, aes(x = age, y = N, fill = Sex)) +
  geom_col() +
  scale_y_continuous(breaks = c(seq(-600, -200, 200), seq(200, 600, 200)), 
                     labels = c(seq(600, 200, -200), seq(200, 600, 200)), 
                     limits = c(-600, 600)) +
  coord_flip() +
  labs(x = "Age in years", y = "Number of individuals")

```

Looks like fast growing & slightly sex biased population (standard free-roaming dog pop characteristics)

## Annie's thesis

I pulled from Annie's thesis to estimate growth/survival. Jess's original code:

```{r}
age_pyramid %>%
  filter(!is.na(age) & !is.na(N)) %>%
  mutate(age_class = case_when(age == 0 ~ 0, # < 1 yr olds
                               age == 1 ~ 1, # 1 - 2 yr olds
                               age > 1 & age < 5 ~ 2, # 2 - 5 yr olds
                               age >= 5 ~ 3)) %>% # 5+ yr olds
  group_by(age_class) %>%
  summarize(N = sum(abs(N))) %>%
  mutate(prop = N/sum(N)) -> n_by_age

# Original model from Annie's thesis
SS <- function(par, number.individuals) {
  ## put chosen parameters into the matrix; and use exponential to
  # ensure doesnt go negative
  psurv <- exp(par[1:2])/(1+exp(par[1:2]))
  ## make the matrix with this value of psurv
  mat1 <- matrix(0,4,4)
  mat1[1,1] <- psurv[1]*0 ## after one year, no one stays a baby
  mat1[2,1] <- psurv[1]*1 ## every baby becomes juvenile
  mat1[2,2] <- psurv[2]*0 ## no one stays a juvenile
  mat1[3,2] <- psurv[2]*1 ## every juvenile becomes adult
  mat1[3,3] <- psurv[2]*(1-1/3) ## 1-1/3 of them stay in that age
  class
  mat1[4,3] <- psurv[2]*1/3 ## 1/3 of them age up
  mat1[4,4] <- psurv[2] ## surviving
  ## add in the fertility
  ## xxj note I've added another value - so I've said that in column 3,
  # we have exp(par[3])*1,
  ### and in column 4 we have exp(par[3])*a probabilty defined by
  # par[4]. This is so it has to be smaller!
    mat1[1,3:4] <- 1+exp(par[c(3)])*c(1,exp(par[4])/(1+exp(par[4])))
  ## calculate what this does to the stable structure
  prop.stable.struct.after.seed1 <-
    Re(eigen(mat1)$vector[1:4,1])/sum(Re(eigen(mat1)$vector[1:4,1]))
  ## get the sum of squares (measure of distance between observed and
  # predicted, which we want to be as small as possible)
  sum.of.squares <- sum((number.individuals[1:4]-
                         prop.stable.struct.after.seed1*sum(number.individuals,na.rm=TRUE))^2)
  ## put it back
  return(sum.of.squares)
}

tmp <- optim(par=c(-0.5, -0.5, log(2.5), log(2.5)), fn = SS, 
             number.individuals = n_by_age$N)

## build matrix and predictions with these optimal results
psurv <- exp(tmp$par[1:2])/(1+exp(tmp$par[1:2]))

## make the matrix with this value of psurv
mat1 <- matrix(0,4,4)
mat1[2,1] <- psurv[1]*1 ## every baby becomes juvenile
mat1[3,2] <- psurv[2]*1 ## every juvenile becomes adult
mat1[3,3] <- psurv[2]*(1-1/3) ## 1-1/3 of them stay in that age class
mat1[4,3] <- psurv[2]*1/3 ## 1/3 of them age up
mat1[4,4] <- psurv[2] ## surviving
p.reduce.fert.old <- exp(tmp$par[4])/(1+exp(tmp$par[4])) 
mat1[1,3:4] <- 1+exp(tmp$par[3])*c(1,p.reduce.fert.old)

## predict the size distribution
optim.prop.stable.struct.after.seed <-
  Re(eigen(mat1)$vector[1:4,1])/Re(sum(eigen(mat1)$vector[1:4,1]))

## and add this one to the plot from before 
plot(1:4, optim.prop.stable.struct.after.seed*sum(n_by_age$N, na.rm=TRUE),
     type = "b", pch=19,
     xlab="Age Stage", ylab="Number of Dogs", main="Canine Population",
     ylim=range(c(0, n_by_age$N), na.rm=TRUE)) # predicted
points(1:4, n_by_age$N[1:4], type="b", col="red") # observed

## and this means, now you can figure out what the population growth
Re(eigen(mat1)$value[1]) # growth rate = 1.35

```

## Adapted

I adapted it to try and estimate pup survival and adult survival, and changed it so
that all dogs > 12 mos. reproduce and fertility doesn't decline (which we think is
how it works (ish) for dogs).

In this case, should fertility be something like avg. litter size * avg. # of litters per yr * proportion pop female? Am I getting these units right?

```{r}
agemod1 <- function(par = c(psurv_pup = -1, 
                            psurv_adult = -1),
                    fert = 3, 
                    n_per_age, type = "fit") {

  list2env(as.list(par), envir = environment())

  # transform parameters to [0, 1]
  p1 <- exp(psurv_pup)/(1 + exp(psurv_pup))
  p2 <- exp(psurv_adult)/(1 + exp(psurv_adult))

  # make the Leslie matrix
  lmat <- matrix(0, 4, 4)
  lmat[2, 1] <- p2 # every surviving baby becomes juvenile
  lmat[3, 2] <- p2 # every surviving juvenile becomes adult
  lmat[3, 3] <- p2 * 2/3 # 2/3 of them stay in this age class (for years 1 & 2 in age bin)
  lmat[4 ,3] <- p2 * 1/3 # 1/3 of them age up (after year 3 in age bin)
  lmat[4, 4] <- p2  # surviving of final age class

  # add in the fertility (same for all adults, i.e. excluding puppies)
  # this is pup survival * birthrate which is not actually identifiable
  # so we look at an average litter size of fert & try and estimate pup survival
  lmat[1, 2:4] <- fert * p1

  # calculate what this does to the stable structure
  prop_stable <- Re(eigen(lmat)$vector[1:4, 1]) / sum(Re(eigen(lmat)$vector[1:4, 1]))

  if(type == "pred") {
    # predicted stable age
    stable_age <- prop_stable*sum(n_per_age)
    growth <- Re(eigen(lmat)$value[1])
    return(list(stable_age = stable_age, growth = growth))
  }

  if(type == "fit") {
    # get the sum of squares (distance between observed and predicted prop in each age class)
    sum_of_squares <- sum((n_per_age - prop_stable*sum(n_per_age))^2)
    return(sum_of_squares)
  }
}

exp_trans <- function(x) exp(x)/(1 + exp(x))

age_ests <- optim(par = list(psurv_pup = -1, psurv_adult = 1),
                  fert = 4,
                  fn = agemod1, n_per_age = n_by_age$N)
exp_trans(age_ests$par)
preds <- agemod1(n_per_age = n_by_age$N, par = age_ests$par, type = "pred")
plot(1:4, preds$stable_age,
     type = "b", pch = 19,
     xlab = "Age Stage", ylab = "Number of Dogs", main = "Canine Population",
     ylim = range(c(0, n_by_age$N, preds$stable_age))) # predicted
points(1:4, n_by_age$N[1:4], type="b", col="red") # observed

preds$growth

```

## Sensitivity to initial parameter values

Optim estimates are highly dependent on starting values:

```{r}
age_ests <- optim(par = list(psurv_pup = -1, psurv_adult = -1),
                  fert = 4,
                  fn = agemod1, n_per_age = n_by_age$N)
exp_trans(age_ests$par)
preds <- agemod1(n_per_age = n_by_age$N, par = age_ests$par, type = "pred")
preds$growth

age_ests <- optim(par = list(psurv_pup = -0.5, psurv_adult = -0.5),
                  fert = 4,
                  fn = agemod1, n_per_age = n_by_age$N)
preds <- agemod1(n_per_age = n_by_age$N, par = age_ests$par, type = "pred")
exp_trans(age_ests$par)
preds$growth

age_ests <- optim(par = list(psurv_pup = 1, psurv_adult = 1),
                  fert = 4,
                  fn = agemod1, n_per_age = n_by_age$N)
preds <- agemod1(n_per_age = n_by_age$N, par = age_ests$par, type = "pred")
exp_trans(age_ests$par)
preds$growth

```

## Brute force

But if you brute force it, and just get SS across par values, 
then there's a ridge along low pup survival vs. higher adult survival (which I guess makes sense), 
and intermediate growth ( 1 - 1.5 %):

```{r}
ll_grid <- expand.grid(psurv_pup = seq(-2, 2, by = 0.1), 
                       psurv_adult = seq(-2, 2, by = 0.1), 
                       fert = seq(2, 5, by = 1))
ll_grid %>%
  rowwise() %>%
  mutate(ss = agemod1(n_per_age = n_by_age$N, 
                      par = c(psurv_pup = psurv_pup, 
                              psurv_adult = psurv_adult),
                      fert = fert, type = "fit"),
         gr = agemod1(n_per_age = n_by_age$N, 
                      par = c(psurv_pup = psurv_pup, 
                              psurv_adult = psurv_adult),
                      fert = fert, type = "pred")$growth) -> ll

ggplot(data = ll) +
  geom_tile(aes(x = factor(round(exp_trans(psurv_pup), 2)), 
                y = factor(round(exp_trans(psurv_adult), 2)), 
                fill = sqrt(ss))) +
  facet_wrap(~ fert) +
  scale_fill_gradient2(midpoint = median(sqrt(ll$ss)), name = "Sqrt SS") +
  labs(x = "Annual survival probability - pups", 
       y = "Annual survival probability - adults")

ggplot(data = ll) +
  geom_tile(aes(x = factor(round(exp_trans(psurv_pup), 2)), 
                y = factor(round(exp_trans(psurv_adult), 2)), 
                fill = gr)) +
  facet_wrap(~ fert) +
  scale_fill_gradient2(midpoint = 1, name = "Growth rate") + 
  labs(x = "Annual survival probability - pups", 
       y = "Annual survival probability - adults")

```

**How to present results? Use initial vals with lowest from brute forcing?**
**Am I just asking too much from this data? I guess just fitting to four vals that could be explained by lots of combos of pars?**

## Vax model

Discrete time approximation:

- Three years of vaccine immunity (discrete)
- Different pup vs. adult survival
- Different campaign strategies
    - Campaign style (annual, target dogs of all ages, what we did in 2018)
    - Puppy vaccination, continiously target newborn pups 
    - Combined strategies
```{r}

# Need to track down an accounting issue where some revax get lost
# but this is approximately right
simvacc <- function(tmax, pup_surv, adult_surv, revacc, pup_cov, 
                    campaign_cov, campaign_times, pup_month_vacc = 1,
                    birthrate, S_adult_0, V_adult_0, 
                    V_pup_0, S_pup_0) {

  # Set-up
  V_adult <- matrix(0, nrow = 36, ncol = tmax)
  S_adult <- rep(0, tmax)
  V_pup <- S_pup <- matrix(0, nrow = 12, ncol = tmax)

  # Initalize
  S_adult[1] <- S_adult_0
  V_adult[1, 1] <- V_adult_0
  V_pup[, 1] <- V_pup_0
  S_pup[, 1] <- S_pup_0

  for(t in 2:tmax) {

    # Births (everyone first has babies and then dies)
    births <- birthrate * (sum(V_adult[1:36, t - 1]) + S_adult[t - 1])

    # balance births (all to susceptibles)
    S_pup[1, t] <- births # get replaced

    # Pup mortality
    S_pup[2:12, t] <- S_pup[1:11, t - 1] * pup_surv
    V_pup[2:12, t] <- V_pup[1:11, t - 1] * pup_surv

    # Balance the adults (ageing in 12 mo old pups)
    S_adult[t] <- adult_surv * S_adult[t - 1] + pup_surv * S_pup[12, t - 1]

    # pups to vaccinate (at age x)
    pups_to_vacc <- ifelse(pup_month_vacc > 1, S_pup[pup_month_vacc, t], 
                           births)

    # do puppy vaccination 
    V_pup[pup_month_vacc, t] <- pups_to_vacc * pup_cov 
    S_pup[pup_month_vacc, t] <- S_pup[pup_month_vacc, t] - pups_to_vacc * pup_cov

    # Do the waning (adults only!)
    V_adult[2:36, t] <- adult_surv * V_adult[1:35, t - 1]
    S_adult[t] <- S_adult[t] + V_adult[36, t - 1]

    # Vaccinate as part of campaign accounting for revax
    cov_now <- ifelse(t %in% campaign_times, campaign_cov, 0)
    camp_vacc_adult <- (S_adult[t] + revacc * sum(V_adult[, t])) * cov_now - revacc * sum(V_adult[, t])
    camp_vacc_adult <- ifelse(camp_vacc_adult < 0, 0, camp_vacc_adult)

    # every vax is revaccinated with a certain revacc prob
    if(camp_vacc_adult > 0) {
      V_adult[1, t] <- camp_vacc_adult + revacc * sum(V_adult[, t])
      V_adult[2:36, t] <- (1 - revacc) * V_adult[2:36, t]
    } else {
      V_adult[1, t] <- camp_vacc_adult # newly vaccinated
    }

    # Balance addtl campaign vax adults
    S_adult[t] <- S_adult[t] - camp_vacc_adult

    # Balance addtl campaign vax pups (assuming that no puppies are revax)
    camp_vacc_pup <- S_pup[pup_month_vacc:12, t] * cov_now 
    V_pup[pup_month_vacc:12, t] <- V_pup[pup_month_vacc:12, t] + camp_vacc_pup
    S_pup[pup_month_vacc:12, t] <- S_pup[pup_month_vacc:12, t] - camp_vacc_pup

    # Add in surviving pups aging to adults so that they can wane appropriately
    t_to_sub <- pup_month_vacc - 1
    V_adult[13 - t_to_sub, t] <- V_adult[13, t] +  pup_surv * V_pup[12, t - 1]

  }

  V <- colSums(V_pup) + colSums(V_adult)
  S <- colSums(S_pup) + S_adult

  return(list(S = S, V = V, V_adult = V_adult, V_pup = V_pup, 
              S_adult = S_adult, S_pup = S_pup))

}

```

### Comparing campaigns vs. continuous pup vax
```{r}
age_ests <- optim(par = list(psurv_pup = -1, psurv_adult = 1),
                  fert = 3,
                  fn = agemod1, n_per_age = n_by_age$N)
preds <- agemod1(n_per_age = n_by_age$N, par = age_ests$par, type = "pred")
exp_trans(age_ests$par)
preds$growth

# annual survival probability to monthly
surv <- as.list(exp_trans(age_ests$par)^(1/12))

# proportion adults
prop <- preds$stable_age/(sum(preds$stable_age))
prop_adults <- 1 - prop[1]

# for population of 1000
# with campaigns only
camp <- simvacc(tmax = 12*5, pup_surv = surv$psurv_pup, adult_surv = surv$psurv_adult, 
        revacc = 1, 
        pup_cov = 0, 
        campaign_cov = 0.7, campaign_times = seq(12, 12*5, by = 12), 
        pup_month_vacc = 1, 
        birthrate = 3/12, # monthly birth rate
        S_adult_0 = prop_adults*1000, 
        S_pup_0 = rep(prop[1]/12, 12)*1000, V_adult_0 = 0, 
        V_pup_0 = rep(0, 12))

# with pup vacc only
pup <- simvacc(tmax = 12*5, pup_surv = surv$psurv_pup, adult_surv = surv$psurv_adult, 
        revacc = 1, 
        pup_cov = 0.7, 
        campaign_cov = 0, campaign_times = 0, 
        pup_month_vacc = 3, 
        birthrate = 3/12, # monthly birth rate
        S_adult_0 = prop_adults*1000, 
        S_pup_0 = rep(prop[1]/12, 12)*1000, V_adult_0 = 0, 
        V_pup_0 = rep(0, 12))

both <- simvacc(tmax = 12*5, pup_surv = surv$psurv_pup, adult_surv = surv$psurv_adult, 
        revacc = 1, 
        pup_cov = 0.7, 
        campaign_cov = 0.7, campaign_times = seq(12, 12*5, by = 12), 
        pup_month_vacc = 1, 
        birthrate = 3/12, # monthly birth rate
        S_adult_0 = prop_adults*1000, 
        S_pup_0 = rep(prop[1]/12, 12)*1000, V_adult_0 = 0, 
        V_pup_0 = rep(0, 12))

vacc_comp <- bind_rows(tibble(month = 1:60, 
                              cov = camp$V/(camp$S + camp$V), type = "Campaign"), 
                       tibble(month = 1:60, 
                              cov = pup$V/(pup$S + pup$V), type = "Puppy vax"), 
                       tibble(month = 1:60,
                              cov = both$V/(both$S + both$V), type = "Both"))

ggplot(vacc_comp) +
  geom_line(aes(x = month, y = cov, color = type)) +
  geom_hline(yintercept = 0.7, linetype = 2, color = "grey")

```

### 2020-12-18 R/Table1.R

@@ -1,0 +1,87 @@
+# Table 1
+
+# pkgs
+library(dplyr)
+library(readr)
+library(tidyr)
+library(scales)
+library(lubridate)
+library(gt)
+
+# data
+vacc_2018 <- read_csv("data/animal_2018.csv")
+vacc_2019 <- read_csv("data/campaign_2019.csv")
+owner_2018 <- read_csv("data/owner_2018.csv")
+
+# filter to dogs
+vacc_2018 %>%
+  filter(Species %in% "A") %>%
+  mutate(year = 2018, 
+         prev_vacc = ifelse(!is.na(PreviousVaccine), 1, 0), 
+         vacc_in_year = ifelse(PreviousVaccine %in% 2017, 1, 0)) %>%
+  select(Fokontany, loc = Commune, Date, year, Sex, prev_vacc, 
+         vacc_in_year) -> animals_2018
+
+vacc_2019 %>%
+  filter(`Dog or Cat` %in% "dog") %>%
+  select(Date, Commune, Fokontany = `Owner Location`, 
+         Sex, prev_vacc = `Previous Vaccination`,
+         vacc_in_year = `Last Vaccination Date`) %>%
+  mutate(year = 2019, loc = "All",
+         vacc_in_year = year(dmy(vacc_in_year)), 
+         prev_vacc = ifelse(prev_vacc %in% "YES", 1, 0), 
+         vacc_in_year = ifelse(vacc_in_year %in% 2018, 1, 0)) %>%
+  bind_rows(animals_2018) -> animals_all
+
+animals_all %>%
+  filter(loc != "Ambohibary") %>%
+  group_by(loc, year) %>%
+  summarize(total = scales::number(n(), big.mark = ","), 
+            prev_vacc = percent(sum(prev_vacc)/n()), 
+            vacc_in_year = percent(sum(vacc_in_year)/n()),
+            perc_m = percent(sum(Sex %in% "M")/n())) %>%
+  ungroup() %>%
+  pivot_longer(total:perc_m) %>%
+  pivot_wider(names_from = c(loc, year), 
+              values_from = value) -> dog_stats
+
+owner_2018 %>%
+  group_by(Commune) %>%
+  summarize(avg_dogs = number(mean(AdultDogs + Puppies, na.rm = TRUE), accuracy = 0.1), 
+            avg_cats = number(mean(Cats + Kittens, na.rm = TRUE),  accuracy = 0.1), 
+            roaming = percent(sum(grepl("mireny", 
+                                        Roaming, 
[…]

### 2020-12-18 R/dem_models.R

@@ -1,0 +1,146 @@
+
+agemod1 <- function(par = c(psurv_pup = -1, 
+                            psurv_adult = -1),
+                    fert = 6, 
+                    n_per_age, type = "fit") {
+  
+  list2env(as.list(par), envir = environment())
+  
+  # transform parameters to [0, 1]
+  p1 <- exp(psurv_pup)/(1 + exp(psurv_pup))
+  p2 <- exp(psurv_adult)/(1 + exp(psurv_adult))
+
+  # make the Leslie matrix
+  lmat <- matrix(0, 4, 4)
+  lmat[2, 1] <- p2 # every surviving baby becomes juvenile
+  lmat[3, 2] <- p2 # every surviving juvenile becomes adult
+  lmat[3, 3] <- p2 * 2/3 # 2/3 of them stay in this age class (for years 1 & 2 in age bin)
+  lmat[4 ,3] <- p2 * 1/3 # 1/3 of them age up (after year 3 in age bin)
+  lmat[4, 4] <- p2  # surviving of final age class
+  
+  # add in the fertility (same for all adults, i.e. excluding puppies)
+  # this is pup survival * birthrate which is not actually identifiable
+  # so we look at an average litter size of 6 / 2 = 3 
+  lmat[1, 2:4] <- fert * p1
+  
+  # calculate what this does to the stable structure
+  prop_stable <- Re(eigen(lmat)$vector[1:4, 1]) / sum(Re(eigen(lmat)$vector[1:4, 1]))
+ 
+  if(type == "pred") {
+    # predicted stable age
+    stable_age <- prop_stable*sum(n_per_age)
+    growth <- Re(eigen(lmat)$value[1])
+    return(list(stable_age = stable_age, growth = growth))
+  }
+  
+  if(type == "fit") {
+    # get the sum of squares (distance between observed and predicted prop in each age class)
+    sum_of_squares <- sum((n_per_age - prop_stable*sum(n_per_age))^2)
+    return(sum_of_squares)
+  }
+}
+
+agemod2 <- function(par = c(psurv_pup = -1, 
+                            psurv_adult = -1),
+                    fert = 6, 
+                    n_per_age, type = "fit") {
+  
+  list2env(as.list(par), envir = environment())
+  
+  # transform parameters to [0, 1]
[…]

### 2020-12-18 R/figure2.R

@@ -1,3 +1,5 @@
+# Figure 2. Comparing costs and willingness to pay
+
 # pkgs
 library(dplyr)
 library(readr)
@@ -81,6 +83,7 @@ willing_to_pay %>%
 ggplot(willingness) +
   geom_line(aes(x = charged/3100, y = 0.6 - reduction*0.6, linetype = Commune), 
             size = 1.25) +
+  xlim(c(0, 3)) +
   scale_linetype_manual(values = c(4, 2, 1), name = "") +
   labs(x = "Amount charged to owner (USD)", y = "Coverage", tag = "C") +
   theme_minimal_hgrid(font_size = 12) -> fig2C
@@ -115,11 +118,11 @@ ggplot() +
              linetype = 2) +
   labs(x = "Amount charged (USD)", y = "Estimated cost per animal", tag = "D") +
   ylim(c(0, 5)) +
-  xlim(c(0, 1.75)) +
+  xlim(c(0, 3)) +
   scale_color_brewer(palette = "Dark2", name = "Year") +
   theme_minimal_grid(font_size = 12) -> fig2D
 
 
 # arrange
-fig2 <- fig2A + fig2B + fig2C + fig2D 
+fig2 <- fig2A + fig2C + fig2B + fig2D 
 ggsave("figs/fig2.jpeg", height = 8, width = 10)

### 2020-12-18 R/figure3.R

@@ -1,0 +1,94 @@
+# Figure 3. Demography and vax
+
+# pkgs
+library(dplyr)
+library(readr)
+library(ggplot2)
+library(tidyr)
+library(patchwork)
+library(cowplot)
+library(bbmle)
+
+# animals
+animals_2018 <- read_csv("data/animal_2018.csv")
+animals_2019 <- read_csv("data/campaign_2019.csv")
+
+# filter to dogs
+animals_2018 %>%
+  filter(Species == "A") %>%
+  mutate(year = 2018, 
+         AgeMonths = replace_na(AgeMonths, 0), 
+         AgeYears = replace_na(AgeYears, 0), 
+         age_mos = AgeYears*12 + AgeMonths) %>%
+  mutate(age_mos = ifelse(age_mos == 0, NA, age_mos)) %>%
+  select(Fokontany, Commune, Date, age_mos, year, Sex) -> animals_age
+
+animals_2019 %>%
+  filter(`Dog or Cat` == "dog") %>%
+  select(Date, age_mos = `Age (months)`, Commune, Fokontany = `Owner Location`, 
+         Sex) %>%
+  mutate(year = 2019) %>%
+  bind_rows(animals_age) -> animals_age
+
+# histogram of ages
+ggplot(animals_age) +
+  geom_histogram(aes(x = age_mos)) +
+  facet_wrap(~year)
+
+animals_age %>%
+  mutate(Sex = toupper(Sex), age = floor(age_mos/12)) %>%
+  group_by(Sex, age) %>%
+  summarize(N  = n()) %>%
+  mutate(N = ifelse(Sex == "M", -N, N)) -> age_pyramid
+
+ggplot(age_pyramid, aes(x = age, y = N, fill = Sex)) +
+  geom_col() +
+  scale_y_continuous(breaks = c(seq(-600, -200, 200), seq(200, 600, 200)), 
+                     labels = c(seq(600, 200, -200), seq(200, 600, 200)), 
+                     limits = c(-600, 600)) +
+  coord_flip() +
+  labs(x = "Age in years", y = "Number of individuals")
+
+age_pyramid %>%
+  filter(!is.na(age) & !is.na(N)) %>%
+  mutate(age_class = case_when(age == 0 ~ 0, 
+                               age == 1 ~ 1, 
+                               age > 1 & age < 5 ~ 2, 
+                               age >= 5 ~ 3)) %>%
+  group_by(age_class) %>%
+  summarize(N = sum(abs(N))) %>%
[…]

### 2020-12-18 R/vacc_models.R

@@ -1,0 +1,140 @@
+
+# annual prob -> monthly prob of surv
+# annual birthrate -> monthly birthrate (Adults * births)
+# Can work in proportions (need to get starting proportion of adults vs. pups aged 1 mo - 12)
+
+# Add in revaccination!
+
+simvacc <- function(tmax, pup_surv, adult_surv, revacc, pup_cov, 
+                    campaign_cov, campaign_times, pup_month_vacc = 1,
+                    birthrate, S_adult_0, V_adult_0, 
+                    V_pup_0, S_pup_0) {
+  
+  # Set-up
+  V_adult <- matrix(0, nrow = 36, ncol = tmax)
+  S_adult <- rep(0, tmax)
+  V_pup <- S_pup <- matrix(0, nrow = 12, ncol = tmax)
+  
+  # Initalize
+  S_adult[1] <- S_adult_0
+  V_adult[1, 1] <- V_adult_0
+  V_pup[, 1] <- V_pup_0
+  S_pup[, 1] <- S_pup_0
+  
+  for(t in 2:tmax) {
+    
+    # Births (everyone first has babies and then dies)
+    births <- birthrate * (sum(V_adult[1:36, t - 1]) + S_adult[t - 1])
+    
+    # balance births (all to susceptibles)
+    S_pup[1, t] <- births # get replaced
+    
+    # Pup mortality
+    S_pup[2:12, t] <- S_pup[1:11, t - 1] * pup_surv
+    V_pup[2:12, t] <- V_pup[1:11, t - 1] * pup_surv
+    
+    # Balance the adults (ageing in 12 mo old pups)
+    S_adult[t] <- adult_surv * S_adult[t - 1] + pup_surv * S_pup[12, t - 1] 
+    
+    # pups to vaccinate (at age x)
+    pups_to_vacc <- ifelse(pup_month_vacc > 1, S_pup[pup_month_vacc, t], 
+                           births)
+    
+    # do puppy vaccination 
+    V_pup[pup_month_vacc, t] <- pups_to_vacc * pup_cov 
+    S_pup[pup_month_vacc, t] <- S_pup[pup_month_vacc, t] - pups_to_vacc * pup_cov 
+    
+    # Do the waning (adults only!)
+    V_adult[2:36, t] <- adult_surv * V_adult[1:35, t - 1]
+    S_adult[t] <- S_adult[t] + V_adult[36, t - 1] 
+    
[…]

### 2020-12-18 manuscript.Rmd
adding rmd draft and formatting

---
title: Lessons learned from pilot canine rabies vaccination campaigns in Moramanga
  District, Madagascar
author: "Malavika Rajeev"
date: "11/19/2020"
output:
  word_document: default
  html_document: default
bibliography: format/references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Abstract

## Introduction

Rabies

Dog vax

Rabies Mada & rabies & vets Dog Vax & cov & challenges [central point campaign vs. door to door style] Our study & aims. Misconceptions abt unowned dogs...

Here we compare two pilot vaccination efforts in the Moramanga District: one NGO-lead campaign with central point vaccination occuring over two weeks in 2018 in two communes (sub-district) vs. one locally-lead campaign with vaccines delivered continuously alongside routine veterinary services by a district veterinarian across the district between 2019 - 2020. The goal was to assess the feasibility, cost, and effectiveness of both campaign types.

## Methods

### Study Area
The Moramanga District is located mid-way between the central highlands and the east coast of Madagascar, at an average altitude of 936 m. It comprises 21 communes and NN fokontany (sub-commune), covering approximately 7150 km^2^ with a human population of x (2018 census prelim data). Previous work in the district has established a high burden of rabies exposures and deaths despite the availability of post-exposure prophylaxis at the district hospital. While Moramanga is relatively close to the capital city of Antananarivo (~ 3 hrs by bus), within the district, travel times between locations are highly variable, with much of the population living in more rural areas with limited access to roads and transportation [cite my other paper]. Before 2018, there were limited animal rabies vaccination services, with most vaccines available in urban communes and owners charged > 15,000 Ar (~ x USD) per vaccine administered. Livestock vaccination campaigns for anthrax as mandated by the government are conducted on an annual basis (owners are also charged, but fees vary by location and veterinarian).

### 2018 Campaign

```{r}
pts <- read_csv("data/campaign_pts_2018.csv")

# join with vacc #s and clean

```

In 2018, we organized a pilot vaccination campaign through a collaboration with two NGOs (the Madagascar Dog Initiative and Traveling Animal Doctors), the Department of Veterinary Services, and the Ministry of Public Health in the District of Moramanga. We focused on two communes in Moramanga, Moramanga Ville (the urban center of the district) and Andasibe (a rural commune surrounding Andasibe National Park), where previously high incidence of suspected rabid exposures (Moramanga Ville) and a high burden of deaths (Andasibe) had been recorded [@rajeev2018].

Between date start - date end, we used central point vaccine distribution, vaccinating at `r table(pts$Commune)["Andasibe"]` points in Andasibe and `r table(pts$Commune)["Moramanga Ville"]` in Moramanga Ville. A week before the campaign dates, for each location we advertised the date with the chief of the fokontany (sub-commune) and provided posters advertising the free vaccine. During the campaign, we used Rabisin `.r` (10 mL vials with 1 mL per dose, Boehringer Ingelheim) to vaccinate both dogs and cats presented that were > 1 month old. We surveyed owners about how many dogs and cats they owned in total (split by > 1 yr vs. < 1 yr in order to avoid language ambiguities that might result in excluding pups and kttens), as well as if their dogs were free roaming (no restrictions on movement by the owner or 'mireny' in Malagasy), or tied ('mifatora') or fenced ('fefy'). Vaccinations were delivered at no cost to owners, but as animal vaccination is generally thought of as a paid service, we asked owners how much in Ariary they would be willing to pay to have one animal vaccinated for rabies. For each animal vaccinated, we recorded the species (cat or dog), sex, approximate age in years, and whether the animal had been previously vaccinated.

To assess coverage, we used post-vaccination coverage surveys[@sambo2017; @gibson2015]. We marked all animals vaccinated with a colored, non-toxic, livestock crayon (ask Jochem for brand) along the top and back of the head. Between 4 - 6 PM on the same day as the campaign in each location, we did two transects in pairs (two people per team), walking for 1 hr starting in opposite directions and accompanied by a local guide to ensure that walking paths did not overlap. We recorded any marked and unmarked dogs we observed, and also recorded their roaming status (whether roaming, inside a fence, or tied), and their approximate age (greater or less than 1 yr of age).

### 2019 Campaign

For the 2019 campaign, instead of a central point campaign strategy, we distributed NN vaccinations (Rabisin `.r`, N vials) and the supplies needed to administer them (needle, syringe, vaccination card for owners) to the district veterinarian. She delivered the vaccination at no cost to owners, but was directly compensated 1,5000 Ar (~ 0.40 USD) per rabies vaccine administered. The campaign lasted from September 6, 2019 to June 19, 2020, with the district veterinarian advertising the vaccines by word of mouth for one week prior to her visiting each location [it would be good to know if she visited these places as part of her routine work or whether she went for the purposes of the dog vaccination]. For each vaccination, we collected the same information as previously (species, age, sex), as well as the approximate distance in meters travelled to receive the vaccination. Researchers communicated with the district veterinarian about progress periodically throughout the campaign, primarily through telephone calls. No other compensation or instructions were provided, and we asked the district veterinarian to administer as many (or as few vaccines) as feasible or wanted. As the vaccinations were delivered continuously, we did not do comporable post-vaccination surveys.

### Analyses

#### Coverage estimates
For the 2018 campaign, we used the transect data to estimate coverage as the proportion of dogs sighted that were marked with a binomial confidence interval for each location and at the commune level. For the 2019 campaign, we used a range of human:dog ratios (HDR) of 8 - 25, based on previous data from Madagascar [cite that tana study] and from the Moramanga District [cite leblanc thesis] to estimate the dog population from human population estimates [world pop] in each commune where the vaccinations were delivered [cite other papers that do this]. Then, coverage was estimated as the number of dogs vaccinated in total in that commune divided by the estimated dog population (the range given high and low HDRs). We used this same method for the 2018 campaign, as well, to compare coverage estimated by  the post-vaccination transects vs. the HDR range.

#### Cost estimates
We documented the overall costs of the two vaccination efforts. For the 2018 campaign, we broke costs down into the following categories: direct vaccine costs (cost for Rabisin `.r`, syringes, needles, vaccination cards), supplies (livestock crayons, muzzles, gloves, alcohol, swabs), transportation costs (gas and associated vehicle costs for transport), food and lodging for NGO personnel and other vaccinators during the campaign, personnel costs (per diems for DSV veterinarian, livestock field officers, local guides, and NGO employees), and advertisement (posters and banners for advertising the campaign). Foreign NGO volunteers expenses for travel to Madagascar were not included in this costs. Vehicles and drivers are also not included in these costs, as the drivers' time and vehicle use were donated to the campaign by volunteers involved in the campaign.

In 2019, costs were split into two categories, direct vaccine costs (for same items as in 2018), and personnel costs (per vaccine fee paid to the district veterinarian), and supplies (a generator and fuel for the veterinarian to maintain the vaccine on a cold chain when there were power outages).

#### Dog Demography
Using the age data we estimated stable age distributions, pup and adult survival in a maximum likelihood framework. Split dogs by these age classes (Anna & Morters). 
Eqs.

Optim. Estimates were sensitive to initial conditions, so we estimated across a range of par vals. And then took the pars where pop was growing (as we believe the dog population is likely tracking the growing pop of the district)

#### Modeling vaccination campaign strategies
We took those estimates and put them into a vaccination model of a hypothetical commune with 1000 dogs.

Eqs.

Vax lasts 3 yrs (based on Nobivac). And puppies were targeted for continous vax. Campaigns that reach 70% for each type. And combined vax. Revaccination of all individuals in between campaigns. But pup vacc is additive.

#### Data and code availability

## Results

###  Summary of 2018 and 2019 campaigns
During the volunteer-led campaign, a total of 3,137 animals were vaccinated (2,057 dogs and 1,080 cats) in two communes in the Moramanga District. We vaccinated 2,057 animals in the urban commune of Moramanga Ville (1,756 dogs and 254 cats) across 14 vaccination points and 528 animals in the rural commune of Andasibe (254 dogs and 235 cats) across seven vaccination points (Figure 1). On average, we vaccinated 148.3 animals per vaccination point, with a maximum of 354 animals vaccinated at one point. Overall, 10.7% of animals (15.1% of dogs and 2 % of cats) had a history of previous vaccination (owner reported), however only 4.7% had been vaccinated in the previous year (6.7% of dogs and <1% of cats). The mean number of dogs owned per household was 1.04 and the mean number of cats per household was 0.69. 55% of owners only had dogs and 27% had only cats (Table 1). Owners reported that most dogs resided inside a fenced-in area (52%) in Moramanga Ville compared to 71% of dogs in Andasibe, who were reported to be free-roaming. However, fenced animals were frequently observed outside fences during the campaign, and thus the majority of animals could be classified as semi-confined in the more urban township of Moramanga Ville, and free-roaming in the rural setting of Andasibe.

During the eight months of the DSV-led campaign, the DSV veterinarian vaccinated a total of 2,384 animals (1,898 dogs, 486 cats) across N communes in the Moramanga District. Overall, 72% of animals (75% of dogs and 20% of cats) had a history of previous vaccination (owner reported), however only 1.1% had been vaccinated in the previous year (0.6% of dogs and 2% of cats). 218 animals were reported as having been previously vaccinated in April 2018, likely during the volunteer-led campaign. Additionally, 23% of all animals vaccinated were spayed or neutered (27% of dogs and 8% of cats). Owners reported travelling an average of 611 meters (SD= 572.7 meters) to have their animals vaccinated (621 meters for dogs and 567 meters for cats). Owners with animals that did not have a history of previous vaccination travelled further on average than those that did (t = 7.747, df=136.1, p< 0.001) and dog owners travelled further to receive the vaccination than cat owners (t = 32.2416, df=1022.5, p= 0.013). (Table 2)

```{r}
knitr::include_graphics("figs/table1.png")
```

### Comparing campaign coverage estimates

In Andasibe Ville, we were only able to did this in one of our campaign locations, due to the sparsely distributed households and roads throughout the area, as such coverage in thi s

### Comparing campaign cost estimates

### Modeling campaigns and coverage

For dogs, the sex ratio was male-biased (55%); while for cats it was female-biased (68% female). For dogs, the sex ratio was male-biased (65%), while for cats it was female-biased (66% female).

```{r this is a code chunk!}

```

## Discussion

Main findings para (feasible, should not charge, costs vs. coverage).

Limitations (owner reports, no solid ests of dog pop although HDR is reasonable and supported) Need to monitor incidence (Rajeev unpublished data anecdotally). Limitations of age data and methods. Limitations of cov data and methods. One limitation here is unowned dogs Better understanding of demography (using campaigns to monitor pop growth etc. costs may rise given increase in dog pops, pop control strategies may also be important for humane implementation of vaccination programs). Compare to LeBlanc + Rila studies Better understanding of transmission (i.e. Bourrhy & Chad studies). Cost not included was PEP for exposures during a campaign (mostly cat scratches). Getting vaccinators prep. Focusing on puppies could also help that, as they are easier to handle. Covid?

More community based delivery (Ahmed's paper, Pivot, accessibility, etc.) Possibly combining different campaign strategies could work per models Turnover is tricky & waning doesn't contribute too much. Thermotolerant & 3 yrs even without booster. (although most dogs are revaccinated and we do this in the model) Volunteer based/NGO based. Push for govt investment and viewing as a public good. The veterinary workforce (relieving pressures on them, challenges with private-public roles).

Conclude with main points & how to move forward in Madagascar

## Acknowledgements
A note abt Annie. Others who helped

### 2021-01-19 manuscript.Rmd
parsimonious model of age distribution

@@
 Thermotolerant & 3 yrs even without booster.
-(although most dogs are revaccinated and we do this in the model) Volunteer based/NGO based.
+(although most dogs are revaccinated and we do this in the model).
+Volunteer based/NGO based.
 Push for govt investment and viewing as a public good.
+Modeling after childhood vax = routine through local offices of puppies + supplemental activities by trained vets annually to catch unvax (feral, oral).
 The veterinary workforce (relieving pressures on them, challenges with private-public roles).

### 2021-01-19 qs_for_jess/demography.Rmd
parsimonious model of age distribution

@@
 library(cowplot)
+library(here)
 
@@
 # Age pyramid
-animals_age <- read_csv("animals_age.csv")
+animals_age <- read_csv(here("qs_for_jess/animals_age.csv"))
 
@@
 ```{r}
+# constraining growth
 agemod1 <- function(par = c(psurv_pup = -1, 
@@
 
+  # Order of steps:
+  # 1. all adults have babies (S + V)
+  # 2. survival probabilities are applies to everyone
+  # 3. Surviving vaccinated indiviudals immunity wanes
+  # 4. Then new vax and revax happends
+
   for(t in 2:tmax) {
@@
 
-    # balance births (all to susceptibles)
-    S_pup[1, t] <- births # get replaced
+    # balance births (all surviving to susceptibles)
+    S_pup[1, t] <- births
 
@@
     # do puppy vaccination 
+    # CHECK: this should be additive no? Or is it always zero? Yes because no vaccs at t yet! Unless others have been vaccinated in campaign?
     V_pup[pup_month_vacc, t] <- pups_to_vacc * pup_cov 
@@
 
-    # Vaccinate as part of campaign accounting for revax
+    # Campaign vaccination of adults accounting for revaccination --------------
+
+    # Do the waning (adults only)
+    V_adult[2:36, t] <- adult_surv * V_adult[1:35, t - 1]
+    # After applying survival probabilities
+    S_adult[t] <- S_adult[t] + V_adult[36, t - 1]
+
+    # Coverage now
     cov_now <- ifelse(t %in% campaign_times, campaign_cov, 0)
-    camp_vacc_adult <- (S_adult[t] + revacc * sum(V_adult[, t])) * cov_now - revacc * sum(V_adult[, t])
-    camp_vacc_adult <- ifelse(camp_vacc_adult < 0, 0, camp_vacc_adult)
 
-    # every vax is revaccinated with a certain revacc prob
-    if(camp_vacc_adult > 0) {
-      V_adult[1, t] <- camp_vacc_adult + revacc * sum(V_adult[, t])
-      V_adult[2:36, t] <- (1 - revacc) * V_adult[2:36, t]
-    } else {
-      V_adult[1, t] <- camp_vacc_adult # newly vaccinated
-    }
+    # First get the number of currently vaccinated doggoes that are revaccinated
+    # Assume that these are distributed evenly across the waning continuum 
+    # i.e. same prop across V_adult[, t]
+    total_vax_adult <- cov_now * (S_adult[t] + sum(V_adult[, t]))
+    revax_adult <- sum(revacc * V_adult[, t])
 
-    # Balance addtl campaign vax adults
-    S_adult[t] <- S_adult[t] - camp_vacc_adult
+    # constrain new vaccination by preferentially revaccinating already vaxed adults
+    new_vax_adult <- ifelse(revax_adult > total_vax_adult, 0, total_vax_adult - revax_adult)
 
-    # Balance addtl campaign vax pups (assuming that no puppies are revax)
+    # constrain revaccination so it cannot exceed coverage level
+    # if greater than total_vax_adult, then adjusted coverage should be total_vax_adult/revac_adult
+    # to adjust to total expected vaccinations
+    adj_cov <- ifelse(revax_adult > total_vax_adult, 
+                      total_vax_adult/revax_adult * V_adult[, t], cov_now * V_adult[, t])
+
+    # subtract these from currently vaccinated adults 
+    V_adult[2:36, t] <- V_adult[2:36, t] - adj_cov
+
+    # and move them into the first waning step along with the newly vaccinated
+    V_adult[1, t] <- sum(adj_cov) + new_vax_adult
+
+    # Balance the susceptibles as well (subtract newly vaccinated)
+    S_adult[t] <- S_adult[t] - new_vax_adult
+
+    # Do the same vax steps for pups
+    # But assuming that puppies aren't revaccinated
     camp_vacc_pup <- S_pup[pup_month_vacc:12, t] * cov_now 
@@
     # Add in surviving pups aging to adults so that they can wane appropriately
-    t_to_sub <- pup_month_vacc - 1
-    V_adult[13 - t_to_sub, t] <- V_adult[13, t] +  pup_surv * V_pup[12, t - 1]
+    # for example if vaccinated at three months, will enter adulthood at waning stage of 10 months
+    # i.e. having spent 9 mos. vaccinated!
+    v_to_sub <- 13 - pup_month_vacc
+    V_adult[v_to_sub, t] <- V_adult[v_to_sub, t] +  pup_surv * V_pup[12, t - 1]
 
@@
 # with campaigns only
-camp <- simvacc(tmax = 12*5, pup_surv = surv$psurv_pup, adult_surv = surv$psurv_adult, 
+camp <- simvacc(tmax = 12*10, pup_surv = surv$psurv_pup, adult_surv = surv$psurv_adult, 
                 revacc = 1, 
                 pup_cov = 0, 
-        campaign_cov = 0.7, campaign_times = seq(12, 12*5, by = 12), 
+                campaign_cov = 0.7, campaign_times = seq(12, 12*10, by = 12), 
                 pup_month_vacc = 1, 
@@
 # with pup vacc only
-pup <- simvacc(tmax = 12*5, pup_surv = surv$psurv_pup, adult_surv = surv$psurv_adult, 
+pup <- simvacc(tmax = 12*10, pup_surv = surv$psurv_pup, adult_surv = surv$psurv_adult, 
                revacc = 1, 
@@
 
-both <- simvacc(tmax = 12*5, pup_surv = surv$psurv_pup, adult_surv = surv$psurv_adult, 
+both <- simvacc(tmax = 12*10, pup_surv = surv$psurv_pup, adult_surv = surv$psurv_adult, 
                 revacc = 1, 
                 pup_cov = 0.7, 
-        campaign_cov = 0.7, campaign_times = seq(12, 12*5, by = 12), 
+                campaign_cov = 0.7, 
+                campaign_times = seq(12, 12*10, by = 12), 
                 pup_month_vacc = 1, 
@@
 
-vacc_comp <- bind_rows(tibble(month = 1:60, 
+vacc_comp <- bind_rows(tibble(month = 1:120, 
                               cov = camp$V/(camp$S + camp$V), type = "Campaign"), 
-                       tibble(month = 1:60, 
+                       tibble(month = 1:120, 
                               cov = pup$V/(pup$S + pup$V), type = "Puppy vax"), 
-                       tibble(month = 1:60,
+                       tibble(month = 1:120,
                               cov = both$V/(both$S + both$V), type = "Both"))

### 2021-01-19 R/dem_models.R

@@ -1,7 +1,7 @@
-
-agemod1 <- function(par = c(psurv_pup = -1, 
-                            psurv_adult = -1),
-                    fert = 6, 
+agemod <- function(par = c(psurv_pup = -1, 
+                           psurv_adult = -1, 
+                           psurv_older = -1),
+                    fert = 3, 
                     n_per_age, type = "fit") {
   
   list2env(as.list(par), envir = environment())
@@ -9,23 +9,25 @@ agemod1 <- function(par = c(psurv_pup = -1,
   # transform parameters to [0, 1]
   p1 <- exp(psurv_pup)/(1 + exp(psurv_pup))
   p2 <- exp(psurv_adult)/(1 + exp(psurv_adult))
-
+  p3 <- exp(psurv_older)/(1 + exp(psurv_older))
+  
   # make the Leslie matrix
-  lmat <- matrix(0, 4, 4)
-  lmat[2, 1] <- p2 # every surviving baby becomes juvenile
-  lmat[3, 2] <- p2 # every surviving juvenile becomes adult
-  lmat[3, 3] <- p2 * 2/3 # 2/3 of them stay in this age class (for years 1 & 2 in age bin)
-  lmat[4 ,3] <- p2 * 1/3 # 1/3 of them age up (after year 3 in age bin)
-  lmat[4, 4] <- p2  # surviving of final age class
+  lmat <- matrix(0, 5, 5)
+  lmat[2, 1] <- p2 # every surviving pup becomes y1
+  lmat[3, 2] <- p2 # every surviving y1 becomes y2
+  lmat[4, 3] <- p2 # every surviving y2 becomes y3-5
+  lmat[4, 4] <- p3 * 1/2 # 2/3 of them stay in this age class (for years 3 and 4 in age bin)
+  lmat[5 ,4] <- p3 * 1/2 # 1/3 of them age up (after year 5 in age bin)
+  lmat[5, 5] <- p3  # surviving to final age class
   
   # add in the fertility (same for all adults, i.e. excluding puppies)
   # this is pup survival * birthrate which is not actually identifiable
-  # so we look at an average litter size of 6 / 2 = 3 
-  lmat[1, 2:4] <- fert * p1
+  # so we look at an average litter size of fert & try and estimate pup survival
+  lmat[1, 2:5] <- fert * p1
   
   # calculate what this does to the stable structure
[…]

### 2021-01-19 R/figure2.R

@@ -9,7 +9,7 @@ library(patchwork)
 library(cowplot)
 
 # data on costs
-costs <- read_csv("data/costs_temp.csv")
+costs <- read_csv("data/costs_summarized.csv")
 willing_to_pay <- read_csv("data/owner_2018.csv")
 
 # animals
@@ -18,22 +18,29 @@ animals_2019 <- read_csv("data/campaign_2019.csv")
 
 # Figure 1A costs by category
 costs %>%
-  tidyr::complete(category, year = 2018:2019, fill = list(prop = 0)) %>%
+  filter(include == "Y") %>%
+  group_by(year, category) %>%
+  summarize(cost_usd = sum(usd)) %>%
+  group_by(year) %>%
+  mutate(prop = cost_usd/sum(cost_usd)) %>%
+  ungroup() %>%
+  tidyr::complete(category, year = 2018:2019, 
+                  fill = list(cost_usd = 0, prop = 0)) %>%
   mutate(type = case_when(category == "Vaccine" ~ "Vaccine", 
-                          TRUE ~ "Implementation")) -> costs
+                   TRUE ~ "Implementation")) -> costs
 
 ggplot(costs) +
-  geom_col(aes(x = category, y = cost, fill = factor(year)), 
+  geom_col(aes(x = category, y = cost_usd, fill = factor(year)), 
            position = position_dodge()) +
   labs(x = "Category", y = "Approximate cost (USD)", tag = "A") +
   scale_fill_brewer(palette = "Dark2", name = "Year") +
   theme_minimal_hgrid(font_size = 12) +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1)) -> fig2A
+  coord_flip() -> fig2A
 
 # Figure 1B Cost per animal vaccinated
 costs %>%
   group_by(year, type) %>%
-  summarize(cost = sum(cost, na.rm = TRUE)) %>%
+  summarize(cost = sum(cost_usd, na.rm = TRUE)) %>%
   mutate(n_animals = case_when(year == 2018 ~ nrow(animals_2018), 
                                   year == 2019 ~ nrow(animals_2019)), 
           cost_per_animal = cost/n_animals) -> costs_per_animal
@@ -42,24 +49,30 @@ costs_per_animal %>%
   group_by(year) %>%
   summarize(total = sum(cost_per_animal)) -> total_per_animal
 
+# jerry rig
+costs_per_animal %>%
[…]

### 2021-01-19 R/vacc_models.R

@@ -1,10 +1,5 @@
-
-# annual prob -> monthly prob of surv
-# annual birthrate -> monthly birthrate (Adults * births)
-# Can work in proportions (need to get starting proportion of adults vs. pups aged 1 mo - 12)
-
-# Add in revaccination!
-
+# Need to track down an accounting issue where some revax get lost
+# but this is approximately right
 simvacc <- function(tmax, pup_surv, adult_surv, revacc, pup_cov, 
                     campaign_cov, campaign_times, pup_month_vacc = 1,
                     birthrate, S_adult_0, V_adult_0, 
@@ -21,13 +16,19 @@ simvacc <- function(tmax, pup_surv, adult_surv, revacc, pup_cov,
   V_pup[, 1] <- V_pup_0
   S_pup[, 1] <- S_pup_0
   
+  # Order of steps:
+  # 1. all adults have babies (S + V)
+  # 2. survival probabilities are applies to everyone
+  # 3. Surviving vaccinated indiviudals immunity wanes
+  # 4. Then new vax and revax happends
+  
   for(t in 2:tmax) {
     
     # Births (everyone first has babies and then dies)
     births <- birthrate * (sum(V_adult[1:36, t - 1]) + S_adult[t - 1])
     
-    # balance births (all to susceptibles)
-    S_pup[1, t] <- births # get replaced
+    # balance births (all surviving to susceptibles)
+    S_pup[1, t] <- births
     
     # Pup mortality
     S_pup[2:12, t] <- S_pup[1:11, t - 1] * pup_surv
@@ -41,6 +42,7 @@ simvacc <- function(tmax, pup_surv, adult_surv, revacc, pup_cov,
                            births)
     
     # do puppy vaccination 
+    # CHECK: this should be additive no? Or is it always zero? Yes because no vaccs at t yet! Unless others have been vaccinated in campaign?
     V_pup[pup_month_vacc, t] <- pups_to_vacc * pup_cov 
     S_pup[pup_month_vacc, t] <- S_pup[pup_month_vacc, t] - pups_to_vacc * pup_cov 
     
@@ -48,30 +50,51 @@ simvacc <- function(tmax, pup_surv, adult_surv, revacc, pup_cov,
     V_adult[2:36, t] <- adult_surv * V_adult[1:35, t - 1]
     S_adult[t] <- S_adult[t] + V_adult[36, t - 1] 
     
[…]

### 2021-01-21 R/dem_models.R
moving to ssh from https

@@ -1,6 +1,64 @@
+# testing gh token issue
+# other opts that you need to be able to do
+# estimating pup_surv and assuming fertility vs. estimating fertility and backing out fertility
+# constraining fertility to be greater than 1 (1 + fert) 
+# can just pass par_trans = (..., ..., ..., function(x) 1 + exp(x))
+# dpois (-ll minimize?) vs. sum of squares
+
+les_mat <- function(n_classes = 5, years_in_class = c(1, 1, 1, 2, 1),
+                    n_per_age, 
+                    par = c(p1 = -1, 
+                            p2 = -1, 
+                            fert = -1),
+                    par_trans = list(exp_trans, exp_trans, exp),
+                    trans_probs = c("p2", "p2", "p2", "p2", "p2"), 
+                    type = "fit") {
+  
+  
+  for(i in 1:length(par)) {
+    par[i] <- par_trans[[i]](par[i])
+  }
+  
+  list2env(as.list(par), envir = environment())
+  
+  lmat <- matrix(0, n_classes, n_classes)
+  
+  for(i in 1:(n_classes - 1)) {
+    lmat[i + 1, i] <- get(trans_probs[i]) * 1 / years_in_class[i]
+    lmat[i, i] <- get(trans_probs[i]) * (1 - 1/years_in_class[i])
+  }
+  
+  # last row / col (age class can be arbitrarily large)
+  lmat[n_classes, n_classes] <- get(trans_probs[n_classes])
+  
+  # Fertility
+   lmat[1, 2:n_classes] <- fert
+  
+  # calculate what this does to the stable structure
+  prop_stable <- Re(eigen(lmat)$vector[1:n_classes, 1]) / 
+    sum(Re(eigen(lmat)$vector[1:n_classes, 1]))
+  
+  if(type == "pred") {
+    # predicted stable age
+    stable_age <- prop_stable*sum(n_per_age)
+    growth <- Re(eigen(lmat)$value[1])
+    return(list(stable_age = stable_age, growth = growth))
+  }
+  
+  if(type == "fit") {
+    # get the sum of squares (distance between observed and predicted prop in each age class)
+    sum_of_squares <- sum((n_per_age - prop_stable*sum(n_per_age))^2)
+    return(sum_of_squares)
+  }
+}
+
+
+# Function for getting data binned by different age classes
+
+
+
[…]

### 2021-01-28 R/run_dem_mod.R
adding vacc and dem sims

@@ -1,0 +1,109 @@
+# Get demographic estimates for a set of bootstrapped data
+library(foreach)
+library(doParallel)
+library(iterators)
+library(dplyr)
+library(readr)
+library(here)
+library(tidyr)
+
+source("R/dem_models.R")
+
+# animals
+animals_2018 <- read_csv("data/animal_2018.csv")
+animals_2019 <- read_csv("data/campaign_2019.csv")
+
+# filter to dogs
+animals_2018 %>%
+  filter(Species == "A") %>%
+  mutate(year = 2018, 
+         AgeMonths = replace_na(AgeMonths, 0), 
+         AgeYears = replace_na(AgeYears, 0), 
+         age_mos = AgeYears*12 + AgeMonths) %>%
+  mutate(age_mos = ifelse(age_mos == 0, NA, age_mos), 
+         Sex = toupper(Sex)) %>%
+  select(Fokontany, Commune, Date, age_mos, year, Sex) -> animals_age
+
+animals_2019 %>%
+  filter(`Dog or Cat` == "dog") %>%
+  select(Date, age_mos = `Age (months)`, Commune, Fokontany = `Owner Location`, 
+         Sex) %>%
+  mutate(year = 2019, Sex = toupper(Sex)) %>%
+  bind_rows(animals_age) -> animals_age
+
+write_csv(animals_age, "out/animals_age.csv")
+
+# Bootstrapped data
+set.seed(1334)
+n_by_age <- (replicate(100, 
+                       ages_to_class(age_yrs = animals_age$age_mos/12, 
+                                     years_in_class = c(1, 1, 4, 1),
+                                     sample = TRUE, N = 1000)))
+n_by_age <- split(n_by_age, rep(1:ncol(n_by_age), each = nrow(n_by_age)))
+
+# original data
+original <- ages_to_class(age_yrs = animals_age$age_mos/12, 
+                          years_in_class = c(1, 1, 4, 1),
+                          sample = FALSE)
+n_by_age$`0` <- original
+  
+# Get expanded grid
+set.seed(1335)
+par_df <- data.frame(p1 = rnorm(100, mean = 0, sd = 1), 
+                     fert = rnorm(100, mean = 0, sd = 1))
+ll_grid <- expand_grid(par_df,  
+                       age_data = 1:length(n_by_age))
+
+# parallelize this (otherwise it takes a while...)
+cl <- makeCluster(detectCores() - 1)
[…]

### 2021-01-28 R/run_vacc_mod.R

@@ -1,0 +1,113 @@
+# Get demographic estimates for a set of bootstrapped data
+library(foreach)
+library(doParallel)
+library(iterators)
+library(dplyr)
+library(readr)
+library(here)
+library(tidyr)
+source("R/dem_models.R")
+source("R/vacc_models.R")
+
+dem_ests <- read_csv("out/dem_ests.csv")
+
+# Sim vaccination with parameters where pop is growing -------------------------
+dem_ests %>%
+  ungroup() %>%
+  filter(pop_growth == "Growing") %>%
+  select(fert_est, psurv_adult_est, growth, prop_est, age_class, sim) %>%
+  pivot_wider(values_from = prop_est, names_from = age_class) %>%
+  select(-sim) %>%
+  distinct() -> dem_ests_all
+
+# sample 1000 of these par values
+set.seed(1222)
+dem_ests_sim <- dem_ests_all[sample(nrow(dem_ests_all), 1000), ]
+
+# Average # of pups born to a dog each year = 4.9 pups * 1 / 0.4
+pups_per_dogyr <- 5 * 1 * 0.4 # from Annie's thesis & Anna's
+
+# parallelize this (otherwise it takes a while...)
+cl <- makeCluster(detectCores() - 1)
+registerDoParallel(cl)
+
+vacc_sims <- 
+  foreach(i = iter(dem_ests_sim, "row"), 
+          .export = c("simvacc", "bind_rows", "tibble"),
+          .combine = 'rbind') %dopar% {
+            
+            # get the prop
+            prop_pup <- as.numeric(i[, "1"])
+            prop_adult <- 1 - prop_pup
+            
+            # take the vals from optim and convert to adult survival
+            pup_surv <- (i$fert_est/pups_per_dogyr)^(1/12) # annual to monthly 
+            adult_surv <- i$psurv_adult_est^(1/12)
+            
+            # with campaigns only
+            camp <- simvacc(tmax = 12*10, 
+                            pup_surv = pup_surv, 
+                            adult_surv = adult_surv, 
+                            revacc = 1, 
+                            pup_cov = 0, 
+                            campaign_cov = 0.7, 
+                            campaign_times = seq(12, 12*10, by = 12), 
+                            pup_month_vacc = 1, 
+                            birthrate = 2.4/12, # monthly birth rate
+                            S_adult_0 = prop_adult*1000, 
+                            S_pup_0 = rep(prop_pup/12, 12)*1000, 
+                            V_adult_0 = 0, 
[…]

### 2021-01-28 R/dem_models.R
cleaning up figure and model src

@@ -31,6 +31,14 @@ exp_trans <- function(x) {
   exp(x)/(1 + exp(x))
 }
 
+apply_trans <- function(par, par_trans) {
+  for(i in 1:length(par)) {
+    par[i] <- par_trans[[i]](par[i])
+  }
+  return(par)
+}
+
+
 # First thing to try
 # estimating fertility and backing out pup survival given litter sizes
 # constraining fertility to be greater than 1 (1 + fert) vs not constrained
@@ -46,10 +54,7 @@ les_mat <- function(n_classes = 5, years_in_class = c(1, 1, 1, 2, 1),
                     trans_probs = c("p2", "p2", "p2", "p2", "p2"), 
                     type = "ss") {
   
-  
-  for(i in 1:length(par)) {
-    par[i] <- par_trans[[i]](par[i])
-  }
+  par <- apply_trans(par, par_trans)
   
   list2env(as.list(par), envir = environment())
   

### 2021-01-28 R/figure3.R

@@ -1,5 +1,4 @@
 # Figure 3. Demography and vax
-
 # pkgs
 library(dplyr)
 library(readr)
@@ -7,88 +6,81 @@ library(ggplot2)
 library(tidyr)
 library(patchwork)
 library(cowplot)
-library(bbmle)
 
-# animals
-animals_2018 <- read_csv("data/animal_2018.csv")
-animals_2019 <- read_csv("data/campaign_2019.csv")
+# bring in data
+animals_age <- read_csv("out/animals_age.csv")
 
-# filter to dogs
-animals_2018 %>%
-  filter(Species == "A") %>%
-  mutate(year = 2018, 
-         AgeMonths = replace_na(AgeMonths, 0), 
-         AgeYears = replace_na(AgeYears, 0), 
-         age_mos = AgeYears*12 + AgeMonths) %>%
-  mutate(age_mos = ifelse(age_mos == 0, NA, age_mos)) %>%
-  select(Fokontany, Commune, Date, age_mos, year, Sex) -> animals_age
-
-animals_2019 %>%
-  filter(`Dog or Cat` == "dog") %>%
-  select(Date, age_mos = `Age (months)`, Commune, Fokontany = `Owner Location`, 
-         Sex) %>%
-  mutate(year = 2019) %>%
-  bind_rows(animals_age) -> animals_age
-
-# histogram of ages
-ggplot(animals_age) +
-  geom_histogram(aes(x = age_mos)) +
-  facet_wrap(~year)
-
+# A original age pyramid
 animals_age %>%
   mutate(Sex = toupper(Sex), age = floor(age_mos/12)) %>%
   group_by(Sex, age) %>%
   summarize(N  = n()) %>%
   mutate(N = ifelse(Sex == "M", -N, N)) -> age_pyramid
 
-ggplot(age_pyramid, aes(x = age, y = N, fill = Sex)) +
+pyramid_A <- 
+  ggplot(age_pyramid, aes(x = age, y = N, fill = Sex)) +
   geom_col() +
   scale_y_continuous(breaks = c(seq(-600, -200, 200), seq(200, 600, 200)), 
                      labels = c(seq(600, 200, -200), seq(200, 600, 200)), 
                      limits = c(-600, 600)) +
   coord_flip() +
-  labs(x = "Age in years", y = "Number of individuals")
+  scale_fill_manual(values = c("#6f50a1", "#0d6d64")) +
+  labs(x = "Age in years", y = "Number of individuals", tag = "A") +
[…]

### 2021-01-28 R/dem_models.R
adding fig 3 in prog

@@ -93,3 +93,7 @@ les_mat <- function(n_classes = 5, years_in_class = c(1, 1, 1, 2, 1),
     return(-sum(dpois(n_per_age, prop_stable * sum(n_per_age), log = TRUE)))
   }
 }
+
+
+
+

### 2021-01-28 R/figure3.R

@@ -1,5 +1,4 @@
 # Figure 3. Demography and vax
-# pkgs
 library(dplyr)
 library(readr)
 library(ggplot2)
@@ -9,8 +8,10 @@ library(cowplot)
 
 # bring in data
 animals_age <- read_csv("out/animals_age.csv")
+dem_ests <- read_csv("out/dem_ests.csv")
+vacc_sims <- read_csv("out/vacc_sims.csv")
 
-# A original age pyramid
+# A: age pyramid ---------------
 animals_age %>%
   mutate(Sex = toupper(Sex), age = floor(age_mos/12)) %>%
   group_by(Sex, age) %>%
@@ -28,59 +29,63 @@ pyramid_A <-
   labs(x = "Age in years", y = "Number of individuals", tag = "A") +
   theme_minimal_hgrid(font_size = 12)
 
-ggplot(dem_ests) +
-  geom_histogram(aes(x = growth, fill = pop_growth))
-ggplot(dem_ests) +
-  geom_histogram(aes(x = fert_est, fill = pop_growth)) 
-ggplot(dem_ests) + 
-  geom_histogram(aes(x = psurv_adult_est, fill = pop_growth)) 
+# B: stable age distributions ---------------
+dem_ests %>%
+  select(sim, growth, age_class, prop_data, prop_est) %>%
+  pivot_longer(prop_data:prop_est, values_to = "prop") -> dem_props
 
-par_est_B <-
-  ggplot(filter(dem_ests), 
-         aes(x = fert_est, y = psurv_adult_est, 
-             shape = pop_growth)) +
-  geom_point() +
-  scale_shape_manual(values = c(1, 16)) +
+
+stable_age_B <- 
+  ggplot(dem_props) +
+  geom_line(aes(x = age_class, y = prop, group = interaction(sim, name), 
+                color = name), 
+            alpha = 0.25) +
+  scale_x_continuous(breaks = 1:4, 
+                     labels = c("0 - 1", "1 - 2", "2 - 6", "6+")) +
+  scale_color_manual(values = c("red", "blue"), 
+                     labels = c("Bootstrapped data", "Model estimate"), 
+                     name = "") +
+  labs(x = "Age class", y = "Proportion of \n population", tag = "B") +
   theme_minimal_hgrid(font_size = 12) +
-  labs(x = "Pup survival (annual prob)", 
-       y = "Adult survival (annual prob)", 
-       tag = "B", 
[…]

### 2021-01-28 R/run_dem_mod.R

@@ -6,7 +6,7 @@ library(dplyr)
 library(readr)
 library(here)
 library(tidyr)
-
+select <- dplyr::select
 source("R/dem_models.R")
 
 # animals
@@ -50,11 +50,11 @@ n_by_age$`0` <- original
 # Get expanded grid
 set.seed(1335)
 par_df <- data.frame(p1 = rnorm(100, mean = 0, sd = 1), 
-                     fert = rnorm(100, mean = 0, sd = 1))
+                     fert = rnorm(100, mean = 1, sd = 1))
 ll_grid <- expand_grid(par_df,  
                        age_data = 1:length(n_by_age))
 
-# parallelize this (otherwise it takes a while...)
+# parallelize this
 cl <- makeCluster(detectCores() - 1)
 registerDoParallel(cl)
 

### 2021-01-28 R/run_vacc_mod.R

@@ -25,7 +25,8 @@ set.seed(1222)
 dem_ests_sim <- dem_ests_all[sample(nrow(dem_ests_all), 1000), ]
 
 # Average # of pups born to a dog each year = 4.9 pups * 1 / 0.4
-pups_per_dogyr <- 5 * 1 * 0.4 # from Annie's thesis & Anna's
+pups_per_dogyr <- c(4, 6) * 1 * 0.4 # from Annie's thesis & Anna's
+dem_ests_sim <- expand_grid(dem_ests_sim, pups_per_dogyr)
 
 # parallelize this (otherwise it takes a while...)
 cl <- makeCluster(detectCores() - 1)
@@ -41,7 +42,7 @@ vacc_sims <-
             prop_adult <- 1 - prop_pup
             
             # take the vals from optim and convert to adult survival
-            pup_surv <- (i$fert_est/pups_per_dogyr)^(1/12) # annual to monthly 
+            pup_surv <- (i$fert_est/i$pups_per_dogyr)^(1/12) # annual to monthly 
             adult_surv <- i$psurv_adult_est^(1/12)
             
             # with campaigns only
@@ -89,19 +90,19 @@ vacc_sims <-
             vacc_comp <- bind_rows(tibble(month = 1:120, 
                                           cov = camp$V/(camp$S + camp$V), 
                                           type = "Campaign", 
-                                          pups_per_dogyr, 
+                                          i$pups_per_dogyr, 
                                           pup_surv = pup_surv, 
                                           adult_surv = adult_surv), 
                                    tibble(month = 1:120, 
                                           cov = pup$V/(pup$S + pup$V), 
                                           type = "Puppy vax", 
-                                          pups_per_dogyr, 
+                                          i$pups_per_dogyr, 
                                           pup_surv = pup_surv, 
                                           adult_surv = adult_surv), 
                                    tibble(month = 1:120,
                                           cov = both$V/(both$S + both$V), 
                                           type = "Both", 
-                                          pups_per_dogyr, 
+                                          i$pups_per_dogyr, 
                                           pup_surv = pup_surv, 
                                           adult_surv = adult_surv))
             vacc_comp

### 2021-01-28 R/Table1.R
finalized figs & tables

@@ -12,13 +12,16 @@ library(gt)
 vacc_2018 <- read_csv("data/animal_2018.csv")
 vacc_2019 <- read_csv("data/campaign_2019.csv")
 owner_2018 <- read_csv("data/owner_2018.csv")
+vacc_pts_2018 <- read_csv("data/campaign_pts_2018.csv")
 
 # filter to dogs
 vacc_2018 %>%
   filter(Species %in% "A") %>%
   mutate(year = 2018, 
          prev_vacc = ifelse(!is.na(PreviousVaccine), 1, 0), 
-         vacc_in_year = ifelse(PreviousVaccine %in% 2017, 1, 0)) %>%
+         vacc_in_year = ifelse(PreviousVaccine %in% 2017, 1, 0), 
+         Commune = ifelse(Commune %in% "Ambohibary", "Moramanga", Commune)
+         ) %>%
   select(Fokontany, loc = Commune, Date, year, Sex, prev_vacc, 
          vacc_in_year) -> animals_2018
 
@@ -34,7 +37,6 @@ vacc_2019 %>%
   bind_rows(animals_2018) -> animals_all
 
 animals_all %>%
-  filter(loc != "Ambohibary") %>%
   group_by(loc, year) %>%
   summarize(total = scales::number(n(), big.mark = ","), 
             prev_vacc = percent(sum(prev_vacc)/n()), 
@@ -45,6 +47,20 @@ animals_all %>%
   pivot_wider(names_from = c(loc, year), 
               values_from = value) -> dog_stats
 
+animals_all %>%
+  filter(year == 2018) %>%
+  mutate(loc = "All") %>%
+  group_by(loc, year) %>%
+  summarize(total = scales::number(n(), big.mark = ","), 
+            prev_vacc = percent(sum(prev_vacc)/n()), 
+            vacc_in_year = percent(sum(vacc_in_year)/n()),
+            perc_m = percent(sum(Sex %in% "M")/n())) %>%
+  pivot_longer(total:perc_m) %>%
+  pivot_wider(names_from = c(loc, year), 
+              values_from = value) %>%
+  left_join(dog_stats) -> dog_stats
+  
+# Get owner stats -------
 owner_2018 %>%
   group_by(Commune) %>%
   summarize(avg_dogs = number(mean(AdultDogs + Puppies, na.rm = TRUE), accuracy = 0.1), 
@@ -57,8 +73,44 @@ owner_2018 %>%
[…]

### 2021-01-28 R/figure3.R

@@ -21,8 +21,8 @@ animals_age %>%
 pyramid_A <- 
   ggplot(age_pyramid, aes(x = age, y = N, fill = Sex)) +
   geom_col() +
-  scale_y_continuous(breaks = c(seq(-600, -200, 200), seq(200, 600, 200)), 
-                     labels = c(seq(600, 200, -200), seq(200, 600, 200)), 
+  scale_y_continuous(breaks = c(seq(-600, -200, 200), 0, seq(200, 600, 200)), 
+                     labels = c(seq(600, 200, -200), "", seq(200, 600, 200)), 
                      limits = c(-600, 600)) +
   coord_flip() +
   scale_fill_manual(values = c("#6f50a1", "#0d6d64")) +
@@ -42,33 +42,33 @@ stable_age_B <-
             alpha = 0.25) +
   scale_x_continuous(breaks = 1:4, 
                      labels = c("0 - 1", "1 - 2", "2 - 6", "6+")) +
-  scale_color_manual(values = c("red", "blue"), 
-                     labels = c("Bootstrapped data", "Model estimate"), 
+  scale_color_manual(values = c("navy", "lightblue"), 
+                     labels = c("Bootstrapped data", "Model estimates"), 
                      name = "") +
   labs(x = "Age class", y = "Proportion of \n population", tag = "B") +
   theme_minimal_hgrid(font_size = 12) +
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  guides(color = guide_legend(override.aes = list(size = 1.1, alpha = 1)))
+
   
 # C: par ests ---------------
-pups_per_dogyr <- 5 * 1 * 0.4 # from Annie's thesis & Anna's
 dem_ests %>%
-  mutate(psurv_pup_est = fert_est/pups_per_dogyr) %>%
-  select(fert_est, psurv_adult_est, psurv_pup_est,
+  select(fert_est, psurv_adult_est,
          growth, pop_growth) %>%
   distinct() %>%
-  pivot_longer(fert_est:psurv_pup_est, values_to = "estimate", 
+  pivot_longer(fert_est:psurv_adult_est, values_to = "estimate", 
                names_to = "par") -> dem_pars
 
 par_labs <- c("fert_est" = "Fertility", 
-              "psurv_adult_est" = "Adult survival", 
-              "psurv_pup_est" = "Pup survival")
[…]

### 2021-01-28 R/run_vacc_mod.R

@@ -25,8 +25,9 @@ set.seed(1222)
 dem_ests_sim <- dem_ests_all[sample(nrow(dem_ests_all), 1000), ]
 
 # Average # of pups born to a dog each year = 4.9 pups * 1 / 0.4
-pups_per_dogyr <- c(4, 6) * 1 * 0.4 # from Annie's thesis & Anna's
-dem_ests_sim <- expand_grid(dem_ests_sim, pups_per_dogyr)
+set.seed(1224)
+litter_size <- c(4, 8) * 1 * 0.4 # from literature (Morters, Czupryna, etc.)
+dem_ests_sim <- expand_grid(dem_ests_sim, litter_size)
 
 # parallelize this (otherwise it takes a while...)
 cl <- makeCluster(detectCores() - 1)
@@ -42,7 +43,7 @@ vacc_sims <-
             prop_adult <- 1 - prop_pup
             
             # take the vals from optim and convert to adult survival
-            pup_surv <- (i$fert_est/i$pups_per_dogyr)^(1/12) # annual to monthly 
+            pup_surv <- (i$fert_est/i$litter_size)^(1/12) # annual to monthly 
             adult_surv <- i$psurv_adult_est^(1/12)
             
             # with campaigns only
@@ -90,19 +91,19 @@ vacc_sims <-
             vacc_comp <- bind_rows(tibble(month = 1:120, 
                                           cov = camp$V/(camp$S + camp$V), 
                                           type = "Campaign", 
-                                          i$pups_per_dogyr, 
+                                          litter_size = i$litter_size, 
                                           pup_surv = pup_surv, 
                                           adult_surv = adult_surv), 
                                    tibble(month = 1:120, 
                                           cov = pup$V/(pup$S + pup$V), 
                                           type = "Puppy vax", 
-                                          i$pups_per_dogyr, 
+                                          litter_size = i$litter_size, 
                                           pup_surv = pup_surv, 
                                           adult_surv = adult_surv), 
                                    tibble(month = 1:120,
                                           cov = both$V/(both$S + both$V), 
                                           type = "Both", 
-                                          i$pups_per_dogyr, 
+                                          litter_size = i$litter_size, 
                                           pup_surv = pup_surv, 
                                           adult_surv = adult_surv))
             vacc_comp

### 2021-01-28 manuscript.Rmd
updating with wip manuscript

@@
 
-Rabies
+*CITE*
 
-Dog vax
+Canine rabies results in an estimated 60,000 human deaths per year globally.
+These deaths are entirely preventable: prompt post-exposure prophylaxis (PEP) of humans is highly effective at preventing death and mass dog vaccination can interrupt transmission in domestic dogs and eventually lead to disease elimination.
+Annual vaccination campaigns achieving atleast 70% coverage are the recommended target for controlling rabies in domestic dog populations.
+However, achieving this coverage target in low and middle income countries where the virus can be challenging due to economic, sociocultural, and political reasons.
 
-Rabies Mada & rabies & vets Dog Vax & cov & challenges [central point campaign vs. door to door style] Our study & aims. Misconceptions abt unowned dogs...
+*Vax challenges and limitations here. Diff between SE Asia & SSA*
 
-Here we compare two pilot vaccination efforts in the Moramanga District: one NGO-lead campaign with central point vaccination occuring over two weeks in 2018 in two communes (sub-district) vs.
-one locally-lead campaign with vaccines delivered continuously alongside routine veterinary services by a district veterinarian across the district between 2019 - 2020.
-The goal was to assess the feasibility, cost, and effectiveness of both campaign types.
+In Madagascar, canine rabies has been endemic for over a century and the Institut Pasteur de Madagascar has provided PEP free-of-charge to patients in the country shortly therafter.
+However, currently, there are only 31 clinics where the vaccine is available, and there is limited dog vaccination due to high costs to owners and lack of availability of the vaccine and veterinary services more generally.
+*Estimates of dog population + Vet sector (WHO report)*.
+Livestock vaccination campaigns for anthrax as mandated by the government are conducted on an annual basis (owners are also charged, but fees vary by location and veterinarian).
+
+Here, we summarize lessons learned through implementation of pilot vaccination programs in the Moramanga District of Madagascar, where previous work has shown high incidence of dog rabies cases and human exposures.
+In 2018 and 2019, we deployed two different vaccination strategies.
+In 2018, we led a larger scale volunteer and NGO led pilot vaccination campaign in two Communes (sub-district level) using a central point strategy.
+In 2019/2020, we provided vaccinations and a per-vaccine fee to the District Veterinarian (DV) to vaccinate animals as part of routine services.
+We compare monetary costs and coverage estimates between the campaigns.
+Using demographic data collected during the campaigns and a vaccination model, we explore different vaccination strategies based on what we learned during implementation.
 
@@
 ### Study Area
-The Moramanga District is located mid-way between the central highlands and the east coast of Madagascar, at an average altitude of 936 m.
+
+The Moramanga District is located midway between the central highlands and the east coast of Madagascar, at an average altitude of 936 m.
 It comprises 21 communes and NN fokontany (sub-commune), covering approximately 7150 km^2^ with a human population of x (2018 census prelim data).
 Previous work in the district has established a high burden of rabies exposures and deaths despite the availability of post-exposure prophylaxis at the district hospital.
-While Moramanga is relatively close to the capital city of Antananarivo (~ 3 hrs by bus), within the district, travel times between locations are highly variable, with much of the population living in more rural areas with limited access to roads and transportation [cite my other paper].
-Before 2018, there were limited animal rabies vaccination services, with most vaccines available in urban communes and owners charged > 15,000 Ar (~ x USD) per vaccine administered.
-Livestock vaccination campaigns for anthrax as mandated by the government are conducted on an annual basis (owners are also charged, but fees vary by location and veterinarian).
+While Moramanga is relatively close to the capital city of Antananarivo (\~ 3 hrs by bus), within the district, travel times between locations are highly variable, with much of the population living in more rural areas with limited access to roads and transportation [cite my other paper].
+Before 2018, there were limited animal rabies vaccination services, with most vaccines available in urban communes and owners charged \> 15,000 Ar (\~ x USD) per vaccine administered.
 
@@
 
-Between date start - date end, we used central point vaccine distribution, vaccinating at `r table(pts$Commune)["Andasibe"]` points in Andasibe and `r table(pts$Commune)["Moramanga Ville"]` in Moramanga Ville.
 A week before the campaign dates, for each location we advertised the date with the chief of the fokontany (sub-commune) and provided posters advertising the free vaccine.
-During the campaign, we used Rabisin `.r` (10 mL vials with 1 mL per dose, Boehringer Ingelheim) to vaccinate both dogs and cats presented that were > 1 month old.
-We surveyed owners about how many dogs and cats they owned in total (split by > 1 yr vs.
-< 1 yr in order to avoid language ambiguities that might result in excluding pups and kttens), as well as if their dogs were free roaming (no restrictions on movement by the owner or 'mireny' in Malagasy), or tied ('mifatora') or fenced ('fefy').
-Vaccinations were delivered at no cost to owners, but as animal vaccination is generally thought of as a paid service, we asked owners how much in Ariary they would be willing to pay to have one animal vaccinated for rabies.
+During the campaign, we used Rabisin `.r` (10 mL vials with 1 mL per dose, Boehringer Ingelheim) to vaccinate both dogs and cats presented that were \> 1 month old.
+We surveyed owners about how many dogs and cats they owned in total (split by \> 1 yr vs.
+\< 1 yr in order to avoid language ambiguities that might result in excluding pups and kttens), as well as if their dogs were free roaming (no restrictions on movement by the owner or 'mireny' in Malagasy), or tied ('mifatora') or fenced ('fefy').
+Vaccinations were delivered at no cost to owners, but as animal vaccination is generally thought of as a paid service in Madagascar, we asked owners how much in Ariary they would be willing to pay to have one animal vaccinated for rabies.
 For each animal vaccinated, we recorded the species (cat or dog), sex, approximate age in years, and whether the animal had been previously vaccinated.
@@
 To assess coverage, we used post-vaccination coverage surveys[@sambo2017; @gibson2015].
-We marked all animals vaccinated with a colored, non-toxic, livestock crayon (ask Jochem for brand) along the top and back of the head.
+We marked all animals vaccinated with a colored, non-toxic, livestock crayon [ask Jochem for brand] along the top or back of the head.
 Between 4 - 6 PM on the same day as the campaign in each location, we did two transects in pairs (two people per team), walking for 1 hr starting in opposite directions and accompanied by a local guide to ensure that walking paths did not overlap.
@@
 
-For the 2019 campaign, instead of a central point campaign strategy, we distributed NN vaccinations (Rabisin `.r`, N vials) and the supplies needed to administer them (needle, syringe, vaccination card for owners) to the district veterinarian.
-She delivered the vaccination at no cost to owners, but was directly compensated 1,5000 Ar (~ 0.40 USD) per rabies vaccine administered.
-The campaign lasted from September 6, 2019 to June 19, 2020, with the district veterinarian advertising the vaccines by word of mouth for one week prior to her visiting each location [it would be good to know if she visited these places as part of her routine work or whether she went for the purposes of the dog vaccination].
-For each vaccination, we collected the same information as previously (species, age, sex), as well as the approximate distance in meters travelled to receive the vaccination.
-Researchers communicated with the district veterinarian about progress periodically throughout the campaign, primarily through telephone calls.
-No other compensation or instructions were provided, and we asked the district veterinarian to administer as many (or as few vaccines) as feasible or wanted.
-As the vaccinations were delivered continuously, we did not do comporable post-vaccination surveys.
+For the 2019 campaign, instead of a central point campaign strategy, we distributed vaccinations (Rabisin `.r`, N vials) and the supplies needed to administer them (needle, syringe, vaccination card for owners) to the District Veterinarian (DV).
+The DV delivered the vaccination at no cost to owners, but was directly compensated 1,5000 Ar (\~ 0.40 USD) per rabies vaccine administered.
+The campaign lasted from September 6, 2019 to June 19, 2020, with the DV advertising the vaccines by word of mouth for one week prior to her visiting each location [it would be good to know if she visited these places as part of her routine work or whether she went for the purposes of the dog vaccination].
+For each animal vaccinated, we collected the age and sex, and the DV also asked owners to approximate the distance they travelled to receive the vaccination.
+Researchers communicated with the DV about progress periodically throughout the campaign, primarily through telephone calls.
+No other compensation or instructions were provided, and we asked the DV to administer as many (or as few vaccines) as feasible or wanted.
+As the vaccinations were delivered continuously, we were unable to do comparable post-vaccination surveys.
 
@@
 #### Coverage estimates
+
 For the 2018 campaign, we used the transect data to estimate coverage as the proportion of dogs sighted that were marked with a binomial confidence interval for each location and at the commune level.
@@
 #### Cost estimates
+
 We documented the overall costs of the two vaccination efforts.
@@
 
+*Willingness to pay*.
+
+*Person days*.
+
+*Vax per point*.
+
 #### Dog Demography
-Using the age data we estimated stable age distributions, pup and adult survival in a maximum likelihood framework. Split dogs by these age classes (Anna & Morters). 
-Eqs.
 
-Optim.
-Estimates were sensitive to initial conditions, so we estimated across a range of par vals.
-And then took the pars where pop was growing (as we believe the dog population is likely tracking the growing pop of the district)
+Using the age data on vaccinated animals collected during both vaccination campaigns, we estimated the proportion of the population in four age classes: pups under the age of 1 year, juveniles 1 - 2 yrs, adults age 2 - 6 years, and older dogs age 6+ years.
+With the assumption that these estimates represent the population at a stable age distribution, we use a Leslie matrix model to estimate annual adult survival probability and fertility using maximum likelihood estimation [@Fujiwara2017].
+Specifically, we assume that the number of individuals in each age class follows a Poisson distribution with the mean predicted by the stable age distribution by the model (the proportion of individuals in each age class at equilibrium, equal to the eigenvector associated with the dominant eigenvalue of the matrix $\nu$) multiplied by the total number of individuals in the population ($N_t$):
+
+$$ N_{a} \sim Pois(\nu N_t) $$
+
+We assume that all individuals older than 1 reproduce and that fertility does not substantially decline (cite).
+To get bootstrapped estimates, we used 100 sub-sampled data sets of 1000 observations from the observed age data to fit the parameters, and also varied initital values used in the optimization (N = 100 initial values sets) for 10000 parameter estimates total.
 
 #### Modeling vaccination campaign strategies
-We took those estimates and put them into a vaccination model of a hypothetical commune with 1000 dogs.
 
-Eqs.
+```{r}
+animals_age <- read_csv("out/animals_age.csv")
+original <- ages_to_class(age_yrs = animals_age$age_mos/12, 
+                          years_in_class = c(1, 1, 4, 1),
+                          sample = FALSE)
+female <- ages_to_class(age_yrs = animals_age$age_mos[animals_age$Sex == "F"]/12, 
+                          years_in_class = c(1, 1, 4, 1),
+                          sample = FALSE)
+```
 
-Vax lasts 3 yrs (based on Nobivac).
-And puppies were targeted for continous vax.
-Campaigns that reach 70% for each type.
-And combined vax.
-Revaccination of all individuals in between campaigns.
-But pup vacc is additive.
+We used the parameter estimates from the demographic model to simulate different vaccination strategies in a hypothetical commune with 1000 dogs.
+We used a discrete time age-structed model with a monthly time step to compare three strategies:
 
-#### Data and code availability
+1)  Annual vaccination campaigns occuring within the same month each year targeting dogs of all ages
+2)  Routine and continuous vaccination of new puppies targeted at the age of 3 months
+3)  A combined approach with annual campaigns and routine puppy vaccination
+
+We split the dog population into pups (\< 1 yr old) and adults based on the stable age distribution estimated from the demographic model.
+To estimate pup survival in year one, we took the fertility estimates (number of new pups per reproducing dog in the pup age class per year) and divided by an estimate of newborn pups per reproducing dog each year based on average litter size (range of 3 - 8 based on estimates from the literature, Czupryna, Annie Yang, Morters, Hampson), average number of litters per female per year (1 on average, Czupryna, Annie Yang), and the proportion of the adult population that is female (estimated from our data).
+We assumed that for the annual campaign strategy, surviving vaccinated adult dogs were revaccinated in subsequent years, but that if a pup had been vaccinated within 9 months of the campaign, it was not revaccinated (cite?).
+A subset of parameter estimates resulted in estimates of population decline, but based on the shape of the age pyramid, the growing human population in the region, and to simulate reasonable campaign scenarios, we filtered to parameter estimates which corresponded to population growth.
+
+#### Data analysis
 
@@
 ### Summary of 2018 and 2019 campaigns
-During the volunteer-led campaign, a total of 3,137 animals were vaccinated (2,057 dogs and 1,080 cats) in two communes in the Moramanga District.
-We vaccinated 2,057 animals in the urban commune of Moramanga Ville (1,756 dogs and 254 cats) across 14 vaccination points and 528 animals in the rural commune of Andasibe (254 dogs and 235 cats) across seven vaccination points (Figure 1).
-On average, we vaccinated 148.3 animals per vaccination point, with a maximum of 354 animals vaccinated at one point.
-Overall, 10.7% of animals (15.1% of dogs and 2 % of cats) had a history of previous vaccination (owner reported), however only 4.7% had been vaccinated in the previous year (6.7% of dogs and <1% of cats).
-The mean number of dogs owned per household was 1.04 and the mean number of cats per household was 0.69.
-55% of owners only had dogs and 27% had only cats (Table 1).
-Owners reported that most dogs resided inside a fenced-in area (52%) in Moramanga Ville compared to 71% of dogs in Andasibe, who were reported to be free-roaming.
-However, fenced animals were frequently observed outside fences during the campaign, and thus the majority of animals could be classified as semi-confined in the more urban township of Moramanga Ville, and free-roaming in the rural setting of Andasibe.
+```{r}
+species_2018 <- table(animals_2018$Species)
+species_2019 <- table(tolower(animals_2019$`Dog or Cat`))
 
-During the eight months of the DSV-led campaign, the DSV veterinarian vaccinated a total of 2,384 animals (1,898 dogs, 486 cats) across N communes in the Moramanga District.
+```
+
+During the 2018 campaign, a total of `nrow(animals_2018)` animals were vaccinated (`r species_2018["A"]` dogs and `r species_2018["A"]` cats) in the Moramanga (urban) and Andasibe (rural) communes over a period of xx days (Table 1).
+We vaccinated at `r table(pts$Commune)["Andasibe"]` points in Andasibe and `r table(pts$Commune)["Moramanga Ville"]` in Moramanga Ville.
+On average, we vaccinated xx animals per vaccination point, but this varied significantly between the two communes, with a maximum of 354 animals vaccinated at one point.
+Overall, 10.7% of animals (15.1% of dogs and 2 % of cats) had a history of previous vaccination (owner reported), however only 4.7% had been vaccinated in the previous year (6.7% of dogs and \<1% of cats).
+While less than 20% of owners reported that animals were not free roaming, fenced animals were frequently observed outside fences during the campaign, and thus the majority of animals could be classified as semi-confined in the more urban township of Moramanga Ville, and free-roaming in the rural setting of Andasibe where almost 70% of owners reported their dogs as free-roaming.
+
+During 2019/2020, the DV vaccinated a total of 2,384 animals (1,898 dogs, 486 cats, Table 1) over a period of xx months and in N communes in the Moramanga District.
 Overall, 72% of animals (75% of dogs and 20% of cats) had a history of previous vaccination (owner reported), however only 1.1% had been vaccinated in the previous year (0.6% of dogs and 2% of cats).
 218 animals were reported as having been previously vaccinated in April 2018, likely during the volunteer-led campaign.
-Additionally, 23% of all animals vaccinated were spayed or neutered (27% of dogs and 8% of cats).
-Owners reported travelling an average of 611 meters (SD= 572.7 meters) to have their animals vaccinated (621 meters for dogs and 567 meters for cats).
-Owners with animals that did not have a history of previous vaccination travelled further on average than those that did (t = 7.747, df=136.1, p< 0.001) and dog owners travelled further to receive the vaccination than cat owners (t = 32.2416, df=1022.5, p= 0.013).
-(Table 2)
+Additionally, 23% of all animals vaccinated were spayed or neutered (27% of dogs and 8% of cats)**.
+The majority of people estimated travelling less than xx meters to bring their animal to the DV.
+
+*Table 1 should describe dogs!! (Take out cat parts & keep dogs only, & person days!)*
 
@@
 
-In Andasibe Ville, we were only able to did this in one of our campaign locations, due to the sparsely distributed households and roads throughout the area, as such coverage in thi s
+In Andasibe Ville, we were only able to did this in one of our campaign locations, due to the sparsely distributed households and roads throughout the area, as such coverage in this
 
-### Comparing campaign cost estimates
+### Comparing campaign costs and willingness to pay
+Costs (different categories)
+Person days + dogs vax per person day (number of vaccination points)
+Willingness to pay
 
-### Modeling campaigns and coverage
+### Dog demography and simulating vaccination strategies
 
@@
 costs may rise given increase in dog pops, pop control strategies may also be important for humane implementation of vaccination programs).
+Estimates were not super identifiable.
+No carrying cap.
+Turnover of the dog pop.
 Compare to LeBlanc + Rila studies Better understanding of transmission (i.e.
@@
 
-More community based delivery (Ahmed's paper, Pivot, accessibility, etc.) Possibly combining different campaign strategies could work per models Turnover is tricky & waning doesn't contribute too much.
+More community based delivery (Ahmed's paper, Pivot, accessibility, etc.).
+Possibly combining different campaign strategies could work per models Turnover is tricky & waning doesn't contribute too much.
 Thermotolerant & 3 yrs even without booster.
@@
 ## Acknowledgements
+
 A note abt Annie. Others who helped

### 2021-01-28 R/Table1.R

@@ -113,12 +113,12 @@ vacc_pts_2018 %>%
   
   
 all_stats %>%
-  mutate(name = case_when(name == "total" ~ "Total # of dogs \n vaccinated", 
+  mutate(name = case_when(name == "total" ~ "Total dogs \n vaccinated", 
                           name == "prev_vacc" ~ "Previous history of vaccination", 
                           name == "vacc_in_year" ~ "Vaccinated within last year", 
                           name == "perc_m" ~ "Percent male", 
-                          name == "avg_dogs" ~ "Average # of dogs per owner", 
-                          name == "avg_cats" ~  "Average # of cats per owner", 
+                          name == "avg_dogs" ~ "Average dogs per owner", 
+                          name == "avg_cats" ~  "Average cats per owner", 
                           name == "roaming" ~ "Percent of owners with \n free-roaming animals", 
                           TRUE ~ name)) %>%
   gt() %>%

### 2021-02-03 manuscript.Rmd
adding intermediate results to out

@@
 ---
-title: Lessons learned from pilot canine rabies vaccination campaigns in Moramanga
-  District, Madagascar
+title: "Lessons learned and paths forward for mass dog vaccination in Madagascar: a case study of pilot canine rabies vaccination campaigns in Moramanga District"
 author: "Malavika Rajeev"
@@
 Annual vaccination campaigns achieving atleast 70% coverage are the recommended target for controlling rabies in domestic dog populations.
-However, achieving this coverage target in low and middle income countries where the virus can be challenging due to economic, sociocultural, and political reasons.
+However, achieving this coverage target in low and middle income countries where the virus ia endemic can be challenging due to economic, ecological sociocultural, and political reasons.
 
@@
 In 2018 and 2019, we deployed two different vaccination strategies.
-In 2018, we led a larger scale volunteer and NGO led pilot vaccination campaign in two Communes (sub-district level) using a central point strategy.
+In 2018, we led a larger scale volunteer-led pilot vaccination campaign in two Communes (sub-district level) using a central point strategy.
 In 2019/2020, we provided vaccinations and a per-vaccine fee to the District Veterinarian (DV) to vaccinate animals as part of routine services.
@@
 
-#### Coverage estimates
-
-For the 2018 campaign, we used the transect data to estimate coverage as the proportion of dogs sighted that were marked with a binomial confidence interval for each location and at the commune level.
-For the 2019 campaign, we used a range of human:dog ratios (HDR) of 8 - 25, based on previous data from Madagascar [cite that tana study] and from the Moramanga District [cite leblanc thesis] to estimate the dog population from human population estimates [world pop] in each commune where the vaccinations were delivered [cite other papers that do this].
-Then, coverage was estimated as the number of dogs vaccinated in total in that commune divided by the estimated dog population (the range given high and low HDRs).
-We used this same method for the 2018 campaign, as well, to compare coverage estimated by the post-vaccination transects vs.
-the HDR range.
-
 #### Cost estimates
@@
 
-*Willingness to pay*.
+*Willingness to pay*. *Person days*. *Vax per point*.
 
-*Person days*.
+#### Coverage estimates
 
-*Vax per point*.
+For the 2018 campaign, we used the transect data to estimate coverage as the proportion of dogs sighted that were marked with a binomial confidence interval for each location and at the commune level.
+For the 2019 campaign, we used a range of human:dog ratios (HDR) of 8 - 25, based on previous data from Madagascar [cite that tana study] and from the Moramanga District [cite leblanc thesis] to estimate the dog population from human population estimates [world pop] in each commune where the vaccinations were delivered [cite other papers that do this].
+Then, coverage was estimated as the number of dogs vaccinated in total in that commune divided by the estimated dog population (the range given high and low HDRs).
+We used this same method for the 2018 campaign, as well, to compare coverage estimated by the post-vaccination transects vs.
+the HDR range.
 
@@
 species_2019 <- table(tolower(animals_2019$`Dog or Cat`))
-
 ```
@@
 
-*Table 1 should describe dogs!! (Take out cat parts & keep dogs only, & person days!)*
+*Table 1 should describe dogs!! (Take out cat parts & keep dogs only, & person days, split by location!)*
 
@@
 
+### Comparing campaign costs and willingness to pay
+
+The 2018 campaign cost more overall and per animal vaccinated than the 2019 campaign, largely due to increased personnel costs (Fig 3A & B, Table 1).
+Reflecting the extra personnel necessary to run the central point campaign, the 2018 campaign also took substantially more person-days per animal vaccinated (Table 1).
+We found that in most contexts, charging owners for vaccination would result in minimal cost recovery (Fig 3C), but that even a nominal fee would significantly reduce vaccination coverage acheived (Fig 3D), particulary in the rural commune of Andasibe.
+
+```{r this is a code chunk!} 
+knitr::include_graphics("figs/fig2.jpeg")
+```
+
 ### Comparing campaign coverage estimates
 
-In Andasibe Ville, we were only able to did this in one of our campaign locations, due to the sparsely distributed households and roads throughout the area, as such coverage in this
+In Andasibe Ville, we were only able to did this in one of our campaign locations, due to the sparsely distributed households and roads throughout the area, as such coverage in this.
+Make sure to mention back calc of pop & hdr is it reasonable?
 
-### Comparing campaign costs and willingness to pay
-Costs (different categories)
-Person days + dogs vax per person day (number of vaccination points)
-Willingness to pay
+```{r this is a code chunk!}
+knitr::include_graphics("figs/fig1.jpeg")
+```
 
 ### Dog demography and simulating vaccination strategies
-
 For dogs, the sex ratio was male-biased (55%); while for cats it was female-biased (68% female). For dogs, the sex ratio was male-biased (65%), while for cats it was female-biased (66% female).
@@
 ```{r this is a code chunk!}
-
+knitr::include_graphics("figs/fig3.jpeg")
 ```
@@
 
+Targeting dogs only?
+Point to lit on willingness to pay / person day comp / cost per animal vaccinated
+
+PrEP and exposures during vax, we had N exposures. Puppies were the easiest to handle for both owners and vaccinators and vax is effective in them. Better dosing guidelines. Longer lasting vax.
+
 More community based delivery (Ahmed's paper, Pivot, accessibility, etc.).
@@
 The veterinary workforce (relieving pressures on them, challenges with private-public roles).
+Paying people is important hard/relieving pressures and making sure it's viewed as a public good (Ahmed's paper). Veterinary students?
 
@@
 
+Policy recommendations:
+- Dog vax as a public good, remove barriers for dog owners & incentivize vets/etc/ remove economic pressures
+- Community based vax delivery particulalry in novel areas (could lower cost while improving coverage)
+- Easier ways to estimate coverage (& coupling with surveillance to assess impact)
+
 ## Acknowledgements
 
-A note abt Annie. Others who helped
+We thank vets livestock officers volunteers fokontany leaders government officials.
+We want to especially acknowledge the important contributions made by Annie Yang, an undergraduate researcher and volunteer with Princeton University and the Madagascar Dog Initiative who tragically passed away in a car accident before this research was published.

### 2021-02-03 R/Table1.R

@@ -1,4 +1,4 @@
-# Table 1
+# Table 1 & other stats
 
 # pkgs
 library(dplyr)
@@ -13,6 +13,20 @@ vacc_2018 <- read_csv("data/animal_2018.csv")
 vacc_2019 <- read_csv("data/campaign_2019.csv")
 owner_2018 <- read_csv("data/owner_2018.csv")
 vacc_pts_2018 <- read_csv("data/campaign_pts_2018.csv")
+
+# Get all animals vaccinated
+vacc_2018 %>%
+  mutate(year = 2018, 
+         location = ifelse(Commune %in% "Ambohibary", "Moramanga", Commune)) %>%
+  group_by(year, location) %>%
+  summarize(n_animals = n()) %>%
+  bind_rows(vacc_2018 %>% 
+              summarize(n_animals = n()) %>% 
+              mutate(year = 2018, location = "All")) %>%
+  bind_rows(vacc_2019 %>% 
+              summarize(n_animals = n()) %>%
+              mutate(year = 2019, 
+                    location = "All")) -> animals_All
 
 # filter to dogs
 vacc_2018 %>%
@@ -64,7 +78,6 @@ animals_all %>%
 owner_2018 %>%
   group_by(Commune) %>%
   summarize(avg_dogs = number(mean(AdultDogs + Puppies, na.rm = TRUE), accuracy = 0.1), 
-            avg_cats = number(mean(Cats + Kittens, na.rm = TRUE),  accuracy = 0.1), 
             roaming = percent(sum(grepl("mireny", 
                                         Roaming, 
                                         ignore.case = TRUE))/n())) %>%
@@ -80,7 +93,6 @@ owner_2018 %>%
   mutate(loc = "All_2018") %>%
   group_by(loc) %>%
   summarize(avg_dogs = number(mean(AdultDogs + Puppies, na.rm = TRUE), accuracy = 0.1), 
-            avg_cats = number(mean(Cats + Kittens, na.rm = TRUE),  accuracy = 0.1), 
             roaming = percent(sum(grepl("mireny", 
                                         Roaming, 
                                         ignore.case = TRUE))/n())) %>%
@@ -90,37 +102,66 @@ owner_2018 %>%
   left_join(owner_stats) %>%
   bind_rows(dog_stats, .) -> all_stats
 
-# Get owner stats -------
-vacc_2019 %>% 
-  select(Date) %>% 
[…]

### 2021-02-03 R/figure2.R

@@ -12,6 +12,9 @@ library(cowplot)
 costs <- read_csv("data/costs_summarized.csv")
 willing_to_pay <- read_csv("data/owner_2018.csv")
 
+# exchange rate (midpoint between 2018 and 2019 est)
+exch_rate <- (3128 + 3500)/2
+
 # animals
 animals_2018 <- read_csv("data/animal_2018.csv")
 animals_2019 <- read_csv("data/campaign_2019.csv")
@@ -36,6 +39,8 @@ ggplot(costs) +
   scale_fill_brewer(palette = "Dark2", name = "Year") +
   theme_minimal_hgrid(font_size = 12) +
   coord_flip() -> fig2A
+
+write_csv(costs, "out/costs_by_category.csv")
 
 # Figure 1B Cost per animal vaccinated
 costs %>%
@@ -70,7 +75,9 @@ ggplot() +
   theme(axis.line.y = element_blank()) +
   guides(color = guide_legend(override.aes = list(fill = NA))) -> fig2B
 
-# Figure 1C Willingness to pay
+write_csv(costs_per_animal, "out/costs_per_animal.csv")
+
+# cost recovery bits
 prop_willing <- function(charge, willing) {
   length(willing[willing >= charge])/length(willing)
 }
@@ -92,18 +99,7 @@ willing_to_pay %>%
   mutate(charged = charge, Commune = "Overall") %>%
   bind_rows(willing_by_loc) -> willingness
 
-ggplot(willingness) +
-  geom_line(aes(x = charged/3100, y = prop_willing, linetype = Commune), 
-            size = 1.2) +
-  xlim(c(0, 3)) +
-  scale_linetype_manual(values = c(3, 2, 1), 
-                        c("Andasibe (rural)", "Moramanga (urban)", 
-                          "Overall"), name = "Location") +
-  labs(x = "Amount charged to \n owner (USD)", y = "Proportion of owners \n willing to pay", tag = "C") +
-  theme_minimal_hgrid(font_size = 12) +
-  guides(linetype = guide_legend(override.aes = list(size = 0.5))) -> fig2C
-
-# Figure 1D cost recovery (given estimate for both years)
+# Figure 1C cost recovery (given estimate for both years)
 cost_tradeoff <- function(charged, base_cost, cost_per_animal, nvacc, 
                           prop_willing) {
   new_vacc <- nvacc - (1 - prop_willing)*nvacc
[…]

### 2021-02-03 R/figure1.R
switching figure order

@@ -101,7 +101,7 @@ cov_commune %>%
 st_geometry(map_cov) <- map_cov$geometry
 
 # Map with midpoint
-fig1A <-
+fig2A <-
   ggplot(map_cov) +
   geom_sf(fill = NA) + 
   stat_sf_coordinates(data = filter(map_cov, vacc != 0), 
@@ -123,7 +123,7 @@ fig1A <-
     facet.lev = 2019, border.size = 0.5
   ) 
 
-# fig1A inset
+# fig2A inset
 mada_out <- st_read("data/shapefile/mada_districts_simple.shp")
 mora_plot <- st_union(filter(mada_out, district == "Moramanga"))
 mada_plot <- st_union(mada_out)
@@ -135,12 +135,12 @@ map_inset <- ggplot() +
   theme(panel.border = element_rect(color = "black", fill = NA))
 
 # Final figure 1 formatted for pos
-layout_A <- fig1A + inset_element(map_inset, left = 0, right = 0.2,
+layout_A <- fig2A + inset_element(map_inset, left = 0, right = 0.2,
                                   top = 0.9, bottom = 0.5, on_top = TRUE, 
                                   align_to = "full")
 
 # Plot with range
-fig1B <- 
+fig2B <- 
   ggplot(cov_commune) +
   geom_pointrange(aes(x = Commune, y = cov_est, 
                       ymin = scales::squish(cov_est_lower, c(0, 1)), 
@@ -156,7 +156,7 @@ fig1B <-
   theme_minimal_hgrid() +
   theme(axis.text.x = element_text(angle = 45, hjust = 1))
 
-fig1 <- (layout_A / fig1B) + plot_layout(heights = c(2, 1), guides = "collect")
+fig2 <- (layout_A / fig2B) + plot_layout(heights = c(2, 1), guides = "collect")
 
-ggsave("figs/fig1.jpeg", height = 6, width = 7)
+ggsave("figs/fig2.jpeg", height = 6, width = 7)
                   

### 2021-02-03 R/figure2.R

@@ -38,7 +38,7 @@ ggplot(costs) +
   labs(x = "Category", y = "Approximate cost (USD)", tag = "A") +
   scale_fill_brewer(palette = "Dark2", name = "Year") +
   theme_minimal_hgrid(font_size = 12) +
-  coord_flip() -> fig2A
+  coord_flip() -> fig1A
 
 write_csv(costs, "out/costs_by_category.csv")
 
@@ -73,7 +73,7 @@ ggplot() +
   coord_flip() +
   theme_cowplot(font_size = 12) + 
   theme(axis.line.y = element_blank()) +
-  guides(color = guide_legend(override.aes = list(fill = NA))) -> fig2B
+  guides(color = guide_legend(override.aes = list(fill = NA))) -> fig1B
 
 write_csv(costs_per_animal, "out/costs_per_animal.csv")
 
@@ -138,7 +138,7 @@ ggplot() +
   facet_wrap(~year) +
   theme_minimal_hgrid(font_size = 12, ) +
   guides(linetype = guide_legend(override.aes = list(size = 0.5))) +
-  theme(panel.border = element_rect(color = "black"))-> fig2C
+  theme(panel.border = element_rect(color = "black"))-> fig1C
 
 write_csv(cost_recovery, "out/cost_recovery.csv")
 
@@ -152,10 +152,10 @@ ggplot(willingness) +
                           "Overall"), name = "Location") +
   labs(x = "Amount charged to \n owner (USD)", y = "Reduction in coverage (%)", tag = "D") +
   theme_minimal_hgrid(font_size = 12) +
-  guides(linetype = guide_legend(override.aes = list(size = 0.5))) -> fig2D
+  guides(linetype = guide_legend(override.aes = list(size = 0.5))) -> fig1D
 
 write_csv(willingness, "out/willingess_to_pay.csv")
 
 # arrange
-fig2 <- (fig2A | fig2B) / (fig2C | fig2D) + plot_layout(guides = "collect")
-ggsave("figs/fig2.jpeg", height = 8, width = 10)
+fig1 <- (fig1A | fig1B) / (fig1C | fig1D) + plot_layout(guides = "collect")
+ggsave("figs/fig1.jpeg", height = 8, width = 10)

### 2021-02-04 manuscript.Rmd
drafting paper

@@
 library(tidyverse)
+library(glue)
+library(magrittr)
+library(lubridate)
+
+# stats
+# From Claire's MSC 
+hdrs <- tribble(~loc, ~range_people_ph, ~range_dogs_ph, ~n_households,
+                "urban", 4.13, 0.24, 170,
+                "rural", 5.654, 0.26, 164)
+# Range of hdrs then
+hdrs %>% 
+  mutate(mora_hdrs = range_people_ph / range_dogs_ph) -> hdrs
+mora_mid_hdr <- median(hdrs$mora_hdrs)
+hdr_backcalcs <- read_csv("out/hdr_backcalcs.csv")
+
+# animals
+animals_2018 <- read_csv("data/animal_2018.csv")
+animals_2019 <- read_csv("data/campaign_2019.csv")
+
 ```
@@
 
-*CITE*
-
 Canine rabies results in an estimated 60,000 human deaths per year globally.
@@
 Annual vaccination campaigns achieving atleast 70% coverage are the recommended target for controlling rabies in domestic dog populations.
-However, achieving this coverage target in low and middle income countries where the virus ia endemic can be challenging due to economic, ecological sociocultural, and political reasons.
+However, achieving this coverage target in low and middle income countries where the virus is endemic can be challenging due to economic, ecological, sociocultural, and political reasons.
 
-*Vax challenges and limitations here. Diff between SE Asia & SSA*
+In sub-Saharan African countries, parenteral vaccination implemented through static point vaccination campaigns have been shown to be cost-effective and feasible.
+While most dogs in these settings are considered free-roaming, they are owned and are accessible for vaccination through these campaigns.
+However, reaching and maintaining high coverage vaccination campaigns requires large scale coordinated efforts, and the challenges in implementation reflect both financial and logistical constraints more than feasibility of vaccination itself.
 
 In Madagascar, canine rabies has been endemic for over a century and the Institut Pasteur de Madagascar has provided PEP free-of-charge to patients in the country shortly therafter.
-However, currently, there are only 31 clinics where the vaccine is available, and there is limited dog vaccination due to high costs to owners and lack of availability of the vaccine and veterinary services more generally.
-*Estimates of dog population + Vet sector (WHO report)*.
-Livestock vaccination campaigns for anthrax as mandated by the government are conducted on an annual basis (owners are also charged, but fees vary by location and veterinarian).
+However, currently, there are only 31 clinics where the vaccine is available, and there is limited dog vaccination due to high costs to owners and lack of availability of the vaccine.
+Recent studies have estimated a high burden of human deaths, masked by limited surveillance for canine rabies cases and human deaths across the country.
+The veterinary sector is largely private, with some veterinarians employed in hybrid private/public employment as a designated district veterinarian.
+There are currently xx district veterinarians situated in xx out of 114 districts in the countries.
+While dog vaccination is rare, livestock officers and veterinarians work together to implement vaccination campaigns for anthrax on an annual basis as mandated by the government (wners are charged a fee per animal for these vaccines which vary by location).
+To date, no regular mass dog vaccinations are conducted on the island, although a few pilot programs have begun in recent years, largely implemented by NGO-goverment partnerships.
 
@@
 In 2019/2020, we provided vaccinations and a per-vaccine fee to the District Veterinarian (DV) to vaccinate animals as part of routine services.
-We compare monetary costs and coverage estimates between the campaigns.
-Using demographic data collected during the campaigns and a vaccination model, we explore different vaccination strategies based on what we learned during implementation.
+We compare time, human resources, costs, and coverage estimates between the campaigns.
+Using demographic data collected during the campaigns and a vaccination model, we also explore different vaccination strategies based on what we learned during implementation.
 
@@
 Previous work in the district has established a high burden of rabies exposures and deaths despite the availability of post-exposure prophylaxis at the district hospital.
-While Moramanga is relatively close to the capital city of Antananarivo (\~ 3 hrs by bus), within the district, travel times between locations are highly variable, with much of the population living in more rural areas with limited access to roads and transportation [cite my other paper].
-Before 2018, there were limited animal rabies vaccination services, with most vaccines available in urban communes and owners charged \> 15,000 Ar (\~ x USD) per vaccine administered.
+While Moramanga is relatively close to the capital city of Antananarivo (~ 3 hrs by bus), within the district, travel times between locations are highly variable, with much of the population living in more rural areas with limited access to roads and transportation [cite my other paper].
+Before 2018, there were limited animal rabies vaccination services, with most vaccines available in urban communes and owners charged > 15,000 Ar (~ x USD) per vaccine administered.
 
@@
 A week before the campaign dates, for each location we advertised the date with the chief of the fokontany (sub-commune) and provided posters advertising the free vaccine.
-During the campaign, we used Rabisin `.r` (10 mL vials with 1 mL per dose, Boehringer Ingelheim) to vaccinate both dogs and cats presented that were \> 1 month old.
-We surveyed owners about how many dogs and cats they owned in total (split by \> 1 yr vs.
-\< 1 yr in order to avoid language ambiguities that might result in excluding pups and kttens), as well as if their dogs were free roaming (no restrictions on movement by the owner or 'mireny' in Malagasy), or tied ('mifatora') or fenced ('fefy').
+During the campaign, we used Rabisin `.r` (10 mL vials with 1 mL per dose, Boehringer Ingelheim) to vaccinate both dogs and cats presented that were > 1 month old.
+We surveyed owners about how many dogs and cats they owned in total (split by > 1 yr vs.
+< 1 yr in order to avoid language ambiguities that might result in excluding pups and kttens), as well as if their dogs were free roaming (no restrictions on movement by the owner or 'mireny' in Malagasy), or tied ('mifatora') or fenced ('fefy').
 Vaccinations were delivered at no cost to owners, but as animal vaccination is generally thought of as a paid service in Madagascar, we asked owners how much in Ariary they would be willing to pay to have one animal vaccinated for rabies.
@@
 For the 2019 campaign, instead of a central point campaign strategy, we distributed vaccinations (Rabisin `.r`, N vials) and the supplies needed to administer them (needle, syringe, vaccination card for owners) to the District Veterinarian (DV).
-The DV delivered the vaccination at no cost to owners, but was directly compensated 1,5000 Ar (\~ 0.40 USD) per rabies vaccine administered.
+The DV delivered the vaccination at no cost to owners, but was directly compensated 1,5000 Ar (~ 0.40 USD) per rabies vaccine administered.
 The campaign lasted from September 6, 2019 to June 19, 2020, with the DV advertising the vaccines by word of mouth for one week prior to her visiting each location [it would be good to know if she visited these places as part of her routine work or whether she went for the purposes of the dog vaccination].
-For each animal vaccinated, we collected the age and sex, and the DV also asked owners to approximate the distance they travelled to receive the vaccination.
+For each animal vaccinated, we collected the age and sex, and the DV also asked owners to approximate the distance they travelled to receive the vaccination in meters.
 Researchers communicated with the DV about progress periodically throughout the campaign, primarily through telephone calls.
@@
 
-#### Cost estimates
+#### Campaign resource and cost comparisons
 
-We documented the overall costs of the two vaccination efforts.
+We documented the overall costs and resources required for the two vaccination efforts.
+We tracked the number of vaccination points, the number of days over which these vaccinations occurred, and the number of person days required (i.e.
+the number of working people per day over the campaign [@mazeri2021]).
+
 For the 2018 campaign, we broke costs down into the following categories: direct vaccine costs (cost for Rabisin `.r`, syringes, needles, vaccination cards), supplies (livestock crayons, muzzles, gloves, alcohol, swabs), transportation costs (gas and associated vehicle costs for transport), food and lodging for NGO personnel and other vaccinators during the campaign, personnel costs (per diems for DSV veterinarian, livestock field officers, local guides, and NGO employees), and advertisement (posters and banners for advertising the campaign).
@@
 Vehicles and drivers are also not included in these costs, as the drivers' time and vehicle use were donated to the campaign by volunteers involved in the campaign.
-
 In 2019, costs were split into two categories, direct vaccine costs (for same items as in 2018), and personnel costs (per vaccine fee paid to the district veterinarian), and supplies (a generator and fuel for the veterinarian to maintain the vaccine on a cold chain when there were power outages).
 
-*Willingness to pay*. *Person days*. *Vax per point*.
+We used the data on owner reported willingness to pay for vaccine to estimate the proportional reduction in animals vaccinated as fees are increased.
+We also estimated how this would impact cost per animal vaccinated by approximating the costs for implementation (i.e.
+those costs that remain fixed) from costs incurred per animal vaccinated (i.e.
+vaccine, syringe, vaccination card, per vaccination fee to DV in 2019), and calculating the balance between the returns from owner payments (i.e.
+increases in cost recovery per animal vaccinated) vs.
+decreasing numbers of animals vaccinated overall.
 
@@
 
-We split the dog population into pups (\< 1 yr old) and adults based on the stable age distribution estimated from the demographic model.
+We split the dog population into pups (< 1 yr old) and adults based on the stable age distribution estimated from the demographic model.
 To estimate pup survival in year one, we took the fertility estimates (number of new pups per reproducing dog in the pup age class per year) and divided by an estimate of newborn pups per reproducing dog each year based on average litter size (range of 3 - 8 based on estimates from the literature, Czupryna, Annie Yang, Morters, Hampson), average number of litters per female per year (1 on average, Czupryna, Annie Yang), and the proportion of the adult population that is female (estimated from our data).
@@
 
-#### Data analysis
+#### Data and ethics statement
+All data were analysed in `r R.version$version.string` [@R2020], largely using the `tidyverse` package suite [@tidyverse].
+Geospatial data were mapped using the sf [@sf] package.
+All data and code are available at [https://github.com/mrajeev08/mora_vax](https://github.com/mrajeev08/mora_vax).
+The vaccinations were part of a public health campaign and routine veterinary service provisioning carried out by the local veterinary officals, the NGOs involved, and in partnership with the Ministry of Public Health and the Department of Veterinary Services at the national level.
+The Madagascar Dog Initiative also maintains institutional approvals (IACUC # at Institution X) and national research permits (MICET permit #) for it's research and volunteer programs.
 
@@
 ```{r}
+summ_tab1 <- read_csv("out/table1_stats.csv")
 species_2018 <- table(animals_2018$Species)
 species_2019 <- table(tolower(animals_2019$`Dog or Cat`))
+n_communes <- n_distinct(animals_2019$Commune)
+animals_2019 %>% 
+  filter(tolower(`Dog or Cat`) %in% "dog") %>%
+  mutate(Commune = ifelse(Commune %in% "Moramanga", "Moramanga", "other"), 
+         prev_vacc = ifelse(is.na(`Previous Vaccination`), "NO", 
+                            `Previous Vaccination`)) %>% 
+  group_by(prev_vacc, Commune) %>% 
+  summarize(n = n()) %>%
+  pivot_wider(values_from = n, names_from = prev_vacc) %>%
+  mutate(total = NO + YES, 
+         perc_vacc = round(YES / total * 100, 2)) -> prev_vacc_2019
+spayed_neutered <- table(animals_2019$`Spayed/Neutered`)["YES"]/nrow(animals_2019)
+
+dist_over_100 <- round(sum(animals_2019$`Distance Travelled (m)` <= 1000, na.rm = TRUE)/sum(!is.na(animals_2019$`Distance Travelled (m)`))*100, 0)
+
 ```
 
-During the 2018 campaign, a total of `nrow(animals_2018)` animals were vaccinated (`r species_2018["A"]` dogs and `r species_2018["A"]` cats) in the Moramanga (urban) and Andasibe (rural) communes over a period of xx days (Table 1).
+During the 2018 campaign, a total of `nrow(animals_2018)` animals were vaccinated (`r species_2018["A"]` dogs and `r species_2018["A"]` cats) in the Moramanga (urban) and Andasibe (rural) communes over a period of 13 days during the month of April (Table 1).
 We vaccinated at `r table(pts$Commune)["Andasibe"]` points in Andasibe and `r table(pts$Commune)["Moramanga Ville"]` in Moramanga Ville.
-On average, we vaccinated xx animals per vaccination point, but this varied significantly between the two communes, with a maximum of 354 animals vaccinated at one point.
-Overall, 10.7% of animals (15.1% of dogs and 2 % of cats) had a history of previous vaccination (owner reported), however only 4.7% had been vaccinated in the previous year (6.7% of dogs and \<1% of cats).
-While less than 20% of owners reported that animals were not free roaming, fenced animals were frequently observed outside fences during the campaign, and thus the majority of animals could be classified as semi-confined in the more urban township of Moramanga Ville, and free-roaming in the rural setting of Andasibe where almost 70% of owners reported their dogs as free-roaming.
+During the 2019 campaign, between `r paste(format.Date(range(dmy(animals_2019$Date)), "%B %Y"), collapse = " - ")`, the DV vaccinated a total of 2,385 animals (1,898 dogs, 487 cats, Table 1) over 48 days in `r n_distinct(animals_2019$Commune)` communes in the Moramanga District.
+While more animals were vaccinated per vaccination point and per vaccination day in 2018 compared to 2019, the number of animals vaccinated per person-day was much higher for the 2019 campaign (Table 1).
+More animals were vaccinated in 2018 vs.
+2019, but this was largely a result of vaccinating more cats during the 2018 campaign.
 
-During 2019/2020, the DV vaccinated a total of 2,384 animals (1,898 dogs, 486 cats, Table 1) over a period of xx months and in N communes in the Moramanga District.
-Overall, 72% of animals (75% of dogs and 20% of cats) had a history of previous vaccination (owner reported), however only 1.1% had been vaccinated in the previous year (0.6% of dogs and 2% of cats).
-218 animals were reported as having been previously vaccinated in April 2018, likely during the volunteer-led campaign.
-Additionally, 23% of all animals vaccinated were spayed or neutered (27% of dogs and 8% of cats)**.
-The majority of people estimated travelling less than xx meters to bring their animal to the DV.
+In 2018, 15% of dogs had a previous history of vaccination (largely in the urban commune of Moramanga Ville) with only 7% of dogs vaccinated within the last year.
+This remained largely the same in 2019, as the DV focused their efforts in other communes.
+The DV did vaccinate 771 dogs in Moramanga Ville in 2019, and of those 25% had been vaccinated in the previous year's campaign, wheras only 5% of dogs vaccinated in all other communes had a previous history of vaccination.
 
-*Table 1 should describe dogs!! (Take out cat parts & keep dogs only, & person days, split by location!)*
+While less than 20% of owners in Moramanga Ville reported that their animals free roaming, fenced animals were frequently observed outside fences during the campaign, and thus the majority of animals could be classified as semi-confined in the more urban township of Moramanga Ville, and free-roaming in the rural setting of Andasibe (approximately 70% of owners reported their dogs as free-roaming, Table 1).
+In 2019, the DV also asked owners how far they travelled approximately in meters, and `r dist_over_100`% of people travelled less than 1 km to reach the vaccination point.
 
+#### Table 1. Summary of 2018 and 2019 campaigns 
 ```{r}
@@
 
-The 2018 campaign cost more overall and per animal vaccinated than the 2019 campaign, largely due to increased personnel costs (Fig 3A & B, Table 1).
+The 2018 campaign cost more overall and per animal vaccinated than the 2019 campaign, largely due to increased personnel costs (Fig 1A & B, Table 1).
 Reflecting the extra personnel necessary to run the central point campaign, the 2018 campaign also took substantially more person-days per animal vaccinated (Table 1).
-We found that in most contexts, charging owners for vaccination would result in minimal cost recovery (Fig 3C), but that even a nominal fee would significantly reduce vaccination coverage acheived (Fig 3D), particulary in the rural commune of Andasibe.
+We found that in most contexts, charging owners for vaccination would result in minimal cost recovery (Fig 1C) and beyond a minimal cost would actually result in increased costs per vaccinated individual.
+More importantly, even a nominal fee would significantly reduce numbers of dogs vaccinated and thus vaccination coverage, particulary in the rural commune of Andasibe (Fig 1D).
 
-```{r this is a code chunk!} 
+```{r fig1} 
+knitr::include_graphics("figs/fig1.jpeg")
+```
+#### Figure 1. Comparing campaign costs and willigness to pay 
+*A) Vaccine costs broken down by category and by year (colors).
+B) Overall cost per animal vaccinated for the two campaign years, split by direct costs of vaccination per animal (i.e.
+vaccine, vaccination card, syringes) and baseline implementation costs (i.e.
+personnel, supplies, lodging and food during the campaign).
+C) Estimated cost per animal for each campaign year given increasing costs charged to the owner, with estimated costs declining due to cost recovery through owner payments and then peaking once owners are no longer willing to pay for the vaccine.
+D) The percent reduction in number of animals vaccinated given owners willingness to pay.
+The curves in C and D are shown based on the overall responses on willingness to pay (solid line) from both Moramanga and Andasibe, and the responses split by Commune (dashed and dotted lines).*
+
+### Comparing campaign coverage estimates
+The 2018 campaign covered two communes, and was estimated to have acheived approximately 60% coverage (Figure 2A).
+The 2019 campaign covered more communes, but likely acheived lower coverage levels at the commune level (Figure 2B).
+We also compared the coverage estimates from the transects with estimates using human-to-dog ratios (circles in Figure 2B), with the midpoint assuming HDR estimates from a recent household survey study in the Moramanga District and the range based on estimates from the literature.
+In the 2018 campaign, we used post-vaccination coverage transects to estimate vaccination coverage, but we were unable to do this in 2019 given the continuous delivery strategy.
+In addition, in Andasibe in 2018, coverage estimates were based on a single transect resulting in more uncertainty.
+However, in both cases, coverage estimated from the transects in 2018 were consistent with the HDR based estimates (for both Andasibe and Moramanga, they fell within the range of the HDR estimates).
+We also back-calculated HDRs given our vaccination coverage estimates, and these were similar to the HDRs calculated from the household survey (`r hdr_backcalcs %>% filter(Commune == "Andasibe") %$% glue("{hdr_cov_est_lower} - {hdr_cov_est_upper}")` for Andasibe compared to `r `hdrs$mora_hdrs[hdrs$loc == "rural"]` in a rural community and `r hdr_backcalcs %>% filter(Commune == "Moramanga") %$% glue("{hdr_cov_est_lower} - {hdr_cov_est_upper}")` for Moramanga Ville compared to `r `hdrs$mora_hdrs[hdrs$loc == "urban"]` in a urban community).
+
+```{r fig2}
 knitr::include_graphics("figs/fig2.jpeg")
 ```
+#### Figure 2. Spatial estimates of coverage achieved by the 2018 and 2019 campaigns.
+*A) The commune level numbers of dogs vaccinated (size of points) and the associated coverage estimates (color of points) for the year 2018 (estimated using post-vaccination transects) and 2019 (estimated using a human-to-dog ratio (HDR) of NN, based on a recent household survey in the Moramanga District).
+The inset shows the location of the Moramanga District in Madagascar.
+B) A comparison of coverage estimates by location and by method of estimation (post-vaccination surveys vs.
+HDR based estimates); for the transect based estimates, the line range shows the 95% exact binomial confidence interval while for the HDR based estimates, the line range shows the range of coverage estimates assuming an HDR range of 8 - 25 based on the literature.*
 
-### Comparing campaign coverage estimates
+### Dog demography and simulating vaccination strategies
+```{r}
+age_data <- read_csv("out/animals_age.csv")
+prop_sex <- round(table(age_data$Sex)/nrow(age_data)*100, 0)
+litter_size <- c(4, 8) * 1 * 0.4 # from literature (Morters, Czupryna, etc.)
+dem_ests <- read_csv("out/dem_ests.csv")
+q_high <- function(x) quantile(x, 0.975)
+q_low <-  function(x) quantile(x, 0.025)
+mean_round <- function(x) round(mean(x), 2)
+dem_ests %>%
+  ungroup() %>%
+  filter(pop_growth == "Growing") %>%
+  select(fert_est, psurv_adult_est, growth, sim) %>%
+  select(-sim) %>%
+  distinct() %>%
+  mutate(pup_surv_upper = fert_est/litter_size[1], 
+         pup_surv_lower = fert_est/litter_size[2]) %>%
+  pivot_longer(everything()) %>%
+  group_by(name) %>%
+  summarize(across(everything(), list(mean = mean_round, upper = q_high, lower = q_low), 
+                   .names = "{.col}.{.fn}")) %>%
+  rowwise() %>%
+  mutate(ci = paste(round(value.lower, 2), " - ", 
+                    round(value.upper, 2))) -> dem_pars
 
-In Andasibe Ville, we were only able to did this in one of our campaign locations, due to the sparsely distributed households and roads throughout the area, as such coverage in this.
-Make sure to mention back calc of pop & hdr is it reasonable?
-
-```{r this is a code chunk!}
-knitr::include_graphics("figs/fig1.jpeg")
 ```
 
-### Dog demography and simulating vaccination strategies
-For dogs, the sex ratio was male-biased (55%); while for cats it was female-biased (68% female). For dogs, the sex ratio was male-biased (65%), while for cats it was female-biased (66% female).
+Demographic data from vaccinated dogs shows a population pyramid with a large base, indicative of a fast growing population, and with a male bias (approximately `r prop_sex["M"]`% Male, Figure 3A).
+We fit these data to an age-structured model, and were able to generate parameter estimates which resulted in stable age distributions consistent with the data (Figure 3B).
+We filter to parameter estimates that are consistent with a growing population resulting in mean adult annual survival probability of `r dem_pars[dem_pars$name == "psurv_adult_est", "value.mean"]` (95% quantile: `r dem_pars[dem_pars$name == "psurv_adult_est", "ci"]`).
+We use estimates of fertility (on average `r dem_pars[dem_pars$name == "fert_est", "value.mean"]`, 95% quantile: `r dem_pars[dem_pars$name == "fert_est", "ci"]`) to back out pup survival, which ranges from `r dem_pars[dem_pars$name == "pup_surv_lower", "value.lower"]` - `r dem_pars[dem_pars$name == "pup_surv_lower", "value.upper"]`.
+We find that given these demographic parameters, annual campaigns that target dogs of all ages result in rapid decline in vaccination coverage in between campaigns, largely due to rapid turnover of the dog population (compared to the impact of waning of vaccine immunity, which you can see in the additional dip at year 3, Fig 3D).
+Continuously targeting 70% of the puppy population for vaccination, while unable to acheive the peak coverage of 70% due to puppy mortality, achieves consistent coverage of about 50% of the dog population.
+A combined strategy maintains the highest and most consistent levels of coverage close to the target of 70%.
 
-```{r this is a code chunk!}
+```{r fig3}
 knitr::include_graphics("figs/fig3.jpeg")
 ```
-
+#### Figure 3. Dog demography and implications for campaign strategies
+*A) The age pyramid for vaccinated dogs by sex.
+B) Bootstrapped estimates of the proportion of the population in each age class (dark blue) compared to estimates from the demographic models fit to these age data.
+C) Parameter estimates for annual fertility rates and adult survival probability, with estimates in orange showing the parameter estimates which result in positive population growth.
+B) Simulated vaccination coverage (N = 1000) using the demographic parameters from (C) in a hypothetical commune with 1000 dogs for three different campaign strategies: 1) annual vaccination campaigns targeting dogs of all ages (purple), 2) Routine vaccination of puppies at 3 months of age, and 3) a combined strategy with campaigns annually and continuous puppy vaccination in between campaigns.*
 ## Discussion
@@
 
-Limitations (owner reports, no solid ests of dog pop although HDR is reasonable and supported) Need to monitor incidence (Rajeev unpublished data anecdotally).
-Limitations of age data and methods.
-Limitations of cov data and methods.
-One limitation here is unowned dogs Better understanding of demography (using campaigns to monitor pop growth etc.
-costs may rise given increase in dog pops, pop control strategies may also be important for humane implementation of vaccination programs).
-Estimates were not super identifiable.
-No carrying cap.
-Turnover of the dog pop.
-Compare to LeBlanc + Rila studies Better understanding of transmission (i.e.
-Bourrhy & Chad studies).
-Cost not included was PEP for exposures during a campaign (mostly cat scratches).
+Through the 2018 and 2019 campaigns implemented in Moramanga District, we learned that mass vaccination of dogs is a feasible strategy for controlling canine rabies in Madagascar.
+There was high demand for vaccination from owners and dogs were accesible and able to be handled efficiently for parenteral vaccination.
+We found that providing the vaccine at no direct cost to owners will be critical to acheiving sufficient coverage, as even with nominal fees, a significant proportion of owners would no longer vaccinate their animals.
+While the 2018 vaccination campaigns acheived higher coverage over a shorter time period, it came at the a higher cost per animal vaccinated, was more limited in geographic scope, and was much more resource intensive in terms of personnel requirements.
+In addition, in the rural setting of Andasibe, the static point campaign strategy seemed to be less effective, reflecting less accessible human populations in this location.
+The 2019 campaign on the other hand, where the District Veterinarian provisioned vaccines through routine services, covered a much larger geographic scope and was significantly less resource intensive, but achieved lower vaccination coverage.
+Finally, we used age data collected and a vaccination model to simulate different vaccination strategies, and found that continuous vaccination targeting puppies in particular may be an effective way to maintain coverage levels, particulalry in rural areas where static point vaccination campaigns may be less effective.
+
+There were several limitations to our analyses.
+Owner reports of willingness to pay, age of animals, and distance travelled to the vaccination point likely all suffer from recall bias and uncertainty.
+However, our findings from these data are consistent with previous literature.
+We estimate that the dog population is male-biased and skewed towards puppies, but that pups likely suffer high mortatlity in the first year of life, consistent with studies from other sub-Saharan African countries.
+Other studies have also shown that implementing fees for vaccination greatly reduces the efficacy of campaigns.
+Similarly, distance to campaign points has also been identified as major barrier to vaccination, and as with other studies we found that the majority of people report traveling less than 1km to reach a vaccination point.
+Dog population estimates were also largely lacking, but we were able to ground our estimates of human-to-dog ratios from a recent household survey study in the distict and used a range that is consistent with the literature.
+*Estimates between the two methods were comparable.* We also make several simplifying assumptions in our model of vaccination strategies: we assume that immunity wanes after three years (simulating vaccination with a long-lasting vaccine such as Nobivac, compared to the 1 year efficacy estimated for the Rabisin vaccine); we assume that vaccinated dogs that survive from year to year are revaccinated in subsequent campaigns; and in our age-structured models we do not account for population carrying capacity, but rather simulate over a short time frame over which growth estimates remain reasonable.
+
+*Costs for the campaign did not include volunteer costs.
+Point to volunteer issues.*  Cost not included was PEP for exposures during a campaign (mostly cat scratches).
 Getting vaccinators prep.
-Focusing on puppies could also help that, as they are easier to handle.
-Covid?
+PrEP and exposures during vax, we had N exposures.
+Puppies were the easiest to handle for both owners and vaccinators and vax is effective in them.
+Better dosing guidelines.
+Longer lasting vax.
+*We weren't that good person per day*
 
-Targeting dogs only?
-Point to lit on willingness to pay / person day comp / cost per animal vaccinated
+One key aspect we do not consider is the feedback loops between vaccination and demography.
+Estimates of the effects of vaccination on dog demography are mixed, but vaccination could potentially increase survival of dogs.
+If dog population growth is driven by survival, then this could mean that increased vaccination results in increased population growth.
+However, if growth is driven more by demand from human communities, then vaccination could stabilize the population and reduce turnover of the population.
+Improving dog population management, encouraging responsible pet ownership practices, and increasing veterinary services could all complement vaccination efforts.
+*Overall high turnover of the dog pop makes it harder to.*
 
-PrEP and exposures during vax, we had N exposures. Puppies were the easiest to handle for both owners and vaccinators and vax is effective in them. Better dosing guidelines. Longer lasting vax.
+In these settings with high dog ownership, moving towards community-based vaccination strategies, could be a highly effective way to achieve sufficient and consistent coverage, particulalry in rural areas where static point campaigns may be less likely to reach.
+During our campaign, we found that puppies at approximately 3 months old were easy to handle for both vaccinators and owners.
+Puppy vaccination could therefore use local officials emmbedded in communities (along the lines of community health workers) to implement vaccination, particularly given recent findings on thermotolerance of rabies vaccines and locally manufactured methods for maintaing temperatures required for sustained vaccine storage (up to 3 months).
+*Covid?
+Accessibility?*
 
-More community based delivery (Ahmed's paper, Pivot, accessibility, etc.).
-Possibly combining different campaign strategies could work per models Turnover is tricky & waning doesn't contribute too much.
-Thermotolerant & 3 yrs even without booster.
-(although most dogs are revaccinated and we do this in the model).
-Volunteer based/NGO based.
-Push for govt investment and viewing as a public good.
+Paying people is important hard/relieving pressures and making sure it's viewed as a public good (Ahmed's paper). Veterinary students?
+Ultimately push for govt investment and viewing as a public good.
+The veterinary workforce (relieving pressures on them, challenges with private-public roles)..
 Modeling after childhood vax = routine through local offices of puppies + supplemental activities by trained vets annually to catch unvax (feral, oral).
-The veterinary workforce (relieving pressures on them, challenges with private-public roles).
-Paying people is important hard/relieving pressures and making sure it's viewed as a public good (Ahmed's paper). Veterinary students?
-
-Conclude with main points & how to move forward in Madagascar
 
@@
 - Dog vax as a public good, remove barriers for dog owners & incentivize vets/etc/ remove economic pressures
-- Community based vax delivery particulalry in novel areas (could lower cost while improving coverage)
+- Community based vax delivery particulalry in rural areas (could lower cost while improving coverage)
 - Easier ways to estimate coverage (& coupling with surveillance to assess impact)

### 2021-02-04 R/figure1.R

@@ -1,162 +1,161 @@
+# Figure 2. Comparing costs and willingness to pay
+
 # pkgs
-library(sf)
 library(dplyr)
 library(readr)
 library(ggplot2)
 library(tidyr)
 library(patchwork)
 library(cowplot)
-library(ggsn)
 
-# From Claire's MSC
-range_people_ph <- c(4.13, 5.654)
-range_dogs_ph <- c(0.24, 0.26)
+# data on costs
+costs <- read_csv("data/costs_summarized.csv")
+willing_to_pay <- read_csv("data/owner_2018.csv")
 
-# Range of hdrs then
-mora_hdrs <- range_people_ph / range_dogs_ph
-mora_mid_hdr <- median(mora_hdrs)
-hdr_high <- 25
-hdr_low <- 8
+# exchange rate (midpoint between 2018 and 2019 est)
+exch_rate <- (3128 + 3500)/2
 
-# function to get upper & lower conf intervals from binomial into a data.frame
-bn_ci <- function(resighted, sighted_total) {
-  out <- binom.test(resighted, sighted_total, resighted/sighted_total,
-                    alternative = "two.sided", 
-                    conf.level = 0.95)
-  return(data.frame(cov_est = resighted/sighted_total, 
-                    cov_est_lower = out$conf.int[1], 
-                    cov_est_upper = out$conf.int[2]))
+# animals
+animals_2018 <- read_csv("data/animal_2018.csv")
+animals_2019 <- read_csv("data/campaign_2019.csv")
+
+# Figure 1A costs by category
+costs %>%
+  filter(include == "Y") %>%
+  group_by(year, category) %>%
+  summarize(cost_usd = sum(usd)) %>%
+  group_by(year) %>%
+  mutate(prop = cost_usd/sum(cost_usd)) %>%
+  ungroup() %>%
+  tidyr::complete(category, year = 2018:2019, 
+                  fill = list(cost_usd = 0, prop = 0)) %>%
+  mutate(type = case_when(category == "Vaccine" ~ "Vaccine", 
+                   TRUE ~ "Implementation")) -> costs
+
+ggplot(costs) +
+  geom_col(aes(x = category, y = cost_usd, fill = factor(year)), 
+           position = position_dodge()) +
+  labs(x = "Category", y = "Approximate cost (USD)", tag = "A") +
+  scale_fill_brewer(palette = "Dark2", name = "Year") +
+  theme_minimal_hgrid(font_size = 12) +
+  coord_flip() -> fig1A
+
[…]

### 2021-02-04 R/figure2.R

@@ -1,161 +1,176 @@
-# Figure 2. Comparing costs and willingness to pay
-
 # pkgs
+library(sf)
 library(dplyr)
 library(readr)
 library(ggplot2)
 library(tidyr)
 library(patchwork)
 library(cowplot)
+library(ggsn)
 
-# data on costs
-costs <- read_csv("data/costs_summarized.csv")
-willing_to_pay <- read_csv("data/owner_2018.csv")
+# From Claire's MSC
+range_people_ph <- c(4.13, 5.654)
+range_dogs_ph <- c(0.24, 0.26)
 
-# exchange rate (midpoint between 2018 and 2019 est)
-exch_rate <- (3128 + 3500)/2
+# Range of hdrs then
+mora_hdrs <- range_people_ph / range_dogs_ph
+mora_mid_hdr <- median(mora_hdrs)
+hdr_high <- 25
+hdr_low <- 8
 
-# animals
-animals_2018 <- read_csv("data/animal_2018.csv")
-animals_2019 <- read_csv("data/campaign_2019.csv")
-
-# Figure 1A costs by category
-costs %>%
-  filter(include == "Y") %>%
-  group_by(year, category) %>%
-  summarize(cost_usd = sum(usd)) %>%
-  group_by(year) %>%
-  mutate(prop = cost_usd/sum(cost_usd)) %>%
-  ungroup() %>%
-  tidyr::complete(category, year = 2018:2019, 
-                  fill = list(cost_usd = 0, prop = 0)) %>%
-  mutate(type = case_when(category == "Vaccine" ~ "Vaccine", 
-                   TRUE ~ "Implementation")) -> costs
-
-ggplot(costs) +
-  geom_col(aes(x = category, y = cost_usd, fill = factor(year)), 
-           position = position_dodge()) +
-  labs(x = "Category", y = "Approximate cost (USD)", tag = "A") +
-  scale_fill_brewer(palette = "Dark2", name = "Year") +
-  theme_minimal_hgrid(font_size = 12) +
-  coord_flip() -> fig1A
-
-write_csv(costs, "out/costs_by_category.csv")
-
-# Figure 1B Cost per animal vaccinated
-costs %>%
-  group_by(year, type) %>%
-  summarize(cost = sum(cost_usd, na.rm = TRUE)) %>%
-  mutate(n_animals = case_when(year == 2018 ~ nrow(animals_2018), 
-                                  year == 2019 ~ nrow(animals_2019)), 
[…]

### 2021-02-04 manuscript.Rmd
adding author list & acknowledgements

@@
 title: "Lessons learned and paths forward for mass dog vaccination in Madagascar: a case study of pilot canine rabies vaccination campaigns in Moramanga District"
-author: "Malavika Rajeev"
-date: "11/19/2020"
+author:
+  - Caitlynn Filla:
+      insitute: [florida]
+      equal_contributor: "yes"
+  - Malavika Rajeev:
+      email: mrajeev@princeton.edu
+      institute: [princeton]
+      correspondence: true
+      equal_contributor: "yes"
+  - Zoavina Randriana:
+      institute: [maddog]
+  - Chantal Hanitriniana:
+      institute: [uoa]
+  - Radoniaina R. Rafaliarison:
+      institute: [maddog]
+  - ?Glenn Edosoa:
+      institute: [oms]
+  - ?Mamitiana Andriamananjara:
+      institute: [dsv]
+  - Fenomanana Daniel:
+      institute: [maddog]
+  - Tara A. Clarke:
+      institute: [maddog]
+  - Zachary Farris:
+      institute: [maddog, appstate]
+  - Terry Stone:
+      institute: [tad]
+  - Jochem Lastdrager:
+      institute: [tad]
+  - Tsiky Rajaonarivelo:
+      institute: [maddog]
+  - Katie Hampson:
+      institute: [glasgow]
+  - C. Jessica E. Metcalf:      
+      institute: [princeton]
+  - Kim Valenta:
+      institute: [maddog, florida]
+institute:
+  - princeton: Department of Ecology and Evolutionary Biology, Princeton University, Princeton, United States
+  - oms: Chargé des Maladies Tropicales Négligées, Organisation mondiale de la Santé Madagascar, Antananarivo, Madagascar
+  - florida: University of Florida, Department of Anthropology, Gainesville, Florida, USA
+  - maddog: The Mad Dog Initiative, Akanin’ny Veterinera Akaikiniarivo, Ambatobe, Antananarivo 101, Madagascar
+  - tad: Travelling Animal Doctors, 919 Crossan Rd, Newark, Delaware, USA 
+  - uoa: Mention Zoologie et Biodiversité Animale, Faculté des Sciences, Université d’Antananarivo, Antananarivo, Madagascar
+  - glasgow: Boyd Orr Centre for Population and Ecosystem Health, Institute of Biodiversity, Animal Health and Comparative Medicine, University of Glasgow, Glasgow, UK
+  - appstate: Appalachian State University, Department of Health and Exercise Science, Boone, North Carolina, USA
+  - dsv: Direction des Services Vétérinaires, Ministère chargé de l’Agriculture et de l’Élevage, Antananarivo, Madagascar
 output:
-  word_document: default
+  bookdown::word_document2:
+    pandoc_args:
+      - --lua-filter=format/scholarly-metadata.lua
+      - --lua-filter=format/author-info-blocks.lua
+      - --lua-filter=format/pagebreak.lua
   html_document: default
@@
 
+Canine rabies causes an estimated 60,000 human deaths per year, but these deaths are preventable through post-exposure prophylaxis of people and vaccination of domestic dogs.
+Dog vaccination campaigns acheiving coverage of 70% at the population level are effective at interrupting transmission and eventually can lead to elimination of the disease regionally.
+Given this evidence, the WHO and it's partners have set a global goal to eliminate deaths due to canine rabies by the year 2030.
+Here, we report on initial lessons learned during pilot dog vaccination campaigns in the Moramanga District of Madagascar, a country with a high estimated burden of rabies and limited control in the dog population.
+We compare two different vaccination strategies: a larger volunteer driven effort to vaccinate dogs in two communes using static point vaccination, and continuous vaccination as part of routine veterinary services in the district by the District Veterinarian.
+We compared costs, time, personnel, and coverage acheived by each of the campaigns.
+We also use age data from the campaigns to estimate key demographic parameters, and use these estimates to simulate different vaccination strategies.
+Overall we found that dog vaccination was feasible and that most dogs were accessible to vaccination.
+We find that there are tradeoffs between the vaccination strategies, with the first campaign acheiving higher coverage, but with more resources and with more limited geographic coverage compared to the second campaign with continuous delivery.
+Our modeling results suggest that   that targeting puppies through community based vaccination efforts could hypothetically acheive coverage levels sufficient to interrupt transmission.
+Moving forward, our results show that mass dog vaccination is feasible and can acheive high coverage in Madagascar, however innovative strategies and an investment in dog vaccination as a public good will be required to maintain the high and sustained levels of coverage needed for disease elimination.
+
 ## Introduction
@@
 These deaths are entirely preventable: prompt post-exposure prophylaxis (PEP) of humans is highly effective at preventing death and mass dog vaccination can interrupt transmission in domestic dogs and eventually lead to disease elimination.
-Annual vaccination campaigns achieving atleast 70% coverage are the recommended target for controlling rabies in domestic dog populations.
+The WHO and its partners have set a goal to eliminate human deaths due to canine rabies by the year 2030 ('ZeroBy30').
+Annual vaccination campaigns achieving atleast 70% coverage are the recommended target for controlling rabies in the domestic dog populations.
 However, achieving this coverage target in low and middle income countries where the virus is endemic can be challenging due to economic, ecological, sociocultural, and political reasons.
@@
 It comprises 21 communes and NN fokontany (sub-commune), covering approximately 7150 km^2^ with a human population of x (2018 census prelim data).
+Urban center of Moramanga (cite Rila's paper).
 Previous work in the district has established a high burden of rabies exposures and deaths despite the availability of post-exposure prophylaxis at the district hospital.
@@
 The vaccinations were part of a public health campaign and routine veterinary service provisioning carried out by the local veterinary officals, the NGOs involved, and in partnership with the Ministry of Public Health and the Department of Veterinary Services at the national level.
-The Madagascar Dog Initiative also maintains institutional approvals (IACUC # at Institution X) and national research permits (MICET permit #) for it's research and volunteer programs.
+The Madagascar Dog Initiative also maintains institutional approvals (IACUC # at Institution X) and national research permits (MICET permit #) for it's research and volunteer programs [Kim to fill in].
 
@@
 
-While less than 20% of owners in Moramanga Ville reported that their animals free roaming, fenced animals were frequently observed outside fences during the campaign, and thus the majority of animals could be classified as semi-confined in the more urban township of Moramanga Ville, and free-roaming in the rural setting of Andasibe (approximately 70% of owners reported their dogs as free-roaming, Table 1).
+While less than 20% of owners in Moramanga Ville reported that their animals were free roaming, fenced animals were frequently observed outside fences during the campaign, and thus the majority of animals could be classified as semi-confined in the more urban township of Moramanga Ville, and free-roaming in the rural setting of Andasibe (approximately 70% of owners reported their dogs as free-roaming, Table 1).
 In 2019, the DV also asked owners how far they travelled approximately in meters, and `r dist_over_100`% of people travelled less than 1 km to reach the vaccination point.
@@
 B) Simulated vaccination coverage (N = 1000) using the demographic parameters from (C) in a hypothetical commune with 1000 dogs for three different campaign strategies: 1) annual vaccination campaigns targeting dogs of all ages (purple), 2) Routine vaccination of puppies at 3 months of age, and 3) a combined strategy with campaigns annually and continuous puppy vaccination in between campaigns.*
+
 ## Discussion
-
-Main findings para (feasible, should not charge, costs vs. coverage).
 
@@
 Similarly, distance to campaign points has also been identified as major barrier to vaccination, and as with other studies we found that the majority of people report traveling less than 1km to reach a vaccination point.
-Dog population estimates were also largely lacking, but we were able to ground our estimates of human-to-dog ratios from a recent household survey study in the distict and used a range that is consistent with the literature.
-*Estimates between the two methods were comparable.* We also make several simplifying assumptions in our model of vaccination strategies: we assume that immunity wanes after three years (simulating vaccination with a long-lasting vaccine such as Nobivac, compared to the 1 year efficacy estimated for the Rabisin vaccine); we assume that vaccinated dogs that survive from year to year are revaccinated in subsequent campaigns; and in our age-structured models we do not account for population carrying capacity, but rather simulate over a short time frame over which growth estimates remain reasonable.
+Dog population estimates were also largely lacking, but we were able to ground our estimates of human-to-dog ratios from a recent household survey study in the distict and used a range that is consistent with the literature, and coverage estimates using HDRs were comparable to estimates from post-vaccination transects.
+We also make several simplifying assumptions in our model of vaccination strategies: we assume that immunity wanes after three years (simulating vaccination with a long-lasting vaccine such as Nobivac, compared to the 1 year efficacy estimated for the Rabisin vaccine); we assume that vaccinated dogs that survive from year to year are revaccinated in subsequent campaigns; and in our age-structured models we do not account for population carrying capacity, but rather simulate over a short time frame over which growth estimates remain reasonable.
 
-*Costs for the campaign did not include volunteer costs.
-Point to volunteer issues.*  Cost not included was PEP for exposures during a campaign (mostly cat scratches).
-Getting vaccinators prep.
-PrEP and exposures during vax, we had N exposures.
-Puppies were the easiest to handle for both owners and vaccinators and vax is effective in them.
-Better dosing guidelines.
-Longer lasting vax.
-*We weren't that good person per day*
+Our estimates of costs per animal vaccinated are in line with recent estimates from other settings, but on the lower end.
+In our estimates, we excluded costs associated with international volunteers (i.e.
+airfare and visa costs) as well as costs of transportation, as these were donated to the campaign.
+We also did not include estimates of pre-exposure prophylaxis as all of our volunteers and vaccinators had received the course prior to the campaigns.
+In a rabies endemic situation, it will be necessary to have both PrEP and PEP readily available for vaccinators.
+While the volunteer-led effort resulted in significant financial and personnel resources being devoted to the campaign, costs were higher per animal and overall.
+In addition, it can be difficult for NGO and volunteer based efforts to sustain these types of efforts given unpredictable funding, time commitments, and turnover in staff.
+The routine vaccination strategy through the DV was less costly, with the majority of the costs directly related to vaccination.
+Transport to locations was not included in these estimates either, as the DV had their own vehicle and largely delivered the vaccinations as part of her routine services.
 
@@
 Improving dog population management, encouraging responsible pet ownership practices, and increasing veterinary services could all complement vaccination efforts.
-*Overall high turnover of the dog pop makes it harder to.*
+Overall, high population turnover of the dog population erodes coverage, therefore strategies to stabilize the population would be beneficial to improving vaccination coverage.
 
@@
 Puppy vaccination could therefore use local officials emmbedded in communities (along the lines of community health workers) to implement vaccination, particularly given recent findings on thermotolerance of rabies vaccines and locally manufactured methods for maintaing temperatures required for sustained vaccine storage (up to 3 months).
-*Covid?
-Accessibility?*
+Incentivizing vaccinators appropriately will be a key challenge, as currently providing no-cost rabies vaccination is not seen as part of routine duties for district veterinarians or for livestock officers, and these officials often required additional per diems.
+Expanding veterinary services across the country and releiving financial pressures on veterinarians and animal health workers could greatly mitigate these issues in Madagascar.
+For volunteer based efforts, veterinary students could be a local volunteer pool, however similarly to international volunteers, they require food, transport, and lodging during vaccination campaigns when not based in the communities where they study or live, adding to overall costs.
 
-Paying people is important hard/relieving pressures and making sure it's viewed as a public good (Ahmed's paper). Veterinary students?
-Ultimately push for govt investment and viewing as a public good.
-The veterinary workforce (relieving pressures on them, challenges with private-public roles)..
-Modeling after childhood vax = routine through local offices of puppies + supplemental activities by trained vets annually to catch unvax (feral, oral).
-
-Policy recommendations:
-- Dog vax as a public good, remove barriers for dog owners & incentivize vets/etc/ remove economic pressures
-- Community based vax delivery particulalry in rural areas (could lower cost while improving coverage)
-- Easier ways to estimate coverage (& coupling with surveillance to assess impact)
+Overall, our results suggest that dog vaccination is a feasible and feasible strategy for controlling canine rabies.
+It will be key to push rabies vaccination as a public good and to remove barriers for dog owners as well as to incentivize veterinarians and other animal health workers to implement vaccination.
+Borrowing strategies from human vaccination efforts, i.e.
+childhood vaccination and using community health workers to provide routine services, could be a way to improve vaccination coverage and reduce costs, particularly in rural communities.
+To monitor the success of these campaigns, it will be critical to develop efficient and effective methods to estimate vaccination coverage and the impact of campaigns on rabies incidence through robust surveillance systems.
+Given it's island context, Madagascar could be a model for rabies elimination, with limited chances for reintroduction from outside the island, and strategies that tackle the context specific challenges could be a path forward for the country to reach ZeroBy30.
 
 ## Acknowledgements
+We thank all the veterinarians, livestock officers, local officials and leaders, NGO staff, and volunteers involved in each of the campaigns.
+We are grateful to staff and officials at the Department of Veterinary Services, the Ministry of Public Health, and the Institut Pasteur de Madagascar for their support and technical assistance.
+In particular, we thank Angelique Ferreira, Jean Hyacinthe Randrianarisoa, Ranaivoarimanana, Fierenantsoa Randriamahatana, Esther Noiarisaona, Cara Brook, and Christian Ranaivoson, Rila Ratovoson, and Claire LeBlanc.
+The campaign costs were supported by donations from Lush Cosmetics and [Kim to insert].
+Parts of this work were funded by grants from the Center for Health and Wellbeing and the Department of Ecology and Evolutionary Biology at Princeton University to MR.
+MR was supported by an NSF Graduate Research Fellowship and a Princeton Insitute for Regional and International Studies Graduate Fellowship.
+KH is supported by the Wellcome Trust (207569/Z/17/Z).
 
-We thank vets livestock officers volunteers fokontany leaders government officials.
-We want to especially acknowledge the important contributions made by Annie Yang, an undergraduate researcher and volunteer with Princeton University and the Madagascar Dog Initiative who tragically passed away in a car accident before this research was published.
+We want to particularly acknowledge the contributions to this work by Annie Yang, an undergraduate researcher and volunteer with Princeton University and the Madagascar Dog Initiative who tragically passed away in a car accident before this research was published.

### 2021-02-05 manuscript.Rmd
updating draft

@@
 ---
-title: "Lessons learned and paths forward for mass dog vaccination in Madagascar: a case study of pilot canine rabies vaccination campaigns in Moramanga District"
+title: "Lessons learned and paths forward for rabies dog vaccination in Madagascar: a case study of pilot vaccination campaigns in Moramanga District"
 author:
@@
 
+<!-- Fix citations without parentheses -->
+<!-- Fix my paper should be 2019 -->
+<!-- Spelling -->
+<!-- Inline stats fixes -->
+
 ```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, 
+                      message=FALSE, out.width="80%")
 library(tidyverse)
@@
 library(lubridate)
+library(sf)
+source("R/dem_models.R")
 
@@
 
+# pop ests
+mora_sf <- st_read("data/shapefile/mora_communes.shp")
+exch_rate <- (3128 + 3500)/2
+
 ```
 
-## Abstract
+## Abstract [comment: Journal reccs]
 
 Canine rabies causes an estimated 60,000 human deaths per year, but these deaths are preventable through post-exposure prophylaxis of people and vaccination of domestic dogs.
-Dog vaccination campaigns acheiving coverage of 70% at the population level are effective at interrupting transmission and eventually can lead to elimination of the disease regionally.
+Dog vaccination campaigns acheiving coverage of 70% at the population level are effective at interrupting transmission and eventually can lead to elimination of the disease.
 Given this evidence, the WHO and it's partners have set a global goal to eliminate deaths due to canine rabies by the year 2030.
@@
 Overall we found that dog vaccination was feasible and that most dogs were accessible to vaccination.
-We find that there are tradeoffs between the vaccination strategies, with the first campaign acheiving higher coverage, but with more resources and with more limited geographic coverage compared to the second campaign with continuous delivery.
-Our modeling results suggest that   that targeting puppies through community based vaccination efforts could hypothetically acheive coverage levels sufficient to interrupt transmission.
+We find that there are tradeoffs between the vaccination strategies, with the first campaign acheiving higher coverage, but with more resources and with more limited geographic scope compared to the second campaign with continuous delivery.
+Our modeling results suggest that targeting puppies through community based vaccination efforts could help acheive coverage levels sufficient to interrupt transmission.
 Moving forward, our results show that mass dog vaccination is feasible and can acheive high coverage in Madagascar, however innovative strategies and an investment in dog vaccination as a public good will be required to maintain the high and sustained levels of coverage needed for disease elimination.
@@
 
-Canine rabies results in an estimated 60,000 human deaths per year globally.
-These deaths are entirely preventable: prompt post-exposure prophylaxis (PEP) of humans is highly effective at preventing death and mass dog vaccination can interrupt transmission in domestic dogs and eventually lead to disease elimination.
-The WHO and its partners have set a goal to eliminate human deaths due to canine rabies by the year 2030 ('ZeroBy30').
-Annual vaccination campaigns achieving atleast 70% coverage are the recommended target for controlling rabies in the domestic dog populations.
-However, achieving this coverage target in low and middle income countries where the virus is endemic can be challenging due to economic, ecological, sociocultural, and political reasons.
+Canine rabies results in an estimated 60,000 human deaths per year globally [@hampson2015].
+These deaths are entirely preventable: prompt post-exposure prophylaxis (PEP) of humans is highly effective at preventing death and mass dog vaccination can interrupt transmission in domestic dogs and eventually lead to disease elimination [@fooks2014].
+The WHO and its partners have set a goal to eliminate human deaths due to canine rabies by the year 2030 ('ZeroBy30')[@abela-ridder2016].
+Annual vaccination campaigns achieving atleast 70% coverage are the recommended target for controlling rabies in the domestic dog population [@worldhealthorganization2013].
+However, achieving this coverage target in low and middle income countries where the burden of human rabies is concentrated can be challenging due to economic, ecological, sociocultural, and political reasons [@fahrion2017].
 
-In sub-Saharan African countries, parenteral vaccination implemented through static point vaccination campaigns have been shown to be cost-effective and feasible.
-While most dogs in these settings are considered free-roaming, they are owned and are accessible for vaccination through these campaigns.
-However, reaching and maintaining high coverage vaccination campaigns requires large scale coordinated efforts, and the challenges in implementation reflect both financial and logistical constraints more than feasibility of vaccination itself.
+In sub-Saharan African countries, parenteral vaccination implemented through static point vaccination campaigns have been shown to be cost-effective and feasible [@borse2018].
+While most dogs in these settings are considered free-roaming, they are owned and are accessible for vaccination through these campaigns [@morters2014, @jibat2015].
+However, reaching and maintaining high coverage vaccination campaigns requires large scale coordinated efforts, and the challenges in implementation reflect both financial and logistical constraints more than feasibility of vaccination itself [@fahrion2017].
 
-In Madagascar, canine rabies has been endemic for over a century and the Institut Pasteur de Madagascar has provided PEP free-of-charge to patients in the country shortly therafter.
-However, currently, there are only 31 clinics where the vaccine is available, and there is limited dog vaccination due to high costs to owners and lack of availability of the vaccine.
-Recent studies have estimated a high burden of human deaths, masked by limited surveillance for canine rabies cases and human deaths across the country.
+In Madagascar, canine rabies has been endemic for over a century and the Institut Pasteur de Madagascar has provided PEP free-of-charge to patients in the country shortly therafter [@reynes2011].
+However, currently, there are only 31 clinics where the vaccine is available, and there is limited dog vaccination due to high costs to owners and lack of availability of the vaccine [@rajeev2018a].
+Recent studies have estimated a high burden of human deaths, masked by limited surveillance for canine rabies cases and human deaths across the country [@rajeev2020; @rajeev2018a].
 The veterinary sector is largely private, with some veterinarians employed in hybrid private/public employment as a designated district veterinarian.
-There are currently xx district veterinarians situated in xx out of 114 districts in the countries.
-While dog vaccination is rare, livestock officers and veterinarians work together to implement vaccination campaigns for anthrax on an annual basis as mandated by the government (wners are charged a fee per animal for these vaccines which vary by location).
-To date, no regular mass dog vaccinations are conducted on the island, although a few pilot programs have begun in recent years, largely implemented by NGO-goverment partnerships.
+While dog vaccination is rare, livestock officers and veterinarians work together to implement vaccination campaigns for anthrax on an annual basis as mandated by the government (wners are charged a fee per animal for these vaccines which vary by location) [@mondiale2017evaluation].
+To date, no regular mass dog vaccinations are conducted on the island, although a few pilot programs have begun in recent years, largely implemented by NGO-government partnerships.
 
@@
 In 2018, we led a larger scale volunteer-led pilot vaccination campaign in two Communes (sub-district level) using a central point strategy.
-In 2019/2020, we provided vaccinations and a per-vaccine fee to the District Veterinarian (DV) to vaccinate animals as part of routine services.
+In 2019/2020, we provided vaccines, all necessary supplies for vaccine administration, and a per-vaccine fee to the District Veterinarian (DV) to vaccinate animals as part of routine services.
 We compare time, human resources, costs, and coverage estimates between the campaigns.
@@
 The Moramanga District is located midway between the central highlands and the east coast of Madagascar, at an average altitude of 936 m.
-It comprises 21 communes and NN fokontany (sub-commune), covering approximately 7150 km^2^ with a human population of x (2018 census prelim data).
-Urban center of Moramanga (cite Rila's paper).
-Previous work in the district has established a high burden of rabies exposures and deaths despite the availability of post-exposure prophylaxis at the district hospital.
-While Moramanga is relatively close to the capital city of Antananarivo (~ 3 hrs by bus), within the district, travel times between locations are highly variable, with much of the population living in more rural areas with limited access to roads and transportation [cite my other paper].
-Before 2018, there were limited animal rabies vaccination services, with most vaccines available in urban communes and owners charged > 15,000 Ar (~ x USD) per vaccine administered.
+It comprises 21 communes, covering approximately 7150 km^2^ with an approximate human population of `r round(sum(mora_sf$pop), -2)`.
+Previous work in the district has established a high burden of rabies exposures and deaths despite the availability of post-exposure prophylaxis at the district hospital [@rajeev2018a].
+While Moramanga is relatively close to the capital city of Antananarivo (~ 3 hrs by bus), within the district, travel times between locations are highly variable, with much of the population living in more rural areas with limited access to roads and transportation [@rajeev2020].
+Before 2018, there were limited animal rabies vaccination services, with most vaccines available in urban communes and owners charged > 15,000 Ariary (~ 4.28 USD) per vaccine administered.
 
@@
 In 2018, we organized a pilot vaccination campaign through a collaboration with two NGOs (the Madagascar Dog Initiative and Traveling Animal Doctors), the Department of Veterinary Services, and the Ministry of Public Health in the District of Moramanga.
-We focused on two communes in Moramanga, Moramanga Ville (the urban center of the district) and Andasibe (a rural commune surrounding Andasibe National Park), where previously high incidence of suspected rabid exposures (Moramanga Ville) and a high burden of deaths (Andasibe) had been recorded [@rajeev2018].
+We focused on two communes in Moramanga, Moramanga Ville and Andasibe (a rural commune surrounding Andasibe National Park), where previously high incidence of suspected rabid exposures (Moramanga Ville) and a high burden of deaths (Andasibe) had been recorded [@rajeev2018].
 
@@
 We surveyed owners about how many dogs and cats they owned in total (split by > 1 yr vs.
-< 1 yr in order to avoid language ambiguities that might result in excluding pups and kttens), as well as if their dogs were free roaming (no restrictions on movement by the owner or 'mireny' in Malagasy), or tied ('mifatora') or fenced ('fefy').
+< 1 yr in order to avoid language ambiguities that might result in excluding pups and kttens), as well as if their dogs were free roaming (no restrictions on movement by the owner/'mireny', tied/'mifatora', or fenced/'fefy').
 Vaccinations were delivered at no cost to owners, but as animal vaccination is generally thought of as a paid service in Madagascar, we asked owners how much in Ariary they would be willing to pay to have one animal vaccinated for rabies.
@@
 
-To assess coverage, we used post-vaccination coverage surveys[@sambo2017; @gibson2015].
-We marked all animals vaccinated with a colored, non-toxic, livestock crayon [ask Jochem for brand] along the top or back of the head.
-Between 4 - 6 PM on the same day as the campaign in each location, we did two transects in pairs (two people per team), walking for 1 hr starting in opposite directions and accompanied by a local guide to ensure that walking paths did not overlap.
+To assess coverage, we used post-vaccination coverage surveys according to previously established methodology[@sambo2017; @gibson2015].
+We marked all animals vaccinated with a colored, non-toxic, livestock crayon [comment: ask Jochem for brand] along the top or back of the head.
+Between 4 - 6 PM when dogs are most active [@sambo2017] and on the same day as the campaign in each location, we did two transects in pairs (two people per team), walking for 1 hr starting in opposite directions and accompanied by a local guide to ensure that walking paths did not overlap.
 We recorded any marked and unmarked dogs we observed, and also recorded their roaming status (whether roaming, inside a fence, or tied), and their approximate age (greater or less than 1 yr of age).
@@
 
-For the 2019 campaign, instead of a central point campaign strategy, we distributed vaccinations (Rabisin `.r`, N vials) and the supplies needed to administer them (needle, syringe, vaccination card for owners) to the District Veterinarian (DV).
+For the 2019 campaign, instead of a central point campaign strategy, we distributed vaccine vials (Rabisin `.r`) and the supplies needed to administer them (needle, syringe, vaccination card for owners) to the District Veterinarian (DV).
 The DV delivered the vaccination at no cost to owners, but was directly compensated 1,5000 Ar (~ 0.40 USD) per rabies vaccine administered.
-The campaign lasted from September 6, 2019 to June 19, 2020, with the DV advertising the vaccines by word of mouth for one week prior to her visiting each location [it would be good to know if she visited these places as part of her routine work or whether she went for the purposes of the dog vaccination].
-For each animal vaccinated, we collected the age and sex, and the DV also asked owners to approximate the distance they travelled to receive the vaccination in meters.
+The campaign lasted from September 6, 2019 to June 19, 2020, with the DV advertising the vaccines by word of mouth for one week prior to her visiting each location.
+For each animal vaccinated, the DV collected the age, sex, and also asked owners to approximate the distance they travelled to receive the vaccination in meters.
 Researchers communicated with the DV about progress periodically throughout the campaign, primarily through telephone calls.
@@
 We tracked the number of vaccination points, the number of days over which these vaccinations occurred, and the number of person days required (i.e.
-the number of working people per day over the campaign [@mazeri2021]).
+the number of working people per day over the campaign [@mazeri2021]), in addition to monetary costs.
+As costs were incurred in both USD and Ariary, and the exhange rate declined rapidly between 2018 and 2019, we used the midpoint between the two years (`r exch_rate` Ar to 1 USD) to do the cost comparisons.
 
-For the 2018 campaign, we broke costs down into the following categories: direct vaccine costs (cost for Rabisin `.r`, syringes, needles, vaccination cards), supplies (livestock crayons, muzzles, gloves, alcohol, swabs), transportation costs (gas and associated vehicle costs for transport), food and lodging for NGO personnel and other vaccinators during the campaign, personnel costs (per diems for DSV veterinarian, livestock field officers, local guides, and NGO employees), and advertisement (posters and banners for advertising the campaign).
+For the 2018 campaign, we broke costs down into the following categories: direct vaccine costs (cost for vaccine `.r`, syringes, needles, vaccination cards), supplies (livestock crayons, muzzles, gloves, alcohol, swabs), food and lodging for NGO personnel and other vaccinators during the campaign, personnel costs (per diems for DSV veterinarian, livestock field officers, local guides, and NGO employees), and advertisement (posters and banners for advertising the campaign).
 Foreign NGO volunteers expenses for travel to Madagascar were not included in this costs.
@@
 In 2019, costs were split into two categories, direct vaccine costs (for same items as in 2018), and personnel costs (per vaccine fee paid to the district veterinarian), and supplies (a generator and fuel for the veterinarian to maintain the vaccine on a cold chain when there were power outages).
+Transportation costs were also not included as the DV used her own vehicle and vaccinated alongside their routine veterinary practice.
 
@@
 
-For the 2018 campaign, we used the transect data to estimate coverage as the proportion of dogs sighted that were marked with a binomial confidence interval for each location and at the commune level.
-For the 2019 campaign, we used a range of human:dog ratios (HDR) of 8 - 25, based on previous data from Madagascar [cite that tana study] and from the Moramanga District [cite leblanc thesis] to estimate the dog population from human population estimates [world pop] in each commune where the vaccinations were delivered [cite other papers that do this].
-Then, coverage was estimated as the number of dogs vaccinated in total in that commune divided by the estimated dog population (the range given high and low HDRs).
+For the 2018 campaign, we used the transect data to estimate coverage as the proportion of dogs sighted that were marked with a binomial confidence interval at the commune level.
+For the 2019 campaign, we estimated vaccination coverage using human-to-dog ratios and human population estimates [@cleaveland2003, @athingo2020].
+We used a range of 8 - 25, based on previous data from Madagascar [@ratsitorahina2009] and based on recent estimates from household surveys in the Moramanga District [@leblancclaireRabiesMadagascarThreepronged2019].
+We set the point estimate using an HDR of `r round(mora_mid_hdr, 1)`, the midpoint between the HDRs estimated for two communities in the District by LeBlanc et al.
+2019.
+We used human population estimates from WorldPop [@linard2012] in each commune where the vaccinations were delivered.
+Coverage was estimated as the number of dogs vaccinated in total in that commune divided by the estimated dog population.
 We used this same method for the 2018 campaign, as well, to compare coverage estimated by the post-vaccination transects vs.
-the HDR range.
+HDRs.
 
@@
 
-Using the age data on vaccinated animals collected during both vaccination campaigns, we estimated the proportion of the population in four age classes: pups under the age of 1 year, juveniles 1 - 2 yrs, adults age 2 - 6 years, and older dogs age 6+ years.
+Using the age data on vaccinated animals collected during both vaccination campaigns, we estimated the proportion of the population in four age classes: pups under the age of 1 year, juveniles 1 - 2 yrs, adults age 2 - 6 years, and older dogs age 6+ years based on .
 With the assumption that these estimates represent the population at a stable age distribution, we use a Leslie matrix model to estimate annual adult survival probability and fertility using maximum likelihood estimation [@Fujiwara2017].
-Specifically, we assume that the number of individuals in each age class follows a Poisson distribution with the mean predicted by the stable age distribution by the model (the proportion of individuals in each age class at equilibrium, equal to the eigenvector associated with the dominant eigenvalue of the matrix $\nu$) multiplied by the total number of individuals in the population ($N_t$):
+Specifically, we assume that the number of individuals in each age class follows a Poisson distribution with the mean predicted by the stable age distribution from the model (the proportion of individuals in each age class at equilibrium, equal to the eigenvector associated with the dominant eigenvalue of the matrix $\nu$) multiplied by the total number of individuals in the population ($N_t$):
 
@@
 
-We assume that all individuals older than 1 reproduce and that fertility does not substantially decline (cite).
-To get bootstrapped estimates, we used 100 sub-sampled data sets of 1000 observations from the observed age data to fit the parameters, and also varied initital values used in the optimization (N = 100 initial values sets) for 10000 parameter estimates total.
+We assume that all individuals older than 1 reproduce and do not estimate declines in fertility given the small proportion of dogs older than age six in the population.
+To get bootstrapped estimates, we used 100 sub-sampled data sets of 1000 observations each from the observed age data to fit the parameters, and also varied initial values used in the optimization (N = 100 initial values sets) for 10000 parameter estimates total.
 
@@
 We split the dog population into pups (< 1 yr old) and adults based on the stable age distribution estimated from the demographic model.
-To estimate pup survival in year one, we took the fertility estimates (number of new pups per reproducing dog in the pup age class per year) and divided by an estimate of newborn pups per reproducing dog each year based on average litter size (range of 3 - 8 based on estimates from the literature, Czupryna, Annie Yang, Morters, Hampson), average number of litters per female per year (1 on average, Czupryna, Annie Yang), and the proportion of the adult population that is female (estimated from our data).
-We assumed that for the annual campaign strategy, surviving vaccinated adult dogs were revaccinated in subsequent years, but that if a pup had been vaccinated within 9 months of the campaign, it was not revaccinated (cite?).
-A subset of parameter estimates resulted in estimates of population decline, but based on the shape of the age pyramid, the growing human population in the region, and to simulate reasonable campaign scenarios, we filtered to parameter estimates which corresponded to population growth.
+To estimate pup survival in year one, we took the fertility estimates (number of new pups per reproducing dog in the pup age class per year) and divided by an estimate of newborn pups per reproducing dog each year based on average litter size, average number of litters per female per year [@hampson2009, @morters2014, @czupryna2016], and the proportion of the adult population that is female (estimated from our data).
+We assumed that for the annual campaign strategy, surviving vaccinated adult dogs were revaccinated in subsequent years [@delriovilas2017], but that if a pup had been vaccinated within 9 months of the campaign, it was not revaccinated.
+A subset of parameter estimates resulted in estimates of population decline, but based on the shape of the age pyramid and to simulate reasonable campaign scenarios, we filtered to parameter estimates which corresponded to positive population growth.
 
 #### Data and ethics statement
-All data were analysed in `r R.version$version.string` [@R2020], largely using the `tidyverse` package suite [@tidyverse].
+All data were analysed in `r R.version$version.string` [@R2020], largely using the `tidyverse` package suite [@wickham2019].
 Geospatial data were mapped using the sf [@sf] package.
@@
 The vaccinations were part of a public health campaign and routine veterinary service provisioning carried out by the local veterinary officals, the NGOs involved, and in partnership with the Ministry of Public Health and the Department of Veterinary Services at the national level.
-The Madagascar Dog Initiative also maintains institutional approvals (IACUC # at Institution X) and national research permits (MICET permit #) for it's research and volunteer programs [Kim to fill in].
+The Madagascar Dog Initiative also maintains institutional approvals (IACUC # at Institution X) and national research permits (MICET permit #) for it's research and volunteer programs [comment: Kim to fill in].
 
@@
   mutate(total = NO + YES, 
-         perc_vacc = round(YES / total * 100, 2)) -> prev_vacc_2019
+         perc_vacc = round(YES / total * 100, 1)) -> prev_vacc_2019
 spayed_neutered <- table(animals_2019$`Spayed/Neutered`)["YES"]/nrow(animals_2019)
@@
 
-During the 2018 campaign, a total of `nrow(animals_2018)` animals were vaccinated (`r species_2018["A"]` dogs and `r species_2018["A"]` cats) in the Moramanga (urban) and Andasibe (rural) communes over a period of 13 days during the month of April (Table 1).
-We vaccinated at `r table(pts$Commune)["Andasibe"]` points in Andasibe and `r table(pts$Commune)["Moramanga Ville"]` in Moramanga Ville.
-During the 2019 campaign, between `r paste(format.Date(range(dmy(animals_2019$Date)), "%B %Y"), collapse = " - ")`, the DV vaccinated a total of 2,385 animals (1,898 dogs, 487 cats, Table 1) over 48 days in `r n_distinct(animals_2019$Commune)` communes in the Moramanga District.
-While more animals were vaccinated per vaccination point and per vaccination day in 2018 compared to 2019, the number of animals vaccinated per person-day was much higher for the 2019 campaign (Table 1).
+During the 2018 campaign, a total of `r nrow(animals_2018)` animals were vaccinated (`r species_2018["A"]` dogs and `r species_2018["S"]` cats) in the Moramanga (urban) and Andasibe (rural) communes over 13 days during the month of April (Table 1).
+We vaccinated at `r table(pts$Commune)["Andasibe"]` points in Andasibe and `r table(pts$Commune)["Moramanga Ville"]` points in Moramanga Ville.
+During the 2019 campaign, between `r paste(format.Date(range(dmy(animals_2019$Date)), "%B %Y"), collapse = " - ")`, the DV vaccinated a total of `r nrow(animals_2019)`animals (`r species_2019["dog"]` dogs and `r species_2019["cat"]`) over 48 days in `r n_distinct(animals_2019$Commune)` communes in the Moramanga District.
+While more animals were vaccinated per vaccination point and per vaccination day in 2018 compared to 2019, the number of animals vaccinated per person-day was much higher for the 2019 campaign.
 More animals were vaccinated in 2018 vs.
-2019, but this was largely a result of vaccinating more cats during the 2018 campaign.
+2019, but this was largely a result of vaccinating more cats during the 2018 campaign (Table 1).
 
-In 2018, 15% of dogs had a previous history of vaccination (largely in the urban commune of Moramanga Ville) with only 7% of dogs vaccinated within the last year.
-This remained largely the same in 2019, as the DV focused their efforts in other communes.
-The DV did vaccinate 771 dogs in Moramanga Ville in 2019, and of those 25% had been vaccinated in the previous year's campaign, wheras only 5% of dogs vaccinated in all other communes had a previous history of vaccination.
+In 2018, `r summ_tab1$All_2018[summ_tab1$prev_vacc]` of dogs had a previous history of vaccination (largely in the urban commune of Moramanga Ville) with only `r summ_tab1$All_2018[summ_tab1$vacc_in_year]` of dogs vaccinated within the last year.
+This remained largely the same in 2019 (~ `r summ_tab1$All_2019[summ_tab1$prev_vacc]`), as the DV focused their efforts in other communes.
+The DV did vaccinate `r prev_vacc_2019$total[prev_vacc_2019$Commune %in% "Moramanga"]` dogs in Moramanga Ville in 2019, and of those `r prev_vacc_2019$perc_vacc[prev_vacc_2019$Commune %in% "Moramanga"]`% had been vaccinated in the previous year's campaign, wheras only `r prev_vacc_2019$perc_vacc[prev_vacc_2019$Commune %in% "Moramanga"]`% of dogs vaccinated in all other communes had any history of previous vaccination.
+In addition, in 2019, `r round(spayed_neutered * 100, 1)`% of animals had been spayed or neutered, reflecting efforts by the Mad Dog Initiative to implemement free spay and neuter clinics in the district.
 
-While less than 20% of owners in Moramanga Ville reported that their animals were free roaming, fenced animals were frequently observed outside fences during the campaign, and thus the majority of animals could be classified as semi-confined in the more urban township of Moramanga Ville, and free-roaming in the rural setting of Andasibe (approximately 70% of owners reported their dogs as free-roaming, Table 1).
+While less than `r summ_tab1$Moramanga_2018[summ_tab1$name %in% "roaming"]`% of owners in Moramanga Ville reported that their animals were free roaming, fenced animals were frequently observed outside fences during the campaign, and thus the majority of animals could be classified as semi-confined in the more urban township of Moramanga Ville, and free-roaming in the rural setting of Andasibe (approximately `r summ_tab1$Andasibe_2018[summ_tab1$name %in% "roaming"]`% of owners reported their dogs as free-roaming, Table 1).
 In 2019, the DV also asked owners how far they travelled approximately in meters, and `r dist_over_100`% of people travelled less than 1 km to reach the vaccination point.
@@
 The 2018 campaign covered two communes, and was estimated to have acheived approximately 60% coverage (Figure 2A).
-The 2019 campaign covered more communes, but likely acheived lower coverage levels at the commune level (Figure 2B).
-We also compared the coverage estimates from the transects with estimates using human-to-dog ratios (circles in Figure 2B), with the midpoint assuming HDR estimates from a recent household survey study in the Moramanga District and the range based on estimates from the literature.
+The 2019 campaign covered `r n_distinct(animals_2019$Commune)` communes, but was estimated to have acheived lower coverage levels (Figure 2).
 In the 2018 campaign, we used post-vaccination coverage transects to estimate vaccination coverage, but we were unable to do this in 2019 given the continuous delivery strategy.
 In addition, in Andasibe in 2018, coverage estimates were based on a single transect resulting in more uncertainty.
-However, in both cases, coverage estimated from the transects in 2018 were consistent with the HDR based estimates (for both Andasibe and Moramanga, they fell within the range of the HDR estimates).
-We also back-calculated HDRs given our vaccination coverage estimates, and these were similar to the HDRs calculated from the household survey (`r hdr_backcalcs %>% filter(Commune == "Andasibe") %$% glue("{hdr_cov_est_lower} - {hdr_cov_est_upper}")` for Andasibe compared to `r `hdrs$mora_hdrs[hdrs$loc == "rural"]` in a rural community and `r hdr_backcalcs %>% filter(Commune == "Moramanga") %$% glue("{hdr_cov_est_lower} - {hdr_cov_est_upper}")` for Moramanga Ville compared to `r `hdrs$mora_hdrs[hdrs$loc == "urban"]` in a urban community).
+However, in both cases for 2018, coverage estimated from the transects in 2018 were consistent with the HDR based estimates (for both Andasibe and Moramanga, they fell within the range of the HDR estimates).
+We also back-calculated HDRs given our vaccination coverage estimates, and these were similar to the HDRs calculated from the household survey (`r hdr_backcalcs %>% filter(Commune == "Andasibe") %$% glue("{hdr_cov_est_lower} - {hdr_cov_est_upper}")` for Andasibe compared to `r `hdrs$mora_hdrs[hdrs$loc == "rural"]` in a rural community and `r hdr_backcalcs %>% filter(Commune == "Moramanga") %$% glue("{hdr_cov_est_lower} - {hdr_cov_est_upper}")` for Moramanga Ville compared to `r hdrs$mora_hdrs[hdrs$loc == "urban"]` in an urban community).
 
@@
 #### Figure 2. Spatial estimates of coverage achieved by the 2018 and 2019 campaigns.
-*A) The commune level numbers of dogs vaccinated (size of points) and the associated coverage estimates (color of points) for the year 2018 (estimated using post-vaccination transects) and 2019 (estimated using a human-to-dog ratio (HDR) of NN, based on a recent household survey in the Moramanga District).
+*A) The commune level numbers of dogs vaccinated (size of points) and the associated coverage estimates (color of points) for the year 2018 (estimated using post-vaccination transects) and 2019 (estimated using a human-to-dog ratio (HDR) of `r round(mora_mid_hdr, 1)`, based on a recent household survey in the Moramanga District).
 The inset shows the location of the Moramanga District in Madagascar.
 B) A comparison of coverage estimates by location and by method of estimation (post-vaccination surveys vs.
-HDR based estimates); for the transect based estimates, the line range shows the 95% exact binomial confidence interval while for the HDR based estimates, the line range shows the range of coverage estimates assuming an HDR range of 8 - 25 based on the literature.*
+HDR based estimates); for the transect based estimates, the line range shows the 95% exact binomial confidence interval while for the HDR based estimates, the line range shows the range of coverage estimates assuming an HDR range of 8 - 25 based on estimates from the literature.*
 
@@
 
-Demographic data from vaccinated dogs shows a population pyramid with a large base, indicative of a fast growing population, and with a male bias (approximately `r prop_sex["M"]`% Male, Figure 3A).
+Demographic data from vaccinated dogs shows a population pyramid with a large base, indicative of a fast growing population, and with a male bias (approximately `r prop_sex["M"]`% male, Figure 3A).
 We fit these data to an age-structured model, and were able to generate parameter estimates which resulted in stable age distributions consistent with the data (Figure 3B).
@@
 We find that given these demographic parameters, annual campaigns that target dogs of all ages result in rapid decline in vaccination coverage in between campaigns, largely due to rapid turnover of the dog population (compared to the impact of waning of vaccine immunity, which you can see in the additional dip at year 3, Fig 3D).
-Continuously targeting 70% of the puppy population for vaccination, while unable to acheive the peak coverage of 70% due to puppy mortality, achieves consistent coverage of about 50% of the dog population.
+Continuously targeting 70% of the puppy population for vaccination, while unable to acheive the peak coverage, achieves consistent coverage of about 50% of the dog population.
 A combined strategy maintains the highest and most consistent levels of coverage close to the target of 70%.
@@
 C) Parameter estimates for annual fertility rates and adult survival probability, with estimates in orange showing the parameter estimates which result in positive population growth.
-B) Simulated vaccination coverage (N = 1000) using the demographic parameters from (C) in a hypothetical commune with 1000 dogs for three different campaign strategies: 1) annual vaccination campaigns targeting dogs of all ages (purple), 2) Routine vaccination of puppies at 3 months of age, and 3) a combined strategy with campaigns annually and continuous puppy vaccination in between campaigns.*
+B) Simulated vaccination coverage (N = 1000) using the demographic parameters from (C) in a hypothetical commune with 1000 dogs for three different campaign strategies: 1) annual vaccination campaigns targeting dogs of all ages (purple), 2) routine vaccination of puppies at 3 months of age, and 3) a combined strategy with campaigns annually and continuous puppy vaccination in between campaigns.*
 
@@
 
-Through the 2018 and 2019 campaigns implemented in Moramanga District, we learned that mass vaccination of dogs is a feasible strategy for controlling canine rabies in Madagascar.
-There was high demand for vaccination from owners and dogs were accesible and able to be handled efficiently for parenteral vaccination.
-We found that providing the vaccine at no direct cost to owners will be critical to acheiving sufficient coverage, as even with nominal fees, a significant proportion of owners would no longer vaccinate their animals.
+Through the 2018 and 2019 campaigns implemented in Moramanga District, we saw high demand for vaccination from owners and found that dogs were accesible and able to be handled efficiently for parenteral vaccination.
+Providing the vaccine at no direct cost to owners will be critical to acheiving sufficient coverage, as even with nominal fees, a significant proportion of owners indicated they would no longer vaccinate their animals.
 While the 2018 vaccination campaigns acheived higher coverage over a shorter time period, it came at the a higher cost per animal vaccinated, was more limited in geographic scope, and was much more resource intensive in terms of personnel requirements.
-In addition, in the rural setting of Andasibe, the static point campaign strategy seemed to be less effective, reflecting less accessible human populations in this location.
-The 2019 campaign on the other hand, where the District Veterinarian provisioned vaccines through routine services, covered a much larger geographic scope and was significantly less resource intensive, but achieved lower vaccination coverage.
-Finally, we used age data collected and a vaccination model to simulate different vaccination strategies, and found that continuous vaccination targeting puppies in particular may be an effective way to maintain coverage levels, particulalry in rural areas where static point vaccination campaigns may be less effective.
+In addition, in the rural setting of Andasibe, the static point campaign strategy seemed to be less effective, reflecting less accessible human communities in this location.
+The 2019 campaign on the other hand where the District Veterinarian provisioned vaccines through routine services, had a much wider geographic scope and was significantly less resource intensive, but achieved lower vaccination coverage.
+Based on the lessons learned through the campaigns, in particular the observation that puppies were relatively easy to handle for vaccinators and owners, we used age data collected and a vaccination model to simulate different vaccination strategies, and found that continuous vaccination targeting puppies in particular may be an effective way to maintain coverage levels given high turnover in dog populations.
 
@@
 Owner reports of willingness to pay, age of animals, and distance travelled to the vaccination point likely all suffer from recall bias and uncertainty.
-However, our findings from these data are consistent with previous literature.
-We estimate that the dog population is male-biased and skewed towards puppies, but that pups likely suffer high mortatlity in the first year of life, consistent with studies from other sub-Saharan African countries.
-Other studies have also shown that implementing fees for vaccination greatly reduces the efficacy of campaigns.
-Similarly, distance to campaign points has also been identified as major barrier to vaccination, and as with other studies we found that the majority of people report traveling less than 1km to reach a vaccination point.
-Dog population estimates were also largely lacking, but we were able to ground our estimates of human-to-dog ratios from a recent household survey study in the distict and used a range that is consistent with the literature, and coverage estimates using HDRs were comparable to estimates from post-vaccination transects.
-We also make several simplifying assumptions in our model of vaccination strategies: we assume that immunity wanes after three years (simulating vaccination with a long-lasting vaccine such as Nobivac, compared to the 1 year efficacy estimated for the Rabisin vaccine); we assume that vaccinated dogs that survive from year to year are revaccinated in subsequent campaigns; and in our age-structured models we do not account for population carrying capacity, but rather simulate over a short time frame over which growth estimates remain reasonable.
+In addition, age of animals brought to vaccination points may not be representative ofthe age structure of the underlying population, and other work has shown that in general puppies (individuals < 1 year) tend to have lower vaccination coverage than adults [@arief2016, @sanchez-soriano2020, @mazeri2019, @minyoo2015].
+However, our analyses based on these data are consistent with previous findings from SSA that demonstrate male-biased populations, skewed towards puppies, and with high mortatlity in the first year of life [@czupryna2016, @taylor2017].
+Studies have also shown that charging fees for vaccines can greatly reduce the efficacy of campaigns and acts as a significant barrier to vaccination for owners [@schildecker2016, @yoak2021, @barbosacosta2020].
+Similarly, distance to campaign points has also been identified as barrier to vaccination, and in most cases owners eport traveling less than 1km to reach a vaccination point [@mazeri2018barriers, @mazeri2019].
+We used human to dog ratios, which can be sensitive to underlying estimates of the dog population and the spatial scale of estimation [@sambo2018estimating, @sambo2017].
+However, we were able to use estimates from a recent household survey study in the districs, and they were also consistent with coverage estimates from post-vaccination transects.
+Finally, in our vaccinaiton model, we  make several simplifying assumptions: we assume that immunity wanes after three years (simulating vaccination with a long-lasting vaccine such as Nobivac, compared to the one year efficacy estimated for the Rabisin vaccines that we used); we assume that vaccinated dogs that survive from year to year are revaccinated in subsequent campaigns; and in our age-structured models we do not account for population carrying capacity, but rather simulate over a short time frame over which growth estimates remain reasonable.
 
-Our estimates of costs per animal vaccinated are in line with recent estimates from other settings, but on the lower end.
+Our estimates of costs per animal vaccinated are in line with recent estimates from other settings [@undurraga2020], although we likely underestimate costs given donation of time and resources by the organizations and indiviudals involved.
 In our estimates, we excluded costs associated with international volunteers (i.e.
@@
 While the volunteer-led effort resulted in significant financial and personnel resources being devoted to the campaign, costs were higher per animal and overall.
-In addition, it can be difficult for NGO and volunteer based efforts to sustain these types of efforts given unpredictable funding, time commitments, and turnover in staff.
+In addition, it can be difficult for NGO and volunteer based campaigns to sustain these efforts given unpredictable funding, time commitments, and turnover in staff [@ferguson2020].
 The routine vaccination strategy through the DV was less costly, with the majority of the costs directly related to vaccination.
@@
 One key aspect we do not consider is the feedback loops between vaccination and demography.
-Estimates of the effects of vaccination on dog demography are mixed, but vaccination could potentially increase survival of dogs.
+Estimates of the effects of vaccination on dog demography are mixed [ @czupryna2016], but vaccination could potentially increase survival of dogs [@conan2015].
 If dog population growth is driven by survival, then this could mean that increased vaccination results in increased population growth.
 However, if growth is driven more by demand from human communities, then vaccination could stabilize the population and reduce turnover of the population.
-Improving dog population management, encouraging responsible pet ownership practices, and increasing veterinary services could all complement vaccination efforts.
-Overall, high population turnover of the dog population erodes coverage, therefore strategies to stabilize the population would be beneficial to improving vaccination coverage.
+Improving dog population management, encouraging responsible pet ownership practices, and increasing veterinary services could all complement vaccination efforts, but has not been demonstrated to result in meaningful rabies control [@taylor2017].
 
-In these settings with high dog ownership, moving towards community-based vaccination strategies, could be a highly effective way to achieve sufficient and consistent coverage, particulalry in rural areas where static point campaigns may be less likely to reach.
+In settings with high dog ownership, moving towards community-based vaccination strategies could be a highly effective way to achieve sufficient and consistent coverage, particulalry in rural areas where static point campaigns may be less effective.
 During our campaign, we found that puppies at approximately 3 months old were easy to handle for both vaccinators and owners.
-Puppy vaccination could therefore use local officials emmbedded in communities (along the lines of community health workers) to implement vaccination, particularly given recent findings on thermotolerance of rabies vaccines and locally manufactured methods for maintaing temperatures required for sustained vaccine storage (up to 3 months).
+Puppy vaccination could therefore use local officials emmbedded in communities (along the lines of community health workers) to implement vaccination, particularly given recent findings on thermotolerance of rabies vaccines and locally manufactured methods for maintaing temperatures required for sustained vaccine storage (up to 3 months)
+[@lugelo2020].
 Incentivizing vaccinators appropriately will be a key challenge, as currently providing no-cost rabies vaccination is not seen as part of routine duties for district veterinarians or for livestock officers, and these officials often required additional per diems.
-Expanding veterinary services across the country and releiving financial pressures on veterinarians and animal health workers could greatly mitigate these issues in Madagascar.
-For volunteer based efforts, veterinary students could be a local volunteer pool, however similarly to international volunteers, they require food, transport, and lodging during vaccination campaigns when not based in the communities where they study or live, adding to overall costs.
+Implementing dog vaccination alongside government mandated livestock vaccination campaigns may also be a stategy to scale up vaccination efforts in a low-cost way.
+Expanding veterinary services across the country and releiving financial pressures on veterinarians and animal health workers could greatly improve veterinary services across Madagascar [@mondiale2017evaluation].
+For volunteer based efforts, focusing on local volunteers (i.e.
+veterinary students) may be a more cost-effective strategy.
+However similar to international volunteers, volunteers require food, transport, and lodging during vaccination campaigns when not based in the communities where they study or live.
 
-Overall, our results suggest that dog vaccination is a feasible and feasible strategy for controlling canine rabies.
-It will be key to push rabies vaccination as a public good and to remove barriers for dog owners as well as to incentivize veterinarians and other animal health workers to implement vaccination.
+Overall, our results suggest that dog vaccination is a feasible and effective strategy for controlling canine rabies in Madagascar.
+It will be key to push rabies vaccination as a public good, to remove barriers for dog owners, and to incentivize veterinarians and other animal health workers to implement vaccination.
 Borrowing strategies from human vaccination efforts, i.e.
-childhood vaccination and using community health workers to provide routine services, could be a way to improve vaccination coverage and reduce costs, particularly in rural communities.
-To monitor the success of these campaigns, it will be critical to develop efficient and effective methods to estimate vaccination coverage and the impact of campaigns on rabies incidence through robust surveillance systems.
+use of community health workers, could be a way to improve vaccination coverage and reduce costs, particularly in rural communities where static point campaigns are less effective.
+In addition, continually refining vaccination strategies based on local contexts and using improved tools and systems, such as mobile phone based data collection, could improve efficacy and lower costs of campaigns [@mazeri2021a, @gibson2016a].
+To monitor the success of these campaigns, it will be critical to develop efficient and effective methods to estimate vaccination coverage and to measure the true impact of campaigns on reducing rabies incidence through robust surveillance systems [@lushasi2020, @ferguson2020].
 Given it's island context, Madagascar could be a model for rabies elimination, with limited chances for reintroduction from outside the island, and strategies that tackle the context specific challenges could be a path forward for the country to reach ZeroBy30.

### 2021-02-05 R/Table1.R

@@ -40,7 +40,7 @@ vacc_2018 %>%
          vacc_in_year) -> animals_2018
 
 vacc_2019 %>%
-  filter(`Dog or Cat` %in% "dog") %>%
+  filter(tolower(`Dog or Cat`) %in% "dog") %>%
   select(Date, Commune, Fokontany = `Owner Location`, 
          Sex, prev_vacc = `Previous Vaccination`,
          vacc_in_year = `Last Vaccination Date`) %>%

### 2021-02-05 R/figure1.R

@@ -144,13 +144,13 @@ write_csv(cost_recovery, "out/cost_recovery.csv")
 
 # Willingess to pay
 ggplot(willingness) +
-  geom_line(aes(x = charged/exch_rate, y = 1 - prop_willing*100, linetype = Commune), 
+  geom_line(aes(x = charged/exch_rate, y = (1 - prop_willing)*100, linetype = Commune), 
             size = 1.2) +
   xlim(c(0, 3)) +
   scale_linetype_manual(values = c(3, 2, 1), 
                         c("Andasibe (rural)", "Moramanga (urban)", 
                           "Overall"), name = "Location") +
-  labs(x = "Amount charged to \n owner (USD)", y = "Reduction in # \n of dogs vaccinated", tag = "D") +
+  labs(x = "Amount charged to \n owner (USD)", y = "% Reduction in dogs vaccinated", tag = "D") +
   theme_minimal_hgrid(font_size = 12) +
   guides(linetype = guide_legend(override.aes = list(size = 0.5))) -> fig1D
 

### 2021-02-05 R/make_paper.R

@@ -1,0 +1,5 @@
+# Make paper and push up to google doc
+rmarkdown::render("manuscript.Rmd", output_format = "bookdown::word_document2")
+
+# upload to google drive for feedback
+googledrive::drive_put("manuscript.docx", 'moramanga_vax')

### 2021-02-12 manuscript.Rmd
updating with revised draft

@@
       institute: [maddog]
-  - ?Glenn Edosoa:
+  - Glenn Edosoa:
       institute: [oms]
-  - ?Mamitiana Andriamananjara:
+  - Mamitiana Andriamananjara:
       institute: [dsv]
+  - Nivohanitra Perle Razafindraibe:
+      institute: [dsv]
+  - José Nely:
+      institute: [slmen]
+  - Angelique Ferreira:
+      insitute: [mdi, tad]
+  - Annie Yang:
+      institute: [princeton, note]
   - Fenomanana Daniel:
@@
 institute:
+  - florida: University of Florida, Department of Anthropology, Gainesville, Florida, USA
   - princeton: Department of Ecology and Evolutionary Biology, Princeton University, Princeton, United States
+  - maddog: The Mad Dog Initiative, Akanin’ny Veterinera Akaikiniarivo, Ambatobe, Antananarivo 101, Madagascar
+  - uoa: Mention Zoologie et Biodiversité Animale, Faculté des Sciences, Université d’Antananarivo, Antananarivo, Madagascar
   - oms: Chargé des Maladies Tropicales Négligées, Organisation mondiale de la Santé Madagascar, Antananarivo, Madagascar
-  - florida: University of Florida, Department of Anthropology, Gainesville, Florida, USA
-  - maddog: The Mad Dog Initiative, Akanin’ny Veterinera Akaikiniarivo, Ambatobe, Antananarivo 101, Madagascar
+  - dsv: Direction des Services Vétérinaires, Ministère chargé de l’Agriculture et de l’Élevage, Antananarivo, Madagascar
+  - slmen: Service contre les Maladies Endémo-épidémiques et Tropicales Négligées, Ministère de la Santé Publique,  Antananarivo, Madagascar
   - tad: Travelling Animal Doctors, 919 Crossan Rd, Newark, Delaware, USA 
-  - uoa: Mention Zoologie et Biodiversité Animale, Faculté des Sciences, Université d’Antananarivo, Antananarivo, Madagascar
+  - appstate: Appalachian State University, Department of Health and Exercise Science, Boone, North Carolina, USA
   - glasgow: Boyd Orr Centre for Population and Ecosystem Health, Institute of Biodiversity, Animal Health and Comparative Medicine, University of Glasgow, Glasgow, UK
-  - appstate: Appalachian State University, Department of Health and Exercise Science, Boone, North Carolina, USA
-  - dsv: Direction des Services Vétérinaires, Ministère chargé de l’Agriculture et de l’Élevage, Antananarivo, Madagascar
+  - note: Deceased.
 output:
@@
 <!-- Inline stats fixes -->
+<!-- Kim V -->
 
@@
 
+# data
+transects <- read_csv("data/transects_2018.csv")
+
 ```
 
-## Abstract [comment: Journal reccs]
+## Abstract
 
 Canine rabies causes an estimated 60,000 human deaths per year, but these deaths are preventable through post-exposure prophylaxis of people and vaccination of domestic dogs.
-Dog vaccination campaigns acheiving coverage of 70% at the population level are effective at interrupting transmission and eventually can lead to elimination of the disease.
+Dog vaccination campaigns targeting coverage of 70% at the population level are effective at interrupting transmission and eventually can lead to elimination of the disease.
 Given this evidence, the WHO and it's partners have set a global goal to eliminate deaths due to canine rabies by the year 2030.
@@
 We compare two different vaccination strategies: a larger volunteer driven effort to vaccinate dogs in two communes using static point vaccination, and continuous vaccination as part of routine veterinary services in the district by the District Veterinarian.
-We compared costs, time, personnel, and coverage acheived by each of the campaigns.
-We also use age data from the campaigns to estimate key demographic parameters, and use these estimates to simulate different vaccination strategies.
+We compared costs, time, personnel, and coverage achieved by each of the campaigns.
+We also use dog age data from the campaigns to estimate key demographic parameters, and use these estimates to simulate different vaccination strategies.
 Overall we found that dog vaccination was feasible and that most dogs were accessible to vaccination.
-We find that there are tradeoffs between the vaccination strategies, with the first campaign acheiving higher coverage, but with more resources and with more limited geographic scope compared to the second campaign with continuous delivery.
-Our modeling results suggest that targeting puppies through community based vaccination efforts could help acheive coverage levels sufficient to interrupt transmission.
-Moving forward, our results show that mass dog vaccination is feasible and can acheive high coverage in Madagascar, however innovative strategies and an investment in dog vaccination as a public good will be required to maintain the high and sustained levels of coverage needed for disease elimination.
+We find that there are tradeoffs between the vaccination strategies, with the first campaign achieving higher coverage, but with more resources and with more limited geographic scope compared to the second campaign with continuous delivery.
+Our modeling results suggest that targeting puppies through community based vaccination efforts could help achieve coverage levels sufficient to interrupt transmission.
+Moving forward, our results show that mass dog vaccination is feasible and can achieve high coverage in Madagascar, however innovative strategies and an investment in dog vaccination as a public good will be required to maintain the sustained levels of coverage needed for disease elimination.
 
@@
 Canine rabies results in an estimated 60,000 human deaths per year globally [@hampson2015].
-These deaths are entirely preventable: prompt post-exposure prophylaxis (PEP) of humans is highly effective at preventing death and mass dog vaccination can interrupt transmission in domestic dogs and eventually lead to disease elimination [@fooks2014].
+These deaths are entirely preventable: prompt post-exposure prophylaxis of humans exposed to rabies is highly effective at preventing death and mass dog vaccination can interrupt transmission in domestic dogs and eventually lead to disease elimination [@fooks2014].
 The WHO and its partners have set a goal to eliminate human deaths due to canine rabies by the year 2030 ('ZeroBy30')[@abela-ridder2016].
-Annual vaccination campaigns achieving atleast 70% coverage are the recommended target for controlling rabies in the domestic dog population [@worldhealthorganization2013].
+Annual dog vaccination campaigns achieving at least 70% coverage are the recommended target for controlling rabies in the domestic dog population [@worldhealthorganization2013].
 However, achieving this coverage target in low and middle income countries where the burden of human rabies is concentrated can be challenging due to economic, ecological, sociocultural, and political reasons [@fahrion2017].
@@
 In sub-Saharan African countries, parenteral vaccination implemented through static point vaccination campaigns have been shown to be cost-effective and feasible [@borse2018].
-While most dogs in these settings are considered free-roaming, they are owned and are accessible for vaccination through these campaigns [@morters2014, @jibat2015].
-However, reaching and maintaining high coverage vaccination campaigns requires large scale coordinated efforts, and the challenges in implementation reflect both financial and logistical constraints more than feasibility of vaccination itself [@fahrion2017].
+While most dogs in these settings are considered free-roaming, they are mostly owned and are accessible for vaccination through these campaigns [@morters2014, @jibat2015].
+However, reaching high coverage and maintaining vaccination campaigns at scale requires sustained investment and coordination, and the challenges in implementation largely reflect financial and logistical constraints more than feasibility of vaccination itself [@fahrion2017].
+Developing clear and context-specific strategies and lowering costs and resources needed could help spur implementation and scaling up of campaigns in these countries.
 
-In Madagascar, canine rabies has been endemic for over a century and the Institut Pasteur de Madagascar has provided PEP free-of-charge to patients in the country shortly therafter [@reynes2011].
-However, currently, there are only 31 clinics where the vaccine is available, and there is limited dog vaccination due to high costs to owners and lack of availability of the vaccine [@rajeev2018a].
-Recent studies have estimated a high burden of human deaths, masked by limited surveillance for canine rabies cases and human deaths across the country [@rajeev2020; @rajeev2018a].
-The veterinary sector is largely private, with some veterinarians employed in hybrid private/public employment as a designated district veterinarian.
-While dog vaccination is rare, livestock officers and veterinarians work together to implement vaccination campaigns for anthrax on an annual basis as mandated by the government (wners are charged a fee per animal for these vaccines which vary by location) [@mondiale2017evaluation].
-To date, no regular mass dog vaccinations are conducted on the island, although a few pilot programs have begun in recent years, largely implemented by NGO-government partnerships.
+In Madagascar, canine rabies has been endemic for over a century and over almost the same time period, the Institut Pasteur de Madagascar has provided post-exposure prophylaxis free-of-charge to bite patients in the country [@reynes2011].
+Currently, there are only 31 clinics where these human vaccines are available, and there is minimal dog vaccination due to high costs to owners and lack of availability of the vaccine [@rajeev2018].
+Recent studies have estimated a high burden of human deaths, masked by weak surveillance across the country [@rajeev2020; @rajeev2018].
 
-Here, we summarize lessons learned through implementation of pilot vaccination programs in the Moramanga District of Madagascar, where previous work has shown high incidence of dog rabies cases and human exposures.
+The veterinary sector is largely private, with some veterinarians employed in hybrid private/public employment as designated district veterinarians [comment: Njara/Nivo].
+While dog vaccination is rare, livestock officers and veterinarians work together to implement cattle vaccination campaigns for anthrax on an annual basis as mandated by the government (owners are charged a fee per animal for these vaccines which vary by location) [@mondiale2017evaluation].
+No routine mass dog vaccinations have been conducted on the island, although a few pilot programs have begun in recent years, largely implemented by NGO-government partnerships.
+
+Here, we summarize lessons learned through implementation of pilot vaccination programs in the Moramanga District of Madagascar, where previous work has shown high incidence of dog rabies cases and human rabies exposures.
 In 2018 and 2019, we deployed two different vaccination strategies.
 In 2018, we led a larger scale volunteer-led pilot vaccination campaign in two Communes (sub-district level) using a central point strategy.
-In 2019/2020, we provided vaccines, all necessary supplies for vaccine administration, and a per-vaccine fee to the District Veterinarian (DV) to vaccinate animals as part of routine services.
-We compare time, human resources, costs, and coverage estimates between the campaigns.
-Using demographic data collected during the campaigns and a vaccination model, we also explore different vaccination strategies based on what we learned during implementation.
+In 2019/2020, we provided vaccines, all necessary supplies for vaccine administration, and a per-vaccine fee to the District Veterinarian to vaccinate animals as part of routine services.
+We compare time, human resources, costs, and coverage estimates between these campaigns, and using a demographic and vaccination model we further explore different vaccination strategies based on what we learned during implementation.
 
@@
 It comprises 21 communes, covering approximately 7150 km^2^ with an approximate human population of `r round(sum(mora_sf$pop), -2)`.
-Previous work in the district has established a high burden of rabies exposures and deaths despite the availability of post-exposure prophylaxis at the district hospital [@rajeev2018a].
+Previous work in the district has established a high burden of rabies exposures and deaths despite the availability of post-exposure prophylaxis at the district hospital [@rajeev2018].
 While Moramanga is relatively close to the capital city of Antananarivo (~ 3 hrs by bus), within the district, travel times between locations are highly variable, with much of the population living in more rural areas with limited access to roads and transportation [@rajeev2020].
-Before 2018, there were limited animal rabies vaccination services, with most vaccines available in urban communes and owners charged > 15,000 Ariary (~ 4.28 USD) per vaccine administered.
+Before 2018, there were limited animal rabies vaccination services, with most animal vaccines available in the urban commune of Moramanga Ville and owners charged > 15,000 Ariary (~ 4.28 USD) per vaccine administered.
 
@@
 
-In 2018, we organized a pilot vaccination campaign through a collaboration with two NGOs (the Madagascar Dog Initiative and Traveling Animal Doctors), the Department of Veterinary Services, and the Ministry of Public Health in the District of Moramanga.
-We focused on two communes in Moramanga, Moramanga Ville and Andasibe (a rural commune surrounding Andasibe National Park), where previously high incidence of suspected rabid exposures (Moramanga Ville) and a high burden of deaths (Andasibe) had been recorded [@rajeev2018].
+In 2018, a pilot vaccination campaign was organized through a collaboration with two NGOs (the Mad Dog Initiative and Traveling Animal Doctors), the Department of Veterinary Services, and the Ministry of Public Health in the District of Moramanga.
+This campaign focused on two communes in Moramanga, Moramanga Ville (the district center) and Andasibe (a rural commune surrounding Andasibe National Park), where previously high incidence of probable rabies exposures (Moramanga Ville) and a high burden of deaths (Andasibe) had been recorded [@rajeev2018].
 
-A week before the campaign dates, for each location we advertised the date with the chief of the fokontany (sub-commune) and provided posters advertising the free vaccine.
-During the campaign, we used Rabisin `.r` (10 mL vials with 1 mL per dose, Boehringer Ingelheim) to vaccinate both dogs and cats presented that were > 1 month old.
-We surveyed owners about how many dogs and cats they owned in total (split by > 1 yr vs.
-< 1 yr in order to avoid language ambiguities that might result in excluding pups and kttens), as well as if their dogs were free roaming (no restrictions on movement by the owner/'mireny', tied/'mifatora', or fenced/'fefy').
-Vaccinations were delivered at no cost to owners, but as animal vaccination is generally thought of as a paid service in Madagascar, we asked owners how much in Ariary they would be willing to pay to have one animal vaccinated for rabies.
+The campaign was planned as a series of central point vaccination stations covering 1 - 3 sub-communes (or fokontany) per day.
+A week before the campaign dates, the vaccination team advertised the date with the chief of the fokontany and provided posters advertising the date of the vaccine and that it would be available at no cost to owners.
+During the campaign, we used Rabisin (10 mL vials with 1 mL per dose, Boehringer Ingelheim) to vaccinate both dogs and cats presented that were at least 1 month old.
+As part of the campaign, owners were surveyed by vaccinators about how many dogs and cats they owned in total (split by > 1 yr vs.
+< 1 yr in order to avoid language ambiguities that might result in excluding pups and kittens), as well as if their dogs were free roaming (no restrictions on movement by the owner/'mireny', tied/'mifatora', or fenced/'fefy').
+Vaccinations were delivered at no cost to owners, but as animal vaccination is generally thought of as a paid service in Madagascar, owners were asked how much in Ariary they would be willing to pay to have one animal vaccinated against rabies.
 For each animal vaccinated, we recorded the species (cat or dog), sex, approximate age in years, and whether the animal had been previously vaccinated.
 
-To assess coverage, we used post-vaccination coverage surveys according to previously established methodology[@sambo2017; @gibson2015].
-We marked all animals vaccinated with a colored, non-toxic, livestock crayon [comment: ask Jochem for brand] along the top or back of the head.
-Between 4 - 6 PM when dogs are most active [@sambo2017] and on the same day as the campaign in each location, we did two transects in pairs (two people per team), walking for 1 hr starting in opposite directions and accompanied by a local guide to ensure that walking paths did not overlap.
-We recorded any marked and unmarked dogs we observed, and also recorded their roaming status (whether roaming, inside a fence, or tied), and their approximate age (greater or less than 1 yr of age).
+To assess coverage, post-vaccination coverage surveys were conducted according to previously established methodology[@sambo2017; @gibson2015].
+All animals vaccinated were concurrently marked with a colored, non-toxic, livestock crayon [comment: ask Jochem for brand] along the top or back of the head.
+Between 4 - 6 PM when dogs are most active [@sambo2017] and on the same day as the campaign, in each vaccination location two transect surveys were done.
+Each team (two people per team) walked a transect for 1 hour starting in opposite directions and accompanied by a local guide to ensure that walking paths did not overlap.
+Teams recorded any marked and unmarked dogs observed, their roaming status (whether roaming, inside a fence, or tied), and their approximate age (greater or less than 1 yr of age).
 
@@
 
-For the 2019 campaign, instead of a central point campaign strategy, we distributed vaccine vials (Rabisin `.r`) and the supplies needed to administer them (needle, syringe, vaccination card for owners) to the District Veterinarian (DV).
-The DV delivered the vaccination at no cost to owners, but was directly compensated 1,5000 Ar (~ 0.40 USD) per rabies vaccine administered.
-The campaign lasted from September 6, 2019 to June 19, 2020, with the DV advertising the vaccines by word of mouth for one week prior to her visiting each location.
-For each animal vaccinated, the DV collected the age, sex, and also asked owners to approximate the distance they travelled to receive the vaccination in meters.
-Researchers communicated with the DV about progress periodically throughout the campaign, primarily through telephone calls.
-No other compensation or instructions were provided, and we asked the DV to administer as many (or as few vaccines) as feasible or wanted.
+For the 2019 campaign, instead of a central point campaign strategy, vaccine vials(Rabisin) and the supplies needed to administer them (needle, syringe, vaccination card for owners) were distributed to the District Veterinarian, who then delivered the vaccination at no cost to owners, but was directly compensated 1,5000 Ar (~ 0.40 USD) per rabies vaccine administered.
+The campaign lasted from September 6, 2019 to June 19, 2020, One week prior to her visiting each location, the District Veterinarian advertised the vaccines by calling ahead to the fokontany leaders and other officials who then advertised to their communities, largely through word-of-mouth.
+For each animal vaccinated, the District Veterinarian collected the age, sex, and also asked owners to approximate the distance they travelled to receive the vaccination in meters.
+Researchers communicated with the District Veterinarian about progress periodically throughout the campaign, primarily through telephone calls.
+No other compensation or instructions were provided, and we asked the District Veterinarian to administer as many (or as few vaccines) as feasible or wanted.
 As the vaccinations were delivered continuously, we were unable to do comparable post-vaccination surveys.
@@
 We documented the overall costs and resources required for the two vaccination efforts.
-We tracked the number of vaccination points, the number of days over which these vaccinations occurred, and the number of person days required (i.e.
+We tracked the number of vaccination points, the number of days over which these vaccinations occurred, and the number of person days required overall (i.e.
 the number of working people per day over the campaign [@mazeri2021]), in addition to monetary costs.
-As costs were incurred in both USD and Ariary, and the exhange rate declined rapidly between 2018 and 2019, we used the midpoint between the two years (`r exch_rate` Ar to 1 USD) to do the cost comparisons.
+As costs were incurred in both USD and Ariary, and the exchange rate declined rapidly between 2018 and 2019, we used the midpoint between the two years (`r exch_rate` Ar to 1 USD) for the cost comparisons.
 
-For the 2018 campaign, we broke costs down into the following categories: direct vaccine costs (cost for vaccine `.r`, syringes, needles, vaccination cards), supplies (livestock crayons, muzzles, gloves, alcohol, swabs), food and lodging for NGO personnel and other vaccinators during the campaign, personnel costs (per diems for DSV veterinarian, livestock field officers, local guides, and NGO employees), and advertisement (posters and banners for advertising the campaign).
-Foreign NGO volunteers expenses for travel to Madagascar were not included in this costs.
-Vehicles and drivers are also not included in these costs, as the drivers' time and vehicle use were donated to the campaign by volunteers involved in the campaign.
-In 2019, costs were split into two categories, direct vaccine costs (for same items as in 2018), and personnel costs (per vaccine fee paid to the district veterinarian), and supplies (a generator and fuel for the veterinarian to maintain the vaccine on a cold chain when there were power outages).
-Transportation costs were also not included as the DV used her own vehicle and vaccinated alongside their routine veterinary practice.
+For the 2018 campaign, we broke costs down into the following categories: direct vaccine costs (cost for vaccine, syringes, needles, vaccination cards), supplies (livestock crayons, muzzles, gloves, alcohol, swabs), food and lodging for NGO personnel and other vaccinators during the campaign, personnel costs (per diems for DSV veterinarian, livestock field officers, local guides, and NGO employees), and advertisement (posters and banners for advertising the campaign).
+Foreign NGO volunteer expenses for travel to Madagascar were not included in these costs.
+Vehicles and drivers are also not included in these costs, as the drivers' time and vehicle use were donated to the campaign by volunteers involved in the campaign [comment: re Katie, transects were really cheap because people vaccinating did them, like ~ 100 bucks for crayons & guides, so will keep it the costs but mention that they are cheap in the discussion].
+In 2019, costs were split into two categories, direct vaccine costs (for same items as in 2018), and personnel costs (per vaccine fee paid to the district veterinarian), and supplies (a generator and fuel for the veterinarian to maintain the vaccine under cold chain during power outages).
+Transportation costs were also not included as the District Veterinarian used her own vehicle and vaccinated a part of their routine veterinary service provisioning.
 
-We used the data on owner reported willingness to pay for vaccine to estimate the proportional reduction in animals vaccinated as fees are increased.
+We used the data on owners' reported willingness to pay for vaccines to estimate the proportional reduction in animals vaccinated as fees are increased.
 We also estimated how this would impact cost per animal vaccinated by approximating the costs for implementation (i.e.
 those costs that remain fixed) from costs incurred per animal vaccinated (i.e.
-vaccine, syringe, vaccination card, per vaccination fee to DV in 2019), and calculating the balance between the returns from owner payments (i.e.
+vaccine, syringe, vaccination card, and per vaccination fee to District Veterinarian in 2019), and calculating the balance between the returns from owner payments (i.e.
 increases in cost recovery per animal vaccinated) vs.
 decreasing numbers of animals vaccinated overall.
-
+[comment maybe add campaign photos (without identifiable people) per Tara]?
 #### Coverage estimates
 
-For the 2018 campaign, we used the transect data to estimate coverage as the proportion of dogs sighted that were marked with a binomial confidence interval at the commune level.
-For the 2019 campaign, we estimated vaccination coverage using human-to-dog ratios and human population estimates [@cleaveland2003, @athingo2020].
+For the 2018 campaign, we used the transect data to estimate coverage as the proportion of dogs sighted that were marked using a binomial confidence interval at the commune level.
+For the 2019 campaign, we estimated vaccination coverage using human-to-dog ratios and human population estimates [@cleaveland2003; @athingo2020].
 We used a range of 8 - 25, based on previous data from Madagascar [@ratsitorahina2009] and based on recent estimates from household surveys in the Moramanga District [@leblancclaireRabiesMadagascarThreepronged2019].
@@
 
-Using the age data on vaccinated animals collected during both vaccination campaigns, we estimated the proportion of the population in four age classes: pups under the age of 1 year, juveniles 1 - 2 yrs, adults age 2 - 6 years, and older dogs age 6+ years based on .
+Using the age data on vaccinated animals collected during both vaccination campaigns, we estimated the proportion of the population in four age classes: pups under the age of 1 year, juveniles 1 - 2 yrs, adults aged 2 - 6 years, and older dogs age 6+ years based on .
 With the assumption that these estimates represent the population at a stable age distribution, we use a Leslie matrix model to estimate annual adult survival probability and fertility using maximum likelihood estimation [@Fujiwara2017].
@@
 
-We assume that all individuals older than 1 reproduce and do not estimate declines in fertility given the small proportion of dogs older than age six in the population.
-To get bootstrapped estimates, we used 100 sub-sampled data sets of 1000 observations each from the observed age data to fit the parameters, and also varied initial values used in the optimization (N = 100 initial values sets) for 10000 parameter estimates total.
+We assume that all individuals older than 1 year of age reproduce, and we do not estimate declines in fertility given the small proportion of dogs older than age six years in the population.
+To get bootstrapped estimates, we used 100 sub-sampled data sets of 1000 observations each from the observed age data to fit the parameters, and also varied initial values used in the optimization (N = 100 initial values sets) for 10,000 parameter estimates total.
 
@@
 We used the parameter estimates from the demographic model to simulate different vaccination strategies in a hypothetical commune with 1000 dogs.
-We used a discrete time age-structed model with a monthly time step to compare three strategies:
+We used a discrete time age-structured model with a monthly time step to compare three strategies:
 
 1)  Annual vaccination campaigns occuring within the same month each year targeting dogs of all ages
-2)  Routine and continuous vaccination of new puppies targeted at the age of 3 months
-3)  A combined approach with annual campaigns and routine puppy vaccination
+2)  Continuous vaccination of new puppies throughout the year targeting puppies that reach the age of 3 months
+3)  A combined approach with annual campaigns (as 1) and routine puppy vaccination (as 2)
 
-We split the dog population into pups (< 1 yr old) and adults based on the stable age distribution estimated from the demographic model.
-To estimate pup survival in year one, we took the fertility estimates (number of new pups per reproducing dog in the pup age class per year) and divided by an estimate of newborn pups per reproducing dog each year based on average litter size, average number of litters per female per year [@hampson2009, @morters2014, @czupryna2016], and the proportion of the adult population that is female (estimated from our data).
+We split the dog population into pups (< 1 year old) and adults based on the stable age distribution estimated from the demographic model.
+To estimate pup survival in year one, we took the fertility estimates (number of new pups per reproducing dog in the pup age class per year) and divided by an estimate of newborn pups per reproducing dog each year based on average litter size, average number of litters per female per year [@hampson2009; @morters2014; @czupryna2016], and the proportion of the adult population that is female (estimated from our data).
 We assumed that for the annual campaign strategy, surviving vaccinated adult dogs were revaccinated in subsequent years [@delriovilas2017], but that if a pup had been vaccinated within 9 months of the campaign, it was not revaccinated.
+We also assumed that vaccine immunity lasted for a discrete period of 3 years (with revaccination resetting this period of waning)[comment for KH abt clarification here].
 A subset of parameter estimates resulted in estimates of population decline, but based on the shape of the age pyramid and to simulate reasonable campaign scenarios, we filtered to parameter estimates which corresponded to positive population growth.
@@
 All data and code are available at [https://github.com/mrajeev08/mora_vax](https://github.com/mrajeev08/mora_vax).
-The vaccinations were part of a public health campaign and routine veterinary service provisioning carried out by the local veterinary officals, the NGOs involved, and in partnership with the Ministry of Public Health and the Department of Veterinary Services at the national level.
-The Madagascar Dog Initiative also maintains institutional approvals (IACUC # at Institution X) and national research permits (MICET permit #) for it's research and volunteer programs [comment: Kim to fill in].
+The vaccinations were part of a public health campaign and routine veterinary service provisioning carried out by the local veterinary officials, the NGOs involved, and in partnership with the Ministry of Public Health and the Department of Veterinary Services at the national level.
+The Madagascar Dog Initiative also maintains institutional approvals (IACUC # at Institution X)[comment: Kim to fill in] and national research permits (MICET permit #130-19/MEDD/SG/DGEF/DGRNE) for its research and volunteer programs .
 
@@
 dist_over_100 <- round(sum(animals_2019$`Distance Travelled (m)` <= 1000, na.rm = TRUE)/sum(!is.na(animals_2019$`Distance Travelled (m)`))*100, 0)
-
+transects_roaming <- round(sum(!(transects$Status %in% c("Fefy", "Mifatotra")))/nrow(transects)*100, 0)
 ```
@@
 We vaccinated at `r table(pts$Commune)["Andasibe"]` points in Andasibe and `r table(pts$Commune)["Moramanga Ville"]` points in Moramanga Ville.
-During the 2019 campaign, between `r paste(format.Date(range(dmy(animals_2019$Date)), "%B %Y"), collapse = " - ")`, the DV vaccinated a total of `r nrow(animals_2019)`animals (`r species_2019["dog"]` dogs and `r species_2019["cat"]`) over 48 days in `r n_distinct(animals_2019$Commune)` communes in the Moramanga District.
+During the 2019 campaign, between `r paste(format.Date(range(dmy(animals_2019$Date)), "%B %Y"), collapse = " - ")`, the District Veterinarian vaccinated a total of `r nrow(animals_2019)`animals (`r species_2019["dog"]` dogs and `r species_2019["cat"]`) over 48 days in `r n_distinct(animals_2019$Commune)` communes in the Moramanga District.
 While more animals were vaccinated per vaccination point and per vaccination day in 2018 compared to 2019, the number of animals vaccinated per person-day was much higher for the 2019 campaign.
@@
 
-In 2018, `r summ_tab1$All_2018[summ_tab1$prev_vacc]` of dogs had a previous history of vaccination (largely in the urban commune of Moramanga Ville) with only `r summ_tab1$All_2018[summ_tab1$vacc_in_year]` of dogs vaccinated within the last year.
-This remained largely the same in 2019 (~ `r summ_tab1$All_2019[summ_tab1$prev_vacc]`), as the DV focused their efforts in other communes.
-The DV did vaccinate `r prev_vacc_2019$total[prev_vacc_2019$Commune %in% "Moramanga"]` dogs in Moramanga Ville in 2019, and of those `r prev_vacc_2019$perc_vacc[prev_vacc_2019$Commune %in% "Moramanga"]`% had been vaccinated in the previous year's campaign, wheras only `r prev_vacc_2019$perc_vacc[prev_vacc_2019$Commune %in% "Moramanga"]`% of dogs vaccinated in all other communes had any history of previous vaccination.
-In addition, in 2019, `r round(spayed_neutered * 100, 1)`% of animals had been spayed or neutered, reflecting efforts by the Mad Dog Initiative to implemement free spay and neuter clinics in the district.
+In 2018, `r summ_tab1$All_2018[summ_tab1$name %in% "prev_vacc"]` of dogs had a previous history of vaccination (largely in the urban commune of Moramanga Ville) with only `r summ_tab1$All_2018[summ_tab1$name %in% "vacc_in_year"]` of dogs vaccinated within the last year.
+This remained largely the same in 2019 (~ `r summ_tab1$All_2019[summ_tab1$name %in% "prev_vacc"]`), as the District Veterinarian focused their efforts in other communes.
+The District Veterinarian did vaccinate `r prev_vacc_2019$total[prev_vacc_2019$Commune %in% "Moramanga"]` dogs in Moramanga Ville in 2019, and of those `r prev_vacc_2019$perc_vacc[prev_vacc_2019$Commune %in% "Moramanga"]`% had been vaccinated in the previous year's campaign, whereas only `r prev_vacc_2019$perc_vacc[prev_vacc_2019$Commune %in% "Moramanga"]`% of dogs vaccinated in all other communes had any history of previous vaccination.
+In addition, in 2019, `r round(spayed_neutered * 100, 1)`% of animals had been spayed or neutered, reflecting efforts by the Mad Dog Initiative to implement free spay and neuter clinics in the district.
 
-While less than `r summ_tab1$Moramanga_2018[summ_tab1$name %in% "roaming"]`% of owners in Moramanga Ville reported that their animals were free roaming, fenced animals were frequently observed outside fences during the campaign, and thus the majority of animals could be classified as semi-confined in the more urban township of Moramanga Ville, and free-roaming in the rural setting of Andasibe (approximately `r summ_tab1$Andasibe_2018[summ_tab1$name %in% "roaming"]`% of owners reported their dogs as free-roaming, Table 1).
-In 2019, the DV also asked owners how far they travelled approximately in meters, and `r dist_over_100`% of people travelled less than 1 km to reach the vaccination point.
+In 2018, `r summ_tab1$Moramanga_2018[summ_tab1$name %in% "roaming"]` of owners reported their animals were free-roaming, but this varied by location (Table 1).
+In addition, while less than `r summ_tab1$Moramanga_2018[summ_tab1$name %in% "roaming"]` of owners in Moramanga Ville reported that their animals were free roaming, the majority of animals observed during the transects (`r transects_roaming`%) were observed outside of fences and not tied, and thus the majority of animals could be classified as semi-confined in the more urban township of Moramanga Ville, and free-roaming in the rural setting of Andasibe (approximately `r summ_tab1$Andasibe_2018[summ_tab1$name %in% "roaming"]` of owners reported their dogs as free-roaming, Table 1).
+In 2019, the District Veterinarian also asked owners to approximate how far they travelled in meters to get their animals vaccinated, and `r dist_over_100`% of people reportedly travelled less than 1 km to reach the vaccination point.
 
-#### Table 1. Summary of 2018 and 2019 campaigns 
+#### Table 1.
+Summary of 2018 and 2019 campaigns: breakdown of of animals vaccinated, prior vaccination history, dog demography, dog ownership, daily and per vaccination rates by year and location (for 2018).
 ```{r}
@@
 
-### Comparing campaign costs and willingness to pay
+### Cost comparison and willingness to pay
 
@@
 Reflecting the extra personnel necessary to run the central point campaign, the 2018 campaign also took substantially more person-days per animal vaccinated (Table 1).
-We found that in most contexts, charging owners for vaccination would result in minimal cost recovery (Fig 1C) and beyond a minimal cost would actually result in increased costs per vaccinated individual.
-More importantly, even a nominal fee would significantly reduce numbers of dogs vaccinated and thus vaccination coverage, particulary in the rural commune of Andasibe (Fig 1D).
+We found that charging owners for vaccination would result in minimal cost recovery (Fig 1C) and beyond a minimal cost would actually result in increased costs per vaccinated individual than free-of-charge campaigns.
+Cost recovery would be more likely given a 2019 style strategy where the majority of the costs are incurred on a per animal basis, compared to the 2018 campaigns where the costs are largely due to setting up the central point vaccination stations (Fig 3A) [comment for KH abt clarification here].
+More importantly, in all cases, even a nominal fee would significantly reduce numbers of dogs vaccinated and thus vaccination coverage, particularly in the rural commune of Andasibe (Fig 1D).
 
@@
 ```
-#### Figure 1. Comparing campaign costs and willigness to pay 
+#### Figure 1. Comparing campaign costs and willingness to pay 
 *A) Vaccine costs broken down by category and by year (colors).
@@
 vaccine, vaccination card, syringes) and baseline implementation costs (i.e.
-personnel, supplies, lodging and food during the campaign).
-C) Estimated cost per animal for each campaign year given increasing costs charged to the owner, with estimated costs declining due to cost recovery through owner payments and then peaking once owners are no longer willing to pay for the vaccine.
+personnel, supplies, subsistence costs for vaccinators during the campaign).
+C) Estimated cost per animal vaccinated under a willingness to pay model for two campaigns examining increasing costs charged to the owner, with estimated costs declining due to cost recovery through owner payments and then peaking once owners report to be no longer willing to pay for the vaccine.
 D) The percent reduction in number of animals vaccinated given owners willingness to pay.
@@
 ### Comparing campaign coverage estimates
-The 2018 campaign covered two communes, and was estimated to have acheived approximately 60% coverage (Figure 2A).
-The 2019 campaign covered `r n_distinct(animals_2019$Commune)` communes, but was estimated to have acheived lower coverage levels (Figure 2).
+The 2018 campaign covered two communes and was estimated to have achieved approximately 60% coverage (Figure 2A).
+The 2019 campaign covered `r n_distinct(animals_2019$Commune)` communes, but was estimated to have achieved lower coverage levels (Figure 2, ranging from 5 - 60%).
 In the 2018 campaign, we used post-vaccination coverage transects to estimate vaccination coverage, but we were unable to do this in 2019 given the continuous delivery strategy.
 In addition, in Andasibe in 2018, coverage estimates were based on a single transect resulting in more uncertainty.
-However, in both cases for 2018, coverage estimated from the transects in 2018 were consistent with the HDR based estimates (for both Andasibe and Moramanga, they fell within the range of the HDR estimates).
-We also back-calculated HDRs given our vaccination coverage estimates, and these were similar to the HDRs calculated from the household survey (`r hdr_backcalcs %>% filter(Commune == "Andasibe") %$% glue("{hdr_cov_est_lower} - {hdr_cov_est_upper}")` for Andasibe compared to `r `hdrs$mora_hdrs[hdrs$loc == "rural"]` in a rural community and `r hdr_backcalcs %>% filter(Commune == "Moramanga") %$% glue("{hdr_cov_est_lower} - {hdr_cov_est_upper}")` for Moramanga Ville compared to `r hdrs$mora_hdrs[hdrs$loc == "urban"]` in an urban community).
+However, coverage estimated from the transects in 2018 were consistent with the HDR based estimates (for both Andasibe and Moramanga, transect-based estimates fell within the range of the HDR estimates).
+We also back-calculated HDRs given our vaccination coverage estimates, and these were similar to the HDRs calculated from the household survey (`r hdr_backcalcs %>% filter(Commune == "Andasibe") %$% glue("{hdr_cov_est_lower} - {hdr_cov_est_upper}")` for Andasibe compared to `r round(hdrs$mora_hdrs[hdrs$loc == "rural"], 1)` in a rural community and `r hdr_backcalcs %>% filter(Commune == "Moramanga") %$% glue("{hdr_cov_est_lower} - {hdr_cov_est_upper}")` for Moramanga Ville compared to `r round(hdrs$mora_hdrs[hdrs$loc == "urban"], 1)` in an urban community).
 
@@
 ```
-#### Figure 2. Spatial estimates of coverage achieved by the 2018 and 2019 campaigns.
+#### Figure 2. Estimates of coverage achieved by the 2018 and 2019 campaigns.
 *A) The commune level numbers of dogs vaccinated (size of points) and the associated coverage estimates (color of points) for the year 2018 (estimated using post-vaccination transects) and 2019 (estimated using a human-to-dog ratio (HDR) of `r round(mora_mid_hdr, 1)`, based on a recent household survey in the Moramanga District).
 The inset shows the location of the Moramanga District in Madagascar.
-B) A comparison of coverage estimates by location and by method of estimation (post-vaccination surveys vs.
-HDR based estimates); for the transect based estimates, the line range shows the 95% exact binomial confidence interval while for the HDR based estimates, the line range shows the range of coverage estimates assuming an HDR range of 8 - 25 based on estimates from the literature.*
+B) A comparison of coverage estimates by location and by method of estimation (post-vaccination transects vs.
+HDR based estimates); for the transect based estimates, the line range shows the 95% exact binomial confidence interval while for the HDR based estimates, the line range shows the range of coverage estimates assuming an HDR range of 8 - 25 according to estimates from the literature.*
 
@@
 
-Demographic data from vaccinated dogs shows a population pyramid with a large base, indicative of a fast growing population, and with a male bias (approximately `r prop_sex["M"]`% male, Figure 3A).
+Demographic data from vaccinated dogs shows a population pyramid with a large base, indicative of a fast-growing population, and with a male bias (approximately `r prop_sex["M"]`% male, Figure 3A).
 We fit these data to an age-structured model, and were able to generate parameter estimates which resulted in stable age distributions consistent with the data (Figure 3B).
 We filter to parameter estimates that are consistent with a growing population resulting in mean adult annual survival probability of `r dem_pars[dem_pars$name == "psurv_adult_est", "value.mean"]` (95% quantile: `r dem_pars[dem_pars$name == "psurv_adult_est", "ci"]`).
-We use estimates of fertility (on average `r dem_pars[dem_pars$name == "fert_est", "value.mean"]`, 95% quantile: `r dem_pars[dem_pars$name == "fert_est", "ci"]`) to back out pup survival, which ranges from `r dem_pars[dem_pars$name == "pup_surv_lower", "value.lower"]` - `r dem_pars[dem_pars$name == "pup_surv_lower", "value.upper"]`.
-We find that given these demographic parameters, annual campaigns that target dogs of all ages result in rapid decline in vaccination coverage in between campaigns, largely due to rapid turnover of the dog population (compared to the impact of waning of vaccine immunity, which you can see in the additional dip at year 3, Fig 3D).
-Continuously targeting 70% of the puppy population for vaccination, while unable to acheive the peak coverage, achieves consistent coverage of about 50% of the dog population.
-A combined strategy maintains the highest and most consistent levels of coverage close to the target of 70%.
+We use estimates of fertility (on average `r dem_pars[dem_pars$name == "fert_est", "value.mean"]`, 95% quantile: `r dem_pars[dem_pars$name == "fert_est", "ci"]`) to back-calculate pup survival, which ranges from `r round(dem_pars[dem_pars$name == "pup_surv_lower", "value.lower"], 2)` - `r round(dem_pars[dem_pars$name == "pup_surv_lower", "value.upper"], 2)`.
+We find that given these demographic parameters, annual campaigns that target dogs of all ages result in rapid decline in vaccination coverage in between campaigns, largely due to rapid turnover of the dog population (compared to the impact of waning immunity assuming a discrete 3 year period of vaccine immunity, evident in the additional dip at year 3, Fig 3D).
+Continuously targeting 70% of the puppy population for vaccination, while unable to achieve the peak coverage, consistently reaches coverage of about 50% of the dog population.
+A combined strategy maintains the highest and most temporally stable levels of coverage close to the target of 70%.
 
@@
 *A) The age pyramid for vaccinated dogs by sex.
-B) Bootstrapped estimates of the proportion of the population in each age class (dark blue) compared to estimates from the demographic models fit to these age data.
-C) Parameter estimates for annual fertility rates and adult survival probability, with estimates in orange showing the parameter estimates which result in positive population growth.
+B) Bootstrapped estimates of the proportion of the population in each age class from the age data (dark blue) compared to estimates from the demographic models fit to these data (light blue).
+C) Parameter estimates for annual fertility rates and adult survival probability, with estimates highlighted in orange showing the parameter estimates which result in positive population growth.
 B) Simulated vaccination coverage (N = 1000) using the demographic parameters from (C) in a hypothetical commune with 1000 dogs for three different campaign strategies: 1) annual vaccination campaigns targeting dogs of all ages (purple), 2) routine vaccination of puppies at 3 months of age, and 3) a combined strategy with campaigns annually and continuous puppy vaccination in between campaigns.*
@@
 
-Through the 2018 and 2019 campaigns implemented in Moramanga District, we saw high demand for vaccination from owners and found that dogs were accesible and able to be handled efficiently for parenteral vaccination.
-Providing the vaccine at no direct cost to owners will be critical to acheiving sufficient coverage, as even with nominal fees, a significant proportion of owners indicated they would no longer vaccinate their animals.
-While the 2018 vaccination campaigns acheived higher coverage over a shorter time period, it came at the a higher cost per animal vaccinated, was more limited in geographic scope, and was much more resource intensive in terms of personnel requirements.
-In addition, in the rural setting of Andasibe, the static point campaign strategy seemed to be less effective, reflecting less accessible human communities in this location.
-The 2019 campaign on the other hand where the District Veterinarian provisioned vaccines through routine services, had a much wider geographic scope and was significantly less resource intensive, but achieved lower vaccination coverage.
-Based on the lessons learned through the campaigns, in particular the observation that puppies were relatively easy to handle for vaccinators and owners, we used age data collected and a vaccination model to simulate different vaccination strategies, and found that continuous vaccination targeting puppies in particular may be an effective way to maintain coverage levels given high turnover in dog populations.
+Through vaccination campaigns implemented in Moramanga District of Madagascar, we saw high demand for vaccination from dog owners and found that dogs were accessible and able to be handled efficiently for parenteral vaccination at a reasonable cost (between 1.3 - 1.8 USD per animal vaccinated).
+We find that providing the vaccine at no direct cost to dog owners will be critical to achieving sufficient coverage, as even with nominal fees, a significant proportion of owners indicated they would no longer vaccinate their animals.
+A central point vaccination strategy achieved higher coverage over a shorter time period in 2018 compared to dog vaccinations conducted by the District Veterinarian as part of routine veterinary service provision in 2019.
+However, it came at a higher cost per animal vaccinated, was more limited in geographic scope, and required more resources in terms of personnel.
+In addition, in the rural setting of Andasibe, the central point campaign strategy achieved lower coverage, reflecting less accessible, hard-to-reach human communities in this location.
+Based on the lessons learned through these campaigns, in particular the observation that puppies were relatively easy to handle for vaccinators and owners, we find that continuous vaccination targeting puppies in particular may be an effective way to maintain vaccination coverage levels given high turnover in dog populations.
 
 There were several limitations to our analyses.
-Owner reports of willingness to pay, age of animals, and distance travelled to the vaccination point likely all suffer from recall bias and uncertainty.
-In addition, age of animals brought to vaccination points may not be representative ofthe age structure of the underlying population, and other work has shown that in general puppies (individuals < 1 year) tend to have lower vaccination coverage than adults [@arief2016, @sanchez-soriano2020, @mazeri2019, @minyoo2015].
-However, our analyses based on these data are consistent with previous findings from SSA that demonstrate male-biased populations, skewed towards puppies, and with high mortatlity in the first year of life [@czupryna2016, @taylor2017].
-Studies have also shown that charging fees for vaccines can greatly reduce the efficacy of campaigns and acts as a significant barrier to vaccination for owners [@schildecker2016, @yoak2021, @barbosacosta2020].
-Similarly, distance to campaign points has also been identified as barrier to vaccination, and in most cases owners eport traveling less than 1km to reach a vaccination point [@mazeri2018barriers, @mazeri2019].
-We used human to dog ratios, which can be sensitive to underlying estimates of the dog population and the spatial scale of estimation [@sambo2018estimating, @sambo2017].
-However, we were able to use estimates from a recent household survey study in the districs, and they were also consistent with coverage estimates from post-vaccination transects.
-Finally, in our vaccinaiton model, we  make several simplifying assumptions: we assume that immunity wanes after three years (simulating vaccination with a long-lasting vaccine such as Nobivac, compared to the one year efficacy estimated for the Rabisin vaccines that we used); we assume that vaccinated dogs that survive from year to year are revaccinated in subsequent campaigns; and in our age-structured models we do not account for population carrying capacity, but rather simulate over a short time frame over which growth estimates remain reasonable.
+Owner reports of willingness to pay, age of animals, and distance travelled to the vaccination point likely all suffer from recall bias and uncertainty, whilst willingness to pay studies are often overestimates of observed practice [@birhane2016].
+In addition, age of animals brought to vaccination points may not be representative of the age structure of the underlying population, and other work has shown that in general puppies (individuals < 1 year) tend to have lower vaccination coverage than adults [@arief2016; @sanchez-soriano2020; @mazeri2019; @minyoo2015].
+We also did not vaccinate animals less than 1 month of age, which may explain why we sometimes estimated declining populations (as part of the age class of 0-1 years was not captured by the data).
+However, our analyses based on these data are consistent with previous findings from SSA that demonstrate male-biased populations, skewed towards puppies, and with high mortality in the first year of life [@czupryna2016; @taylor2017].
+Studies have also shown that charging fees for vaccines can greatly reduce the efficacy of campaigns and are a significant barrier to vaccination for owners [@schildecker2016; @yoak2021; @barbosacosta2020].
+Similarly, distance to campaign points has been identified as a barrier to vaccination, and in most cases owners eport traveling less than 1km to reach a vaccination point [@mazeri2018barriers; @mazeri2019].
+We used HDRs, which can be sensitive to underlying estimates of the human population and the spatial scale of estimation [@sambo2018estimating; @sambo2017].
+However, we used HDR estimates from a recent household survey study in the district, and found these to give coverage estimates consistent with those from post-vaccination transects.
+Finally, in our vaccination model, we  make several simplifying assumptions: we assume that immunity wanes after three years (simulating vaccination with a long-lasting vaccine such as Nobivac, compared to the manufacturer stated one year efficacy for Rabisin [comment for KH BI states this on their vials]); we assume that vaccinated dogs that survive to the following year are revaccinated in subsequent campaigns; and in our age-structured models we do not account for population carrying capacity, but rather simulate over a short time frame over which population growth remains reasonable.
 
-Our estimates of costs per animal vaccinated are in line with recent estimates from other settings [@undurraga2020], although we likely underestimate costs given donation of time and resources by the organizations and indiviudals involved.
-In our estimates, we excluded costs associated with international volunteers (i.e.
-airfare and visa costs) as well as costs of transportation, as these were donated to the campaign.
-We also did not include estimates of pre-exposure prophylaxis as all of our volunteers and vaccinators had received the course prior to the campaigns.
-In a rabies endemic situation, it will be necessary to have both PrEP and PEP readily available for vaccinators.
-While the volunteer-led effort resulted in significant financial and personnel resources being devoted to the campaign, costs were higher per animal and overall.
-In addition, it can be difficult for NGO and volunteer based campaigns to sustain these efforts given unpredictable funding, time commitments, and turnover in staff [@ferguson2020].
-The routine vaccination strategy through the DV was less costly, with the majority of the costs directly related to vaccination.
-Transport to locations was not included in these estimates either, as the DV had their own vehicle and largely delivered the vaccinations as part of her routine services.
+Our estimates of costs per animal vaccinated are in line with recent estimates from other settings [@undurraga2020], although we likely underestimate costs given the donation of time and resources by the organizations and individuals involved (i.e.
+costs associated with international volunteers including airfare and visa costs as well as costs of transportation donated to the campaign).
+We also did not include costs of pre-exposure prophylaxis as all of our volunteers and vaccinators had been vaccinated prior to the campaigns.
+Both pre- and post-exposure vaccines should however always be readily available for vaccinators and be included in vaccination program budgets.
+While the volunteer-led effort resulted in significant financial and personnel resources being devoted to the campaign, costs were higher per animal vaccinated and overall.
+In addition, NGO and volunteer based campaigns may be difficult to sustain given unpredictable funding, time commitments, and turnover in staff [@ferguson2020].
+Dog vaccination delivered through the District Veterinarian was less costly, with the majority of the costs directly related to vaccination.
 
-One key aspect we do not consider is the feedback loops between vaccination and demography.
-Estimates of the effects of vaccination on dog demography are mixed [ @czupryna2016], but vaccination could potentially increase survival of dogs [@conan2015].
+One key aspect we do not consider is potential feedback loops between vaccination and demography.
+Estimates of the effects of vaccination on dog demography are mixed [@czupryna2016], but vaccination may increase dog survival [@conan2015].
 If dog population growth is driven by survival, then this could mean that increased vaccination results in increased population growth.
 However, if growth is driven more by demand from human communities, then vaccination could stabilize the population and reduce turnover of the population.
-Improving dog population management, encouraging responsible pet ownership practices, and increasing veterinary services could all complement vaccination efforts, but has not been demonstrated to result in meaningful rabies control [@taylor2017].
+Improving dog population management, encouraging responsible pet ownership practices, and increasing veterinary services could all complement vaccination efforts, but have not been demonstrated to result in meaningful rabies control without parallel dog vaccination [@taylor2017].
 
-In settings with high dog ownership, moving towards community-based vaccination strategies could be a highly effective way to achieve sufficient and consistent coverage, particulalry in rural areas where static point campaigns may be less effective.
+In settings with high dog ownership, moving towards community-based vaccination strategies could be an effective way to achieve sufficient and consistent coverage, particulary in hard-to-reach communities.
 During our campaign, we found that puppies at approximately 3 months old were easy to handle for both vaccinators and owners.
-Puppy vaccination could therefore use local officials emmbedded in communities (along the lines of community health workers) to implement vaccination, particularly given recent findings on thermotolerance of rabies vaccines and locally manufactured methods for maintaing temperatures required for sustained vaccine storage (up to 3 months)
+Puppy vaccination could be carried out by local officials embedded in communities (along the lines of community health workers), particularly given recent findings on thermotolerance of rabies vaccines and locally manufactured methods for maintaining temperatures required for sustained vaccine storage (up to 3 months)
 [@lugelo2020].
-Incentivizing vaccinators appropriately will be a key challenge, as currently providing no-cost rabies vaccination is not seen as part of routine duties for district veterinarians or for livestock officers, and these officials often required additional per diems.
-Implementing dog vaccination alongside government mandated livestock vaccination campaigns may also be a stategy to scale up vaccination efforts in a low-cost way.
-Expanding veterinary services across the country and releiving financial pressures on veterinarians and animal health workers could greatly improve veterinary services across Madagascar [@mondiale2017evaluation].
+Incentivizing vaccinators appropriately will be a key challenge, as currently providing no-cost rabies vaccination is not seen as part of routine duties for district veterinarians or for livestock officers.
+Implementing dog vaccination alongside government mandated livestock vaccination campaigns may also be a strategy to scale up vaccination efforts at relatively low-cost.
+Expanding veterinary services across the country and relieving financial pressures on veterinarians and animal health workers through appropriate compensation could greatly improve veterinary services across Madagascar [@mondiale2017evaluation].
 For volunteer based efforts, focusing on local volunteers (i.e.
 veterinary students) may be a more cost-effective strategy.
-However similar to international volunteers, volunteers require food, transport, and lodging during vaccination campaigns when not based in the communities where they study or live.
+However similar to international volunteers, volunteers require subsistence during vaccination campaigns when not based in the communities where they study or live.
 
 Overall, our results suggest that dog vaccination is a feasible and effective strategy for controlling canine rabies in Madagascar.
-It will be key to push rabies vaccination as a public good, to remove barriers for dog owners, and to incentivize veterinarians and other animal health workers to implement vaccination.
+However, rabies vaccination must be recognized as a public good, and strategies to remove barriers for dog owners and to incentivize veterinarians and other animal health workers to implement vaccination will be key to campaign success.
 Borrowing strategies from human vaccination efforts, i.e.
-use of community health workers, could be a way to improve vaccination coverage and reduce costs, particularly in rural communities where static point campaigns are less effective.
-In addition, continually refining vaccination strategies based on local contexts and using improved tools and systems, such as mobile phone based data collection, could improve efficacy and lower costs of campaigns [@mazeri2021a, @gibson2016a].
-To monitor the success of these campaigns, it will be critical to develop efficient and effective methods to estimate vaccination coverage and to measure the true impact of campaigns on reducing rabies incidence through robust surveillance systems [@lushasi2020, @ferguson2020].
+use of community health workers, could be a way to deliver vaccinations and reduce costs in the hardest-to-reach communities.
+In addition, refining vaccination strategies to local contexts and using improved tools and systems, such as mobile phone based data collection, could improve efficacy and lower costs of campaigns [@mazeri2021; @gibson2016a].
+To monitor the success of these campaigns, it will be critical to develop efficient and effective methods to estimate vaccination coverage and to measure their impact on reducing rabies incidence through robust surveillance [@lushasi2020; @ferguson2020].
 Given it's island context, Madagascar could be a model for rabies elimination, with limited chances for reintroduction from outside the island, and strategies that tackle the context specific challenges could be a path forward for the country to reach ZeroBy30.
@@
 We are grateful to staff and officials at the Department of Veterinary Services, the Ministry of Public Health, and the Institut Pasteur de Madagascar for their support and technical assistance.
-In particular, we thank Angelique Ferreira, Jean Hyacinthe Randrianarisoa, Ranaivoarimanana, Fierenantsoa Randriamahatana, Esther Noiarisaona, Cara Brook, and Christian Ranaivoson, Rila Ratovoson, and Claire LeBlanc.
-The campaign costs were supported by donations from Lush Cosmetics and [Kim to insert].
+In particular, we thank Carmen Barba Claassens, Jean Hyacinthe Randrianarisoa, Ranaivoarimanana, Fierenantsoa Randriamahatana, Esther Noiarisaona, Cara Brook, and Christian Ranaivoson, Rila Ratovoson, and Claire LeBlanc.
+The campaign costs were supported by donations from Lush Cosmetics  Charity Pot.
 Parts of this work were funded by grants from the Center for Health and Wellbeing and the Department of Ecology and Evolutionary Biology at Princeton University to MR.
-MR was supported by an NSF Graduate Research Fellowship and a Princeton Insitute for Regional and International Studies Graduate Fellowship.
+MR was supported by an NSF Graduate Research Fellowship and a Princeton Institute for Regional and International Studies Graduate Fellowship.
 KH is supported by the Wellcome Trust (207569/Z/17/Z).
@@
 We want to particularly acknowledge the contributions to this work by Annie Yang, an undergraduate researcher and volunteer with Princeton University and the Madagascar Dog Initiative who tragically passed away in a car accident before this research was published.
+
+## References

### 2021-02-12 data-raw/process_data.R

@@ -1,0 +1,30 @@
+# Getting spatial data 
+library(sf)
+library(fasterize)
+library(raster)
+library(dplyr)
+
+# Shapefile: filter to Moramanga District
+mada_communes <- read_sf("data-raw/mada_communes/mdg_admbnda_adm3_BNGRC_OCHA_20181031.shp")
+mora_communes <- filter(mada_communes, ADM2_EN == "Moramanga")
+mora_communes$id <- 1:nrow(mora_communes)
+
+# pop raster: mask to Moramanga District
+mada_pop <- raster("data-raw/wp_2015_1x1.tif")
+mora_pop <- crop(mada_pop, mora_communes)
+mora_pop <- mask(mora_pop, mora_communes)
+
+# fasterize & aggregate 
+mora_ids <- fasterize(mora_communes, mora_pop, field = "id")[]
+pop_data <- data.frame(mora_ids, pop = mora_pop[])
+pop_data %>%
+  group_by(mora_ids) %>%
+  summarize(pop = sum(pop, na.rm = TRUE)) -> pop_data
+
+# join to sf & fix names 
+mora_communes %>%
+  left_join(pop_data, by = c("id" = "mora_ids")) %>%
+  select(district = ADM2_EN, commune = ADM3_EN, commcode = ADM3_PCODE, pop) -> pop_sf
+
+# write out to data
+st_write(pop_sf, "data/shapefile/mora_communes.shp")

### 2021-02-20 manuscript.Rmd
cleaning up scripts and incorporating feedback

@@
   - Tara A. Clarke:
-      institute: [maddog]
+      institute: [maddog, ncstate]
   - Zachary Farris:
@@
   - tad: Travelling Animal Doctors, 919 Crossan Rd, Newark, Delaware, USA 
+  - ncstate: North Carolina State University, Department of Sociology and Anthropology, Raleigh, North Carolina, USA
   - appstate: Appalachian State University, Department of Health and Exercise Science, Boone, North Carolina, USA
@@
 output:
+  html_document: default
   bookdown::word_document2:
+    reference_docx: format/reference_main.docx
     pandoc_args:
@@
       - --lua-filter=format/pagebreak.lua
-  html_document: default
+params:
+  outformat: "pdf"
 bibliography: format/references.bib
 ---
-
-<!-- Fix citations without parentheses -->
-<!-- Fix my paper should be 2019 -->
-<!-- Spelling -->
-<!-- Inline stats fixes -->
-<!-- Kim V -->
 
@@
 library(sf)
+library(gt)
 source("R/dem_models.R")
@@
 Canine rabies causes an estimated 60,000 human deaths per year, but these deaths are preventable through post-exposure prophylaxis of people and vaccination of domestic dogs.
-Dog vaccination campaigns targeting coverage of 70% at the population level are effective at interrupting transmission and eventually can lead to elimination of the disease.
-Given this evidence, the WHO and it's partners have set a global goal to eliminate deaths due to canine rabies by the year 2030.
-Here, we report on initial lessons learned during pilot dog vaccination campaigns in the Moramanga District of Madagascar, a country with a high estimated burden of rabies and limited control in the dog population.
-We compare two different vaccination strategies: a larger volunteer driven effort to vaccinate dogs in two communes using static point vaccination, and continuous vaccination as part of routine veterinary services in the district by the District Veterinarian.
-We compared costs, time, personnel, and coverage achieved by each of the campaigns.
-We also use dog age data from the campaigns to estimate key demographic parameters, and use these estimates to simulate different vaccination strategies.
-Overall we found that dog vaccination was feasible and that most dogs were accessible to vaccination.
-We find that there are tradeoffs between the vaccination strategies, with the first campaign achieving higher coverage, but with more resources and with more limited geographic scope compared to the second campaign with continuous delivery.
-Our modeling results suggest that targeting puppies through community based vaccination efforts could help achieve coverage levels sufficient to interrupt transmission.
-Moving forward, our results show that mass dog vaccination is feasible and can achieve high coverage in Madagascar, however innovative strategies and an investment in dog vaccination as a public good will be required to maintain the sustained levels of coverage needed for disease elimination.
+Dog vaccination campaigns targeting 70% of the population are effective at interrupting transmission.
+Here, we report on lessons learned during pilot dog vaccination campaigns in the Moramanga District of Madagascar.
+We compare two different vaccination strategies: a volunteer driven effort to vaccinate dogs in two communes using static point vaccination, and continuous vaccination as part of routine veterinary services.
+We used dog age data from the campaigns to estimate key demographic parameters and to simulate different vaccination strategies.
+Overall, we found that dog vaccination was feasible and that most dogs were accessible to vaccination.
+The static-point campaign achieved higher coverage, but required more resources and had a limited geographic scope compared to the continuous delivery campaign.
+Our modeling results suggest that targeting puppies through community-based vaccination efforts could improve coverage.
+We found that mass dog vaccination is feasible and can achieve high coverage in Madagascar, however context-specific strategies and an investment in dog vaccination as a public good will be required to move the country towards elimination.
 
@@
 These deaths are entirely preventable: prompt post-exposure prophylaxis of humans exposed to rabies is highly effective at preventing death and mass dog vaccination can interrupt transmission in domestic dogs and eventually lead to disease elimination [@fooks2014].
-The WHO and its partners have set a goal to eliminate human deaths due to canine rabies by the year 2030 ('ZeroBy30')[@abela-ridder2016].
-Annual dog vaccination campaigns achieving at least 70% coverage are the recommended target for controlling rabies in the domestic dog population [@worldhealthorganization2013].
-However, achieving this coverage target in low and middle income countries where the burden of human rabies is concentrated can be challenging due to economic, ecological, sociocultural, and political reasons [@fahrion2017].
+The World Health Organization (WHO) and its partners have set a goal to eliminate human deaths due to canine rabies by the year 2030 ('ZeroBy30')[@abela-ridder2016].
+Annual dog vaccination campaigns achieving at least 70% coverage are the recommended target for controlling rabies in domestic dog populations [@worldhealthorganization2013].
+However, achieving this coverage target in low and middle income countries where the burden of human rabies is highest can be challenging due to economic, ecological, sociocultural, and political barriers [@fahrion2017].
 
-In sub-Saharan African countries, parenteral vaccination implemented through static point vaccination campaigns have been shown to be cost-effective and feasible [@borse2018].
-While most dogs in these settings are considered free-roaming, they are mostly owned and are accessible for vaccination through these campaigns [@morters2014, @jibat2015].
-However, reaching high coverage and maintaining vaccination campaigns at scale requires sustained investment and coordination, and the challenges in implementation largely reflect financial and logistical constraints more than feasibility of vaccination itself [@fahrion2017].
-Developing clear and context-specific strategies and lowering costs and resources needed could help spur implementation and scaling up of campaigns in these countries.
+In sub-Saharan African countries, parenteral vaccinations implemented through static point campaigns have been shown to be cost-effective and feasible [@borse2018].
+While most dogs in these settings are considered free roaming, they are mostly owned and are accessible for vaccination through these campaigns [@morters2014; @jibat2015].
+However, reaching high coverage and maintaining vaccination campaigns at scale requires sustained investment and coordination, and the challenges in implementation largely reflect financial and logistical constraints more than the feasibility of vaccination itself [@fahrion2017].
+Developing clear and context-specific strategies and lowering costs and resources needed could help spur the implementation and scaling up of campaigns in these countries.
 
-In Madagascar, canine rabies has been endemic for over a century and over almost the same time period, the Institut Pasteur de Madagascar has provided post-exposure prophylaxis free-of-charge to bite patients in the country [@reynes2011].
-Currently, there are only 31 clinics where these human vaccines are available, and there is minimal dog vaccination due to high costs to owners and lack of availability of the vaccine [@rajeev2018].
+In Madagascar, canine rabies has been endemic for over a century, and for most of that period, the Institut Pasteur de Madagascar has provided post-exposure prophylaxis free-of-charge to bite patients in the country [@reynes2011].
+Currently, there are only 31 clinics where these human vaccines are available, and there is minimal dog vaccination due to high costs to owners and lack of vaccine availability [@rajeev2018].
 Recent studies have estimated a high burden of human deaths, masked by weak surveillance across the country [@rajeev2020; @rajeev2018].
 
-The veterinary sector is largely private, with some veterinarians employed in hybrid private/public employment as designated district veterinarians [comment: Njara/Nivo].
+The veterinary sector is largely private, with 204 veterinarians employed in hybrid private/public employment as designated district veterinarians.
 While dog vaccination is rare, livestock officers and veterinarians work together to implement cattle vaccination campaigns for anthrax on an annual basis as mandated by the government (owners are charged a fee per animal for these vaccines which vary by location) [@mondiale2017evaluation].
@@
 
-Here, we summarize lessons learned through implementation of pilot vaccination programs in the Moramanga District of Madagascar, where previous work has shown high incidence of dog rabies cases and human rabies exposures.
+Here, we summarize lessons learned through the implementation of pilot vaccination programs in the Moramanga District of Madagascar, where previous work has shown high incidence of dog rabies cases and human rabies exposures.
 In 2018 and 2019, we deployed two different vaccination strategies.
-In 2018, we led a larger scale volunteer-led pilot vaccination campaign in two Communes (sub-district level) using a central point strategy.
-In 2019/2020, we provided vaccines, all necessary supplies for vaccine administration, and a per-vaccine fee to the District Veterinarian to vaccinate animals as part of routine services.
+In 2018, we carried out a larger scale volunteer-led pilot vaccination campaign in two communes (sub-district level) using a static point strategy.
+In 2019/2020, we provided vaccines, all necessary supplies for vaccine administration, and a per-vaccine fee to the District Veterinarian to vaccinate animals as part of routine services delivered continuously.
 We compare time, human resources, costs, and coverage estimates between these campaigns, and using a demographic and vaccination model we further explore different vaccination strategies based on what we learned during implementation.
@@
 The Moramanga District is located midway between the central highlands and the east coast of Madagascar, at an average altitude of 936 m.
-It comprises 21 communes, covering approximately 7150 km^2^ with an approximate human population of `r round(sum(mora_sf$pop), -2)`.
+It comprises 21 communes, covering approximately 7150 km^2^ with an approximate human population of `r format(round(sum(mora_sf$pop), -3), big.mark = ",")`[@instat2021].
 Previous work in the district has established a high burden of rabies exposures and deaths despite the availability of post-exposure prophylaxis at the district hospital [@rajeev2018].
-While Moramanga is relatively close to the capital city of Antananarivo (~ 3 hrs by bus), within the district, travel times between locations are highly variable, with much of the population living in more rural areas with limited access to roads and transportation [@rajeev2020].
-Before 2018, there were limited animal rabies vaccination services, with most animal vaccines available in the urban commune of Moramanga Ville and owners charged > 15,000 Ariary (~ 4.28 USD) per vaccine administered.
+While Moramanga is relatively close to the capital city of Antananarivo (~ 3 hours by bus), within the district, travel times between locations are highly variable, with much of the population living in more rural areas with limited access to roads and transportation [@rajeev2020].
+Before 2018, there were limited animal rabies vaccination services, with most animal vaccines available in the urban commune of Moramanga Ville where owners were often charged > 15,000 Ariary (~ 4.28 USD) per vaccine administered.
 
@@
 pts <- read_csv("data/campaign_pts_2018.csv")
-
-# join with vacc #s and clean
-
 ```
 
-In 2018, a pilot vaccination campaign was organized through a collaboration with two NGOs (the Mad Dog Initiative and Traveling Animal Doctors), the Department of Veterinary Services, and the Ministry of Public Health in the District of Moramanga.
+In 2018, two NGOs (the Mad Dog Initiative (MDI) and Traveling Animal Doctors (TAD)) organized a pilot vaccination campaign in collaboration with the Department of Veterinary Services, and the Ministry of Public Health in the District of Moramanga.
 This campaign focused on two communes in Moramanga, Moramanga Ville (the district center) and Andasibe (a rural commune surrounding Andasibe National Park), where previously high incidence of probable rabies exposures (Moramanga Ville) and a high burden of deaths (Andasibe) had been recorded [@rajeev2018].
 
-The campaign was planned as a series of central point vaccination stations covering 1 - 3 sub-communes (or fokontany) per day.
-A week before the campaign dates, the vaccination team advertised the date with the chief of the fokontany and provided posters advertising the date of the vaccine and that it would be available at no cost to owners.
+The campaign was planned as a series of static point vaccination stations covering 1 - 3 fokontany (i.e., sub-communes) per day.
+A week before the campaign dates, the vaccination team informed the chief of the fokontany about the campaign and provided posters advertising the date of the vaccine and that it would be available at no cost to owners.
 During the campaign, we used Rabisin (10 mL vials with 1 mL per dose, Boehringer Ingelheim) to vaccinate both dogs and cats presented that were at least 1 month old.
-As part of the campaign, owners were surveyed by vaccinators about how many dogs and cats they owned in total (split by > 1 yr vs.
-< 1 yr in order to avoid language ambiguities that might result in excluding pups and kittens), as well as if their dogs were free roaming (no restrictions on movement by the owner/'mireny', tied/'mifatora', or fenced/'fefy').
+As part of the campaign, owners were surveyed by vaccinators about how many dogs and cats they owned in total (split by > 1 year vs.
+< 1 year in order to avoid language ambiguities that might result in excluding puppies and kittens), as well as if their dogs were free roaming (no restrictions on movement by the owner/'mirenyreny', tied/'mifatora', or fenced/'mifefy').
 Vaccinations were delivered at no cost to owners, but as animal vaccination is generally thought of as a paid service in Madagascar, owners were asked how much in Ariary they would be willing to pay to have one animal vaccinated against rabies.
@@
 To assess coverage, post-vaccination coverage surveys were conducted according to previously established methodology[@sambo2017; @gibson2015].
-All animals vaccinated were concurrently marked with a colored, non-toxic, livestock crayon [comment: ask Jochem for brand] along the top or back of the head.
+All animals vaccinated were concurrently marked with a colored, non-toxic, livestock crayon along the top of the head or back.
 Between 4 - 6 PM when dogs are most active [@sambo2017] and on the same day as the campaign, in each vaccination location two transect surveys were done.
 Each team (two people per team) walked a transect for 1 hour starting in opposite directions and accompanied by a local guide to ensure that walking paths did not overlap.
-Teams recorded any marked and unmarked dogs observed, their roaming status (whether roaming, inside a fence, or tied), and their approximate age (greater or less than 1 yr of age).
+Teams recorded any marked and unmarked dogs observed, their roaming status (whether roaming, inside a fence, or tied), and their approximate age (greater or less than 1 year of age).
+
+```{r fig1} 
+knitr::include_graphics("figs/fig1.jpeg")
+```
 
@@
 
-For the 2019 campaign, instead of a central point campaign strategy, vaccine vials(Rabisin) and the supplies needed to administer them (needle, syringe, vaccination card for owners) were distributed to the District Veterinarian, who then delivered the vaccination at no cost to owners, but was directly compensated 1,5000 Ar (~ 0.40 USD) per rabies vaccine administered.
+For the 2019 campaign, instead of a static point campaign strategy, vaccine vials(Rabisin) and the supplies needed to administer them (needle, syringe, vaccination card for owners) were distributed to the District Veterinarian, who then delivered the vaccination at no cost to owners, but was directly compensated 1,500 Ar (~ 0.40 USD) per rabies vaccine administered.
 The campaign lasted from September 6, 2019 to June 19, 2020, One week prior to her visiting each location, the District Veterinarian advertised the vaccines by calling ahead to the fokontany leaders and other officials who then advertised to their communities, largely through word-of-mouth.
-For each animal vaccinated, the District Veterinarian collected the age, sex, and also asked owners to approximate the distance they travelled to receive the vaccination in meters.
+For each animal vaccinated, the District Veterinarian collected the animal's age and sex, and also asked owners to approximate the distance they travelled to receive the vaccination in meters.
 Researchers communicated with the District Veterinarian about progress periodically throughout the campaign, primarily through telephone calls.
-No other compensation or instructions were provided, and we asked the District Veterinarian to administer as many (or as few vaccines) as feasible or wanted.
+No other compensation or instructions were provided, and we asked the District Veterinarian to administer as many (or as few) as feasible or wanted.
 As the vaccinations were delivered continuously, we were unable to do comparable post-vaccination surveys.
@@
 
-For the 2018 campaign, we broke costs down into the following categories: direct vaccine costs (cost for vaccine, syringes, needles, vaccination cards), supplies (livestock crayons, muzzles, gloves, alcohol, swabs), food and lodging for NGO personnel and other vaccinators during the campaign, personnel costs (per diems for DSV veterinarian, livestock field officers, local guides, and NGO employees), and advertisement (posters and banners for advertising the campaign).
+For the 2018 campaign, we broke costs down into the following categories: direct vaccine costs (cost for vaccine, syringes, needles, vaccination cards), supplies (livestock crayons, muzzles, gloves, alcohol, swabs), food and lodging for NGO personnel and other vaccinators during the campaign, personnel costs (per diems for the district veterinarian, livestock field officers, local guides, and NGO employees), and advertisement (posters and banners for advertising the campaign).
 Foreign NGO volunteer expenses for travel to Madagascar were not included in these costs.
-Vehicles and drivers are also not included in these costs, as the drivers' time and vehicle use were donated to the campaign by volunteers involved in the campaign [comment: re Katie, transects were really cheap because people vaccinating did them, like ~ 100 bucks for crayons & guides, so will keep it the costs but mention that they are cheap in the discussion].
+Vehicles and drivers are also not included in these costs, as the drivers' time and vehicle use were donated to the campaign by volunteers involved in the campaign.
 In 2019, costs were split into two categories, direct vaccine costs (for same items as in 2018), and personnel costs (per vaccine fee paid to the district veterinarian), and supplies (a generator and fuel for the veterinarian to maintain the vaccine under cold chain during power outages).
-Transportation costs were also not included as the District Veterinarian used her own vehicle and vaccinated a part of their routine veterinary service provisioning.
+Transportation costs were also not included as the District Veterinarian used her own vehicle and vaccinated as part of their routine veterinary service provisioning.
 
@@
 decreasing numbers of animals vaccinated overall.
-[comment maybe add campaign photos (without identifiable people) per Tara]?
+
 #### Coverage estimates
 
-For the 2018 campaign, we used the transect data to estimate coverage as the proportion of dogs sighted that were marked using a binomial confidence interval at the commune level.
-For the 2019 campaign, we estimated vaccination coverage using human-to-dog ratios and human population estimates [@cleaveland2003; @athingo2020].
+For the 2018 campaign, we used the transect data to estimate vaccination coverage as the proportion of dogs sighted that were marked using a binomial confidence interval at the commune level.
+For the 2019 campaign, we estimated vaccination coverage using human-to-dog ratios (HDRs) and human population estimates [@cleaveland2003; @athingo2020].
 We used a range of 8 - 25, based on previous data from Madagascar [@ratsitorahina2009] and based on recent estimates from household surveys in the Moramanga District [@leblancclaireRabiesMadagascarThreepronged2019].
@@
 2019.
-We used human population estimates from WorldPop [@linard2012] in each commune where the vaccinations were delivered.
+We used human population estimates from the 2018 national census in each commune where the vaccinations were delivered [@instat2021].
 Coverage was estimated as the number of dogs vaccinated in total in that commune divided by the estimated dog population.
@@
 
-Using the age data on vaccinated animals collected during both vaccination campaigns, we estimated the proportion of the population in four age classes: pups under the age of 1 year, juveniles 1 - 2 yrs, adults aged 2 - 6 years, and older dogs age 6+ years based on .
+Using the age data on vaccinated animals collected during both vaccination campaigns, we estimated the proportion of the population in four age classes: puppies under the age of 1 year, juveniles aged 1 - 2 yrs, adults aged 2 - 6 years, and older dogs aged 6+ years based on .
 With the assumption that these estimates represent the population at a stable age distribution, we use a Leslie matrix model to estimate annual adult survival probability and fertility using maximum likelihood estimation [@Fujiwara2017].
@@
 
-1)  Annual vaccination campaigns occuring within the same month each year targeting dogs of all ages
+1)  Annual vaccination campaigns occurring within the same month each year targeting dogs of all ages
 2)  Continuous vaccination of new puppies throughout the year targeting puppies that reach the age of 3 months
@@
 
-We split the dog population into pups (< 1 year old) and adults based on the stable age distribution estimated from the demographic model.
-To estimate pup survival in year one, we took the fertility estimates (number of new pups per reproducing dog in the pup age class per year) and divided by an estimate of newborn pups per reproducing dog each year based on average litter size, average number of litters per female per year [@hampson2009; @morters2014; @czupryna2016], and the proportion of the adult population that is female (estimated from our data).
+We split the dog population into puppies (< 1 year old) and adults based on the stable age distribution estimated from the demographic model.
+To estimate pup survival in year one, we took the fertility estimates (i.e., number of new puppies per reproducing dog observed in the pup age class) and divided by an estimate of newborn pups per reproducing dog each year based on average litter size, average number of litters per female per year [@hampson2009; @morters2014; @czupryna2016], and the proportion of the adult population that is female (estimated from our data).
 We assumed that for the annual campaign strategy, surviving vaccinated adult dogs were revaccinated in subsequent years [@delriovilas2017], but that if a pup had been vaccinated within 9 months of the campaign, it was not revaccinated.
-We also assumed that vaccine immunity lasted for a discrete period of 3 years (with revaccination resetting this period of waning)[comment for KH abt clarification here].
+We also assumed that vaccine immunity lasted for a discrete period of 3 years (with revaccination resetting immunity).
 A subset of parameter estimates resulted in estimates of population decline, but based on the shape of the age pyramid and to simulate reasonable campaign scenarios, we filtered to parameter estimates which corresponded to positive population growth.
@@
 The vaccinations were part of a public health campaign and routine veterinary service provisioning carried out by the local veterinary officials, the NGOs involved, and in partnership with the Ministry of Public Health and the Department of Veterinary Services at the national level.
-The Madagascar Dog Initiative also maintains institutional approvals (IACUC # at Institution X)[comment: Kim to fill in] and national research permits (MICET permit #130-19/MEDD/SG/DGEF/DGRNE) for its research and volunteer programs .
+MDI also maintains national research permits (MICET permit: #130-19/MEDD/SG/DGEF/DGRNE) for its research and volunteer programs.
 
@@
 We vaccinated at `r table(pts$Commune)["Andasibe"]` points in Andasibe and `r table(pts$Commune)["Moramanga Ville"]` points in Moramanga Ville.
-During the 2019 campaign, between `r paste(format.Date(range(dmy(animals_2019$Date)), "%B %Y"), collapse = " - ")`, the District Veterinarian vaccinated a total of `r nrow(animals_2019)`animals (`r species_2019["dog"]` dogs and `r species_2019["cat"]`) over 48 days in `r n_distinct(animals_2019$Commune)` communes in the Moramanga District.
+During the 2019 campaign, between `r paste(format.Date(range(dmy(animals_2019$Date)), "%B %Y"), collapse = " - ")`, the District Veterinarian vaccinated a total of `r nrow(animals_2019)`animals (`r species_2019["dog"]` dogs and `r species_2019["cat"]` cats) over 48 days in seven communes in the Moramanga District.
 While more animals were vaccinated per vaccination point and per vaccination day in 2018 compared to 2019, the number of animals vaccinated per person-day was much higher for the 2019 campaign.
@@
 This remained largely the same in 2019 (~ `r summ_tab1$All_2019[summ_tab1$name %in% "prev_vacc"]`), as the District Veterinarian focused their efforts in other communes.
-The District Veterinarian did vaccinate `r prev_vacc_2019$total[prev_vacc_2019$Commune %in% "Moramanga"]` dogs in Moramanga Ville in 2019, and of those `r prev_vacc_2019$perc_vacc[prev_vacc_2019$Commune %in% "Moramanga"]`% had been vaccinated in the previous year's campaign, whereas only `r prev_vacc_2019$perc_vacc[prev_vacc_2019$Commune %in% "Moramanga"]`% of dogs vaccinated in all other communes had any history of previous vaccination.
+The District Veterinarian did vaccinate `r prev_vacc_2019$total[prev_vacc_2019$Commune %in% "Moramanga"]` dogs in Moramanga Ville in 2019, and of those `r prev_vacc_2019$perc_vacc[prev_vacc_2019$Commune %in% "Moramanga"]`% had been vaccinated in the previous year's campaign, whereas only `r prev_vacc_2019$perc_vacc[prev_vacc_2019$Commune %in% "other"]`% of dogs vaccinated in all other communes had any history of previous vaccination.
 In addition, in 2019, `r round(spayed_neutered * 100, 1)`% of animals had been spayed or neutered, reflecting efforts by the Mad Dog Initiative to implement free spay and neuter clinics in the district.
@@
 ```{r}
+if(params$outformat %in% "docx") {
   knitr::include_graphics("figs/table1.png")
+} else {
+  tab1 <- readRDS("out/table1.rds")
+
+  tab1 %>% 
+    gt::as_latex() 
+}
 ```
@@
 
-The 2018 campaign cost more overall and per animal vaccinated than the 2019 campaign, largely due to increased personnel costs (Fig 1A & B, Table 1).
-Reflecting the extra personnel necessary to run the central point campaign, the 2018 campaign also took substantially more person-days per animal vaccinated (Table 1).
-We found that charging owners for vaccination would result in minimal cost recovery (Fig 1C) and beyond a minimal cost would actually result in increased costs per vaccinated individual than free-of-charge campaigns.
-Cost recovery would be more likely given a 2019 style strategy where the majority of the costs are incurred on a per animal basis, compared to the 2018 campaigns where the costs are largely due to setting up the central point vaccination stations (Fig 3A) [comment for KH abt clarification here].
-More importantly, in all cases, even a nominal fee would significantly reduce numbers of dogs vaccinated and thus vaccination coverage, particularly in the rural commune of Andasibe (Fig 1D).
+The 2018 campaign cost more overall and per animal vaccinated than the 2019 campaign, largely due to increased personnel costs (fig 2A & B, Table 1).
+Reflecting the extra personnel necessary to run the static point campaign, the 2018 campaign also took substantially more person-days per animal vaccinated (Table 1).
+We found that charging owners for vaccination would result in minimal cost recovery (fig 2C) and beyond a minimal cost would actually result in increased costs per vaccinated individual than free-of-charge campaigns.
+Cost recovery would be more likely given a 2019 style strategy where the majority of the costs are incurred on a per animal basis, compared to the 2018 campaigns where the costs are largely due to setting up the static point vaccination stations (fig 4A).
+More importantly, in all cases, even a nominal fee would significantly reduce numbers of dogs vaccinated and thus vaccination coverage, particularly in the rural commune of Andasibe (fig 2D).
 
-```{r fig1} 
-knitr::include_graphics("figs/fig1.jpeg")
+```{r fig2} 
+knitr::include_graphics("figs/fig2.jpeg")
 ```
-#### Figure 1. Comparing campaign costs and willingness to pay 
+
+#### Figure 2. Comparing campaign costs and willingness to pay 
 *A) Vaccine costs broken down by category and by year (colors).
@@
 ### Comparing campaign coverage estimates
-The 2018 campaign covered two communes and was estimated to have achieved approximately 60% coverage (Figure 2A).
-The 2019 campaign covered `r n_distinct(animals_2019$Commune)` communes, but was estimated to have achieved lower coverage levels (Figure 2, ranging from 5 - 60%).
+The 2018 campaign covered two communes and was estimated to have achieved approximately 60% coverage (fig 3A).
+The 2019 campaign covered seven communes, but was estimated to have achieved lower coverage levels (fig 3, ranging from 5 - 60%).
 In the 2018 campaign, we used post-vaccination coverage transects to estimate vaccination coverage, but we were unable to do this in 2019 given the continuous delivery strategy.
 In addition, in Andasibe in 2018, coverage estimates were based on a single transect resulting in more uncertainty.
-However, coverage estimated from the transects in 2018 were consistent with the HDR based estimates (for both Andasibe and Moramanga, transect-based estimates fell within the range of the HDR estimates).
+However, coverage estimates from the transects in 2018 were consistent with the HDR based estimates (for both Andasibe and Moramanga, transect-based estimates fell within the range of the HDR estimates).
 We also back-calculated HDRs given our vaccination coverage estimates, and these were similar to the HDRs calculated from the household survey (`r hdr_backcalcs %>% filter(Commune == "Andasibe") %$% glue("{hdr_cov_est_lower} - {hdr_cov_est_upper}")` for Andasibe compared to `r round(hdrs$mora_hdrs[hdrs$loc == "rural"], 1)` in a rural community and `r hdr_backcalcs %>% filter(Commune == "Moramanga") %$% glue("{hdr_cov_est_lower} - {hdr_cov_est_upper}")` for Moramanga Ville compared to `r round(hdrs$mora_hdrs[hdrs$loc == "urban"], 1)` in an urban community).
 
-```{r fig2}
-knitr::include_graphics("figs/fig2.jpeg")
+```{r fig3}
+knitr::include_graphics("figs/fig3.jpeg")
 ```
-#### Figure 2. Estimates of coverage achieved by the 2018 and 2019 campaigns.
+
+#### Figure 3. Estimates of coverage achieved by the 2018 and 2019 campaigns.
 *A) The commune level numbers of dogs vaccinated (size of points) and the associated coverage estimates (color of points) for the year 2018 (estimated using post-vaccination transects) and 2019 (estimated using a human-to-dog ratio (HDR) of `r round(mora_mid_hdr, 1)`, based on a recent household survey in the Moramanga District).
@@
 
-Demographic data from vaccinated dogs shows a population pyramid with a large base, indicative of a fast-growing population, and with a male bias (approximately `r prop_sex["M"]`% male, Figure 3A).
-We fit these data to an age-structured model, and were able to generate parameter estimates which resulted in stable age distributions consistent with the data (Figure 3B).
+Demographic data from vaccinated dogs shows a population pyramid with a large base, indicative of a fast-growing population, and with a male bias (approximately `r prop_sex["M"]`% male, Fig 4A).
+We fit these data to an age-structured model, and were able to generate parameter estimates which resulted in stable age distributions consistent with the data (Fig 4B).
 We filter to parameter estimates that are consistent with a growing population resulting in mean adult annual survival probability of `r dem_pars[dem_pars$name == "psurv_adult_est", "value.mean"]` (95% quantile: `r dem_pars[dem_pars$name == "psurv_adult_est", "ci"]`).
 We use estimates of fertility (on average `r dem_pars[dem_pars$name == "fert_est", "value.mean"]`, 95% quantile: `r dem_pars[dem_pars$name == "fert_est", "ci"]`) to back-calculate pup survival, which ranges from `r round(dem_pars[dem_pars$name == "pup_surv_lower", "value.lower"], 2)` - `r round(dem_pars[dem_pars$name == "pup_surv_lower", "value.upper"], 2)`.
-We find that given these demographic parameters, annual campaigns that target dogs of all ages result in rapid decline in vaccination coverage in between campaigns, largely due to rapid turnover of the dog population (compared to the impact of waning immunity assuming a discrete 3 year period of vaccine immunity, evident in the additional dip at year 3, Fig 3D).
+We find that given these demographic parameters, annual campaigns that target dogs of all ages result in rapid decline in vaccination coverage in between campaigns, largely due to rapid turnover of the dog population (compared to the impact of waning immunity assuming a discrete 3 year period of vaccine immunity, evident in the additional dip at year 3, fig 4D).
 Continuously targeting 70% of the puppy population for vaccination, while unable to achieve the peak coverage, consistently reaches coverage of about 50% of the dog population.
@@
 
-```{r fig3}
-knitr::include_graphics("figs/fig3.jpeg")
+```{r fig4}
+knitr::include_graphics("figs/fig4.jpeg")
 ```
-#### Figure 3. Dog demography and implications for campaign strategies
+
+#### figure 4. Dog demography and implications for campaign strategies
 *A) The age pyramid for vaccinated dogs by sex.
@@
 
-Through vaccination campaigns implemented in Moramanga District of Madagascar, we saw high demand for vaccination from dog owners and found that dogs were accessible and able to be handled efficiently for parenteral vaccination at a reasonable cost (between 1.3 - 1.8 USD per animal vaccinated).
+Through vaccination campaigns implemented in Moramanga District of Madagascar, we saw high demand for vaccination from dog owners and found that dogs were accessible and able to be handled safely and efficiently for parenteral vaccination at a reasonable cost (between 1.3 - 1.8 USD per animal vaccinated).
 We find that providing the vaccine at no direct cost to dog owners will be critical to achieving sufficient coverage, as even with nominal fees, a significant proportion of owners indicated they would no longer vaccinate their animals.
-A central point vaccination strategy achieved higher coverage over a shorter time period in 2018 compared to dog vaccinations conducted by the District Veterinarian as part of routine veterinary service provision in 2019.
+A static point vaccination strategy achieved higher coverage over a shorter time period in 2018 compared to dog vaccinations conducted by the District Veterinarian as part of routine veterinary service provision in 2019.
 However, it came at a higher cost per animal vaccinated, was more limited in geographic scope, and required more resources in terms of personnel.
-In addition, in the rural setting of Andasibe, the central point campaign strategy achieved lower coverage, reflecting less accessible, hard-to-reach human communities in this location.
+In addition, in the rural setting of Andasibe, the static point campaign strategy achieved lower coverage, reflecting less accessible, hard-to-reach human communities in this location.
 Based on the lessons learned through these campaigns, in particular the observation that puppies were relatively easy to handle for vaccinators and owners, we find that continuous vaccination targeting puppies in particular may be an effective way to maintain vaccination coverage levels given high turnover in dog populations.
@@
 Owner reports of willingness to pay, age of animals, and distance travelled to the vaccination point likely all suffer from recall bias and uncertainty, whilst willingness to pay studies are often overestimates of observed practice [@birhane2016].
-In addition, age of animals brought to vaccination points may not be representative of the age structure of the underlying population, and other work has shown that in general puppies (individuals < 1 year) tend to have lower vaccination coverage than adults [@arief2016; @sanchez-soriano2020; @mazeri2019; @minyoo2015].
-We also did not vaccinate animals less than 1 month of age, which may explain why we sometimes estimated declining populations (as part of the age class of 0-1 years was not captured by the data).
+In addition, the age of animals brought to vaccination points may not be representative of the age structure of the underlying population, and other work has shown that in general puppies (individuals < 1 year) tend to have lower vaccination coverage than adults [@arief2016; @sanchez-soriano2020; @mazeri2019; @minyoo2015].
+Additionally, we did not vaccinate animals less than 1 month of age, which may explain why we sometimes estimated declining populations (as part of the age class of 0-1 years was not captured by the data).
 However, our analyses based on these data are consistent with previous findings from SSA that demonstrate male-biased populations, skewed towards puppies, and with high mortality in the first year of life [@czupryna2016; @taylor2017].
 Studies have also shown that charging fees for vaccines can greatly reduce the efficacy of campaigns and are a significant barrier to vaccination for owners [@schildecker2016; @yoak2021; @barbosacosta2020].
-Similarly, distance to campaign points has been identified as a barrier to vaccination, and in most cases owners eport traveling less than 1km to reach a vaccination point [@mazeri2018barriers; @mazeri2019].
+Similarly, distance to campaign points has been identified as a barrier to vaccination, and in most cases owners report traveling less than 1km to reach a vaccination point [@mazeri2018barriers; @mazeri2019].
 We used HDRs, which can be sensitive to underlying estimates of the human population and the spatial scale of estimation [@sambo2018estimating; @sambo2017].
 However, we used HDR estimates from a recent household survey study in the district, and found these to give coverage estimates consistent with those from post-vaccination transects.
-Finally, in our vaccination model, we  make several simplifying assumptions: we assume that immunity wanes after three years (simulating vaccination with a long-lasting vaccine such as Nobivac, compared to the manufacturer stated one year efficacy for Rabisin [comment for KH BI states this on their vials]); we assume that vaccinated dogs that survive to the following year are revaccinated in subsequent campaigns; and in our age-structured models we do not account for population carrying capacity, but rather simulate over a short time frame over which population growth remains reasonable.
+Finally, in our vaccination model, we make several simplifying assumptions: we assume that immunity wanes after three years (simulating vaccination with a long-lasting vaccine such as Nobivac), but we also assume that vaccinated dogs that survive to the following year are revaccinated in subsequent campaigns (effectively assuming boosting per manufacturer recommendations for Rabisin).
+Overall waning vaccinal immunity plays a lesser role in declining vaccination coverage given the high population turnover in this context.
+In our age-structured models, we also do not account for population carrying capacity, which likely overestimates growth of the dog population.
 
-Our estimates of costs per animal vaccinated are in line with recent estimates from other settings [@undurraga2020], although we likely underestimate costs given the donation of time and resources by the organizations and individuals involved (i.e.
+Our estimates of costs per animal vaccinated are in line with recent estimates from other countries [@undurraga2020], although we likely underestimate costs given the donation of time and resources by the organizations and individuals involved (i.e.
 costs associated with international volunteers including airfare and visa costs as well as costs of transportation donated to the campaign).
 We also did not include costs of pre-exposure prophylaxis as all of our volunteers and vaccinators had been vaccinated prior to the campaigns.
-Both pre- and post-exposure vaccines should however always be readily available for vaccinators and be included in vaccination program budgets.
-While the volunteer-led effort resulted in significant financial and personnel resources being devoted to the campaign, costs were higher per animal vaccinated and overall.
+Both pre- and post-exposure vaccines should however always be readily available for vaccinators and should be included in vaccination program budgets.
+While the volunteer-led effort resulted in significant financial and personnel resources being devoted to the campaign, costs were higher overall and per animal vaccinated.
 In addition, NGO and volunteer based campaigns may be difficult to sustain given unpredictable funding, time commitments, and turnover in staff [@ferguson2020].
 Dog vaccination delivered through the District Veterinarian was less costly, with the majority of the costs directly related to vaccination.
+Although we included costs of implementing transect based coverage estimates, these were negligible (less than 0.10 USD per dog vaccinated), in line with recent studies which have shown this strategy is a cheap and effective way to estimate coverage @sambo2017; @gibson2015].
 
@@
 If dog population growth is driven by survival, then this could mean that increased vaccination results in increased population growth.
-However, if growth is driven more by demand from human communities, then vaccination could stabilize the population and reduce turnover of the population.
+However, if growth is driven more by demand from human communities, then vaccination could stabilize the population and reduce population turnover.
 Improving dog population management, encouraging responsible pet ownership practices, and increasing veterinary services could all complement vaccination efforts, but have not been demonstrated to result in meaningful rabies control without parallel dog vaccination [@taylor2017].
 
-In settings with high dog ownership, moving towards community-based vaccination strategies could be an effective way to achieve sufficient and consistent coverage, particulary in hard-to-reach communities.
+In settings with high dog ownership, moving towards community-based vaccination strategies could be an effective way to achieve sufficient and consistent coverage, particularly in hard-to-reach communities.
 During our campaign, we found that puppies at approximately 3 months old were easy to handle for both vaccinators and owners.
-Puppy vaccination could be carried out by local officials embedded in communities (along the lines of community health workers), particularly given recent findings on thermotolerance of rabies vaccines and locally manufactured methods for maintaining temperatures required for sustained vaccine storage (up to 3 months)
+Puppy vaccination could be carried out by local officials embedded in communities (along the lines of community health workers), especially given recent findings on thermotolerance of rabies vaccines and locally manufactured methods for maintaining temperatures required for sustained vaccine storage (up to 3 months)
 [@lugelo2020].
@@
 Overall, our results suggest that dog vaccination is a feasible and effective strategy for controlling canine rabies in Madagascar.
-However, rabies vaccination must be recognized as a public good, and strategies to remove barriers for dog owners and to incentivize veterinarians and other animal health workers to implement vaccination will be key to campaign success.
+However, rabies vaccination must be recognized as a public good.
+Strategies to remove barriers for dog owners and to incentivize veterinarians and other animal health workers to implement vaccination will be key to long-term campaign success.
 Borrowing strategies from human vaccination efforts, i.e.
-use of community health workers, could be a way to deliver vaccinations and reduce costs in the hardest-to-reach communities.
+deploying community health workers, could be a way to deliver vaccinations and reduce costs in the hardest-to-reach communities.
 In addition, refining vaccination strategies to local contexts and using improved tools and systems, such as mobile phone based data collection, could improve efficacy and lower costs of campaigns [@mazeri2021; @gibson2016a].
 To monitor the success of these campaigns, it will be critical to develop efficient and effective methods to estimate vaccination coverage and to measure their impact on reducing rabies incidence through robust surveillance [@lushasi2020; @ferguson2020].
-Given it's island context, Madagascar could be a model for rabies elimination, with limited chances for reintroduction from outside the island, and strategies that tackle the context specific challenges could be a path forward for the country to reach ZeroBy30.
+With limited chances for reintroduction from outside the island, Madagascar could be a model for rabies elimination, and strategies that tackle the context specific challenges could be a path forward for the country to reach ZeroBy30.
 
@@
 
-We want to particularly acknowledge the contributions to this work by Annie Yang, an undergraduate researcher and volunteer with Princeton University and the Madagascar Dog Initiative who tragically passed away in a car accident before this research was published.
+We want to particularly acknowledge the contributions to this work by Annie Yang, an undergraduate researcher and volunteer with Princeton University and the Mad Dog Initiative who tragically passed away in a car accident before this research was published.
 

### 2021-02-20 R/figure2.R

@@ -88,9 +88,6 @@ vacc_2018 %>%
          type = "hdr") %>%
   bind_rows(rename(cov_commune_2018, Commune = commune)) -> cov_commune_2018
 
-# back calculate hdr from cov est
-cov_commune
-
 # 2019 coverage estimates
 vacc_2019 %>%
   filter(`Dog or Cat` == "dog") %>%

### 2021-02-20 R/run_analyses_all.R

@@ -1,0 +1,17 @@
+# Main script to run all analyses
+
+# helper function to source files with invisible output
+run_script <- function(path) {
+  source(here::here(path), local = environment())
+  return(paste(path, "is done"))
+}
+
+# First run models
+run_script("R/run_dem_mod.R") # will take a few minutes!
+run_script("R/run_vacc_mod.R")
+
+# Then make the figures & tables
+run_script("R/figure1.R")
+run_script("R/Table1.R")
+run_script("R/figure2.R")
+run_script("R/figure3.R")

### 2021-02-20 data-raw/process_data.R

@@ -1,7 +1,5 @@
 # Getting spatial data 
 library(sf)
-library(fasterize)
-library(raster)
 library(dplyr)
 
 # Shapefile: filter to Moramanga District
@@ -9,22 +7,40 @@ mada_communes <- read_sf("data-raw/mada_communes/mdg_admbnda_adm3_BNGRC_OCHA_201
 mora_communes <- filter(mada_communes, ADM2_EN == "Moramanga")
 mora_communes$id <- 1:nrow(mora_communes)
 
-# pop raster: mask to Moramanga District
-mada_pop <- raster("data-raw/wp_2015_1x1.tif")
-mora_pop <- crop(mada_pop, mora_communes)
-mora_pop <- mask(mora_pop, mora_communes)
+# pop data from census
+# link to release: https://www.instat.mg/recensement-general-de-la-population-et-de-lhabitat-rgph/resultats-definitifs-rgph-3/
+# and tome 2 pg. 67: https://www.instat.mg/wp-content/uploads/Resultat-globaux-RGPH3-Tome-02.pdf
+pop <- c(14836L, 6788L, 15835L, 17545L, 8823L, 13206L, 35923L, 7566L, 7679L, 17418L, 
+         9219L, 12755L, 14493L, 7339L, 10374L, 9023L, 20380L, 
+         21609L, 57084L, 19656L, 6036L, 13601L, 3536L)
+census_name <- c("Fierenana", "Moramanga", "Mandialaza", "Morarano Gara", 
+                 "Ampasipotsy Mandialaza", "Ambohidronono", "Ambohibary", 
+                 "Belavabary", "Anosibe Ifody", "Sabotsy Anjiro", 
+                 "Vodiriana", "Antanandava", "Mangarivotra", 
+                 "Ambatovola", "Andasibe", "Ampasipotsy Gara", 
+                 "Beforona", "Lakato", "Moramanga",
+                 "Andaingo", "Bembary", "Amboasary", "Analasoa")  
 
-# fasterize & aggregate 
-mora_ids <- fasterize(mora_communes, mora_pop, field = "id")[]
-pop_data <- data.frame(mora_ids, pop = mora_pop[])
+# census report error on page 68 (second, smaller Moramanga is likely Antanaditra)
+# also reported Analasoa as a separate commune but considered in shapefile as part of Morarano Gara
+# similarly for Bembary considered a part of Andaingo
[…]

### 2021-02-20 paper/mdpi_paper.Rmd
adding rticles formatting

---
title: Full title of the paper (Capitalized)
author:
  - name: Dominik Leutnant
    affil: 1,2,\ddagger,*
    orcid: 0000-0003-3293-2315
  - name: John Doe
    affil: 2, \dagger, \ddagger
affiliation:
  - num: 1
    address: |
      Muenster University of Applied Sciences - 
      Institute for Infrastructure, Water, Resources, Environment
      Correnstr. 25, 48149 Muenster, Germany
    email: leutnant@fh-muenster.de
  - num: 2
    address: |
      Your department
      Street, City, Country
    email: mail@mail.com
# firstnote to eighthnote
firstnote: |
  Current address: Updated affiliation
secondnote: |
  These authors contributed equally to this work.
correspondence: |
  leutnant@fh-muenster.de; Tel.: +XX-000-00-0000.
journal: water
type: article
status: submit
bibliography: mybibfile.bib
appendix: appendix.tex
simplesummary: |
  A Simple summary goes here.
abstract: |
  Canine rabies causes an estimated 60,000 human deaths per year, but these 
  deaths are preventable through post-exposure prophylaxis of people and 
  vaccination of domestic dogs. Dog vaccination campaigns targeting 70% of 
  the population are effective at interrupting transmission. Here, we report on 
  lessons learned during pilot dog vaccination campaigns in the Moramanga 
  District of Madagascar. We compare two different vaccination strategies: 
  a volunteer driven effort to vaccinate dogs in two communes using static 
  point vaccination, and continuous vaccination as part of routine veterinary 
  services. We used dog age data from the campaigns to estimate key demographic 
  parameters and to simulate different vaccination strategies. Overall, 
  we found that dog vaccination was feasible and that most dogs were accessible to vaccination. 
  The static-point campaign achieved higher coverage, but required more 
  resources and had a limited geographic scope compared to the continuous 
  delivery campaign. Our modeling results suggest that targeting puppies 
  through community-based vaccination efforts could improve coverage. We 
  found that mass dog vaccination is feasible and can achieve high coverage 
  in Madagascar, however context-specific strategies and an investment in dog 
  vaccination as a public good will be required to move the country towards 
  elimination. 
keywords: |
  keyword 1; keyword 2; keyword 3 (list three to ten pertinent keywords specific 
  to the article, yet reasonably common within the subject discipline.).
acknowledgement: |
  All sources of funding of the study should be disclosed. Please clearly 
  indicate grants that you have received in support of your research work. 
  Clearly state if you received funds for covering the costs to publish in open 
  access.
authorcontributions: |
  For research articles with several authors, a short paragraph specifying their 
  individual contributions must be provided. The following statements should be 
  used ``X.X. and Y.Y. conceive and designed the experiments; X.X. performed the 
  experiments; X.X. and Y.Y. analyzed the data; W.W. contributed 
  reagents/materials/analysis tools; Y.Y. wrote the paper.'' Authorship must be
  limited to those who have contributed substantially to the work reported.
conflictsofinterest: |
  Declare conflicts of interest or state 'The authors declare no conflict of 
  interest.' Authors must identify and declare any personal circumstances or
  interest that may be perceived as inappropriately influencing the
  representation or interpretation of reported research results. Any role of the
  funding sponsors in the design of the study; in the collection, analyses or 
  interpretation of data in the writing of the manuscript, or in the decision to 
  publish the results must be declared in this section. If there is no role, 
  please state 'The founding sponsors had no role in the design of the study; 
  in the collection, analyses, or interpretation of data; in the writing of the 
  manuscript, an in the decision to publish the results'.
sampleavailability: |
  Samples of the compounds ...... are available from the authors.
abbreviations:
  - short: MDPI
    long: Multidisciplinary Digital Publishing Institute
  - short: DOAJ
    long: Directory of open access journals
  - short: TLA
    long: Three letter acronym
  - short: LD 
    long: linear dichroism
output: rticles::mdpi_article
---

# Version

This Rmd-skeleton uses the mdpi Latex template published 2019/02. 
However, the official template gets more frequently updated than the 'rticles'
package. Therefore, please make sure prior to paper submission, that you're 
using the most recent .cls, .tex and .bst files 
(available [here](http://www.mdpi.com/authors/latex)).

# Introduction

The introduction should briefly place the study in a broad context and highlight why it is important. It should define the purpose of the work and its significance. The current state of the research field should be reviewed carefully and key publications cited. Please highlight controversial and diverging hypotheses when necessary. Finally, briefly mention the main aim of the work and highlight the principal conclusions. As far as possible, please keep the introduction comprehensible to scientists outside your particular field of research. Citing a journal paper [@bertrand-krajewski_distribution_1998; @leutnant_stormwater_2016]. And now citing a book reference @gujer_systems_2008. Please use the command [@leutnant_stormwater_2016] for the following MDPI journals, which use author-date citation: Administrative Sciences, Arts, Econometrics, Economies, Genealogy, Humanities, IJFS, JRFM, Laws, Religions, Risks, Social Sciences.

# Materials and Methods

Materials and Methods should be described with sufficient details to allow others to replicate and build on published results. Please note that publication of your manuscript implicates that you must make all materials, data, computer code, and protocols associated with the publication available to readers. Please disclose at the submission stage any restrictions on the availability of materials or information. New methods and protocols should be described in detail while well-established methods can be briefly described and appropriately cited.

Research manuscripts reporting large datasets that are deposited in a publicly available database should specify where the data have been deposited and provide the relevant accession numbers. If the accession numbers have not yet been obtained at the time of submission, please state that they will be provided during review. They must be provided prior to publication.

Interventionary studies involving animals or humans, and other studies require ethical approval must list the authority that provided approval and the corresponding ethical approval code.

# Results

This section may be divided by subheadings. It should provide a concise and precise description of the experimental results, their interpretation as well as the experimental conclusions that can be drawn.

## Subsection Heading Here

Subsection text here.

### Subsubsection Heading Here

Bulleted lists look like this:

* First bullet
* Second bullet
* Third bullet

Numbered lists can be added as follows:

1. First item
2. Second item
3. Third item

The text continues here.

All figures and tables should be cited in the main text as Figure 1, Table 1, etc.

\begin{figure}[H]
\centering
\includegraphics[width=3 cm]{logo-mdpi}
\caption{This is a figure, Schemes follow the same formatting. If there are multiple panels, they should be listed as: (\textbf{a}) Description of what is contained in the first panel. (\textbf{b}) Description of what is contained in the second panel. Figures should be placed in the main text near to the first time they are cited. A caption on a single line should be centered.}
\end{figure}

\begin{table}[H]
\caption{This is a table caption. Tables should be placed in the main text near to the first time they are cited.}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{ccc}
\toprule
\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}\\
\midrule
entry 1		& data			& data\\
entry 2		& data			& data\\
\bottomrule
\end{tabular}
\end{table}

This is an example of an equation:

\begin{equation}
\mathbb{S}
\end{equation}
<!-- If the documentclass option "submit" is chosen, please insert a blank line before and after any math environment (equation and eqnarray environments). This ensures correct linenumbering. The blank line should be removed when the documentclass option is changed to "accept" because the text following an equation should not be a new paragraph. -->

<!-- Please punctuate equations as regular text. Theorem-type environments (including propositions, lemmas, corollaries etc.) can be formatted as follows: -->
Example of a theorem:
\begin{Theorem}
Example text of a theorem.
\end{Theorem}

The text continues here. Proofs must be formatted as follows:

Example of a proof:
\begin{proof}[Proof of Theorem 1]
Text of the proof. Note that the phrase `of Theorem 1' is optional if it is clear which theorem is being referred to.
\end{proof}
The text continues here.

# Discussion

Authors should discuss the results and how they can be interpreted in perspective of previous studies and of the working hypotheses. The findings and their implications should be discussed in the broadest context possible. Future research directions may also be highlighted.

# Conclusion

This section is not mandatory, but can be added to the manuscript if the discussion is unusually long or complex.

# Patents

This section is not mandatory, but may be added if there are patents resulting from the work reported in this manuscript.

### 2021-02-20 paper/mdpi_paper.tex
adding rticles formatting

%  LaTeX support: latex@mdpi.com
%  In case you need support, please attach all files that are necessary for compiling as well as the log file, and specify the details of your LaTeX setup (which operating system and LaTeX version / tools you are using).

%=================================================================
\documentclass[water,article,submit,moreauthors,pdftex]{mdpi}

% If you would like to post an early version of this manuscript as a preprint, you may use preprint as the journal and change 'submit' to 'accept'. The document class line would be, e.g., \documentclass[preprints,article,accept,moreauthors,pdftex]{mdpi}. This is especially recommended for submission to arXiv, where line numbers should be removed before posting. For preprints.org, the editorial staff will make this change immediately prior to posting.

%% Some pieces required from the pandoc template
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{4pt}}
\setlist[itemize]{leftmargin=*,labelsep=5.8mm}
\setlist[enumerate]{leftmargin=*,labelsep=4.9mm}

\usepackage{longtable}

% see https://stackoverflow.com/a/47122900

%--------------------
% Class Options:
%--------------------
%----------
% journal
%----------
% Choose between the following MDPI journals:
% acoustics, actuators, addictions, admsci, aerospace, agriculture, agriengineering, agronomy, algorithms, animals, antibiotics, antibodies, antioxidants, applsci, arts, asc, asi, atmosphere, atoms, axioms, batteries, bdcc, behavsci , beverages, bioengineering, biology, biomedicines, biomimetics, biomolecules, biosensors, brainsci , buildings, cancers, carbon , catalysts, cells, ceramics, challenges, chemengineering, chemistry, chemosensors, children, cleantechnol, climate, clockssleep, cmd, coatings, colloids, computation, computers, condensedmatter, cosmetics, cryptography, crystals, dairy, data, dentistry, designs , diagnostics, diseases, diversity, drones, econometrics, economies, education, electrochem, electronics, energies, entropy, environments, epigenomes, est, fermentation, fibers, fire, fishes, fluids, foods, forecasting, forests, fractalfract, futureinternet, futurephys, galaxies, games, gastrointestdisord, gels, genealogy, genes, geohazards, geosciences, geriatrics, hazardousmatters, healthcare, heritage, highthroughput, horticulturae, humanities, hydrology, ijerph, ijfs, ijgi, ijms, ijns, ijtpp, informatics, information, infrastructures, inorganics, insects, instruments, inventions, iot, j, jcdd, jcm, jcp, jcs, jdb, jfb, jfmk, jimaging, jintelligence, jlpea, jmmp, jmse, jnt, jof, joitmc, jpm, jrfm, jsan, land, languages, laws, life, literature, logistics, lubricants, machines, magnetochemistry, make, marinedrugs, materials, mathematics, mca, medicina, medicines, medsci, membranes, metabolites, metals, microarrays, micromachines, microorganisms, minerals, modelling, molbank, molecules, mps, mti, nanomaterials, ncrna, neuroglia, nitrogen, notspecified, nutrients, ohbm, particles, pathogens, pharmaceuticals, pharmaceutics, pharmacy, philosophies, photonics, physics, plants, plasma, polymers, polysaccharides, preprints , proceedings, processes, proteomes, psych, publications, quantumrep, quaternary, qubs, reactions, recycling, religions, remotesensing, reports, resources, risks, robotics, safety, sci, scipharm, sensors, separations, sexes, signals, sinusitis, smartcities, sna, societies, socsci, soilsystems, sports, standards, stats, surfaces, surgeries, sustainability, symmetry, systems, technologies, test, toxics, toxins, tropicalmed, universe, urbansci, vaccines, vehicles, vetsci, vibration, viruses, vision, water, wem, wevj

%---------
% article
%---------
% The default type of manuscript is "article", but can be replaced by:
% abstract, addendum, article, benchmark, book, bookreview, briefreport, casereport, changes, comment, commentary, communication, conceptpaper, conferenceproceedings, correction, conferencereport, expressionofconcern, extendedabstract, meetingreport, creative, datadescriptor, discussion, editorial, essay, erratum, hypothesis, interestingimages, letter, meetingreport, newbookreceived, obituary, opinion, projectreport, reply, retraction, review, perspective, protocol, shortnote, supfile, technicalnote, viewpoint
% supfile = supplementary materials

%----------
% submit
%----------
% The class option "submit" will be changed to "accept" by the Editorial Office when the paper is accepted. This will only make changes to the frontpage (e.g., the logo of the journal will get visible), the headings, and the copyright information. Also, line numbering will be removed. Journal info and pagination for accepted papers will also be assigned by the Editorial Office.

%------------------
% moreauthors
%------------------
% If there is only one author the class option oneauthor should be used. Otherwise use the class option moreauthors.

%---------
% pdftex
%---------
% The option pdftex is for use with pdfLaTeX. If eps figures are used, remove the option pdftex and use LaTeX and dvi2pdf.

%=================================================================
\firstpage{1}
\makeatletter
\setcounter{page}{\@firstpage}
\makeatother
\pubvolume{xx}
\issuenum{1}
\articlenumber{5}
\pubyear{2019}
\copyrightyear{2019}
%\externaleditor{Academic Editor: name}
\history{Received: date; Accepted: date; Published: date}
\updates{yes} % If there is an update available, un-comment this line

%% MDPI internal command: uncomment if new journal that already uses continuous page numbers
%\continuouspages{yes}

%------------------------------------------------------------------
% The following line should be uncommented if the LaTeX file is uploaded to arXiv.org
%\pdfoutput=1

%=================================================================
% Add packages and commands here. The following packages are loaded in our class file: fontenc, calc, indentfirst, fancyhdr, graphicx, lastpage, ifthen, lineno, float, amsmath, setspace, enumitem, mathpazo, booktabs, titlesec, etoolbox, amsthm, hyphenat, natbib, hyperref, footmisc, geometry, caption, url, mdframed, tabto, soul, multirow, microtype, tikz

%=================================================================
%% Please use the following mathematics environments: Theorem, Lemma, Corollary, Proposition, Characterization, Property, Problem, Example, ExamplesandDefinitions, Hypothesis, Remark, Definition
%% For proofs, please use the proof environment (the amsthm package is loaded by the MDPI class).

%=================================================================
% Full title of the paper (Capitalized)
\Title{Full title of the paper (Capitalized)}

% Authors, for the paper (add full first names)
\Author{Dominik
Leutnant$^{1,2,\ddagger,*}$\href{https://orcid.org/0000-0003-3293-2315}{\orcidicon}, John
Doe$^{2, \dagger, \ddagger}$}

% Authors, for metadata in PDF
\AuthorNames{Dominik Leutnant, John Doe}

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address{%
$^{1}$ \quad Muenster University of Applied Sciences - Institute for
Infrastructure, Water, Resources, Environment Correnstr. 25, 48149
Muenster,
Germany; \href{mailto:leutnant@fh-muenster.de}{\nolinkurl{leutnant@fh-muenster.de}}\\
$^{2}$ \quad Your department Street, City,
Country; \href{mailto:mail@mail.com}{\nolinkurl{mail@mail.com}}\\
}
% Contact information of the corresponding author
\corres{Correspondence: \href{mailto:leutnant@fh-muenster.de}{\nolinkurl{leutnant@fh-muenster.de}};
Tel.: +XX-000-00-0000.}

% Current address and/or shared authorship
\firstnote{Current address: Updated affiliation}
\secondnote{These authors contributed equally to this work.}

% The commands \thirdnote{} till \eighthnote{} are available for further notes

% Simple summary
\simplesumm{A Simple summary goes here.}

% Abstract (Do not insert blank lines, i.e. \\)
\abstract{Canine rabies causes an estimated 60,000 human deaths per
year, but these deaths are preventable through post-exposure prophylaxis
of people and vaccination of domestic dogs. Dog vaccination campaigns
targeting 70\% of the population are effective at interrupting
transmission. Here, we report on lessons learned during pilot dog
vaccination campaigns in the Moramanga District of Madagascar. We
compare two different vaccination strategies: a volunteer driven effort
to vaccinate dogs in two communes using static point vaccination, and
continuous vaccination as part of routine veterinary services. We used
dog age data from the campaigns to estimate key demographic parameters
and to simulate different vaccination strategies. Overall, we found that
dog vaccination was feasible and that most dogs were accessible to
vaccination. The static-point campaign achieved higher coverage, but
required more resources and had a limited geographic scope compared to
the continuous delivery campaign. Our modeling results suggest that
targeting puppies through community-based vaccination efforts could
improve coverage. We found that mass dog vaccination is feasible and can
achieve high coverage in Madagascar, however context-specific strategies
and an investment in dog vaccination as a public good will be required
to move the country towards elimination.}

% Keywords
\keyword{keyword 1; keyword 2; keyword 3 (list three to ten pertinent
keywords specific to the article, yet reasonably common within the
subject discipline.).}

% The fields PACS, MSC, and JEL may be left empty or commented out if not applicable
%\PACS{J0101}
%\MSC{}
%\JEL{}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Diversity
%\LSID{\url{http://}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Applied Sciences:
%\featuredapplication{Authors are encouraged to provide a concise description of the specific application or a potential application of the work. This section is not mandatory.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Data:
%\dataset{DOI number or link to the deposited data set in cases where the data set is published or set to be published separately. If the data set is submitted and will be published as a supplement to this paper in the journal Data, this field will be filled by the editors of the journal. In this case, please make sure to submit the data set as a supplement when entering your manuscript into our manuscript editorial system.}

%\datasetlicense{license under which the data set is made available (CC0, CC-BY, CC-BY-SA, CC-BY-NC, etc.)}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Only for the journal Toxins
%\keycontribution{The breakthroughs or highlights of the manuscript. Authors can write one or two sentences to describe the most important part of the paper.}

%\setcounter{secnumdepth}{4}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Pandoc citation processing

\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\hypertarget{version}{%
\section{Version}\label{version}}

This Rmd-skeleton uses the mdpi Latex template published 2019/02.
However, the official template gets more frequently updated than the
`rticles' package. Therefore, please make sure prior to paper
submission, that you're using the most recent .cls, .tex and .bst files
(available \href{http://www.mdpi.com/authors/latex}{here}).

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

The introduction should briefly place the study in a broad context and
highlight why it is important. It should define the purpose of the work
and its significance. The current state of the research field should be
reviewed carefully and key publications cited. Please highlight
controversial and diverging hypotheses when necessary. Finally, briefly
mention the main aim of the work and highlight the principal
conclusions. As far as possible, please keep the introduction
comprehensible to scientists outside your particular field of research.
Citing a journal paper
\citep{bertrand-krajewski_distribution_1998, leutnant_stormwater_2016}.
And now citing a book reference \citet{gujer_systems_2008}. Please use
the command \citep{leutnant_stormwater_2016} for the following MDPI
journals, which use author-date citation: Administrative Sciences, Arts,
Econometrics, Economies, Genealogy, Humanities, IJFS, JRFM, Laws,
Religions, Risks, Social Sciences.

\hypertarget{materials-and-methods}{%
\section{Materials and Methods}\label{materials-and-methods}}

Materials and Methods should be described with sufficient details to
allow others to replicate and build on published results. Please note
that publication of your manuscript implicates that you must make all
materials, data, computer code, and protocols associated with the
publication available to readers. Please disclose at the submission
stage any restrictions on the availability of materials or information.
New methods and protocols should be described in detail while
well-established methods can be briefly described and appropriately
cited.

Research manuscripts reporting large datasets that are deposited in a
publicly available database should specify where the data have been
deposited and provide the relevant accession numbers. If the accession
numbers have not yet been obtained at the time of submission, please
state that they will be provided during review. They must be provided
prior to publication.

Interventionary studies involving animals or humans, and other studies
require ethical approval must list the authority that provided approval
and the corresponding ethical approval code.

\hypertarget{results}{%
\section{Results}\label{results}}

This section may be divided by subheadings. It should provide a concise
and precise description of the experimental results, their
interpretation as well as the experimental conclusions that can be
drawn.

\hypertarget{subsection-heading-here}{%
\subsection{Subsection Heading Here}\label{subsection-heading-here}}

Subsection text here.

\hypertarget{subsubsection-heading-here}{%
\subsubsection{Subsubsection Heading
Here}\label{subsubsection-heading-here}}

Bulleted lists look like this:

\begin{itemize}
\tightlist
\item
  First bullet
\item
  Second bullet
\item
  Third bullet
\end{itemize}

Numbered lists can be added as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  First item
\item
  Second item
\item
  Third item
\end{enumerate}

The text continues here.

All figures and tables should be cited in the main text as Figure 1,
Table 1, etc.

\begin{figure}[H]
\centering
\includegraphics[width=3 cm]{logo-mdpi}
\caption{This is a figure, Schemes follow the same formatting. If there are multiple panels, they should be listed as: (\textbf{a}) Description of what is contained in the first panel. (\textbf{b}) Description of what is contained in the second panel. Figures should be placed in the main text near to the first time they are cited. A caption on a single line should be centered.}
\end{figure}

\begin{table}[H]
\caption{This is a table caption. Tables should be placed in the main text near to the first time they are cited.}
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{ccc}
\toprule
\textbf{Title 1}    & \textbf{Title 2}  & \textbf{Title 3}\\
\midrule
entry 1     & data          & data\\
entry 2     & data          & data\\
\bottomrule
\end{tabular}
\end{table}

This is an example of an equation:

\begin{equation}
\mathbb{S}
\end{equation}

Example of a theorem:

\begin{Theorem}
Example text of a theorem.
\end{Theorem}

The text continues here. Proofs must be formatted as follows:

Example of a proof:

\begin{proof}[Proof of Theorem 1]
Text of the proof. Note that the phrase `of Theorem 1' is optional if it is clear which theorem is being referred to.
\end{proof}

The text continues here.

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

Authors should discuss the results and how they can be interpreted in
perspective of previous studies and of the working hypotheses. The
findings and their implications should be discussed in the broadest
context possible. Future research directions may also be highlighted.

\hypertarget{conclusion}{%
\section{Conclusion}\label{conclusion}}

This section is not mandatory, but can be added to the manuscript if the
discussion is unusually long or complex.

\hypertarget{patents}{%
\section{Patents}\label{patents}}

This section is not mandatory, but may be added if there are patents
resulting from the work reported in this manuscript.

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% %% optional
% \supplementary{The following are available online at www.mdpi.com/link, Figure S1: title, Table S1: title, Video S1: title.}
%
% % Only for the journal Methods and Protocols:
% % If you wish to submit a video article, please do so with any other supplementary material.
% % \supplementary{The following are available at www.mdpi.com/link: Figure S1: title, Table S1: title, Video S1: title. A supporting video article is available at doi: link.}

\vspace{6pt}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\acknowledgments{All sources of funding of the study should be
disclosed. Please clearly indicate grants that you have received in
support of your research work. Clearly state if you received funds for
covering the costs to publish in open access.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\authorcontributions{For research articles with several authors, a short
paragraph specifying their individual contributions must be provided.
The following statements should be used ``X.X. and Y.Y. conceive and
designed the experiments; X.X. performed the experiments; X.X. and Y.Y.
analyzed the data; W.W. contributed reagents/materials/analysis tools;
Y.Y. wrote the paper.'\,' Authorship must be limited to those who have
contributed substantially to the work reported.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\conflictsofinterest{Declare conflicts of interest or state `The authors
declare no conflict of interest.' Authors must identify and declare any
personal circumstances or interest that may be perceived as
inappropriately influencing the representation or interpretation of
reported research results. Any role of the funding sponsors in the
design of the study; in the collection, analyses or interpretation of
data in the writing of the manuscript, or in the decision to publish the
results must be declared in this section. If there is no role, please
state `The founding sponsors had no role in the design of the study; in
the collection, analyses, or interpretation of data; in the writing of
the manuscript, an in the decision to publish the results'.}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional
\abbreviations{The following abbreviations are used in this manuscript:\\

\noindent
\begin{tabular}{@{}ll}
MDPI & Multidisciplinary Digital Publishing Institute \\
DOAJ & Directory of open access journals \\
TLA & Three letter acronym \\
LD & linear dichroism \\
\end{tabular}}

\input{"appendix.tex"}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Citations and References in Supplementary files are permitted provided that they also appear in the reference list here.

%=====================================
% References, variant A: internal bibliography
%=====================================
%\reftitle{References}
%\begin{thebibliography}{999}
% Reference 1
%\bibitem[Author1(year)]{ref-journal}
%Author1, T. The title of the cited article. {\em Journal Abbreviation} {\bf 2008}, {\em 10}, 142--149.
% Reference 2
%\bibitem[Author2(year)]{ref-book}
%Author2, L. The title of the cited contribution. In {\em The Book Title}; Editor1, F., Editor2, A., Eds.; Publishing House: City, Country, 2007; pp. 32--58.
%\end{thebibliography}

% The following MDPI journals use author-date citation: Arts, Econometrics, Economies, Genealogy, Humanities, IJFS, JRFM, Laws, Religions, Risks, Social Sciences. For those journals, please follow the formatting guidelines on http://www.mdpi.com/authors/references
% To cite two works by the same author: \citeauthor{ref-journal-1a} (\citeyear{ref-journal-1a}, \citeyear{ref-journal-1b}). This produces: Whittaker (1967, 1975)
% To cite two works by the same author with specific pages: \citeauthor{ref-journal-3a} (\citeyear{ref-journal-3a}, p. 328; \citeyear{ref-journal-3b}, p.475). This produces: Wong (1999, p. 328; 2000, p. 475)

%=====================================
% References, variant B: external bibliography
%=====================================
\reftitle{References}
\externalbibliography{yes}
\bibliography{mybibfile.bib}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% optional
\sampleavailability{Samples of the compounds \ldots\ldots{} are
available from the authors.}

%% for journal Sci
%\reviewreports{\\
%Reviewer 1 comments and authors’ response\\
%Reviewer 2 comments and authors’ response\\
%Reviewer 3 comments and authors’ response
%}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}

### 2021-02-20 R/figure1.R
changing figs + 1

@@ -1,161 +1,0 @@
-# Figure 2. Comparing costs and willingness to pay
-
-# pkgs
-library(dplyr)
-library(readr)
-library(ggplot2)
-library(tidyr)
-library(patchwork)
-library(cowplot)
-
-# data on costs
-costs <- read_csv("data/costs_summarized.csv")
-willing_to_pay <- read_csv("data/owner_2018.csv")
-
-# exchange rate (midpoint between 2018 and 2019 est)
-exch_rate <- (3128 + 3500)/2
-
-# animals
-animals_2018 <- read_csv("data/animal_2018.csv")
-animals_2019 <- read_csv("data/campaign_2019.csv")
-
-# Figure 1A costs by category
-costs %>%
-  filter(include == "Y") %>%
-  group_by(year, category) %>%
-  summarize(cost_usd = sum(usd)) %>%
-  group_by(year) %>%
-  mutate(prop = cost_usd/sum(cost_usd)) %>%
-  ungroup() %>%
-  tidyr::complete(category, year = 2018:2019, 
-                  fill = list(cost_usd = 0, prop = 0)) %>%
-  mutate(type = case_when(category == "Vaccine" ~ "Vaccine", 
-                   TRUE ~ "Implementation")) -> costs
-
-ggplot(costs) +
-  geom_col(aes(x = category, y = cost_usd, fill = factor(year)), 
-           position = position_dodge()) +
-  labs(x = "Category", y = "Approximate cost (USD)", tag = "A") +
-  scale_fill_brewer(palette = "Dark2", name = "Year") +
-  theme_minimal_hgrid(font_size = 12) +
-  coord_flip() -> fig1A
-
-write_csv(costs, "out/costs_by_category.csv")
-
-# Figure 1B Cost per animal vaccinated
-costs %>%
-  group_by(year, type) %>%
-  summarize(cost = sum(cost_usd, na.rm = TRUE)) %>%
-  mutate(n_animals = case_when(year == 2018 ~ nrow(animals_2018), 
-                                  year == 2019 ~ nrow(animals_2019)), 
-          cost_per_animal = cost/n_animals) -> costs_per_animal
-
-costs_per_animal %>%
-  group_by(year) %>%
-  summarize(total = sum(cost_per_animal)) -> total_per_animal
-
-# jerry rig
-costs_per_animal %>%
-  filter(type == "Vaccine") %>%
-  mutate(border = cost_per_animal + 0.04) -> bord_2b
-
[…]

### 2021-02-20 R/figure2.R

@@ -1,173 +1,161 @@
+# Figure 2. Comparing costs and willingness to pay
+
 # pkgs
-library(sf)
 library(dplyr)
 library(readr)
 library(ggplot2)
 library(tidyr)
 library(patchwork)
 library(cowplot)
-library(ggsn)
 
-# From Claire's MSC
-range_people_ph <- c(4.13, 5.654)
-range_dogs_ph <- c(0.24, 0.26)
+# data on costs
+costs <- read_csv("data/costs_summarized.csv")
+willing_to_pay <- read_csv("data/owner_2018.csv")
 
-# Range of hdrs then
-mora_hdrs <- range_people_ph / range_dogs_ph
-mora_mid_hdr <- median(mora_hdrs)
-hdr_high <- 25
-hdr_low <- 8
+# exchange rate (midpoint between 2018 and 2019 est)
+exch_rate <- (3128 + 3500)/2
 
-# function to get upper & lower conf intervals from binomial into a data.frame
-bn_ci <- function(resighted, sighted_total) {
-  out <- binom.test(resighted, sighted_total, resighted/sighted_total,
-                    alternative = "two.sided", 
-                    conf.level = 0.95)
-  return(data.frame(cov_est = resighted/sighted_total, 
-                    cov_est_lower = out$conf.int[1], 
-                    cov_est_upper = out$conf.int[2]))
+# animals
+animals_2018 <- read_csv("data/animal_2018.csv")
+animals_2019 <- read_csv("data/campaign_2019.csv")
+
+# Figure 1A costs by category
+costs %>%
+  filter(include == "Y") %>%
+  group_by(year, category) %>%
+  summarize(cost_usd = sum(usd)) %>%
+  group_by(year) %>%
+  mutate(prop = cost_usd/sum(cost_usd)) %>%
+  ungroup() %>%
+  tidyr::complete(category, year = 2018:2019, 
+                  fill = list(cost_usd = 0, prop = 0)) %>%
+  mutate(type = case_when(category == "Vaccine" ~ "Vaccine", 
+                   TRUE ~ "Implementation")) -> costs
+
+ggplot(costs) +
+  geom_col(aes(x = category, y = cost_usd, fill = factor(year)), 
+           position = position_dodge()) +
+  labs(x = "Category", y = "Approximate cost (USD)", tag = "A") +
+  scale_fill_brewer(palette = "Dark2", name = "Year") +
+  theme_minimal_hgrid(font_size = 12) +
+  coord_flip() -> fig2A
+
[…]

### 2021-02-20 R/figure3.R

@@ -1,101 +1,173 @@
-# Figure 3. Demography and vax
+# pkgs
+library(sf)
 library(dplyr)
 library(readr)
 library(ggplot2)
 library(tidyr)
 library(patchwork)
 library(cowplot)
+library(ggsn)
 
-# bring in data
-animals_age <- read_csv("out/animals_age.csv")
-dem_ests <- read_csv("out/dem_ests.csv")
-vacc_sims <- read_csv("out/vacc_sims.csv")
+# From Claire's MSC
+range_people_ph <- c(4.13, 5.654)
+range_dogs_ph <- c(0.24, 0.26)
 
-# A: age pyramid ---------------
-animals_age %>%
-  mutate(Sex = toupper(Sex), age = floor(age_mos/12)) %>%
-  group_by(Sex, age) %>%
-  summarize(N  = n()) %>%
-  mutate(N = ifelse(Sex == "M", -N, N)) -> age_pyramid
+# Range of hdrs then
+mora_hdrs <- range_people_ph / range_dogs_ph
+mora_mid_hdr <- median(mora_hdrs)
+hdr_high <- 25
+hdr_low <- 8
 
-pyramid_A <- 
-  ggplot(age_pyramid, aes(x = age, y = N, fill = Sex)) +
-  geom_col() +
-  scale_y_continuous(breaks = c(seq(-600, -200, 200), 0, seq(200, 600, 200)), 
-                     labels = c(seq(600, 200, -200), "", seq(200, 600, 200)), 
-                     limits = c(-600, 600)) +
-  coord_flip() +
-  scale_fill_manual(values = c("#6f50a1", "#0d6d64")) +
-  labs(x = "Age in years", y = "Number of individuals", tag = "A") +
-  theme_minimal_hgrid(font_size = 12)
+# function to get upper & lower conf intervals from binomial into a data.frame
+bn_ci <- function(resighted, sighted_total) {
+  out <- binom.test(resighted, sighted_total, resighted/sighted_total,
+                    alternative = "two.sided", 
+                    conf.level = 0.95)
+  return(data.frame(cov_est = resighted/sighted_total, 
+                    cov_est_lower = out$conf.int[1], 
+                    cov_est_upper = out$conf.int[2]))
+}
 
-# B: stable age distributions ---------------
-dem_ests %>%
-  select(sim, growth, age_class, prop_data, prop_est) %>%
-  pivot_longer(prop_data:prop_est, values_to = "prop") -> dem_props
[…]

### 2021-02-20 R/figure4.R

@@ -1,0 +1,101 @@
+# Figure 3. Demography and vax
+library(dplyr)
+library(readr)
+library(ggplot2)
+library(tidyr)
+library(patchwork)
+library(cowplot)
+
+# bring in data
+animals_age <- read_csv("out/animals_age.csv")
+dem_ests <- read_csv("out/dem_ests.csv")
+vacc_sims <- read_csv("out/vacc_sims.csv")
+
+# A: age pyramid ---------------
+animals_age %>%
+  mutate(Sex = toupper(Sex), age = floor(age_mos/12)) %>%
+  group_by(Sex, age) %>%
+  summarize(N  = n()) %>%
+  mutate(N = ifelse(Sex == "M", -N, N)) -> age_pyramid
+
+pyramid_A <- 
+  ggplot(age_pyramid, aes(x = age, y = N, fill = Sex)) +
+  geom_col() +
+  scale_y_continuous(breaks = c(seq(-600, -200, 200), 0, seq(200, 600, 200)), 
+                     labels = c(seq(600, 200, -200), "", seq(200, 600, 200)), 
+                     limits = c(-600, 600)) +
+  coord_flip() +
+  scale_fill_manual(values = c("#6f50a1", "#0d6d64")) +
+  labs(x = "Age in years", y = "Number of individuals", tag = "A") +
+  theme_minimal_hgrid(font_size = 12)
+
+# B: stable age distributions ---------------
+dem_ests %>%
+  select(sim, growth, age_class, prop_data, prop_est) %>%
+  pivot_longer(prop_data:prop_est, values_to = "prop") -> dem_props
+
+
+stable_age_B <- 
+  ggplot(dem_props) +
+  geom_line(aes(x = age_class, y = prop, group = interaction(sim, name), 
+                color = name), 
+            alpha = 0.25) +
+  scale_x_continuous(breaks = 1:4, 
+                     labels = c("0 - 1", "1 - 2", "2 - 6", "6+")) +
+  scale_color_manual(values = c("navy", "lightblue"), 
+                     labels = c("Bootstrapped data", "Model estimates"), 
+                     name = "") +
+  labs(x = "Age class", y = "Proportion of \n population", tag = "B") +
+  theme_minimal_hgrid(font_size = 12) +
+  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
+  guides(color = guide_legend(override.aes = list(size = 1.1, alpha = 1)))
+
+  
[…]

### 2021-02-20 paper/mdpi_paper.Rmd
udpated with rticles mdpi template

@@
 ---
-title: Full title of the paper (Capitalized)
+title: "Lessons learned and paths forward for rabies dog vaccination in Madagascar: a case study of pilot vaccination campaigns in Moramanga District"
 author:
-  - name: Dominik Leutnant
-    affil: 1,2,\ddagger,*
-    orcid: 0000-0003-3293-2315
-  - name: John Doe
-    affil: 2, \dagger, \ddagger
+  - name: Caitlynn Filla
+    affil: 1, *
+  - name: Malavika Rajeev
+    affil: 2, *
+    orcid: 0000-0001-6030-4995
+  - name: Zoavina Randriana
+    affil: 3
+  - name: Chantal Hanitriniana
+    affil: 4
+  - name: Radoniaina R. Rafaliarison
+    affil: 3
+  - name: Glenn Edosoa
+    affil: 5
+  - name: Mamitiana Andriamananjara
+    affil: 6
+  - name: Nivohanitra P. Razafindraibe
+    affil: 6
+  - name: José Nely
+    affil: 7
+  - name: Angelique Ferreira
+    affil: 3, 8
+  - name: Annie Yang
+    affil: 2, \ddagger
+  - name: Fenomanana Daniel
+    affil: 3
+  - name: Tara A. Clarke
+    affil: 3, 9
+  - name: Zachary Farris
+    affil: 3, 10
+  - name: Terry Stone
+    affil: 8
+  - name: Jochem Lastdrager
+    affil: 8
+  - name: Tsiky Rajaonarivelo
+    affil: 3
+  - name: Katie Hampson
+    affil: 11
+  - name: C. Jessica E. Metcalf    
+    affil: 2
+  - name: Kim Valenta
+    affil: 1, 3
 affiliation:
@@
     address: |
-      Muenster University of Applied Sciences - 
-      Institute for Infrastructure, Water, Resources, Environment
-      Correnstr. 25, 48149 Muenster, Germany
-    email: leutnant@fh-muenster.de
+      University of Florida, Department of Anthropology
+      Gainesville, Florida, USA
   - num: 2
     address: | 
-      Your department
-      Street, City, Country
-    email: mail@mail.com
+      Department of Ecology and Evolutionary Biology, Princeton University
+      Princeton, NJ, United States
+  - num: 3
+    address: |
+      The Mad Dog Initiative
+      Akanin’ny Veterinera, Akaikiniarivo, Ambatobe, Antananarivo 101, Madagascar
+  - num: 4 
+    address: |
+      Mention Zoologie et Biodiversité Animale, Faculté des Sciences
+      Université d’Antananarivo, Antananarivo, Madagascar
+  - num: 5 
+    address: | 
+      Chargé des Maladies Tropicales Négligées
+      Organisation mondiale de la Santé Madagascar, Antananarivo, Madagascar
+  - num: 6 
+    address: | 
+      Direction des Services Vétérinaires 
+      Ministère chargé de l’Agriculture et de l’Élevage, Antananarivo, Madagascar
+  - num: 7 
+    address: |
+      Service contre les Maladies Endémo-épidémiques et Tropicales Négligées
+      Ministère de la Santé Publique,  Antananarivo, Madagascar
+  - num: 8
+    address: | 
+      Travelling Animal Doctors
+      919 Crossan Rd, Newark, Delaware, USA 
+  - num: 9
+    address: | 
+      North Carolina State University, Department of Sociology and Anthropology
+      Raleigh, North Carolina, USA
+  - num: 10
+    address: | 
+      Appalachian State University, Department of Health and Exercise Science
+      Boone, North Carolina, USA
+  - num: 11
+    address: | 
+      Boyd Orr Centre for Population and Ecosystem Health
+      Institute of Biodiversity, Animal Health and Comparative Medicine
+      University of Glasgow, Glasgow, UK
 # firstnote to eighthnote
 firstnote: |
-  Current address: Updated affiliation
+  These authors contributed equally to this work.
 secondnote: |
-  These authors contributed equally to this work.
+  Deceased.
 correspondence: |
-  leutnant@fh-muenster.de; Tel.: +XX-000-00-0000.
-journal: water
+  mrajeev@princeton.edu
+journal: tropicalmed
 type: article
@@
 bibliography: mybibfile.bib
-appendix: appendix.tex
-simplesummary: |
-  A Simple summary goes here.
 abstract: |
@@
 keywords: |
-  keyword 1; keyword 2; keyword 3 (list three to ten pertinent keywords specific 
-  to the article, yet reasonably common within the subject discipline.).
+  canine rabies; mass dog vaccination; central point vaccination; puppy vaccination; Zeroby30;
 acknowledgement: |
-  All sources of funding of the study should be disclosed. Please clearly 
-  indicate grants that you have received in support of your research work. 
-  Clearly state if you received funds for covering the costs to publish in open 
-  access.
+   We thank all the veterinarians, livestock officers, local officials and leaders,
+   NGO staff, and volunteers involved in each of the campaigns. We are grateful 
+   to staff and officials at the Department of Veterinary Services,
+   the Ministry of Public Health, and the Institut Pasteur de Madagascar for 
+   their support and technical assistance. 
+   In particular, we thank Carmen Barba Claassens, Jean Hyacinthe Randrianarisoa, 
+   Ranaivoarimanana, Fierenantsoa Randriamahatana, Esther Noiarisaona, Cara Brook, 
+   and Christian Ranaivoson, Rila Ratovoson, and Claire LeBlanc. The campaign 
+   costs were supported by donations from Lush Cosmetics  Charity Pot. 
+   Parts of this work were funded by grants from the Center for Health and 
+   Wellbeing and the Department of Ecology and Evolutionary Biology at 
+   Princeton University to MR. MR was supported by an NSF Graduate Research 
+   Fellowship and a Princeton Institute for Regional and International Studies 
+   Graduate Fellowship. KH is supported by the Wellcome Trust (207569/Z/17/Z).
+   We want to particularly acknowledge the contributions to this work by Annie Yang, 
+   an undergraduate researcher and volunteer with Princeton University and the 
+   Mad Dog Initiative who tragically passed away in a car accident before this 
+   research was published.
 authorcontributions: |
-  For research articles with several authors, a short paragraph specifying their 
-  individual contributions must be provided. The following statements should be 
-  used ``X.X. and Y.Y. conceive and designed the experiments; X.X. performed the 
-  experiments; X.X. and Y.Y. analyzed the data; W.W. contributed 
-  reagents/materials/analysis tools; Y.Y. wrote the paper.'' Authorship must be
-  limited to those who have contributed substantially to the work reported.
+  Conceptualization, MR, KV, JL, TS, ZF, GE; methodology, MR, KV, JL, TS, ZF, GE, KH, CJEM, AY; data collection: MR, ZR, CH, RRR, GE, MA, NPR, JN, AF, TR, FD; formal analysis, MR, CF, AY; writing—original draft preparation, CF, MR; writing—review and editing, CF, MR, CJEM, TC, ZF, KV, KH, GE, CH; visualization, MR; supervision, KV, KH, CJEM, ZF; project administration, MR, ZR, CH, RRR, GE, MA, NPR, JN, AF, TR, FD, KV, ZF; funding acquisition, MR, KV, JL, TS, ZF, TC.
 conflictsofinterest: |
-  Declare conflicts of interest or state 'The authors declare no conflict of 
-  interest.' Authors must identify and declare any personal circumstances or
-  interest that may be perceived as inappropriately influencing the
-  representation or interpretation of reported research results. Any role of the
-  funding sponsors in the design of the study; in the collection, analyses or 
-  interpretation of data in the writing of the manuscript, or in the decision to 
-  publish the results must be declared in this section. If there is no role, 
-  please state 'The founding sponsors had no role in the design of the study; 
-  in the collection, analyses, or interpretation of data; in the writing of the 
-  manuscript, an in the decision to publish the results'.
-sampleavailability: |
-  Samples of the compounds ...... are available from the authors.
-abbreviations:
-  - short: MDPI
-    long: Multidisciplinary Digital Publishing Institute
-  - short: DOAJ
-    long: Directory of open access journals
-  - short: TLA
-    long: Three letter acronym
-  - short: LD 
-    long: linear dichroism
+  The authors declare no conflict of interest. The funders had no role in the
+  design of the study; in the collection, analyses, or interpretation of data; 
+  in the writing of the manuscript, or in the decision to publish the results.
 output: rticles::mdpi_article
+params:
+  outformat: "pdf"
 ---
 
-# Version
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, 
+                      message=FALSE, out.width="80%")
+library(tidyverse)
+library(glue)
+library(magrittr)
+library(lubridate)
+library(sf)
+library(gt)
+library(here)
+source("R/dem_models.R")
 
-This Rmd-skeleton uses the mdpi Latex template published 2019/02. 
-However, the official template gets more frequently updated than the 'rticles'
-package. Therefore, please make sure prior to paper submission, that you're 
-using the most recent .cls, .tex and .bst files 
-(available [here](http://www.mdpi.com/authors/latex)).
+# stats
+# From Claire's MSC 
+hdrs <- tribble(~loc, ~range_people_ph, ~range_dogs_ph, ~n_households,
+                "urban", 4.13, 0.24, 170,
+                "rural", 5.654, 0.26, 164)
+# Range of hdrs then
+hdrs %>% 
+  mutate(mora_hdrs = range_people_ph / range_dogs_ph) -> hdrs
+mora_mid_hdr <- median(hdrs$mora_hdrs)
+hdr_backcalcs <- read_csv(here("out/hdr_backcalcs.csv"))
+
+# animals
+animals_2018 <- read_csv(here("data/animal_2018.csv"))
+animals_2019 <- read_csv(here("data/campaign_2019.csv"))
+
+# pop ests
+mora_sf <- st_read(here("data/shapefile/mora_communes.shp"))
+exch_rate <- (3128 + 3500)/2
+
+# data
+transects <- read_csv(here("data/transects_2018.csv"))
+
+```
 
@@
 
-The introduction should briefly place the study in a broad context and highlight why it is important.
-It should define the purpose of the work and its significance.
-The current state of the research field should be reviewed carefully and key publications cited.
-Please highlight controversial and diverging hypotheses when necessary.
-Finally, briefly mention the main aim of the work and highlight the principal conclusions.
-As far as possible, please keep the introduction comprehensible to scientists outside your particular field of research.
-Citing a journal paper [@bertrand-krajewski_distribution_1998; @leutnant_stormwater_2016].
-And now citing a book reference @gujer_systems_2008.
-Please use the command [@leutnant_stormwater_2016] for the following MDPI journals, which use author-date citation: Administrative Sciences, Arts, Econometrics, Economies, Genealogy, Humanities, IJFS, JRFM, Laws, Religions, Risks, Social Sciences.
+Canine rabies results in an estimated 60,000 human deaths per year globally [@hampson2015].
+These deaths are entirely preventable: prompt post-exposure prophylaxis of humans exposed to rabies is highly effective at preventing death and mass dog vaccination can interrupt transmission in domestic dogs and eventually lead to disease elimination [@fooks2014].
+The World Health Organization (WHO) and its partners have set a goal to eliminate human deaths due to canine rabies by the year 2030 ('ZeroBy30')[@abela-ridder2016].
+Annual dog vaccination campaigns achieving at least 70% coverage are the recommended target for controlling rabies in domestic dog populations [@worldhealthorganization2013].
+However, achieving this coverage target in low and middle income countries where the burden of human rabies is highest can be challenging due to economic, ecological, sociocultural, and political barriers [@fahrion2017].
 
-# Materials and Methods
+In sub-Saharan African countries, parenteral vaccinations implemented through static point campaigns have been shown to be cost-effective and feasible [@borse2018].
+While most dogs in these settings are considered free roaming, they are mostly owned and are accessible for vaccination through these campaigns [@morters2014; @jibat2015].
+However, reaching high coverage and maintaining vaccination campaigns at scale requires sustained investment and coordination, and the challenges in implementation largely reflect financial and logistical constraints more than the feasibility of vaccination itself [@fahrion2017].
+Developing clear and context-specific strategies and lowering costs and resources needed could help spur the implementation and scaling up of campaigns in these countries.
 
-Materials and Methods should be described with sufficient details to allow others to replicate and build on published results.
-Please note that publication of your manuscript implicates that you must make all materials, data, computer code, and protocols associated with the publication available to readers.
-Please disclose at the submission stage any restrictions on the availability of materials or information.
-New methods and protocols should be described in detail while well-established methods can be briefly described and appropriately cited.
+In Madagascar, canine rabies has been endemic for over a century, and for most of that period, the Institut Pasteur de Madagascar has provided post-exposure prophylaxis free-of-charge to bite patients in the country [@reynes2011].
+Currently, there are only 31 clinics where these human vaccines are available, and there is minimal dog vaccination due to high costs to owners and lack of vaccine availability [@rajeev2018].
+Recent studies have estimated a high burden of human deaths, masked by weak surveillance across the country [@rajeev2020; @rajeev2018].
 
-Research manuscripts reporting large datasets that are deposited in a publicly available database should specify where the data have been deposited and provide the relevant accession numbers.
-If the accession numbers have not yet been obtained at the time of submission, please state that they will be provided during review.
-They must be provided prior to publication.
+The veterinary sector is largely private, with 204 veterinarians employed in hybrid private/public employment as designated district veterinarians.
+While dog vaccination is rare, livestock officers and veterinarians work together to implement cattle vaccination campaigns for anthrax on an annual basis as mandated by the government (owners are charged a fee per animal for these vaccines which vary by location) [@mondiale2017evaluation].
+No routine mass dog vaccinations have been conducted on the island, although a few pilot programs have begun in recent years, largely implemented by NGO-government partnerships.
 
-Interventionary studies involving animals or humans, and other studies require ethical approval must list the authority that provided approval and the corresponding ethical approval code.
+Here, we summarize lessons learned through the implementation of pilot vaccination programs in the Moramanga District of Madagascar, where previous work has shown high incidence of dog rabies cases and human rabies exposures.
+In 2018 and 2019, we deployed two different vaccination strategies.
+In 2018, we carried out a larger scale volunteer-led pilot vaccination campaign in two communes (sub-district level) using a static point strategy.
+In 2019/2020, we provided vaccines, all necessary supplies for vaccine administration, and a per-vaccine fee to the District Veterinarian to vaccinate animals as part of routine services delivered continuously.
+We compare time, human resources, costs, and coverage estimates between these campaigns, and using a demographic and vaccination model we further explore different vaccination strategies based on what we learned during implementation.
+
+# Methods
+
+## Study Area
+
+The Moramanga District is located midway between the central highlands and the east coast of Madagascar, at an average altitude of 936 m.
+It comprises 21 communes, covering approximately 7150 km^2^ with an approximate human population of `r format(round(sum(mora_sf$pop), -3), big.mark = ",")`[@instat2021].
+Previous work in the district has established a high burden of rabies exposures and deaths despite the availability of post-exposure prophylaxis at the district hospital [@rajeev2018].
+While Moramanga is relatively close to the capital city of Antananarivo (~ 3 hours by bus), within the district, travel times between locations are highly variable, with much of the population living in more rural areas with limited access to roads and transportation [@rajeev2020].
+Before 2018, there were limited animal rabies vaccination services, with most animal vaccines available in the urban commune of Moramanga Ville where owners were often charged > 15,000 Ariary (~ 4.28 USD) per vaccine administered.
+
+## 2018 Campaign
+
+```{r}
+pts <- read_csv(here("data/campaign_pts_2018.csv"))
+```
+
+In 2018, two NGOs (the Mad Dog Initiative (MDI) and Traveling Animal Doctors (TAD)) organized a pilot vaccination campaign in collaboration with the Department of Veterinary Services, and the Ministry of Public Health in the District of Moramanga.
+This campaign focused on two communes in Moramanga, Moramanga Ville (the district center) and Andasibe (a rural commune surrounding Andasibe National Park), where previously high incidence of probable rabies exposures (Moramanga Ville) and a high burden of deaths (Andasibe) had been recorded [@rajeev2018].
+
+The campaign was planned as a series of static point vaccination stations covering 1 - 3 fokontany (i.e., sub-communes) per day.
+A week before the campaign dates, the vaccination team informed the chief of the fokontany about the campaign and provided posters advertising the date of the vaccine and that it would be available at no cost to owners.
+During the campaign, we used Rabisin (10 mL vials with 1 mL per dose, Boehringer Ingelheim) to vaccinate both dogs and cats presented that were at least 1 month old.
+As part of the campaign, owners were surveyed by vaccinators about how many dogs and cats they owned in total (split by > 1 year vs.
+< 1 year in order to avoid language ambiguities that might result in excluding puppies and kittens), as well as if their dogs were free roaming (no restrictions on movement by the owner/'mirenyreny', tied/'mifatora', or fenced/'mifefy').
+Vaccinations were delivered at no cost to owners, but as animal vaccination is generally thought of as a paid service in Madagascar, owners were asked how much in Ariary they would be willing to pay to have one animal vaccinated against rabies.
+For each animal vaccinated, we recorded the species (cat or dog), sex, approximate age in years, and whether the animal had been previously vaccinated.
+
+To assess coverage, post-vaccination coverage surveys were conducted according to previously established methodology[@sambo2017; @gibson2015].
+All animals vaccinated were concurrently marked with a colored, non-toxic, livestock crayon along the top of the head or back.
+Between 4 - 6 PM when dogs are most active [@sambo2017] and on the same day as the campaign, in each vaccination location two transect surveys were done.
+Each team (two people per team) walked a transect for 1 hour starting in opposite directions and accompanied by a local guide to ensure that walking paths did not overlap.
+Teams recorded any marked and unmarked dogs observed, their roaming status (whether roaming, inside a fence, or tied), and their approximate age (greater or less than 1 year of age).
+
+```{r fig1} 
+knitr::include_graphics(here("figs/fig1.jpeg"))
+```
+
+#### **Figure 1.
+Photos from the 2018 campaign**.
+*Topleft: advertisement for the campaign posted on the door of the fokontany office as the campaign starts; Topright: a dog post-vaccination marked with a crayon; Bottomleft: a basket of puppies brought for vaccination; Bottomright: a line of owners and dogs waiting for vaccination.*
+
+## 2019 Campaign
+
+For the 2019 campaign, instead of a static point campaign strategy, vaccine vials(Rabisin) and the supplies needed to administer them (needle, syringe, vaccination card for owners) were distributed to the District Veterinarian, who then delivered the vaccination at no cost to owners, but was directly compensated 1,500 Ar (~ 0.40 USD) per rabies vaccine administered.
+The campaign lasted from September 6, 2019 to June 19, 2020, One week prior to her visiting each location, the District Veterinarian advertised the vaccines by calling ahead to the fokontany leaders and other officials who then advertised to their communities, largely through word-of-mouth.
+For each animal vaccinated, the District Veterinarian collected the animal's age and sex, and also asked owners to approximate the distance they travelled to receive the vaccination in meters.
+Researchers communicated with the District Veterinarian about progress periodically throughout the campaign, primarily through telephone calls.
+No other compensation or instructions were provided, and we asked the District Veterinarian to administer as many (or as few) as feasible or wanted.
+As the vaccinations were delivered continuously, we were unable to do comparable post-vaccination surveys.
+
+## Analyses
+
+### Campaign resource and cost comparisons
+
+We documented the overall costs and resources required for the two vaccination efforts.
+We tracked the number of vaccination points, the number of days over which these vaccinations occurred, and the number of person days required overall (i.e.
+the number of working people per day over the campaign [@mazeri2021]), in addition to monetary costs.
+As costs were incurred in both USD and Ariary, and the exchange rate declined rapidly between 2018 and 2019, we used the midpoint between the two years (`r exch_rate` Ar to 1 USD) for the cost comparisons.
+
+For the 2018 campaign, we broke costs down into the following categories: direct vaccine costs (cost for vaccine, syringes, needles, vaccination cards), supplies (livestock crayons, muzzles, gloves, alcohol, swabs), food and lodging for NGO personnel and other vaccinators during the campaign, personnel costs (per diems for the district veterinarian, livestock field officers, local guides, and NGO employees), and advertisement (posters and banners for advertising the campaign).
+Foreign NGO volunteer expenses for travel to Madagascar were not included in these costs.
+Vehicles and drivers are also not included in these costs, as the drivers' time and vehicle use were donated to the campaign by volunteers involved in the campaign.
+In 2019, costs were split into two categories, direct vaccine costs (for same items as in 2018), and personnel costs (per vaccine fee paid to the district veterinarian), and supplies (a generator and fuel for the veterinarian to maintain the vaccine under cold chain during power outages).
+Transportation costs were also not included as the District Veterinarian used her own vehicle and vaccinated as part of their routine veterinary service provisioning.
+
+We used the data on owners' reported willingness to pay for vaccines to estimate the proportional reduction in animals vaccinated as fees are increased.
+We also estimated how this would impact cost per animal vaccinated by approximating the costs for implementation (i.e.
+those costs that remain fixed) from costs incurred per animal vaccinated (i.e.
+vaccine, syringe, vaccination card, and per vaccination fee to District Veterinarian in 2019), and calculating the balance between the returns from owner payments (i.e.
+increases in cost recovery per animal vaccinated) vs.
+decreasing numbers of animals vaccinated overall.
+
+### Coverage estimates
+
+For the 2018 campaign, we used the transect data to estimate vaccination coverage as the proportion of dogs sighted that were marked using a binomial confidence interval at the commune level.
+For the 2019 campaign, we estimated vaccination coverage using human-to-dog ratios (HDRs) and human population estimates [@cleaveland2003; @athingo2020].
+We used a range of 8 - 25, based on previous data from Madagascar [@ratsitorahina2009] and based on recent estimates from household surveys in the Moramanga District [@leblancclaireRabiesMadagascarThreepronged2019].
+We set the point estimate using an HDR of `r round(mora_mid_hdr, 1)`, the midpoint between the HDRs estimated for two communities in the District by LeBlanc et al.
+2019.
+We used human population estimates from the 2018 national census in each commune where the vaccinations were delivered [@instat2021].
+Coverage was estimated as the number of dogs vaccinated in total in that commune divided by the estimated dog population.
+We used this same method for the 2018 campaign, as well, to compare coverage estimated by the post-vaccination transects vs.
+HDRs.
+
+### Dog Demography
+
+Using the age data on vaccinated animals collected during both vaccination campaigns, we estimated the proportion of the population in four age classes: puppies under the age of 1 year, juveniles aged 1 - 2 yrs, adults aged 2 - 6 years, and older dogs aged 6+ years based on .
+With the assumption that these estimates represent the population at a stable age distribution, we use a Leslie matrix model to estimate annual adult survival probability and fertility using maximum likelihood estimation [@Fujiwara2017].
+Specifically, we assume that the number of individuals in each age class follows a Poisson distribution with the mean predicted by the stable age distribution from the model (the proportion of individuals in each age class at equilibrium, equal to the eigenvector associated with the dominant eigenvalue of the matrix $\nu$) multiplied by the total number of individuals in the population ($N_t$):
+
+$$ N_{a} \sim Pois(\nu N_t) $$
+
+We assume that all individuals older than 1 year of age reproduce, and we do not estimate declines in fertility given the small proportion of dogs older than age six years in the population.
+To get bootstrapped estimates, we used 100 sub-sampled data sets of 1000 observations each from the observed age data to fit the parameters, and also varied initial values used in the optimization (N = 100 initial values sets) for 10,000 parameter estimates total.
+
+### Modeling vaccination campaign strategies
+
+```{r}
+animals_age <- read_csv(here("out/animals_age.csv"))
+original <- ages_to_class(age_yrs = animals_age$age_mos/12, 
+                          years_in_class = c(1, 1, 4, 1),
+                          sample = FALSE)
+female <- ages_to_class(age_yrs = animals_age$age_mos[animals_age$Sex == "F"]/12, 
+                          years_in_class = c(1, 1, 4, 1),
+                          sample = FALSE)
+```
+
+We used the parameter estimates from the demographic model to simulate different vaccination strategies in a hypothetical commune with 1000 dogs.
+We used a discrete time age-structured model with a monthly time step to compare three strategies:
+
+1)  Annual vaccination campaigns occurring within the same month each year targeting dogs of all ages
+2)  Continuous vaccination of new puppies throughout the year targeting puppies that reach the age of 3 months
+3)  A combined approach with annual campaigns (as 1) and routine puppy vaccination (as 2)
+
+We split the dog population into puppies (< 1 year old) and adults based on the stable age distribution estimated from the demographic model.
+To estimate pup survival in year one, we took the fertility estimates (i.e., number of new puppies per reproducing dog observed in the pup age class) and divided by an estimate of newborn pups per reproducing dog each year based on average litter size, average number of litters per female per year [@hampson2009; @morters2014; @czupryna2016], and the proportion of the adult population that is female (estimated from our data).
+We assumed that for the annual campaign strategy, surviving vaccinated adult dogs were revaccinated in subsequent years [@delriovilas2017], but that if a pup had been vaccinated within 9 months of the campaign, it was not revaccinated.
+We also assumed that vaccine immunity lasted for a discrete period of 3 years (with revaccination resetting immunity).
+A subset of parameter estimates resulted in estimates of population decline, but based on the shape of the age pyramid and to simulate reasonable campaign scenarios, we filtered to parameter estimates which corresponded to positive population growth.
+
+#### Data and ethics statement
+All data were analysed in `r R.version$version.string` [@R2020], largely using the `tidyverse` package suite [@wickham2019].
+Geospatial data were mapped using the sf [@sf] package.
+All data and code are available at [https://github.com/mrajeev08/mora_vax](https://github.com/mrajeev08/mora_vax).
+The vaccinations were part of a public health campaign and routine veterinary service provisioning carried out by the local veterinary officials, the NGOs involved, and in partnership with the Ministry of Public Health and the Department of Veterinary Services at the national level.
+MDI also maintains national research permits (MICET permit: #130-19/MEDD/SG/DGEF/DGRNE) for its research and volunteer programs.
 
@@
 
-This section may be divided by subheadings.
-It should provide a concise and precise description of the experimental results, their interpretation as well as the experimental conclusions that can be drawn.
+## Summary of 2018 and 2019 campaigns
+```{r}
+summ_tab1 <- read_csv(here("out/table1_stats.csv"))
+species_2018 <- table(animals_2018$Species)
+species_2019 <- table(tolower(animals_2019$`Dog or Cat`))
+n_communes <- n_distinct(animals_2019$Commune)
+animals_2019 %>% 
+  filter(tolower(`Dog or Cat`) %in% "dog") %>%
+  mutate(Commune = ifelse(Commune %in% "Moramanga", "Moramanga", "other"), 
+         prev_vacc = ifelse(is.na(`Previous Vaccination`), "NO", 
+                            `Previous Vaccination`)) %>% 
+  group_by(prev_vacc, Commune) %>% 
+  summarize(n = n()) %>%
+  pivot_wider(values_from = n, names_from = prev_vacc) %>%
+  mutate(total = NO + YES, 
+         perc_vacc = round(YES / total * 100, 1)) -> prev_vacc_2019
+spayed_neutered <- table(animals_2019$`Spayed/Neutered`)["YES"]/nrow(animals_2019)
 
-## Subsection Heading Here
+dist_over_100 <- round(sum(animals_2019$`Distance Travelled (m)` <= 1000, na.rm = TRUE)/sum(!is.na(animals_2019$`Distance Travelled (m)`))*100, 0)
+transects_roaming <- round(sum(!(transects$Status %in% c("Fefy", "Mifatotra")))/nrow(transects)*100, 0)
+```
 
-Subsection text here.
+During the 2018 campaign, a total of `r nrow(animals_2018)` animals were vaccinated (`r species_2018["A"]` dogs and `r species_2018["S"]` cats) in the Moramanga (urban) and Andasibe (rural) communes over 13 days during the month of April (Table 1).
+We vaccinated at `r table(pts$Commune)["Andasibe"]` points in Andasibe and `r table(pts$Commune)["Moramanga Ville"]` points in Moramanga Ville.
+During the 2019 campaign, between `r paste(format.Date(range(dmy(animals_2019$Date)), "%B %Y"), collapse = " - ")`, the District Veterinarian vaccinated a total of `r nrow(animals_2019)`animals (`r species_2019["dog"]` dogs and `r species_2019["cat"]` cats) over 48 days in seven communes in the Moramanga District.
+While more animals were vaccinated per vaccination point and per vaccination day in 2018 compared to 2019, the number of animals vaccinated per person-day was much higher for the 2019 campaign.
+More animals were vaccinated in 2018 vs.
+2019, but this was largely a result of vaccinating more cats during the 2018 campaign (Table 1).
 
-### Subsubsection Heading Here
+In 2018, `r summ_tab1$All_2018[summ_tab1$name %in% "prev_vacc"]` of dogs had a previous history of vaccination (largely in the urban commune of Moramanga Ville) with only `r summ_tab1$All_2018[summ_tab1$name %in% "vacc_in_year"]` of dogs vaccinated within the last year.
+This remained largely the same in 2019 (~ `r summ_tab1$All_2019[summ_tab1$name %in% "prev_vacc"]`), as the District Veterinarian focused their efforts in other communes.
+The District Veterinarian did vaccinate `r prev_vacc_2019$total[prev_vacc_2019$Commune %in% "Moramanga"]` dogs in Moramanga Ville in 2019, and of those `r prev_vacc_2019$perc_vacc[prev_vacc_2019$Commune %in% "Moramanga"]`% had been vaccinated in the previous year's campaign, whereas only `r prev_vacc_2019$perc_vacc[prev_vacc_2019$Commune %in% "other"]`% of dogs vaccinated in all other communes had any history of previous vaccination.
+In addition, in 2019, `r round(spayed_neutered * 100, 1)`% of animals had been spayed or neutered, reflecting efforts by the Mad Dog Initiative to implement free spay and neuter clinics in the district.
 
-Bulleted lists look like this:
+In 2018, `r summ_tab1$Moramanga_2018[summ_tab1$name %in% "roaming"]` of owners reported their animals were free-roaming, but this varied by location (Table 1).
+In addition, while less than `r summ_tab1$Moramanga_2018[summ_tab1$name %in% "roaming"]` of owners in Moramanga Ville reported that their animals were free roaming, the majority of animals observed during the transects (`r transects_roaming`%) were observed outside of fences and not tied, and thus the majority of animals could be classified as semi-confined in the more urban township of Moramanga Ville, and free-roaming in the rural setting of Andasibe (approximately `r summ_tab1$Andasibe_2018[summ_tab1$name %in% "roaming"]` of owners reported their dogs as free-roaming, Table 1).
+In 2019, the District Veterinarian also asked owners to approximate how far they travelled in meters to get their animals vaccinated, and `r dist_over_100`% of people reportedly travelled less than 1 km to reach the vaccination point.
 
-* First bullet
-* Second bullet
-* Third bullet
+#### **Table 1.
+Summary of 2018 and 2019 campaigns**.
+*Breakdown of of animals vaccinated, prior vaccination history, dog demography, dog ownership, daily and per vaccination rates by year and location (for 2018).
+Photo credict: Jochem Lastdrager, Traveling Animal Doctors.*
+```{r}
+if(params$outformat %in% "docx") {
+  knitr::include_graphics(here("figs/table1.png"))
+} else {
+  tab1 <- readRDS(here("out/table1.rds"))
 
-Numbered lists can be added as follows:
+  tab1 %>% 
+    gt::as_latex() 
+}
+```
 
-1. First item
-2. Second item
-3. Third item
+## Cost comparison and willingness to pay
 
-The text continues here.
+The 2018 campaign cost more overall and per animal vaccinated than the 2019 campaign, largely due to increased personnel costs (fig 2A & B, Table 1).
+Reflecting the extra personnel necessary to run the static point campaign, the 2018 campaign also took substantially more person-days per animal vaccinated (Table 1).
+We found that charging owners for vaccination would result in minimal cost recovery (fig 2C) and beyond a minimal cost would actually result in increased costs per vaccinated individual than free-of-charge campaigns.
+Cost recovery would be more likely given a 2019 style strategy where the majority of the costs are incurred on a per animal basis, compared to the 2018 campaigns where the costs are largely due to setting up the static point vaccination stations (fig 4A).
+More importantly, in all cases, even a nominal fee would significantly reduce numbers of dogs vaccinated and thus vaccination coverage, particularly in the rural commune of Andasibe (fig 2D).
 
-All figures and tables should be cited in the main text as Figure 1, Table 1, etc.
+```{r fig2} 
+knitr::include_graphics(here("figs/fig2.jpeg"))
+```
 
-\begin{figure}[H]
-\centering
-\includegraphics[width=3 cm]{logo-mdpi}
-\caption{This is a figure, Schemes follow the same formatting.
-If there are multiple panels, they should be listed as: (\textbf{a}) Description of what is contained in the first panel.
-(\textbf{b}) Description of what is contained in the second panel.
-Figures should be placed in the main text near to the first time they are cited.
-A caption on a single line should be centered.}
-\end{figure}
+#### **Figure 2.
+Comparing campaign costs and willingness to pay**.
+*A) Vaccine costs broken down by category and by year (colors).
+B) Overall cost per animal vaccinated for the two campaign years, split by direct costs of vaccination per animal (i.e.
+vaccine, vaccination card, syringes) and baseline implementation costs (i.e.
+personnel, supplies, subsistence costs for vaccinators during the campaign).
+C) Estimated cost per animal vaccinated under a willingness to pay model for two campaigns examining increasing costs charged to the owner, with estimated costs declining due to cost recovery through owner payments and then peaking once owners report to be no longer willing to pay for the vaccine.
+D) The percent reduction in number of animals vaccinated given owners willingness to pay.
+The curves in C and D are shown based on the overall responses on willingness to pay (solid line) from both Moramanga and Andasibe, and the responses split by Commune (dashed and dotted lines).*
 
-\begin{table}[H]
-\caption{This is a table caption. Tables should be placed in the main text near to the first time they are cited.}
-\centering
-%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
-\begin{tabular}{ccc}
-\toprule
-\textbf{Title 1}	& \textbf{Title 2}	& \textbf{Title 3}\\
-\midrule
-entry 1		& data			& data\\
-entry 2		& data			& data\\
-\bottomrule
-\end{tabular}
-\end{table}
+## Comparing campaign coverage estimates
+The 2018 campaign covered two communes and was estimated to have achieved approximately 60% coverage (fig 3A).
+The 2019 campaign covered seven communes, but was estimated to have achieved lower coverage levels (fig 3, ranging from 5 - 60%).
+In the 2018 campaign, we used post-vaccination coverage transects to estimate vaccination coverage, but we were unable to do this in 2019 given the continuous delivery strategy.
+In addition, in Andasibe in 2018, coverage estimates were based on a single transect resulting in more uncertainty.
+However, coverage estimates from the transects in 2018 were consistent with the HDR based estimates (for both Andasibe and Moramanga, transect-based estimates fell within the range of the HDR estimates).
+We also back-calculated HDRs given our vaccination coverage estimates, and these were similar to the HDRs calculated from the household survey (`r hdr_backcalcs %>% filter(Commune == "Andasibe") %$% glue("{hdr_cov_est_lower} - {hdr_cov_est_upper}")` for Andasibe compared to `r round(hdrs$mora_hdrs[hdrs$loc == "rural"], 1)` in a rural community and `r hdr_backcalcs %>% filter(Commune == "Moramanga") %$% glue("{hdr_cov_est_lower} - {hdr_cov_est_upper}")` for Moramanga Ville compared to `r round(hdrs$mora_hdrs[hdrs$loc == "urban"], 1)` in an urban community).
 
-This is an example of an equation:
+```{r fig3}
+knitr::include_graphics(here("figs/fig3.jpeg"))
+```
 
-\begin{equation}
-\mathbb{S}
-\end{equation}
-<!-- If the documentclass option "submit" is chosen, please insert a blank line before and after any math environment (equation and eqnarray environments).
-This ensures correct linenumbering.
-The blank line should be removed when the documentclass option is changed to "accept" because the text following an equation should not be a new paragraph.
--->
+#### **Figure 3.
+Estimates of coverage achieved by the 2018 and 2019 campaigns**.
+*A) The commune level numbers of dogs vaccinated (size of points) and the associated coverage estimates (color of points) for the year 2018 (estimated using post-vaccination transects) and 2019 (estimated using a human-to-dog ratio (HDR) of `r round(mora_mid_hdr, 1)`, based on a recent household survey in the Moramanga District).
+The inset shows the location of the Moramanga District in Madagascar.
+B) A comparison of coverage estimates by location and by method of estimation (post-vaccination transects vs.
+HDR based estimates); for the transect based estimates, the line range shows the 95% exact binomial confidence interval while for the HDR based estimates, the line range shows the range of coverage estimates assuming an HDR range of 8 - 25 according to estimates from the literature.*
 
-<!-- Please punctuate equations as regular text. Theorem-type environments (including propositions, lemmas, corollaries etc.) can be formatted as follows: -->
-Example of a theorem:
-\begin{Theorem}
-Example text of a theorem.
-\end{Theorem}
+## Dog demography and simulating vaccination strategies
+```{r}
+age_data <- read_csv(here("out/animals_age.csv"))
+prop_sex <- round(table(age_data$Sex)/nrow(age_data)*100, 0)
+litter_size <- c(4, 8) * 1 * 0.4 # from literature (Morters, Czupryna, etc.)
+dem_ests <- read_csv(here("out/dem_ests.csv"))
+q_high <- function(x) quantile(x, 0.975)
+q_low <-  function(x) quantile(x, 0.025)
+mean_round <- function(x) round(mean(x), 2)
+dem_ests %>%
+  ungroup() %>%
+  filter(pop_growth == "Growing") %>%
+  select(fert_est, psurv_adult_est, growth, sim) %>%
+  select(-sim) %>%
+  distinct() %>%
+  mutate(pup_surv_upper = fert_est/litter_size[1], 
+         pup_surv_lower = fert_est/litter_size[2]) %>%
+  pivot_longer(everything()) %>%
+  group_by(name) %>%
+  summarize(across(everything(), list(mean = mean_round, upper = q_high, lower = q_low), 
+                   .names = "{.col}.{.fn}")) %>%
+  rowwise() %>%
+  mutate(ci = paste(round(value.lower, 2), " - ", 
+                    round(value.upper, 2))) -> dem_pars
 
-The text continues here. Proofs must be formatted as follows:
+```
 
-Example of a proof:
-\begin{proof}[Proof of Theorem 1]
-Text of the proof. Note that the phrase `of Theorem 1' is optional if it is clear which theorem is being referred to.
-\end{proof}
-The text continues here.
+Demographic data from vaccinated dogs shows a population pyramid with a large base, indicative of a fast-growing population, and with a male bias (approximately `r prop_sex["M"]`% male, Fig 4A).
+We fit these data to an age-structured model, and were able to generate parameter estimates which resulted in stable age distributions consistent with the data (Fig 4B).
+We filter to parameter estimates that are consistent with a growing population resulting in mean adult annual survival probability of `r dem_pars[dem_pars$name == "psurv_adult_est", "value.mean"]` (95% quantile: `r dem_pars[dem_pars$name == "psurv_adult_est", "ci"]`).
+We use estimates of fertility (on average `r dem_pars[dem_pars$name == "fert_est", "value.mean"]`, 95% quantile: `r dem_pars[dem_pars$name == "fert_est", "ci"]`) to back-calculate pup survival, which ranges from `r round(dem_pars$value.mean[dem_pars$name == "pup_surv_lower"], 2)` - `r round(dem_pars$value.mean[dem_pars$name == "pup_surv_upper"], 2)`.
+We find that given these demographic parameters, annual campaigns that target dogs of all ages result in rapid decline in vaccination coverage between campaigns, largely due to rapid turnover of the dog population (compared to the impact of waning immunity assuming a discrete 3 year period of vaccine immunity, evident in the additional dip at year 3, fig 4D).
+Continuously targeting 70% of the puppy population for vaccination, while unable to achieve the peak coverage, consistently reaches coverage of about 50% of the dog population.
+A combined strategy maintains the highest and most temporally stable levels of coverage close to the target of 70%.
+
+```{r fig4}
+knitr::include_graphics(here("figs/fig4.jpeg"))
+```
+
+#### **Figure 4.
+Dog demography and implications for campaign strategies**.
+*A) The age pyramid for vaccinated dogs by sex.
+B) Bootstrapped estimates of the proportion of the population in each age class from the age data (dark blue) compared to estimates from the demographic models fit to these data (light blue).
+C) Parameter estimates for annual fertility rates and adult survival probability, with estimates highlighted in orange showing the parameter estimates which result in positive population growth.
+B) Simulated vaccination coverage (N = 1000) using the demographic parameters from (C) in a hypothetical commune with 1000 dogs for three different campaign strategies: 1) annual vaccination campaigns targeting dogs of all ages (purple), 2) routine vaccination of puppies at 3 months of age, and 3) a combined strategy with campaigns annually and continuous puppy vaccination in between campaigns.*
 
@@
 
-Authors should discuss the results and how they can be interpreted in perspective of previous studies and of the working hypotheses.
-The findings and their implications should be discussed in the broadest context possible.
-Future research directions may also be highlighted.
+Through vaccination campaigns implemented in Moramanga District of Madagascar, we saw high demand for vaccination from dog owners and found that dogs were accessible and able to be handled safely and efficiently for parenteral vaccination at a reasonable cost (between 1.3 - 1.8 USD per animal vaccinated).
+We find that providing the vaccine at no direct cost to dog owners will be critical to achieving sufficient coverage, as even with nominal fees, a significant proportion of owners indicated they would no longer vaccinate their animals.
+A static point vaccination strategy achieved higher coverage over a shorter time period in 2018 compared to dog vaccinations conducted by the District Veterinarian as part of routine veterinary service provision in 2019.
+However, it came at a higher cost per animal vaccinated, was more limited in geographic scope, and required more resources in terms of personnel.
+In addition, in the rural setting of Andasibe, the static point campaign strategy achieved lower coverage, reflecting less accessible, hard-to-reach human communities in this location.
+Based on the lessons learned through these campaigns, in particular the observation that puppies were relatively easy to handle for vaccinators and owners, we find that continuous vaccination targeting puppies in particular may be an effective way to maintain vaccination coverage levels given high turnover in dog populations.
 
-# Conclusion
+There were several limitations to our analyses.
+Owner reports of willingness to pay, age of animals, and distance travelled to the vaccination point likely all suffer from recall bias and uncertainty, whilst willingness to pay studies are often overestimates of observed practice [@birhane2016].
+In addition, the age of animals brought to vaccination points may not be representative of the age structure of the underlying population, and other work has shown that in general puppies (individuals < 1 year) tend to have lower vaccination coverage than adults [@arief2016; @sanchez-soriano2020; @mazeri2019; @minyoo2015].
+Additionally, we did not vaccinate animals less than 1 month of age, which may explain why we sometimes estimated declining populations (as part of the age class of 0-1 years was not captured by the data).
+However, our analyses based on these data are consistent with previous findings from SSA that demonstrate male-biased populations, skewed towards puppies, and with high mortality in the first year of life [@czupryna2016; @taylor2017].
+Studies have also shown that charging fees for vaccines can greatly reduce the efficacy of campaigns and are a significant barrier to vaccination for owners [@schildecker2016; @yoak2021; @barbosacosta2020].
+Similarly, distance to campaign points has been identified as a barrier to vaccination, and in most cases owners report traveling less than 1km to reach a vaccination point [@mazeri2018barriers; @mazeri2019].
+We used HDRs, which can be sensitive to underlying estimates of the human population and the spatial scale of estimation [@sambo2018estimating; @sambo2017].
+However, we used HDR estimates from a recent household survey study in the district, and found these to give coverage estimates consistent with those from post-vaccination transects.
+Finally, in our vaccination model, we make several simplifying assumptions: we assume that immunity wanes after three years (simulating vaccination with a long-lasting vaccine such as Nobivac), but we also assume that vaccinated dogs that survive to the following year are revaccinated in subsequent campaigns (effectively assuming boosting per manufacturer recommendations for Rabisin).
+Overall waning vaccinal immunity plays a lesser role in declining vaccination coverage given the high population turnover in this context.
+In our age-structured models, we also do not account for population carrying capacity, which likely overestimates growth of the dog population.
 
-This section is not mandatory, but can be added to the manuscript if the discussion is unusually long or complex.
+Our estimates of costs per animal vaccinated are in line with recent estimates from other countries [@undurraga2020], although we likely underestimate costs given the donation of time and resources by the organizations and individuals involved (i.e.
+costs associated with international volunteers including airfare and visa costs as well as costs of transportation donated to the campaign).
+We also did not include costs of pre-exposure prophylaxis as all of our volunteers and vaccinators had been vaccinated prior to the campaigns.
+Both pre- and post-exposure vaccines should however always be readily available for vaccinators and should be included in vaccination program budgets.
+While the volunteer-led effort resulted in significant financial and personnel resources being devoted to the campaign, costs were higher overall and per animal vaccinated.
+In addition, NGO and volunteer based campaigns may be difficult to sustain given unpredictable funding, time commitments, and turnover in staff [@ferguson2020].
+Dog vaccination delivered through the District Veterinarian was less costly, with the majority of the costs directly related to vaccination.
+Although we included costs of implementing transect based coverage estimates, these were negligible (less than 0.10 USD per dog vaccinated), in line with recent studies which have shown this strategy is a cheap and effective way to estimate coverage @sambo2017; @gibson2015].
+
+One key aspect we do not consider is potential feedback loops between vaccination and demography.
+Estimates of the effects of vaccination on dog demography are mixed [@czupryna2016], but vaccination may increase dog survival [@conan2015].
+If dog population growth is driven by survival, then this could mean that increased vaccination results in increased population growth.
+However, if growth is driven more by demand from human communities, then vaccination could stabilize the population and reduce population turnover.
+Improving dog population management, encouraging responsible pet ownership practices, and increasing veterinary services could all complement vaccination efforts, but have not been demonstrated to result in meaningful rabies control without parallel dog vaccination [@taylor2017].
+
+In settings with high dog ownership, moving towards community-based vaccination strategies could be an effective way to achieve sufficient and consistent coverage, particularly in hard-to-reach communities.
+During our campaign, we found that puppies at approximately 3 months old were easy to handle for both vaccinators and owners.
+Puppy vaccination could be carried out by local officials embedded in communities (along the lines of community health workers), especially given recent findings on thermotolerance of rabies vaccines and locally manufactured methods for maintaining temperatures required for sustained vaccine storage (up to 3 months)
+[@lugelo2020].
+Incentivizing vaccinators appropriately will be a key challenge, as currently providing no-cost rabies vaccination is not seen as part of routine duties for district veterinarians or for livestock officers.
+Implementing dog vaccination alongside government mandated livestock vaccination campaigns may also be a strategy to scale up vaccination efforts at relatively low-cost.
+Expanding veterinary services across the country and relieving financial pressures on veterinarians and animal health workers through appropriate compensation could greatly improve veterinary services across Madagascar [@mondiale2017evaluation].
+For volunteer based efforts, focusing on local volunteers (i.e.
+veterinary students) may be a more cost-effective strategy.
+However similar to international volunteers, volunteers require subsistence during vaccination campaigns when not based in the communities where they study or live.
+
+Overall, our results suggest that dog vaccination is a feasible and effective strategy for controlling canine rabies in Madagascar.
+However, rabies vaccination must be recognized as a public good.
+Strategies to remove barriers for dog owners and to incentivize veterinarians and other animal health workers to implement vaccination will be key to long-term campaign success.
+Borrowing strategies from human vaccination efforts, i.e.
+deploying community health workers, could be a way to deliver vaccinations and reduce costs in the hardest-to-reach communities.
+In addition, refining vaccination strategies to local contexts and using improved tools and systems, such as mobile phone based data collection, could improve efficacy and lower costs of campaigns [@mazeri2021; @gibson2016a].
+To monitor the success of these campaigns, it will be critical to develop efficient and effective methods to estimate vaccination coverage and to measure their impact on reducing rabies incidence through robust surveillance [@lushasi2020; @ferguson2020].
+With limited chances for reintroduction from outside the island, Madagascar could be a model for rabies elimination, and strategies that tackle the context specific challenges could be a path forward for the country to reach ZeroBy30.
 
 # Patents
-
-This section is not mandatory, but may be added if there are patents resulting from the work reported in this manuscript.

### 2021-02-20 paper/mdpi_paper.tex
adding subfiles for trop med

@@
 %=================================================================
-\documentclass[water,article,submit,moreauthors,pdftex]{mdpi}
+\documentclass[tropicalmed,article,submit,moreauthors,pdftex]{mdpi}
 
@@
 % Full title of the paper (Capitalized)
-\Title{Full title of the paper (Capitalized)}
+\Title{Lessons learned and paths forward for rabies dog vaccination in
+Madagascar: a case study of pilot vaccination campaigns in Moramanga
+District}
 
 % Authors, for the paper (add full first names)
-\Author{Dominik
-Leutnant$^{1,2,\ddagger,*}$\href{https://orcid.org/0000-0003-3293-2315}{\orcidicon}, John
-Doe$^{2, \dagger, \ddagger}$}
+\Author{Caitlynn Filla$^{1, \dagger}$, Malavika Rajeev$^{2,
+\dagger}$\href{https://orcid.org/0000-0001-6030-4995}{\orcidicon}, Zoavina
+Randriana$^{3}$, Chantal Hanitriniana$^{4}$, Radoniaina R.
+Rafaliarison$^{3}$, Glenn Edosoa$^{5}$, Mamitiana
+Andriamananjara$^{6}$, Nivohanitra P. Razafindraibe$^{6}$, José
+Nely$^{7}$, Angelique Ferreira$^{3, 8}$, Annie Yang$^{2,
+\ddagger}$, Fenomanana Daniel$^{3}$, Tara A. Clarke$^{3, 9}$, Zachary
+Farris$^{3, 10}$, Terry Stone$^{8}$, Jochem Lastdrager$^{8}$, Tsiky
+Rajaonarivelo$^{3}$, Katie Hampson$^{11}$, C. Jessica E.
+Metcalf$^{2}$, Kim Valenta$^{1, 3}$}
 
 % Authors, for metadata in PDF
-\AuthorNames{Dominik Leutnant, John Doe}
+\AuthorNames{Caitlynn Filla, Malavika Rajeev, Zoavina Randriana, Chantal
+Hanitriniana, Radoniaina R. Rafaliarison, Glenn Edosoa, Mamitiana
+Andriamananjara, Nivohanitra P. Razafindraibe, José Nely, Angelique
+Ferreira, Annie Yang, Fenomanana Daniel, Tara A. Clarke, Zachary
+Farris, Terry Stone, Jochem Lastdrager, Tsiky Rajaonarivelo, Katie
+Hampson, C. Jessica E. Metcalf, Kim Valenta}
 
@@
 \address{%
-$^{1}$ \quad Muenster University of Applied Sciences - Institute for
-Infrastructure, Water, Resources, Environment Correnstr. 25, 48149
-Muenster,
-Germany; \href{mailto:leutnant@fh-muenster.de}{\nolinkurl{leutnant@fh-muenster.de}}\\
-$^{2}$ \quad Your department Street, City,
-Country; \href{mailto:mail@mail.com}{\nolinkurl{mail@mail.com}}\\
+$^{1}$ \quad University of Florida, Department of Anthropology
+Gainesville, Florida, USA; \\
+$^{2}$ \quad Department of Ecology and Evolutionary Biology, Princeton
+University Princeton, NJ, United States; \\
+$^{3}$ \quad The Mad Dog Initiative Akanin'ny Veterinera, Akaikiniarivo,
+Ambatobe, Antananarivo 101, Madagascar; \\
+$^{4}$ \quad Mention Zoologie et Biodiversité Animale, Faculté des
+Sciences Université d'Antananarivo, Antananarivo, Madagascar; \\
+$^{5}$ \quad Chargé des Maladies Tropicales Négligées Organisation
+mondiale de la Santé Madagascar, Antananarivo, Madagascar; \\
+$^{6}$ \quad Direction des Services Vétérinaires Ministère chargé de
+l'Agriculture et de l'Élevage, Antananarivo, Madagascar; \\
+$^{7}$ \quad Service contre les Maladies Endémo-épidémiques et
+Tropicales Négligées Ministère de la Santé Publique, Antananarivo,
+Madagascar; \\
+$^{8}$ \quad Travelling Animal Doctors 919 Crossan Rd, Newark, Delaware,
+USA; \\
+$^{9}$ \quad North Carolina State University, Department of Sociology
+and Anthropology Raleigh, North Carolina, USA; \\
+$^{10}$ \quad Appalachian State University, Department of Health and
+Exercise Science Boone, North Carolina, USA; \\
+$^{11}$ \quad Boyd Orr Centre for Population and Ecosystem Health
+Institute of Biodiversity, Animal Health and Comparative Medicine
+University of Glasgow, Glasgow, UK; \\
 }
 % Contact information of the corresponding author
-\corres{Correspondence: \href{mailto:leutnant@fh-muenster.de}{\nolinkurl{leutnant@fh-muenster.de}};
-Tel.: +XX-000-00-0000.}
+\corres{Correspondence: \href{mailto:mrajeev@princeton.edu}{\nolinkurl{mrajeev@princeton.edu}}}
 
 % Current address and/or shared authorship
-\firstnote{Current address: Updated affiliation}
-\secondnote{These authors contributed equally to this work.}
+\firstnote{These authors contributed equally to this work.}
+\secondnote{Deceased.}
 
@@
 % Simple summary
-\simplesumm{A Simple summary goes here.}
 
@@
 % Keywords
-\keyword{keyword 1; keyword 2; keyword 3 (list three to ten pertinent
-keywords specific to the article, yet reasonably common within the
-subject discipline.).}
+\keyword{canine rabies; mass dog vaccination; central point vaccination;
+puppy vaccination; Zeroby30;}
 
@@
 
+\usepackage{amsmath}
+\usepackage{booktabs}
+\usepackage{caption}
+\usepackage{longtable}
+
 \begin{document}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\hypertarget{version}{%
-\section{Version}\label{version}}
-
-This Rmd-skeleton uses the mdpi Latex template published 2019/02.
-However, the official template gets more frequently updated than the
-`rticles' package. Therefore, please make sure prior to paper
-submission, that you're using the most recent .cls, .tex and .bst files
-(available \href{http://www.mdpi.com/authors/latex}{here}).
 
@@
 
-The introduction should briefly place the study in a broad context and
-highlight why it is important. It should define the purpose of the work
-and its significance. The current state of the research field should be
-reviewed carefully and key publications cited. Please highlight
-controversial and diverging hypotheses when necessary. Finally, briefly
-mention the main aim of the work and highlight the principal
-conclusions. As far as possible, please keep the introduction
-comprehensible to scientists outside your particular field of research.
-Citing a journal paper
-\citep{bertrand-krajewski_distribution_1998, leutnant_stormwater_2016}.
-And now citing a book reference \citet{gujer_systems_2008}. Please use
-the command \citep{leutnant_stormwater_2016} for the following MDPI
-journals, which use author-date citation: Administrative Sciences, Arts,
-Econometrics, Economies, Genealogy, Humanities, IJFS, JRFM, Laws,
-Religions, Risks, Social Sciences.
+Canine rabies results in an estimated 60,000 human deaths per year
+globally \citep{hampson2015}. These deaths are entirely preventable:
+prompt post-exposure prophylaxis of humans exposed to rabies is highly
+effective at preventing death and mass dog vaccination can interrupt
+transmission in domestic dogs and eventually lead to disease elimination
+\citep{fooks2014}. The World Health Organization (WHO) and its partners
+have set a goal to eliminate human deaths due to canine rabies by the
+year 2030 (`ZeroBy30')\citep{abela-ridder2016}. Annual dog vaccination
+campaigns achieving at least 70\% coverage are the recommended target
+for controlling rabies in domestic dog populations
+\citep{worldhealthorganization2013}. However, achieving this coverage
+target in low and middle income countries where the burden of human
+rabies is highest can be challenging due to economic, ecological,
+sociocultural, and political barriers \citep{fahrion2017}.
 
-\hypertarget{materials-and-methods}{%
-\section{Materials and Methods}\label{materials-and-methods}}
+In sub-Saharan African countries, parenteral vaccinations implemented
+through static point campaigns have been shown to be cost-effective and
+feasible \citep{borse2018}. While most dogs in these settings are
+considered free roaming, they are mostly owned and are accessible for
+vaccination through these campaigns \citep{morters2014, jibat2015}.
+However, reaching high coverage and maintaining vaccination campaigns at
+scale requires sustained investment and coordination, and the challenges
+in implementation largely reflect financial and logistical constraints
+more than the feasibility of vaccination itself \citep{fahrion2017}.
+Developing clear and context-specific strategies and lowering costs and
+resources needed could help spur the implementation and scaling up of
+campaigns in these countries.
 
-Materials and Methods should be described with sufficient details to
-allow others to replicate and build on published results. Please note
-that publication of your manuscript implicates that you must make all
-materials, data, computer code, and protocols associated with the
-publication available to readers. Please disclose at the submission
-stage any restrictions on the availability of materials or information.
-New methods and protocols should be described in detail while
-well-established methods can be briefly described and appropriately
-cited.
+In Madagascar, canine rabies has been endemic for over a century, and
+for most of that period, the Institut Pasteur de Madagascar has provided
+post-exposure prophylaxis free-of-charge to bite patients in the country
+\citep{reynes2011}. Currently, there are only 31 clinics where these
+human vaccines are available, and there is minimal dog vaccination due
+to high costs to owners and lack of vaccine availability
+\citep{rajeev2018}. Recent studies have estimated a high burden of human
+deaths, masked by weak surveillance across the country
+\citep{rajeev2020, rajeev2018}.
 
-Research manuscripts reporting large datasets that are deposited in a
-publicly available database should specify where the data have been
-deposited and provide the relevant accession numbers. If the accession
-numbers have not yet been obtained at the time of submission, please
-state that they will be provided during review. They must be provided
-prior to publication.
+The veterinary sector is largely private, with 204 veterinarians
+employed in hybrid private/public employment as designated district
+veterinarians. While dog vaccination is rare, livestock officers and
+veterinarians work together to implement cattle vaccination campaigns
+for anthrax on an annual basis as mandated by the government (owners are
+charged a fee per animal for these vaccines which vary by location)
+\citep{mondiale2017evaluation}. No routine mass dog vaccinations have
+been conducted on the island, although a few pilot programs have begun
+in recent years, largely implemented by NGO-government partnerships.
 
-Interventionary studies involving animals or humans, and other studies
-require ethical approval must list the authority that provided approval
-and the corresponding ethical approval code.
+Here, we summarize lessons learned through the implementation of pilot
+vaccination programs in the Moramanga District of Madagascar, where
+previous work has shown high incidence of dog rabies cases and human
+rabies exposures. In 2018 and 2019, we deployed two different
+vaccination strategies. In 2018, we carried out a larger scale
+volunteer-led pilot vaccination campaign in two communes (sub-district
+level) using a static point strategy. In 2019/2020, we provided
+vaccines, all necessary supplies for vaccine administration, and a
+per-vaccine fee to the District Veterinarian to vaccinate animals as
+part of routine services delivered continuously. We compare time, human
+resources, costs, and coverage estimates between these campaigns, and
+using a demographic and vaccination model we further explore different
+vaccination strategies based on what we learned during implementation.
+
+\hypertarget{methods}{%
+\section{Methods}\label{methods}}
+
+\hypertarget{study-area}{%
+\subsection{Study Area}\label{study-area}}
+
+The Moramanga District is located midway between the central highlands
+and the east coast of Madagascar, at an average altitude of 936 m. It
+comprises 21 communes, covering approximately 7150 km\textsuperscript{2}
+with an approximate human population of 347,000\citep{instat2021}.
+Previous work in the district has established a high burden of rabies
+exposures and deaths despite the availability of post-exposure
+prophylaxis at the district hospital \citep{rajeev2018}. While Moramanga
+is relatively close to the capital city of Antananarivo
+(\textasciitilde{} 3 hours by bus), within the district, travel times
+between locations are highly variable, with much of the population
+living in more rural areas with limited access to roads and
+transportation \citep{rajeev2020}. Before 2018, there were limited
+animal rabies vaccination services, with most animal vaccines available
+in the urban commune of Moramanga Ville where owners were often charged
+\textgreater{} 15,000 Ariary (\textasciitilde{} 4.28 USD) per vaccine
+administered.
+
+\hypertarget{campaign}{%
+\subsection{2018 Campaign}\label{campaign}}
+
+In 2018, two NGOs (the Mad Dog Initiative (MDI) and Traveling Animal
+Doctors (TAD)) organized a pilot vaccination campaign in collaboration
+with the Department of Veterinary Services, and the Ministry of Public
+Health in the District of Moramanga. This campaign focused on two
+communes in Moramanga, Moramanga Ville (the district center) and
+Andasibe (a rural commune surrounding Andasibe National Park), where
+previously high incidence of probable rabies exposures (Moramanga Ville)
+and a high burden of deaths (Andasibe) had been recorded
+\citep{rajeev2018}.
+
+The campaign was planned as a series of static point vaccination
+stations covering 1 - 3 fokontany (i.e., sub-communes) per day. A week
+before the campaign dates, the vaccination team informed the chief of
+the fokontany about the campaign and provided posters advertising the
+date of the vaccine and that it would be available at no cost to owners.
+During the campaign, we used Rabisin (10 mL vials with 1 mL per dose,
+Boehringer Ingelheim) to vaccinate both dogs and cats presented that
+were at least 1 month old. As part of the campaign, owners were surveyed
+by vaccinators about how many dogs and cats they owned in total (split
+by \textgreater{} 1 year vs.~\textless{} 1 year in order to avoid
+language ambiguities that might result in excluding puppies and
+kittens), as well as if their dogs were free roaming (no restrictions on
+movement by the owner/`mirenyreny', tied/`mifatora', or
+fenced/`mifefy'). Vaccinations were delivered at no cost to owners, but
+as animal vaccination is generally thought of as a paid service in
+Madagascar, owners were asked how much in Ariary they would be willing
+to pay to have one animal vaccinated against rabies. For each animal
+vaccinated, we recorded the species (cat or dog), sex, approximate age
+in years, and whether the animal had been previously vaccinated.
+
+To assess coverage, post-vaccination coverage surveys were conducted
+according to previously established
+methodology\citep{sambo2017, gibson2015}. All animals vaccinated were
+concurrently marked with a colored, non-toxic, livestock crayon along
+the top of the head or back. Between 4 - 6 PM when dogs are most active
+\citep{sambo2017} and on the same day as the campaign, in each
+vaccination location two transect surveys were done. Each team (two
+people per team) walked a transect for 1 hour starting in opposite
+directions and accompanied by a local guide to ensure that walking paths
+did not overlap. Teams recorded any marked and unmarked dogs observed,
+their roaming status (whether roaming, inside a fence, or tied), and
+their approximate age (greater or less than 1 year of age).
+
+\includegraphics[width=0.8\linewidth]{/Users/mrajeev/Documents/Projects/mora_vax/figs/fig1}
+
+\hypertarget{figure-1.-photos-from-the-2018-campaign.-topleft-advertisement-for-the-campaign-posted-on-the-door-of-the-fokontany-office-as-the-campaign-starts-topright-a-dog-post-vaccination-marked-with-a-crayon-bottomleft-a-basket-of-puppies-brought-for-vaccination-bottomright-a-line-of-owners-and-dogs-waiting-for-vaccination.}{%
+\paragraph{\texorpdfstring{\textbf{Figure 1. Photos from the 2018
+campaign}. \emph{Topleft: advertisement for the campaign posted on the
+door of the fokontany office as the campaign starts; Topright: a dog
+post-vaccination marked with a crayon; Bottomleft: a basket of puppies
+brought for vaccination; Bottomright: a line of owners and dogs waiting
+for
+vaccination.}}{Figure 1.
+Photos from the 2018 campaign.
+Topleft: advertisement for the campaign posted on the door of the fokontany office as the campaign starts; Topright: a dog post-vaccination marked with a crayon; Bottomleft: a basket of puppies brought for vaccination; Bottomright: a line of owners and dogs waiting for vaccination.}}\label{figure-1.-photos-from-the-2018-campaign.-topleft-advertisement-for-the-campaign-posted-on-the-door-of-the-fokontany-office-as-the-campaign-starts-topright-a-dog-post-vaccination-marked-with-a-crayon-bottomleft-a-basket-of-puppies-brought-for-vaccination-bottomright-a-line-of-owners-and-dogs-waiting-for-vaccination.}}
+
+\hypertarget{campaign-1}{%
+\subsection{2019 Campaign}\label{campaign-1}}
+
+For the 2019 campaign, instead of a static point campaign strategy,
+vaccine vials(Rabisin) and the supplies needed to administer them
+(needle, syringe, vaccination card for owners) were distributed to the
+District Veterinarian, who then delivered the vaccination at no cost to
+owners, but was directly compensated 1,500 Ar (\textasciitilde{} 0.40
+USD) per rabies vaccine administered. The campaign lasted from September
+6, 2019 to June 19, 2020, One week prior to her visiting each location,
+the District Veterinarian advertised the vaccines by calling ahead to
+the fokontany leaders and other officials who then advertised to their
+communities, largely through word-of-mouth. For each animal vaccinated,
+the District Veterinarian collected the animal's age and sex, and also
+asked owners to approximate the distance they travelled to receive the
+vaccination in meters. Researchers communicated with the District
+Veterinarian about progress periodically throughout the campaign,
+primarily through telephone calls. No other compensation or instructions
+were provided, and we asked the District Veterinarian to administer as
+many (or as few) as feasible or wanted. As the vaccinations were
+delivered continuously, we were unable to do comparable post-vaccination
+surveys.
+
+\hypertarget{analyses}{%
+\subsection{Analyses}\label{analyses}}
+
+\hypertarget{campaign-resource-and-cost-comparisons}{%
+\subsubsection{Campaign resource and cost
+comparisons}\label{campaign-resource-and-cost-comparisons}}
+
+We documented the overall costs and resources required for the two
+vaccination efforts. We tracked the number of vaccination points, the
+number of days over which these vaccinations occurred, and the number of
+person days required overall (i.e.~the number of working people per day
+over the campaign \citep{mazeri2021}), in addition to monetary costs. As
+costs were incurred in both USD and Ariary, and the exchange rate
+declined rapidly between 2018 and 2019, we used the midpoint between the
+two years (3314 Ar to 1 USD) for the cost comparisons.
+
+For the 2018 campaign, we broke costs down into the following
+categories: direct vaccine costs (cost for vaccine, syringes, needles,
+vaccination cards), supplies (livestock crayons, muzzles, gloves,
+alcohol, swabs), food and lodging for NGO personnel and other
+vaccinators during the campaign, personnel costs (per diems for the
+district veterinarian, livestock field officers, local guides, and NGO
+employees), and advertisement (posters and banners for advertising the
+campaign). Foreign NGO volunteer expenses for travel to Madagascar were
+not included in these costs. Vehicles and drivers are also not included
+in these costs, as the drivers' time and vehicle use were donated to the
+campaign by volunteers involved in the campaign. In 2019, costs were
+split into two categories, direct vaccine costs (for same items as in
+2018), and personnel costs (per vaccine fee paid to the district
+veterinarian), and supplies (a generator and fuel for the veterinarian
+to maintain the vaccine under cold chain during power outages).
+Transportation costs were also not included as the District Veterinarian
+used her own vehicle and vaccinated as part of their routine veterinary
+service provisioning.
+
+We used the data on owners' reported willingness to pay for vaccines to
+estimate the proportional reduction in animals vaccinated as fees are
+increased. We also estimated how this would impact cost per animal
+vaccinated by approximating the costs for implementation (i.e.~those
+costs that remain fixed) from costs incurred per animal vaccinated
+(i.e.~vaccine, syringe, vaccination card, and per vaccination fee to
+District Veterinarian in 2019), and calculating the balance between the
+returns from owner payments (i.e.~increases in cost recovery per animal
+vaccinated) vs.~decreasing numbers of animals vaccinated overall.
+
+\hypertarget{coverage-estimates}{%
+\subsubsection{Coverage estimates}\label{coverage-estimates}}
+
+For the 2018 campaign, we used the transect data to estimate vaccination
+coverage as the proportion of dogs sighted that were marked using a
+binomial confidence interval at the commune level. For the 2019
+campaign, we estimated vaccination coverage using human-to-dog ratios
+(HDRs) and human population estimates
+\citep{cleaveland2003, athingo2020}. We used a range of 8 - 25, based on
+previous data from Madagascar \citep{ratsitorahina2009} and based on
+recent estimates from household surveys in the Moramanga District
+\citep{leblancclaireRabiesMadagascarThreepronged2019}. We set the point
+estimate using an HDR of 19.5, the midpoint between the HDRs estimated
+for two communities in the District by LeBlanc et al.~2019. We used
+human population estimates from the 2018 national census in each commune
+where the vaccinations were delivered \citep{instat2021}. Coverage was
+estimated as the number of dogs vaccinated in total in that commune
+divided by the estimated dog population. We used this same method for
+the 2018 campaign, as well, to compare coverage estimated by the
+post-vaccination transects vs.~HDRs.
+
+\hypertarget{dog-demography}{%
+\subsubsection{Dog Demography}\label{dog-demography}}
+
+Using the age data on vaccinated animals collected during both
+vaccination campaigns, we estimated the proportion of the population in
+four age classes: puppies under the age of 1 year, juveniles aged 1 - 2
+yrs, adults aged 2 - 6 years, and older dogs aged 6+ years based on .
+With the assumption that these estimates represent the population at a
+stable age distribution, we use a Leslie matrix model to estimate annual
+adult survival probability and fertility using maximum likelihood
+estimation \citep{Fujiwara2017}. Specifically, we assume that the number
+of individuals in each age class follows a Poisson distribution with the
+mean predicted by the stable age distribution from the model (the
+proportion of individuals in each age class at equilibrium, equal to the
+eigenvector associated with the dominant eigenvalue of the matrix
+\(\nu\)) multiplied by the total number of individuals in the population
+(\(N_t\)):
+
+\[ N_{a} \sim Pois(\nu N_t) \]
+
+We assume that all individuals older than 1 year of age reproduce, and
+we do not estimate declines in fertility given the small proportion of
+dogs older than age six years in the population. To get bootstrapped
+estimates, we used 100 sub-sampled data sets of 1000 observations each
+from the observed age data to fit the parameters, and also varied
+initial values used in the optimization (N = 100 initial values sets)
+for 10,000 parameter estimates total.
+
+\hypertarget{modeling-vaccination-campaign-strategies}{%
+\subsubsection{Modeling vaccination campaign
+strategies}\label{modeling-vaccination-campaign-strategies}}
+
+We used the parameter estimates from the demographic model to simulate
+different vaccination strategies in a hypothetical commune with 1000
+dogs. We used a discrete time age-structured model with a monthly time
+step to compare three strategies:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi})}
+\tightlist
+\item
+  Annual vaccination campaigns occurring within the same month each year
+  targeting dogs of all ages
+\item
+  Continuous vaccination of new puppies throughout the year targeting
+  puppies that reach the age of 3 months
+\item
+  A combined approach with annual campaigns (as 1) and routine puppy
+  vaccination (as 2)
+\end{enumerate}
+
+We split the dog population into puppies (\textless{} 1 year old) and
+adults based on the stable age distribution estimated from the
+demographic model. To estimate pup survival in year one, we took the
+fertility estimates (i.e., number of new puppies per reproducing dog
+observed in the pup age class) and divided by an estimate of newborn
+pups per reproducing dog each year based on average litter size, average
+number of litters per female per year
+\citep{hampson2009, morters2014, czupryna2016}, and the proportion of
+the adult population that is female (estimated from our data). We
+assumed that for the annual campaign strategy, surviving vaccinated
+adult dogs were revaccinated in subsequent years
+\citep{delriovilas2017}, but that if a pup had been vaccinated within 9
+months of the campaign, it was not revaccinated. We also assumed that
+vaccine immunity lasted for a discrete period of 3 years (with
+revaccination resetting immunity). A subset of parameter estimates
+resulted in estimates of population decline, but based on the shape of
+the age pyramid and to simulate reasonable campaign scenarios, we
+filtered to parameter estimates which corresponded to positive
+population growth.
+
+\hypertarget{data-and-ethics-statement}{%
+\paragraph{Data and ethics statement}\label{data-and-ethics-statement}}
+
+All data were analysed in R version 4.0.2 (2020-06-22) \citep{R2020},
+largely using the \texttt{tidyverse} package suite \citep{wickham2019}.
+Geospatial data were mapped using the sf \citep{sf} package. All data
+and code are available at \url{https://github.com/mrajeev08/mora_vax}.
+The vaccinations were part of a public health campaign and routine
+veterinary service provisioning carried out by the local veterinary
+officials, the NGOs involved, and in partnership with the Ministry of
+Public Health and the Department of Veterinary Services at the national
+level. MDI also maintains national research permits (MICET permit:
+\#130-19/MEDD/SG/DGEF/DGRNE) for its research and volunteer programs.
 
@@
 
-This section may be divided by subheadings. It should provide a concise
-and precise description of the experimental results, their
-interpretation as well as the experimental conclusions that can be
-drawn.
+\hypertarget{summary-of-2018-and-2019-campaigns}{%
+\subsection{Summary of 2018 and 2019
+campaigns}\label{summary-of-2018-and-2019-campaigns}}
 
-\hypertarget{subsection-heading-here}{%
-\subsection{Subsection Heading Here}\label{subsection-heading-here}}
+During the 2018 campaign, a total of 3137 animals were vaccinated (2057
+dogs and 1080 cats) in the Moramanga (urban) and Andasibe (rural)
+communes over 13 days during the month of April (Table 1). We vaccinated
+at 7 points in Andasibe and 14 points in Moramanga Ville. During the
+2019 campaign, between September 2019 - June 2020, the District
+Veterinarian vaccinated a total of 2385animals (1898 dogs and 486 cats)
+over 48 days in seven communes in the Moramanga District. While more
+animals were vaccinated per vaccination point and per vaccination day in
+2018 compared to 2019, the number of animals vaccinated per person-day
+was much higher for the 2019 campaign. More animals were vaccinated in
+2018 vs.~2019, but this was largely a result of vaccinating more cats
+during the 2018 campaign (Table 1).
 
-Subsection text here.
+In 2018, 15\% of dogs had a previous history of vaccination (largely in
+the urban commune of Moramanga Ville) with only 7\% of dogs vaccinated
+within the last year. This remained largely the same in 2019
+(\textasciitilde{} 13\%), as the District Veterinarian focused their
+efforts in other communes. The District Veterinarian did vaccinate 771
+dogs in Moramanga Ville in 2019, and of those 24.9\% had been vaccinated
+in the previous year's campaign, whereas only 4.9\% of dogs vaccinated
+in all other communes had any history of previous vaccination. In
+addition, in 2019, 2.1\% of animals had been spayed or neutered,
+reflecting efforts by the Mad Dog Initiative to implement free spay and
+neuter clinics in the district.
 
-\hypertarget{subsubsection-heading-here}{%
-\subsubsection{Subsubsection Heading
-Here}\label{subsubsection-heading-here}}
+In 2018, 19\% of owners reported their animals were free-roaming, but
+this varied by location (Table 1). In addition, while less than 19\% of
+owners in Moramanga Ville reported that their animals were free roaming,
+the majority of animals observed during the transects (77\%) were
+observed outside of fences and not tied, and thus the majority of
+animals could be classified as semi-confined in the more urban township
+of Moramanga Ville, and free-roaming in the rural setting of Andasibe
+(approximately 67\% of owners reported their dogs as free-roaming, Table
+1). In 2019, the District Veterinarian also asked owners to approximate
+how far they travelled in meters to get their animals vaccinated, and
+94\% of people reportedly travelled less than 1 km to reach the
+vaccination point.
 
-Bulleted lists look like this:
+\hypertarget{table-1.-summary-of-2018-and-2019-campaigns.-breakdown-of-of-animals-vaccinated-prior-vaccination-history-dog-demography-dog-ownership-daily-and-per-vaccination-rates-by-year-and-location-for-2018.-photo-credict-jochem-lastdrager-traveling-animal-doctors.}{%
+\paragraph{\texorpdfstring{\textbf{Table 1. Summary of 2018 and 2019
+campaigns}. \emph{Breakdown of of animals vaccinated, prior vaccination
+history, dog demography, dog ownership, daily and per vaccination rates
+by year and location (for 2018). Photo credict: Jochem Lastdrager,
+Traveling Animal
+Doctors.}}{Table 1.
+Summary of 2018 and 2019 campaigns.
+Breakdown of of animals vaccinated, prior vaccination history, dog demography, dog ownership, daily and per vaccination rates by year and location (for 2018).
+Photo credict: Jochem Lastdrager, Traveling Animal Doctors.}}\label{table-1.-summary-of-2018-and-2019-campaigns.-breakdown-of-of-animals-vaccinated-prior-vaccination-history-dog-demography-dog-ownership-daily-and-per-vaccination-rates-by-year-and-location-for-2018.-photo-credict-jochem-lastdrager-traveling-animal-doctors.}}
 
-\begin{itemize}
-\tightlist
-\item
-  First bullet
-\item
-  Second bullet
-\item
-  Third bullet
-\end{itemize}
+\captionsetup[table]{labelformat=empty,skip=1pt}
+\begin{longtable}{lllll}
+\toprule
+& \multicolumn{3}{c}{2018} & \\ 
+ \cmidrule(lr){2-4}
+ & Andasibe & Moramanga Ville & All Communes & All Communes \\ 
+\midrule
+Total animals 
+ vaccinated & 528 & 2609 & 3137 & 2385 \\ 
+Total dogs 
+ vaccinated & 254 & 1,803 & 2,057 & 1,898 \\ 
+Dogs with history of vaccination & 5\% & 16\% & 15\% & 13\% \\ 
+Dogs vaccinated within last year & 5\% & 7\% & 7\% & 13\% \\ 
+Percent male dogs & 55\% & 56\% & 56\% & 65\% \\ 
+Average dogs per owner & 0.8 & 1.1 & 1.0 & NA \\ 
+Percent of owners with 
+ free-roaming dogs & 67\% & 19\% & 28\% & NA \\ 
+Animals vaccinated per day (total days) & 88 (6) & 372.7 (7) & 241.3 (13) & 49.7 (48) \\ 
+Animals vaccinated per 
+ vaccination point (total points) & 75.4 (7) & 186.4 (14) & 149.4 (21) & 37.3 (64) \\ 
+Animals vaccinated per 
+ person day (total person days) & 11.7 (45) & 21.6 (121) & 18.9 (166) & 49.7 (48) \\ 
+\bottomrule
+\end{longtable}
 
-Numbered lists can be added as follows:
+\hypertarget{cost-comparison-and-willingness-to-pay}{%
+\subsection{Cost comparison and willingness to
+pay}\label{cost-comparison-and-willingness-to-pay}}
 
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  First item
-\item
-  Second item
-\item
-  Third item
-\end{enumerate}
+The 2018 campaign cost more overall and per animal vaccinated than the
+2019 campaign, largely due to increased personnel costs (fig 2A \& B,
+Table 1). Reflecting the extra personnel necessary to run the static
+point campaign, the 2018 campaign also took substantially more
+person-days per animal vaccinated (Table 1). We found that charging
+owners for vaccination would result in minimal cost recovery (fig 2C)
+and beyond a minimal cost would actually result in increased costs per
+vaccinated individual than free-of-charge campaigns. Cost recovery would
+be more likely given a 2019 style strategy where the majority of the
+costs are incurred on a per animal basis, compared to the 2018 campaigns
+where the costs are largely due to setting up the static point
+vaccination stations (fig 4A). More importantly, in all cases, even a
+nominal fee would significantly reduce numbers of dogs vaccinated and
+thus vaccination coverage, particularly in the rural commune of Andasibe
+(fig 2D).
 
-The text continues here.
+\includegraphics[width=0.8\linewidth]{/Users/mrajeev/Documents/Projects/mora_vax/figs/fig2}
 
-All figures and tables should be cited in the main text as Figure 1,
-Table 1, etc.
+\hypertarget{figure-2.-comparing-campaign-costs-and-willingness-to-pay.-a-vaccine-costs-broken-down-by-category-and-by-year-colors.-b-overall-cost-per-animal-vaccinated-for-the-two-campaign-years-split-by-direct-costs-of-vaccination-per-animal-i.e.-vaccine-vaccination-card-syringes-and-baseline-implementation-costs-i.e.-personnel-supplies-subsistence-costs-for-vaccinators-during-the-campaign.-c-estimated-cost-per-animal-vaccinated-under-a-willingness-to-pay-model-for-two-campaigns-examining-increasing-costs-charged-to-the-owner-with-estimated-costs-declining-due-to-cost-recovery-through-owner-payments-and-then-peaking-once-owners-report-to-be-no-longer-willing-to-pay-for-the-vaccine.-d-the-percent-reduction-in-number-of-animals-vaccinated-given-owners-willingness-to-pay.-the-curves-in-c-and-d-are-shown-based-on-the-overall-responses-on-willingness-to-pay-solid-line-from-both-moramanga-and-andasibe-and-the-responses-split-by-commune-dashed-and-dotted-lines.}{%
+\paragraph{\texorpdfstring{\textbf{Figure 2. Comparing campaign costs
+and willingness to pay}. \emph{A) Vaccine costs broken down by category
+and by year (colors). B) Overall cost per animal vaccinated for the two
+campaign years, split by direct costs of vaccination per animal
+(i.e.~vaccine, vaccination card, syringes) and baseline implementation
+costs (i.e.~personnel, supplies, subsistence costs for vaccinators
+during the campaign). C) Estimated cost per animal vaccinated under a
+willingness to pay model for two campaigns examining increasing costs
+charged to the owner, with estimated costs declining due to cost
+recovery through owner payments and then peaking once owners report to
+be no longer willing to pay for the vaccine. D) The percent reduction in
+number of animals vaccinated given owners willingness to pay. The curves
+in C and D are shown based on the overall responses on willingness to
+pay (solid line) from both Moramanga and Andasibe, and the responses
+split by Commune (dashed and dotted
+lines).}}{Figure 2.
+Comparing campaign costs and willingness to pay.
+A) Vaccine costs broken down by category and by year (colors).
+B) Overall cost per animal vaccinated for the two campaign years, split by direct costs of vaccination per animal (i.e.~vaccine, vaccination card, syringes) and baseline implementation costs (i.e.~personnel, supplies, subsistence costs for vaccinators during the campaign).
+C) Estimated cost per animal vaccinated under a willingness to pay model for two campaigns examining increasing costs charged to the owner, with estimated costs declining due to cost recovery through owner payments and then peaking once owners report to be no longer willing to pay for the vaccine.
+D) The percent reduction in number of animals vaccinated given owners willingness to pay.
+The curves in C and D are shown based on the overall responses on willingness to pay (solid line) from both Moramanga and Andasibe, and the responses split by Commune (dashed and dotted lines).}}\label{figure-2.-comparing-campaign-costs-and-willingness-to-pay.-a-vaccine-costs-broken-down-by-category-and-by-year-colors.-b-overall-cost-per-animal-vaccinated-for-the-two-campaign-years-split-by-direct-costs-of-vaccination-per-animal-i.e.-vaccine-vaccination-card-syringes-and-baseline-implementation-costs-i.e.-personnel-supplies-subsistence-costs-for-vaccinators-during-the-campaign.-c-estimated-cost-per-animal-vaccinated-under-a-willingness-to-pay-model-for-two-campaigns-examining-increasing-costs-charged-to-the-owner-with-estimated-costs-declining-due-to-cost-recovery-through-owner-payments-and-then-peaking-once-owners-report-to-be-no-longer-willing-to-pay-for-the-vaccine.-d-the-percent-reduction-in-number-of-animals-vaccinated-given-owners-willingness-to-pay.-the-curves-in-c-and-d-are-shown-based-on-the-overall-responses-on-willingness-to-pay-solid-line-from-both-moramanga-and-andasibe-and-the-responses-split-by-commune-dashed-and-dotted-lines.}}
 
-\begin{figure}[H]
-\centering
-\includegraphics[width=3 cm]{logo-mdpi}
-\caption{This is a figure, Schemes follow the same formatting.
-If there are multiple panels, they should be listed as: (\textbf{a}) Description of what is contained in the first panel.
-(\textbf{b}) Description of what is contained in the second panel.
-Figures should be placed in the main text near to the first time they are cited.
-A caption on a single line should be centered.}
-\end{figure}
+\hypertarget{comparing-campaign-coverage-estimates}{%
+\subsection{Comparing campaign coverage
+estimates}\label{comparing-campaign-coverage-estimates}}
 
-\begin{table}[H]
-\caption{This is a table caption. Tables should be placed in the main text near to the first time they are cited.}
-\centering
-%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
-\begin{tabular}{ccc}
-\toprule
-\textbf{Title 1}    & \textbf{Title 2}  & \textbf{Title 3}\\
-\midrule
-entry 1     & data          & data\\
-entry 2     & data          & data\\
-\bottomrule
-\end{tabular}
-\end{table}
+The 2018 campaign covered two communes and was estimated to have
+achieved approximately 60\% coverage (fig 3A). The 2019 campaign covered
+seven communes, but was estimated to have achieved lower coverage levels
+(fig 3, ranging from 5 - 60\%). In the 2018 campaign, we used
+post-vaccination coverage transects to estimate vaccination coverage,
+but we were unable to do this in 2019 given the continuous delivery
+strategy. In addition, in Andasibe in 2018, coverage estimates were
+based on a single transect resulting in more uncertainty. However,
+coverage estimates from the transects in 2018 were consistent with the
+HDR based estimates (for both Andasibe and Moramanga, transect-based
+estimates fell within the range of the HDR estimates). We also
+back-calculated HDRs given our vaccination coverage estimates, and these
+were similar to the HDRs calculated from the household survey (15.7 -
+32.8 for Andasibe compared to 21.7 in a rural community and 17.8 - 21.2
+for Moramanga Ville compared to 17.2 in an urban community).
 
-This is an example of an equation:
+\includegraphics[width=0.8\linewidth]{/Users/mrajeev/Documents/Projects/mora_vax/figs/fig3}
 
-\begin{equation}
-\mathbb{S}
-\end{equation}
+\hypertarget{figure-3.-estimates-of-coverage-achieved-by-the-2018-and-2019-campaigns.-a-the-commune-level-numbers-of-dogs-vaccinated-size-of-points-and-the-associated-coverage-estimates-color-of-points-for-the-year-2018-estimated-using-post-vaccination-transects-and-2019-estimated-using-a-human-to-dog-ratio-hdr-of-19.5-based-on-a-recent-household-survey-in-the-moramanga-district.-the-inset-shows-the-location-of-the-moramanga-district-in-madagascar.-b-a-comparison-of-coverage-estimates-by-location-and-by-method-of-estimation-post-vaccination-transects-vs.-hdr-based-estimates-for-the-transect-based-estimates-the-line-range-shows-the-95-exact-binomial-confidence-interval-while-for-the-hdr-based-estimates-the-line-range-shows-the-range-of-coverage-estimates-assuming-an-hdr-range-of-8---25-according-to-estimates-from-the-literature.}{%
+\paragraph{\texorpdfstring{\textbf{Figure 3. Estimates of coverage
+achieved by the 2018 and 2019 campaigns}. \emph{A) The commune level
+numbers of dogs vaccinated (size of points) and the associated coverage
+estimates (color of points) for the year 2018 (estimated using
+post-vaccination transects) and 2019 (estimated using a human-to-dog
+ratio (HDR) of 19.5, based on a recent household survey in the Moramanga
+District). The inset shows the location of the Moramanga District in
+Madagascar. B) A comparison of coverage estimates by location and by
+method of estimation (post-vaccination transects vs.~HDR based
+estimates); for the transect based estimates, the line range shows the
+95\% exact binomial confidence interval while for the HDR based
+estimates, the line range shows the range of coverage estimates assuming
+an HDR range of 8 - 25 according to estimates from the
+literature.}}{Figure 3.
+Estimates of coverage achieved by the 2018 and 2019 campaigns.
+A) The commune level numbers of dogs vaccinated (size of points) and the associated coverage estimates (color of points) for the year 2018 (estimated using post-vaccination transects) and 2019 (estimated using a human-to-dog ratio (HDR) of 19.5, based on a recent household survey in the Moramanga District).
+The inset shows the location of the Moramanga District in Madagascar.
+B) A comparison of coverage estimates by location and by method of estimation (post-vaccination transects vs.~HDR based estimates); for the transect based estimates, the line range shows the 95\% exact binomial confidence interval while for the HDR based estimates, the line range shows the range of coverage estimates assuming an HDR range of 8 - 25 according to estimates from the literature.}}\label{figure-3.-estimates-of-coverage-achieved-by-the-2018-and-2019-campaigns.-a-the-commune-level-numbers-of-dogs-vaccinated-size-of-points-and-the-associated-coverage-estimates-color-of-points-for-the-year-2018-estimated-using-post-vaccination-transects-and-2019-estimated-using-a-human-to-dog-ratio-hdr-of-19.5-based-on-a-recent-household-survey-in-the-moramanga-district.-the-inset-shows-the-location-of-the-moramanga-district-in-madagascar.-b-a-comparison-of-coverage-estimates-by-location-and-by-method-of-estimation-post-vaccination-transects-vs.-hdr-based-estimates-for-the-transect-based-estimates-the-line-range-shows-the-95-exact-binomial-confidence-interval-while-for-the-hdr-based-estimates-the-line-range-shows-the-range-of-coverage-estimates-assuming-an-hdr-range-of-8---25-according-to-estimates-from-the-literature.}}
 
-Example of a theorem:
+\hypertarget{dog-demography-and-simulating-vaccination-strategies}{%
+\subsection{Dog demography and simulating vaccination
+strategies}\label{dog-demography-and-simulating-vaccination-strategies}}
 
-\begin{Theorem}
-Example text of a theorem.
-\end{Theorem}
+Demographic data from vaccinated dogs shows a population pyramid with a
+large base, indicative of a fast-growing population, and with a male
+bias (approximately 60\% male, Fig 4A). We fit these data to an
+age-structured model, and were able to generate parameter estimates
+which resulted in stable age distributions consistent with the data (Fig
+4B). We filter to parameter estimates that are consistent with a growing
+population resulting in mean adult annual survival probability of 0.77
+(95\% quantile: 0.59 - 0.97). We use estimates of fertility (on average
+1.09, 95\% quantile: 0.82 - 1.41) to back-calculate pup survival, which
+ranges from 0.34 - 0.68. We find that given these demographic
+parameters, annual campaigns that target dogs of all ages result in
+rapid decline in vaccination coverage between campaigns, largely due to
+rapid turnover of the dog population (compared to the impact of waning
+immunity assuming a discrete 3 year period of vaccine immunity, evident
+in the additional dip at year 3, fig 4D). Continuously targeting 70\% of
+the puppy population for vaccination, while unable to achieve the peak
+coverage, consistently reaches coverage of about 50\% of the dog
+population. A combined strategy maintains the highest and most
+temporally stable levels of coverage close to the target of 70\%.
 
-The text continues here. Proofs must be formatted as follows:
+\includegraphics[width=0.8\linewidth]{/Users/mrajeev/Documents/Projects/mora_vax/figs/fig4}
 
-Example of a proof:
-
-\begin{proof}[Proof of Theorem 1]
-Text of the proof. Note that the phrase `of Theorem 1' is optional if it is clear which theorem is being referred to.
-\end{proof}
-
-The text continues here.
+\hypertarget{figure-4.-dog-demography-and-implications-for-campaign-strategies.-a-the-age-pyramid-for-vaccinated-dogs-by-sex.-b-bootstrapped-estimates-of-the-proportion-of-the-population-in-each-age-class-from-the-age-data-dark-blue-compared-to-estimates-from-the-demographic-models-fit-to-these-data-light-blue.-c-parameter-estimates-for-annual-fertility-rates-and-adult-survival-probability-with-estimates-highlighted-in-orange-showing-the-parameter-estimates-which-result-in-positive-population-growth.-b-simulated-vaccination-coverage-n-1000-using-the-demographic-parameters-from-c-in-a-hypothetical-commune-with-1000-dogs-for-three-different-campaign-strategies-1-annual-vaccination-campaigns-targeting-dogs-of-all-ages-purple-2-routine-vaccination-of-puppies-at-3-months-of-age-and-3-a-combined-strategy-with-campaigns-annually-and-continuous-puppy-vaccination-in-between-campaigns.}{%
+\paragraph{\texorpdfstring{\textbf{Figure 4. Dog demography and
+implications for campaign strategies}. \emph{A) The age pyramid for
+vaccinated dogs by sex. B) Bootstrapped estimates of the proportion of
+the population in each age class from the age data (dark blue) compared
+to estimates from the demographic models fit to these data (light blue).
+C) Parameter estimates for annual fertility rates and adult survival
+probability, with estimates highlighted in orange showing the parameter
+estimates which result in positive population growth. B) Simulated
+vaccination coverage (N = 1000) using the demographic parameters from
+(C) in a hypothetical commune with 1000 dogs for three different
+campaign strategies: 1) annual vaccination campaigns targeting dogs of
+all ages (purple), 2) routine vaccination of puppies at 3 months of age,
+and 3) a combined strategy with campaigns annually and continuous puppy
+vaccination in between
+campaigns.}}{Figure 4.
+Dog demography and implications for campaign strategies.
+A) The age pyramid for vaccinated dogs by sex.
+B) Bootstrapped estimates of the proportion of the population in each age class from the age data (dark blue) compared to estimates from the demographic models fit to these data (light blue).
+C) Parameter estimates for annual fertility rates and adult survival probability, with estimates highlighted in orange showing the parameter estimates which result in positive population growth.
+B) Simulated vaccination coverage (N = 1000) using the demographic parameters from (C) in a hypothetical commune with 1000 dogs for three different campaign strategies: 1) annual vaccination campaigns targeting dogs of all ages (purple), 2) routine vaccination of puppies at 3 months of age, and 3) a combined strategy with campaigns annually and continuous puppy vaccination in between campaigns.}}\label{figure-4.-dog-demography-and-implications-for-campaign-strategies.-a-the-age-pyramid-for-vaccinated-dogs-by-sex.-b-bootstrapped-estimates-of-the-proportion-of-the-population-in-each-age-class-from-the-age-data-dark-blue-compared-to-estimates-from-the-demographic-models-fit-to-these-data-light-blue.-c-parameter-estimates-for-annual-fertility-rates-and-adult-survival-probability-with-estimates-highlighted-in-orange-showing-the-parameter-estimates-which-result-in-positive-population-growth.-b-simulated-vaccination-coverage-n-1000-using-the-demographic-parameters-from-c-in-a-hypothetical-commune-with-1000-dogs-for-three-different-campaign-strategies-1-annual-vaccination-campaigns-targeting-dogs-of-all-ages-purple-2-routine-vaccination-of-puppies-at-3-months-of-age-and-3-a-combined-strategy-with-campaigns-annually-and-continuous-puppy-vaccination-in-between-campaigns.}}
 
@@
 
-Authors should discuss the results and how they can be interpreted in
-perspective of previous studies and of the working hypotheses. The
-findings and their implications should be discussed in the broadest
-context possible. Future research directions may also be highlighted.
+Through vaccination campaigns implemented in Moramanga District of
+Madagascar, we saw high demand for vaccination from dog owners and found
+that dogs were accessible and able to be handled safely and efficiently
+for parenteral vaccination at a reasonable cost (between 1.3 - 1.8 USD
+per animal vaccinated). We find that providing the vaccine at no direct
+cost to dog owners will be critical to achieving sufficient coverage, as
+even with nominal fees, a significant proportion of owners indicated
+they would no longer vaccinate their animals. A static point vaccination
+strategy achieved higher coverage over a shorter time period in 2018
+compared to dog vaccinations conducted by the District Veterinarian as
+part of routine veterinary service provision in 2019. However, it came
+at a higher cost per animal vaccinated, was more limited in geographic
+scope, and required more resources in terms of personnel. In addition,
+in the rural setting of Andasibe, the static point campaign strategy
+achieved lower coverage, reflecting less accessible, hard-to-reach human
+communities in this location. Based on the lessons learned through these
+campaigns, in particular the observation that puppies were relatively
+easy to handle for vaccinators and owners, we find that continuous
+vaccination targeting puppies in particular may be an effective way to
+maintain vaccination coverage levels given high turnover in dog
+populations.
 
-\hypertarget{conclusion}{%
-\section{Conclusion}\label{conclusion}}
+There were several limitations to our analyses. Owner reports of
+willingness to pay, age of animals, and distance travelled to the
+vaccination point likely all suffer from recall bias and uncertainty,
+whilst willingness to pay studies are often overestimates of observed
+practice \citep{birhane2016}. In addition, the age of animals brought to
+vaccination points may not be representative of the age structure of the
+underlying population, and other work has shown that in general puppies
+(individuals \textless{} 1 year) tend to have lower vaccination coverage
+than adults
+\citep{arief2016, sanchez-soriano2020, mazeri2019, minyoo2015}.
+Additionally, we did not vaccinate animals less than 1 month of age,
+which may explain why we sometimes estimated declining populations (as
+part of the age class of 0-1 years was not captured by the data).
+However, our analyses based on these data are consistent with previous
+findings from SSA that demonstrate male-biased populations, skewed
+towards puppies, and with high mortality in the first year of life
+\citep{czupryna2016, taylor2017}. Studies have also shown that charging
+fees for vaccines can greatly reduce the efficacy of campaigns and are a
+significant barrier to vaccination for owners
+\citep{schildecker2016, yoak2021, barbosacosta2020}. Similarly, distance
+to campaign points has been identified as a barrier to vaccination, and
+in most cases owners report traveling less than 1km to reach a
+vaccination point \citep{mazeri2018barriers, mazeri2019}. We used HDRs,
+which can be sensitive to underlying estimates of the human population
+and the spatial scale of estimation
+\citep{sambo2018estimating, sambo2017}. However, we used HDR estimates
+from a recent household survey study in the district, and found these to
+give coverage estimates consistent with those from post-vaccination
+transects. Finally, in our vaccination model, we make several
+simplifying assumptions: we assume that immunity wanes after three years
+(simulating vaccination with a long-lasting vaccine such as Nobivac),
+but we also assume that vaccinated dogs that survive to the following
+year are revaccinated in subsequent campaigns (effectively assuming
+boosting per manufacturer recommendations for Rabisin). Overall waning
+vaccinal immunity plays a lesser role in declining vaccination coverage
+given the high population turnover in this context. In our
+age-structured models, we also do not account for population carrying
+capacity, which likely overestimates growth of the dog population.
 
-This section is not mandatory, but can be added to the manuscript if the
-discussion is unusually long or complex.
+Our estimates of costs per animal vaccinated are in line with recent
+estimates from other countries \citep{undurraga2020}, although we likely
+underestimate costs given the donation of time and resources by the
+organizations and individuals involved (i.e.~costs associated with
+international volunteers including airfare and visa costs as well as
+costs of transportation donated to the campaign). We also did not
+include costs of pre-exposure prophylaxis as all of our volunteers and
+vaccinators had been vaccinated prior to the campaigns. Both pre- and
+post-exposure vaccines should however always be readily available for
+vaccinators and should be included in vaccination program budgets. While
+the volunteer-led effort resulted in significant financial and personnel
+resources being devoted to the campaign, costs were higher overall and
+per animal vaccinated. In addition, NGO and volunteer based campaigns
+may be difficult to sustain given unpredictable funding, time
+commitments, and turnover in staff \citep{ferguson2020}. Dog vaccination
+delivered through the District Veterinarian was less costly, with the
+majority of the costs directly related to vaccination. Although we
+included costs of implementing transect based coverage estimates, these
+were negligible (less than 0.10 USD per dog vaccinated), in line with
+recent studies which have shown this strategy is a cheap and effective
+way to estimate coverage \citet{sambo2017}; \citet{gibson2015}{]}.
 
-\hypertarget{patents}{%
-\section{Patents}\label{patents}}
+One key aspect we do not consider is potential feedback loops between
+vaccination and demography. Estimates of the effects of vaccination on
+dog demography are mixed \citep{czupryna2016}, but vaccination may
+increase dog survival \citep{conan2015}. If dog population growth is
+driven by survival, then this could mean that increased vaccination
+results in increased population growth. However, if growth is driven
+more by demand from human communities, then vaccination could stabilize
+the population and reduce population turnover. Improving dog population
+management, encouraging responsible pet ownership practices, and
+increasing veterinary services could all complement vaccination efforts,
+but have not been demonstrated to result in meaningful rabies control
+without parallel dog vaccination \citep{taylor2017}.
 
-This section is not mandatory, but may be added if there are patents
-resulting from the work reported in this manuscript.
+In settings with high dog ownership, moving towards community-based
+vaccination strategies could be an effective way to achieve sufficient
+and consistent coverage, particularly in hard-to-reach communities.
+During our campaign, we found that puppies at approximately 3 months old
+were easy to handle for both vaccinators and owners. Puppy vaccination
+could be carried out by local officials embedded in communities (along
+the lines of community health workers), especially given recent findings
+on thermotolerance of rabies vaccines and locally manufactured methods
+for maintaining temperatures required for sustained vaccine storage (up
+to 3 months) \citep{lugelo2020}. Incentivizing vaccinators appropriately
+will be a key challenge, as currently providing no-cost rabies
+vaccination is not seen as part of routine duties for district
+veterinarians or for livestock officers. Implementing dog vaccination
+alongside government mandated livestock vaccination campaigns may also
+be a strategy to scale up vaccination efforts at relatively low-cost.
+Expanding veterinary services across the country and relieving financial
+pressures on veterinarians and animal health workers through appropriate
+compensation could greatly improve veterinary services across Madagascar
+\citep{mondiale2017evaluation}. For volunteer based efforts, focusing on
+local volunteers (i.e.~veterinary students) may be a more cost-effective
+strategy. However similar to international volunteers, volunteers
+require subsistence during vaccination campaigns when not based in the
+communities where they study or live.
+
+Overall, our results suggest that dog vaccination is a feasible and
+effective strategy for controlling canine rabies in Madagascar. However,
+rabies vaccination must be recognized as a public good. Strategies to
+remove barriers for dog owners and to incentivize veterinarians and
+other animal health workers to implement vaccination will be key to
+long-term campaign success. Borrowing strategies from human vaccination
+efforts, i.e.~deploying community health workers, could be a way to
+deliver vaccinations and reduce costs in the hardest-to-reach
+communities. In addition, refining vaccination strategies to local
+contexts and using improved tools and systems, such as mobile phone
+based data collection, could improve efficacy and lower costs of
+campaigns \citep{mazeri2021, gibson2016a}. To monitor the success of
+these campaigns, it will be critical to develop efficient and effective
+methods to estimate vaccination coverage and to measure their impact on
+reducing rabies incidence through robust surveillance
+\citep{lushasi2020, ferguson2020}. With limited chances for
+reintroduction from outside the island, Madagascar could be a model for
+rabies elimination, and strategies that tackle the context specific
+challenges could be a path forward for the country to reach ZeroBy30.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\acknowledgments{All sources of funding of the study should be
-disclosed. Please clearly indicate grants that you have received in
-support of your research work. Clearly state if you received funds for
-covering the costs to publish in open access.}
+\acknowledgments{We thank all the veterinarians, livestock officers,
+local officials and leaders, NGO staff, and volunteers involved in each
+of the campaigns. We are grateful to staff and officials at the
+Department of Veterinary Services, the Ministry of Public Health, and
+the Institut Pasteur de Madagascar for their support and technical
+assistance. In particular, we thank Carmen Barba Claassens, Jean
+Hyacinthe Randrianarisoa, Ranaivoarimanana, Fierenantsoa
+Randriamahatana, Esther Noiarisaona, Cara Brook, and Christian
+Ranaivoson, Rila Ratovoson, and Claire LeBlanc. The campaign costs were
+supported by donations from Lush Cosmetics Charity Pot. Parts of this
+work were funded by grants from the Center for Health and Wellbeing and
+the Department of Ecology and Evolutionary Biology at Princeton
+University to MR. MR was supported by an NSF Graduate Research
+Fellowship and a Princeton Institute for Regional and International
+Studies Graduate Fellowship. KH is supported by the Wellcome Trust
+(207569/Z/17/Z). We want to particularly acknowledge the contributions
+to this work by Annie Yang, an undergraduate researcher and volunteer
+with Princeton University and the Mad Dog Initiative who tragically
+passed away in a car accident before this research was published.}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\authorcontributions{For research articles with several authors, a short
-paragraph specifying their individual contributions must be provided.
-The following statements should be used ``X.X. and Y.Y. conceive and
-designed the experiments; X.X. performed the experiments; X.X. and Y.Y.
-analyzed the data; W.W. contributed reagents/materials/analysis tools;
-Y.Y. wrote the paper.'\,' Authorship must be limited to those who have
-contributed substantially to the work reported.}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\conflictsofinterest{Declare conflicts of interest or state `The authors
-declare no conflict of interest.' Authors must identify and declare any
-personal circumstances or interest that may be perceived as
-inappropriately influencing the representation or interpretation of
-reported research results. Any role of the funding sponsors in the
-design of the study; in the collection, analyses or interpretation of
-data in the writing of the manuscript, or in the decision to publish the
-results must be declared in this section. If there is no role, please
-state `The founding sponsors had no role in the design of the study; in
-the collection, analyses, or interpretation of data; in the writing of
-the manuscript, an in the decision to publish the results'.}
+\conflictsofinterest{The authors declare no conflict of interest. The
+funders had no role in the design of the study; in the collection,
+analyses, or interpretation of data; in the writing of the manuscript,
+or in the decision to publish the results.}
 
@@
 %% optional
-\abbreviations{The following abbreviations are used in this manuscript:\\
-
-\noindent
-\begin{tabular}{@{}ll}
-MDPI & Multidisciplinary Digital Publishing Institute \\
-DOAJ & Directory of open access journals \\
-TLA & Three letter acronym \\
-LD & linear dichroism \\
-\end{tabular}}
-
-\input{"appendix.tex"}
 
@@
 %% optional
-\sampleavailability{Samples of the compounds \ldots\ldots{} are
-available from the authors.}
 

### 2021-02-20 R/run_analyses_all.R

@@ -11,7 +11,10 @@ run_script("R/run_dem_mod.R") # will take a few minutes!
 run_script("R/run_vacc_mod.R")
 
 # Then make the figures & tables
-run_script("R/figure1.R")
-run_script("R/Table1.R")
 run_script("R/figure2.R")
+run_script("R/Table1.R")
 run_script("R/figure3.R")
+run_script("R/figure4.R")
+
+# Make the mdpi manuscript
+rmarkdown::render("paper/mdpi_paper.Rmd", output_format = "rticles::mdpi_article")

### 2021-02-20 paper/mdpi_paper.Rmd
updated output files

@@
   - name: Caitlynn Filla
-    affil: 1, *
+    affil: 1, \dagger
   - name: Malavika Rajeev
-    affil: 2, *
+    affil: 2, \dagger
     orcid: 0000-0001-6030-4995
@@
    research was published.
-authorcontributions: |
-  Conceptualization, MR, KV, JL, TS, ZF, GE; methodology, MR, KV, JL, TS, ZF, GE, KH, CJEM, AY; data collection: MR, ZR, CH, RRR, GE, MA, NPR, JN, AF, TR, FD; formal analysis, MR, CF, AY; writing—original draft preparation, CF, MR; writing—review and editing, CF, MR, CJEM, TC, ZF, KV, KH, GE, CH; visualization, MR; supervision, KV, KH, CJEM, ZF; project administration, MR, ZR, CH, RRR, GE, MA, NPR, JN, AF, TR, FD, KV, ZF; funding acquisition, MR, KV, JL, TS, ZF, TC.
 conflictsofinterest: |
@@
 library(here)
-source("R/dem_models.R")
+source(here("R/dem_models.R"))
 

### 2021-04-03 paper/mdpi_paper.Rmd
updating with final proofed files

@@
   - name: Malavika Rajeev
-    affil: 2, \dagger
+    affil: 2, \dagger, \*
     orcid: 0000-0001-6030-4995
@@
     affil: 3
-  - name: Glenn Edosoa
+  - name: Glenn Torrencelli Edosoa
     affil: 5
@@
       University of Florida, Department of Anthropology
-      Gainesville, Florida, USA
+      Gainesville, Florida, 32611 USA
   - num: 2
@@
       Department of Ecology and Evolutionary Biology, Princeton University
-      Princeton, NJ, United States
+      Princeton, NJ, 08544 United States
   - num: 3
@@
    research was published.
+authorcontributions: |
+  Conceptualization, MR, GTE, MA, JN, ZF, TS, JL and KV; Data curation, CF, MR, ZR, CH, RR, AF, AY, FD, JL and TR; Formal      analysis, CF, MR, AY and CJEM; Funding acquisition, MR, ZF, TS, JL, CJEM and KV; Investigation, MR, ZR, CH, RR, AF, TAC and   TR; Methodology, MR, ZR, CH, GTE, MA, AF, AY, ZF, JL, TR, KH, CJEM and KV; Project administration, ZR, CH, RR, GTE, MA,      NPR, JN, AF, FD, TAC, ZF, TS, JL, TR and KV; Supervision, GTE, MA, NPR, JN, ZF, KH, CJEM and KV; Visualization, MR;          Writing---original draft, CF and KV; Writing---review & editing, CF, MR, ZR, CH, GTE, TAC, ZF, KH, CJEM and KV.
 conflictsofinterest: |
@@
 The World Health Organization (WHO) and its partners have set a goal to eliminate human deaths due to canine rabies by the year 2030 ('ZeroBy30')[@abela-ridder2016].
-Annual dog vaccination campaigns achieving at least 70% coverage are the recommended target for controlling rabies in domestic dog populations [@worldhealthorganization2013].
+Annual dog vaccination campaigns achieving at least 70% coverage are the recommended target for controlling rabies in domestic dog populations [@world2018expert].
 However, achieving this coverage target in low and middle income countries where the burden of human rabies is highest can be challenging due to economic, ecological, sociocultural, and political barriers [@fahrion2017].
@@
 In Madagascar, canine rabies has been endemic for over a century, and for most of that period, the Institut Pasteur de Madagascar has provided post-exposure prophylaxis free-of-charge to bite patients in the country [@reynes2011].
-Currently, there are only 31 clinics where these human vaccines are available, and there is minimal dog vaccination due to high costs to owners and lack of vaccine availability [@rajeev2018].
-Recent studies have estimated a high burden of human deaths, masked by weak surveillance across the country [@rajeev2020; @rajeev2018].
+Currently, there are only 31 clinics where these human vaccines are available, with one clinic serving on average greater than 700,000 persons and over 20,000 bite patients treated annually [@rajeev2020].
+There is minimal dog vaccination due to high costs to owners and lack of vaccine availability [@rajeev2018].
+Recent studies have estimated a high burden of human rabies deaths (approximately 1000 deaths annually), masked by weak surveillance across the country [@rajeev2020; @rajeev2018].
 
-The veterinary sector is largely private, with 204 veterinarians employed in hybrid private/public employment as designated district veterinarians.
+The veterinary sector is largely private and practices are largely limited to urban areas, but approximately 204 veterinarians are employed in hybrid private/public employment as designated district veterinarians by the national government.
 While dog vaccination is rare, livestock officers and veterinarians work together to implement cattle vaccination campaigns for anthrax on an annual basis as mandated by the government (owners are charged a fee per animal for these vaccines which vary by location) [@mondiale2017evaluation].
@@
 In 2018 and 2019, we deployed two different vaccination strategies.
-In 2018, we carried out a larger scale volunteer-led pilot vaccination campaign in two communes (sub-district level) using a static point strategy.
-In 2019/2020, we provided vaccines, all necessary supplies for vaccine administration, and a per-vaccine fee to the District Veterinarian to vaccinate animals as part of routine services delivered continuously.
+In 2018, we carried out a larger scale volunteer-led pilot vaccination campaign in two communes (sub-district level) using a static point strategy where owners brought their animals to a fixed location for vaccination.
+In 2019/2020, we provided vaccines, all necessary supplies for vaccine administration, and a per-vaccine fee to the District Veterinarian to vaccinate animals as part of continuous vaccination strategy, where vaccines were delivered alongside routine services provided by the veterinarian.
 We compare time, human resources, costs, and coverage estimates between these campaigns, and using a demographic and vaccination model we further explore different vaccination strategies based on what we learned during implementation.
@@
 It comprises 21 communes, covering approximately 7150 km^2^ with an approximate human population of `r format(round(sum(mora_sf$pop), -3), big.mark = ",")`[@instat2021].
-Previous work in the district has established a high burden of rabies exposures and deaths despite the availability of post-exposure prophylaxis at the district hospital [@rajeev2018].
+Previous work in the district has established a high burden of rabies exposures (42 - 110 per 100,000 persons) and deaths (1 - 3 deaths per 100,000 persons) despite the availability of post-exposure prophylaxis at the district hospital [@rajeev2018].
 While Moramanga is relatively close to the capital city of Antananarivo (~ 3 hours by bus), within the district, travel times between locations are highly variable, with much of the population living in more rural areas with limited access to roads and transportation [@rajeev2020].
@@
 
-In 2018, two NGOs (the Mad Dog Initiative (MDI) and Traveling Animal Doctors (TAD)) organized a pilot vaccination campaign in collaboration with the Department of Veterinary Services, and the Ministry of Public Health in the District of Moramanga.
+In 2018, two NGOs (the Mad Dog Initiative (MDI) and Traveling Animal Doctors (TAD)) organized a pilot vaccination campaign in collaboration with the Department of Veterinary Services, and the Ministry of Public Health in the District of Moramanga (fig 2).
 This campaign focused on two communes in Moramanga, Moramanga Ville (the district center) and Andasibe (a rural commune surrounding Andasibe National Park), where previously high incidence of probable rabies exposures (Moramanga Ville) and a high burden of deaths (Andasibe) had been recorded [@rajeev2018].
@@
 The campaign was planned as a series of static point vaccination stations covering 1 - 3 fokontany (i.e., sub-communes) per day.
-A week before the campaign dates, the vaccination team informed the chief of the fokontany about the campaign and provided posters advertising the date of the vaccine and that it would be available at no cost to owners.
-During the campaign, we used Rabisin (10 mL vials with 1 mL per dose, Boehringer Ingelheim) to vaccinate both dogs and cats presented that were at least 1 month old.
+A week before the campaign dates, the vaccination team informed the chief of the fokontany about the campaign and provided fliers advertising the date of the vaccine and that it would be available at no cost to owners (fig 2).
+During the campaign, we used Rabisin (10 mL vials with 1 mL per dose, Boehringer Ingelheim) to vaccinate both dogs and cats presented that were at least 1 month old based on current WHO recommendations for endemic settings [@morters2015; @world2018expert].
+Rabisin has a manufacturor stated duration of protection of 1 year given one dose, and of an additional three years if an additional dose is given approximately one year after the first dose.
 As part of the campaign, owners were surveyed by vaccinators about how many dogs and cats they owned in total (split by > 1 year vs.
 < 1 year in order to avoid language ambiguities that might result in excluding puppies and kittens), as well as if their dogs were free roaming (no restrictions on movement by the owner/'mirenyreny', tied/'mifatora', or fenced/'mifefy').
-Vaccinations were delivered at no cost to owners, but as animal vaccination is generally thought of as a paid service in Madagascar, owners were asked how much in Ariary they would be willing to pay to have one animal vaccinated against rabies.
-For each animal vaccinated, we recorded the species (cat or dog), sex, approximate age in years, and whether the animal had been previously vaccinated.
+Vaccinations were delivered at no cost to owners, but as animal vaccination is generally thought of as a paid service in Madagascar, owners were asked how much in Ariary they would be willing to pay to have one animal vaccinated against rabies (after being informed that the current vaccination was being free).
+For each animal vaccinated, we recorded the species (cat or dog), sex, approximate age in years as reported by the owner, and whether the animal had been previously vaccinated.
 
@@
 All animals vaccinated were concurrently marked with a colored, non-toxic, livestock crayon along the top of the head or back.
-Between 4 - 6 PM when dogs are most active [@sambo2017] and on the same day as the campaign, in each vaccination location two transect surveys were done.
-Each team (two people per team) walked a transect for 1 hour starting in opposite directions and accompanied by a local guide to ensure that walking paths did not overlap.
-Teams recorded any marked and unmarked dogs observed, their roaming status (whether roaming, inside a fence, or tied), and their approximate age (greater or less than 1 year of age).
+At the end of each campaign day between 1600-1800 hours, when dogs are most active [@sambo2017], two transect surveys were conducted on vaccination campaign days in each vaccination location by two teams (consisting of two volunteers and one local guide) for one hour on separate paths and in opposite directions.
+Marked and unmarked dogs were recorded as well as their roaming status (i.e., roaming, inside a fence, or tied), and their approximate age (>1 year or < 1 year of age).
 
-```{r fig1} 
+```{r fig1, fig.cap="Photos from the 2018 campaign.
+Topleft: advertisement for the campaign posted on the door of the fokontany office as the campaign starts; Topright: a dog post-vaccination marked with a crayon; Bottomleft: a basket of puppies brought for vaccination; Bottomright: a line of owners and dogs waiting for vaccination.
+Photo credit: Jochem Lastdrager, Traveling Animal Doctors."}
+
 knitr::include_graphics(here("figs/fig1.jpeg"))
 ```
-
-#### **Figure 1.
-Photos from the 2018 campaign**.
-*Topleft: advertisement for the campaign posted on the door of the fokontany office as the campaign starts; Topright: a dog post-vaccination marked with a crayon; Bottomleft: a basket of puppies brought for vaccination; Bottomright: a line of owners and dogs waiting for vaccination.*
 
@@
 For the 2019 campaign, we estimated vaccination coverage using human-to-dog ratios (HDRs) and human population estimates [@cleaveland2003; @athingo2020].
-We used a range of 8 - 25, based on previous data from Madagascar [@ratsitorahina2009] and based on recent estimates from household surveys in the Moramanga District [@leblancclaireRabiesMadagascarThreepronged2019].
+We used a ratio range of 8 - 25 humans-to-dogs, based on previous data from Madagascar [@ratsitorahina2009] and based on recent estimates from household surveys in the Moramanga District [@leblancclaireRabiesMadagascarThreepronged2019].
 We set the point estimate using an HDR of `r round(mora_mid_hdr, 1)`, the midpoint between the HDRs estimated for two communities in the District by LeBlanc et al.
@@
 
-Using the age data on vaccinated animals collected during both vaccination campaigns, we estimated the proportion of the population in four age classes: puppies under the age of 1 year, juveniles aged 1 - 2 yrs, adults aged 2 - 6 years, and older dogs aged 6+ years based on .
+Using the age data on vaccinated animals collected during both vaccination campaigns, we estimated the proportion of the population in four age classes: puppies under the age of 1 year, juveniles aged 1 - 2 years, adults aged 2 - 6 years, and older dogs aged 6+ years based on broad patterns of survival in comparable dog populations (i.e.
+low survival in the first year of life, followed by plateauing survival probabilites, [@czupryna2016; @davlin2012]).
 With the assumption that these estimates represent the population at a stable age distribution, we use a Leslie matrix model to estimate annual adult survival probability and fertility using maximum likelihood estimation [@Fujiwara2017].
@@
 
-#### Data and ethics statement
+## Data and ethics statement
 All data were analysed in `r R.version$version.string` [@R2020], largely using the `tidyverse` package suite [@wickham2019].
@@
 MDI also maintains national research permits (MICET permit: #130-19/MEDD/SG/DGEF/DGRNE) for its research and volunteer programs.
+Prior to vaccination, verbal informed consent was obtained from animal owners, and owners could opt out of answering any questions or services provided.
+No personally identifiable information was collected at any point during the campaigns.
 
@@
 We vaccinated at `r table(pts$Commune)["Andasibe"]` points in Andasibe and `r table(pts$Commune)["Moramanga Ville"]` points in Moramanga Ville.
-During the 2019 campaign, between `r paste(format.Date(range(dmy(animals_2019$Date)), "%B %Y"), collapse = " - ")`, the District Veterinarian vaccinated a total of `r nrow(animals_2019)`animals (`r species_2019["dog"]` dogs and `r species_2019["cat"]` cats) over 48 days in seven communes in the Moramanga District.
+During the 2019 campaign, between `r paste(format.Date(range(dmy(animals_2019$Date)), "%B %Y"), collapse = " - ")`, the District Veterinarian vaccinated a total of `r nrow(animals_2019)` animals (`r species_2019["dog"]` dogs and `r species_2019["cat"]` cats) over 48 days in seven communes in the Moramanga District.
 While more animals were vaccinated per vaccination point and per vaccination day in 2018 compared to 2019, the number of animals vaccinated per person-day was much higher for the 2019 campaign.
@@
 
-#### **Table 1.
-Summary of 2018 and 2019 campaigns**.
-*Breakdown of of animals vaccinated, prior vaccination history, dog demography, dog ownership, daily and per vaccination rates by year and location (for 2018).
-Photo credict: Jochem Lastdrager, Traveling Animal Doctors.*
-```{r}
-if(params$outformat %in% "docx") {
-  knitr::include_graphics(here("figs/table1.png"))
-} else {
-  tab1 <- readRDS(here("out/table1.rds"))
-
-  tab1 %>% 
-    gt::as_latex() 
-}
-```
+\begin{specialtable}[H] 
+\caption{Summary of 2018 and 2019 campaigns.
+Breakdown of of animals vaccinated, prior vaccination history, dog demography, dog ownership, daily and per vaccination rates by year and location (for 2018).\label{tab1}}
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.12}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
+\toprule
+& & 2018 & & 2019 \\ \addlinespace
+\midrule
+\endhead
+& Andasibe & Moramanga Ville & All Communes & All
+Communes \\ \addlinespace
+Total animals vaccinated & 528 & 2609 & 3137 & 2385 \\ \addlinespace
+Total dogs vaccinated & 254 & 1,803 & 2,057 & 1,898 \\ \addlinespace
+Dogs with history of vaccination & 5\% & 16\% & 15\% &
+13\% \\ \addlinespace
+Dogs vaccinated within last year & 5\% & 7\% & 7\% &
+13\% \\ \addlinespace
+Percent male dogs & 55\% & 56\% & 56\% & 65\% \\ \addlinespace
+Average dogs per owner & 0.8 & 1.1 & 1.0 & -- \\ \addlinespace
+Percent of owners with free-roaming dogs & 67\% & 19\% & 28\% &
+-- \\ \addlinespace
+Animals vaccinated per day (total days) & 88 (6) & 372.7 (7) & 241.3
+(13) & 49.7 (48) \\ \addlinespace
+Animals vaccinated per vaccination point (total points) & 75.4 (7) &
+186.4 (14) & 149.4 (21) & 37.3 (64) \\ \addlinespace
+Animals vaccinated per person day (total person days) & 11.7 (45) & 21.6
+(121) & 18.9 (166) & 49.7 (48) \\ \addlinespace
+\bottomrule
+\end{longtable}
+\end{specialtable}
 
@@
 We found that charging owners for vaccination would result in minimal cost recovery (fig 2C) and beyond a minimal cost would actually result in increased costs per vaccinated individual than free-of-charge campaigns.
-Cost recovery would be more likely given a 2019 style strategy where the majority of the costs are incurred on a per animal basis, compared to the 2018 campaigns where the costs are largely due to setting up the static point vaccination stations (fig 4A).
+Cost recovery would be more likely given a 2019 style strategy where the majority of the costs are incurred on a per animal basis, compared to the 2018 campaigns where the costs were largely due to setting up the static point vaccination stations (fig 2A & B).
 More importantly, in all cases, even a nominal fee would significantly reduce numbers of dogs vaccinated and thus vaccination coverage, particularly in the rural commune of Andasibe (fig 2D).
 
-```{r fig2} 
-knitr::include_graphics(here("figs/fig2.jpeg"))
-```
-
-#### **Figure 2.
-Comparing campaign costs and willingness to pay**.
-*A) Vaccine costs broken down by category and by year (colors).
+```{r fig2, fig.cap="Comparing campaign costs and willingness to pay.
+A) Vaccine costs broken down by category and by year (colors).
 B) Overall cost per animal vaccinated for the two campaign years, split by direct costs of vaccination per animal (i.e.
@@
 D) The percent reduction in number of animals vaccinated given owners willingness to pay.
-The curves in C and D are shown based on the overall responses on willingness to pay (solid line) from both Moramanga and Andasibe, and the responses split by Commune (dashed and dotted lines).*
+The curves in C and D are shown based on the overall responses on willingness to pay (solid line) from both Moramanga and Andasibe, and the responses split by Commune (dashed and dotted lines)."}
+knitr::include_graphics(here("figs/fig2.jpeg"))
+```
 
@@
 
-```{r fig3}
+```{r fig3, fig.cap='Estimates of coverage achieved by the 2018 and 2019 campaigns.
+A) The commune level numbers of dogs vaccinated (size of points) and the associated coverage estimates (color of points) for the year 2018 (squares, estimated using post-vaccination transects) and 2019 (circles, estimated using a human-to-dog ratio (HDR) of 19.5, based on a recent household survey in the Moramanga District).
+The inset shows the location of the Moramanga District in Madagascar.
+B) A comparison of coverage estimates by location and by method of estimation (shape of points correspond to post-vaccination transects vs.
+HDR based estimates); for the transect based estimates, the line range shows the 95\\% exact binomial confidence interval while for the HDR based estimates, the line range shows the range of coverage estimates assuming an HDR range of 8 - 25 according to estimates from the literature.'}
 knitr::include_graphics(here("figs/fig3.jpeg"))
 ```
-
-#### **Figure 3.
-Estimates of coverage achieved by the 2018 and 2019 campaigns**.
-*A) The commune level numbers of dogs vaccinated (size of points) and the associated coverage estimates (color of points) for the year 2018 (estimated using post-vaccination transects) and 2019 (estimated using a human-to-dog ratio (HDR) of `r round(mora_mid_hdr, 1)`, based on a recent household survey in the Moramanga District).
-The inset shows the location of the Moramanga District in Madagascar.
-B) A comparison of coverage estimates by location and by method of estimation (post-vaccination transects vs.
-HDR based estimates); for the transect based estimates, the line range shows the 95% exact binomial confidence interval while for the HDR based estimates, the line range shows the range of coverage estimates assuming an HDR range of 8 - 25 according to estimates from the literature.*
 
@@
 
-Demographic data from vaccinated dogs shows a population pyramid with a large base, indicative of a fast-growing population, and with a male bias (approximately `r prop_sex["M"]`% male, Fig 4A).
+Demographic data from vaccinated dogs showed a population pyramid with a large base, indicative of a fast-growing population, and with a male bias (approximately `r prop_sex["M"]`% male, Fig 4A).
 We fit these data to an age-structured model, and were able to generate parameter estimates which resulted in stable age distributions consistent with the data (Fig 4B).
-We filter to parameter estimates that are consistent with a growing population resulting in mean adult annual survival probability of `r dem_pars[dem_pars$name == "psurv_adult_est", "value.mean"]` (95% quantile: `r dem_pars[dem_pars$name == "psurv_adult_est", "ci"]`).
-We use estimates of fertility (on average `r dem_pars[dem_pars$name == "fert_est", "value.mean"]`, 95% quantile: `r dem_pars[dem_pars$name == "fert_est", "ci"]`) to back-calculate pup survival, which ranges from `r round(dem_pars$value.mean[dem_pars$name == "pup_surv_lower"], 2)` - `r round(dem_pars$value.mean[dem_pars$name == "pup_surv_upper"], 2)`.
-We find that given these demographic parameters, annual campaigns that target dogs of all ages result in rapid decline in vaccination coverage between campaigns, largely due to rapid turnover of the dog population (compared to the impact of waning immunity assuming a discrete 3 year period of vaccine immunity, evident in the additional dip at year 3, fig 4D).
-Continuously targeting 70% of the puppy population for vaccination, while unable to achieve the peak coverage, consistently reaches coverage of about 50% of the dog population.
+We filtered to parameter estimates that are consistent with a growing population resulting in mean adult annual survival probability of `r dem_pars[dem_pars$name == "psurv_adult_est", "value.mean"]` (95% quantile: `r dem_pars[dem_pars$name == "psurv_adult_est", "ci"]`).
+We used estimates of fertility (on average `r dem_pars[dem_pars$name == "fert_est", "value.mean"]`, 95% quantile: `r dem_pars[dem_pars$name == "fert_est", "ci"]`) to back-calculate pup survival, which ranged from `r round(dem_pars$value.mean[dem_pars$name == "pup_surv_lower"], 2)` - `r round(dem_pars$value.mean[dem_pars$name == "pup_surv_upper"], 2)`.
+We found that given these demographic parameters, annual campaigns that target dogs of all ages result in rapid decline in vaccination coverage between campaigns, largely due to rapid turnover of the dog population (compared to the impact of waning immunity assuming a discrete 3 year period of vaccine immunity, evident in the additional dip at year 3, fig 4D).
+Continuously targeting 70% of the puppy population for vaccination, while unable to achieve the peak coverage, consistently reached coverage of about 50% of the dog population.
 A combined strategy maintains the highest and most temporally stable levels of coverage close to the target of 70%.
 
-```{r fig4}
+```{r fig4, fig.cap="Dog demography and implications for campaign strategies.
+A) The age pyramid for vaccinated dogs by sex.
+B) Bootstrapped estimates of the proportion of the population in each age class from the age data (dark blue) compared to estimates from the demographic models fit to these data (light blue).
+C) Parameter estimates for annual fertility rates and adult survival probability, with estimates highlighted in orange showing the parameter estimates which result in positive population growth.
+D) Simulated vaccination coverage (N = 1000) using the demographic parameters from (C) in a hypothetical commune with 1000 dogs for three different campaign strategies: 1) annual vaccination campaigns targeting dogs of all ages (purple), 2) routine vaccination of puppies at 3 months of age, and 3) a combined strategy with campaigns annually and continuous puppy vaccination in between campaigns."}
 knitr::include_graphics(here("figs/fig4.jpeg"))
 ```
-
-#### **Figure 4.
-Dog demography and implications for campaign strategies**.
-*A) The age pyramid for vaccinated dogs by sex.
-B) Bootstrapped estimates of the proportion of the population in each age class from the age data (dark blue) compared to estimates from the demographic models fit to these data (light blue).
-C) Parameter estimates for annual fertility rates and adult survival probability, with estimates highlighted in orange showing the parameter estimates which result in positive population growth.
-B) Simulated vaccination coverage (N = 1000) using the demographic parameters from (C) in a hypothetical commune with 1000 dogs for three different campaign strategies: 1) annual vaccination campaigns targeting dogs of all ages (purple), 2) routine vaccination of puppies at 3 months of age, and 3) a combined strategy with campaigns annually and continuous puppy vaccination in between campaigns.*
 
@@
 In addition, in the rural setting of Andasibe, the static point campaign strategy achieved lower coverage, reflecting less accessible, hard-to-reach human communities in this location.
-Based on the lessons learned through these campaigns, in particular the observation that puppies were relatively easy to handle for vaccinators and owners, we find that continuous vaccination targeting puppies in particular may be an effective way to maintain vaccination coverage levels given high turnover in dog populations.
+Based on the lessons learned through these campaigns, in particular the observation that puppies were relatively easy to handle for vaccinators and owners (fig 2), we find that continuous vaccination targeting puppies in particular may be an effective way to maintain vaccination coverage levels given high turnover in dog populations.
 
 There were several limitations to our analyses.
-Owner reports of willingness to pay, age of animals, and distance travelled to the vaccination point likely all suffer from recall bias and uncertainty, whilst willingness to pay studies are often overestimates of observed practice [@birhane2016].
-In addition, the age of animals brought to vaccination points may not be representative of the age structure of the underlying population, and other work has shown that in general puppies (individuals < 1 year) tend to have lower vaccination coverage than adults [@arief2016; @sanchez-soriano2020; @mazeri2019; @minyoo2015].
+Owner reports of willingness to pay, age of animals, and distance travelled to the vaccination point likely all suffer from recall bias and uncertainty.
+Owner based estimates of age are very coarse, but given the broad age classes we use, likely are of sufficient precision to capture broad patterns in age structure.
+However,, the age of animals brought to vaccination points may not be representative of the age structure of the underlying population.
+Previous work has shown that in general puppies (individuals < 1 year) tend to have lower vaccination coverage than adults [@arief2016; @sanchez-soriano2020; @mazeri2019; @minyoo2015].
 Additionally, we did not vaccinate animals less than 1 month of age, which may explain why we sometimes estimated declining populations (as part of the age class of 0-1 years was not captured by the data).
-However, our analyses based on these data are consistent with previous findings from SSA that demonstrate male-biased populations, skewed towards puppies, and with high mortality in the first year of life [@czupryna2016; @taylor2017].
-Studies have also shown that charging fees for vaccines can greatly reduce the efficacy of campaigns and are a significant barrier to vaccination for owners [@schildecker2016; @yoak2021; @barbosacosta2020].
+Despite these issues, our analyses based on these data are consistent with previous findings from sub-Saharan Africa that demonstrate male-biased populations, skewed towards puppies, and with high mortality in the first year of life [@czupryna2016; @taylor2017].
+
+Willingness to pay studies have been done previously for rabies vaccination, and have consistently shown that cost recovery is minimal given the price dog owners are willing to pay for vaccination [@schildecker2016; @durr2008; @sambo2014; @barbosacosta2020; @yoak2021].
+In fact, owners generally overstate the amounts they are willing to pay when compared to observed practice [@birhane2016], and thus our analysis could underestimate the impact of charging owners on coverage reductions.
 Similarly, distance to campaign points has been identified as a barrier to vaccination, and in most cases owners report traveling less than 1km to reach a vaccination point [@mazeri2018barriers; @mazeri2019].
+In the context of Madagascar, these findings are of particular relevance, as animal vaccinations can be mandated by the government, but at cost to animal owners (for example, for the anthrax vaccine in cattle).
+Importantly, dog owners in the district do believe that vaccination can prevent rabies transmission [@leblancclaireRabiesMadagascarThreepronged2019], and given the observed demand during the campaigns, are amenable to vaccination of their animals.
+Our results confirm that implementing dog rabies vaccinations as a public health measure and removing as many barriers as possible to dog vaccination will be important to the success of control programs.
+
 We used HDRs, which can be sensitive to underlying estimates of the human population and the spatial scale of estimation [@sambo2018estimating; @sambo2017].
 However, we used HDR estimates from a recent household survey study in the district, and found these to give coverage estimates consistent with those from post-vaccination transects.
-Finally, in our vaccination model, we make several simplifying assumptions: we assume that immunity wanes after three years (simulating vaccination with a long-lasting vaccine such as Nobivac), but we also assume that vaccinated dogs that survive to the following year are revaccinated in subsequent campaigns (effectively assuming boosting per manufacturer recommendations for Rabisin).
-Overall waning vaccinal immunity plays a lesser role in declining vaccination coverage given the high population turnover in this context.
-In our age-structured models, we also do not account for population carrying capacity, which likely overestimates growth of the dog population.
-
-Our estimates of costs per animal vaccinated are in line with recent estimates from other countries [@undurraga2020], although we likely underestimate costs given the donation of time and resources by the organizations and individuals involved (i.e.
-costs associated with international volunteers including airfare and visa costs as well as costs of transportation donated to the campaign).
-We also did not include costs of pre-exposure prophylaxis as all of our volunteers and vaccinators had been vaccinated prior to the campaigns.
-Both pre- and post-exposure vaccines should however always be readily available for vaccinators and should be included in vaccination program budgets.
-While the volunteer-led effort resulted in significant financial and personnel resources being devoted to the campaign, costs were higher overall and per animal vaccinated.
-In addition, NGO and volunteer based campaigns may be difficult to sustain given unpredictable funding, time commitments, and turnover in staff [@ferguson2020].
-Dog vaccination delivered through the District Veterinarian was less costly, with the majority of the costs directly related to vaccination.
-Although we included costs of implementing transect based coverage estimates, these were negligible (less than 0.10 USD per dog vaccinated), in line with recent studies which have shown this strategy is a cheap and effective way to estimate coverage @sambo2017; @gibson2015].
+Finally, in our vaccination model, we make several simplifying assumptions: we assume that the protective effect of vaccination is lost after three years (simulating vaccination with a long-lasting vaccine such as Nobivac), but we also assume that vaccinated dogs that survive to the following year are revaccinated in subsequent campaigns (effectively assuming boosting per manufacturer recommendations for Rabisin).
+Overall loss of vaccine induced immunity plays a lesser role in declining vaccination coverage given the high population turnover in this context.
+In our age-structured models, we also do not account for population carrying capacity, likely resulting in overestimates of growth of the dog population.
 
@@
 
+Our estimates of costs per animal vaccinated are in line with recent estimates from other countries [@undurraga2020], although these are likely underestimates given the donation of time and resources by the organizations and individuals involved (i.e.
+costs associated with international volunteers including airfare and visa costs as well as costs of transportation donated to the campaign).
+We also did not include costs of pre-exposure prophylaxis as all of our volunteers and vaccinators had been vaccinated prior to the campaigns.
+Both pre- and post-exposure vaccines should however always be readily available for vaccinators and should be included in vaccination program budgets.
+While the volunteer-led effort resulted in significant financial and personnel resources being devoted to the campaign, costs were higher overall and per animal vaccinated.
+In addition, NGO and volunteer based campaigns may be difficult to sustain given unpredictable funding, time commitments, and turnover in staff [@ferguson2020].
+For volunteer based efforts, focusing on local volunteers (i.e.
+veterinary students) may be a more cost-effective strategy.
+However similar to international volunteers, volunteers require subsistence during vaccination campaigns when not based in the communities where they study or live.
+Although we included costs of implementing transect based coverage estimates, these were negligible (less than 0.10 USD per dog vaccinated), in line with recent studies which have shown this strategy is a cheap and effective way to estimate coverage @sambo2017; @gibson2015].
+
+Dog vaccination delivered through the District Veterinarian was less costly, with the majority of the costs directly related to vaccination.
 In settings with high dog ownership, moving towards community-based vaccination strategies could be an effective way to achieve sufficient and consistent coverage, particularly in hard-to-reach communities.
-During our campaign, we found that puppies at approximately 3 months old were easy to handle for both vaccinators and owners.
-Puppy vaccination could be carried out by local officials embedded in communities (along the lines of community health workers), especially given recent findings on thermotolerance of rabies vaccines and locally manufactured methods for maintaining temperatures required for sustained vaccine storage (up to 3 months)
+During our campaign, we found that puppies (aged approximately 1 - 6 months) were easier to handle compared to adult dogs for both vaccinators and owners.
+Puppy vaccination could be carried out by local officials embedded in communities (along the lines of community health workers who may not have full veterinary qualifications), especially given recent findings on thermotolerance of rabies vaccines and locally manufactured methods for maintaining temperatures required for sustained vaccine storage (up to 3 months)
 [@lugelo2020].
@@
 Expanding veterinary services across the country and relieving financial pressures on veterinarians and animal health workers through appropriate compensation could greatly improve veterinary services across Madagascar [@mondiale2017evaluation].
-For volunteer based efforts, focusing on local volunteers (i.e.
-veterinary students) may be a more cost-effective strategy.
-However similar to international volunteers, volunteers require subsistence during vaccination campaigns when not based in the communities where they study or live.
 
-Overall, our results suggest that dog vaccination is a feasible and effective strategy for controlling canine rabies in Madagascar.
+Overall, our results suggest that dog vaccination is a feasible strategy for controlling canine rabies in Madagascar.
 However, rabies vaccination must be recognized as a public good.
-Strategies to remove barriers for dog owners and to incentivize veterinarians and other animal health workers to implement vaccination will be key to long-term campaign success.
+Removing barriers for dog owners and  incentivizing veterinarians and other animal health workers to implement vaccination will be key to long-term campaign success.
 Borrowing strategies from human vaccination efforts, i.e.
 deploying community health workers, could be a way to deliver vaccinations and reduce costs in the hardest-to-reach communities.
-In addition, refining vaccination strategies to local contexts and using improved tools and systems, such as mobile phone based data collection, could improve efficacy and lower costs of campaigns [@mazeri2021; @gibson2016a].
+In addition, refining vaccination strategies to local contexts and using improved tools and systems, such as mobile phone based data collection, could improve efficacy and coverage levels reached [@mazeri2021; @gibson2016a].
 To monitor the success of these campaigns, it will be critical to develop efficient and effective methods to estimate vaccination coverage and to measure their impact on reducing rabies incidence through robust surveillance [@lushasi2020; @ferguson2020].
-With limited chances for reintroduction from outside the island, Madagascar could be a model for rabies elimination, and strategies that tackle the context specific challenges could be a path forward for the country to reach ZeroBy30.
-
-# Patents
+With limited chances for reintroduction from outside the island, implementing community based mass dog vacciantion campaigns could be a path forward for Madagascar to reach ZeroBy30.

### 2021-04-03 paper/mdpi_paper.tex
updating with final proofed files

@@
 %  LaTeX support: latex@mdpi.com
-%  In case you need support, please attach all files that are necessary for compiling as well as the log file, and specify the details of your LaTeX setup (which operating system and LaTeX version / tools you are using).
+%  For support, please attach all files needed for compiling as well as the log file, and specify your operating system, LaTeX version, and LaTeX editor.
 
@@
 
-% If you would like to post an early version of this manuscript as a preprint, you may use preprint as the journal and change 'submit' to 'accept'.
+% For posting an early version of this manuscript as a preprint, you may use "preprints" as the journal and change "submit" to "accept".
 The document class line would be, e.g., \documentclass[preprints,article,accept,moreauthors,pdftex]{mdpi}.
@@
 
+%% https://tex.stackexchange.com/questions/161431/how-to-solve-longtable-is-not-in-1-column-mode-error
+\usepackage{xtab}
+
+\makeatletter
+\def\longtable{\@ifnextchar[\newlongtable@i \newlongtable@ii}
+\def\newlongtable@i[#1]{%
+\renewcommand{\endhead}{\ignorespaces}
+\xtabular[#1]}
+\def\newlongtable@ii{%
+\renewcommand{\endhead}{\ignorespaces}
+\xtabular}
+\def\endlongtable{\endxtabular}
+\makeatother
+
 % see https://stackoverflow.com/a/47122900
@@
 % Choose between the following MDPI journals:
-% acoustics, actuators, addictions, admsci, aerospace, agriculture, agriengineering, agronomy, algorithms, animals, antibiotics, antibodies, antioxidants, applsci, arts, asc, asi, atmosphere, atoms, axioms, batteries, bdcc, behavsci , beverages, bioengineering, biology, biomedicines, biomimetics, biomolecules, biosensors, brainsci , buildings, cancers, carbon , catalysts, cells, ceramics, challenges, chemengineering, chemistry, chemosensors, children, cleantechnol, climate, clockssleep, cmd, coatings, colloids, computation, computers, condensedmatter, cosmetics, cryptography, crystals, dairy, data, dentistry, designs , diagnostics, diseases, diversity, drones, econometrics, economies, education, electrochem, electronics, energies, entropy, environments, epigenomes, est, fermentation, fibers, fire, fishes, fluids, foods, forecasting, forests, fractalfract, futureinternet, futurephys, galaxies, games, gastrointestdisord, gels, genealogy, genes, geohazards, geosciences, geriatrics, hazardousmatters, healthcare, heritage, highthroughput, horticulturae, humanities, hydrology, ijerph, ijfs, ijgi, ijms, ijns, ijtpp, informatics, information, infrastructures, inorganics, insects, instruments, inventions, iot, j, jcdd, jcm, jcp, jcs, jdb, jfb, jfmk, jimaging, jintelligence, jlpea, jmmp, jmse, jnt, jof, joitmc, jpm, jrfm, jsan, land, languages, laws, life, literature, logistics, lubricants, machines, magnetochemistry, make, marinedrugs, materials, mathematics, mca, medicina, medicines, medsci, membranes, metabolites, metals, microarrays, micromachines, microorganisms, minerals, modelling, molbank, molecules, mps, mti, nanomaterials, ncrna, neuroglia, nitrogen, notspecified, nutrients, ohbm, particles, pathogens, pharmaceuticals, pharmaceutics, pharmacy, philosophies, photonics, physics, plants, plasma, polymers, polysaccharides, preprints , proceedings, processes, proteomes, psych, publications, quantumrep, quaternary, qubs, reactions, recycling, religions, remotesensing, reports, resources, risks, robotics, safety, sci, scipharm, sensors, separations, sexes, signals, sinusitis, smartcities, sna, societies, socsci, soilsystems, sports, standards, stats, surfaces, surgeries, sustainability, symmetry, systems, technologies, test, toxics, toxins, tropicalmed, universe, urbansci, vaccines, vehicles, vetsci, vibration, viruses, vision, water, wem, wevj
+% acoustics, actuators, addictions, admsci, adolescents, aerospace, agriculture, agriengineering, agronomy, ai, algorithms, allergies, analytica, animals, antibiotics, antibodies, antioxidants, appliedchem, applmech, applmicrobiol, applnano, applsci, arts, asi, atmosphere, atoms, audiolres, automation, axioms, batteries, bdcc, behavsci, beverages, biochem, bioengineering, biologics, biology, biomechanics, biomedicines, biomedinformatics, biomimetics, biomolecules, biophysica, biosensors, biotech, birds, bloods, brainsci, buildings, businesses, cancers, carbon, cardiogenetics, catalysts, cells, ceramics, challenges, chemengineering, chemistry, chemosensors, chemproc, children, civileng, cleantechnol, climate, clinpract, clockssleep, cmd, coatings, colloids, compounds, computation, computers, condensedmatter, conservation, constrmater, cosmetics, crops, cryptography, crystals, curroncol, cyber, dairy, data, dentistry, dermato, dermatopathology, designs, diabetology, diagnostics, digital, disabilities, diseases, diversity, dna, drones, dynamics, earth, ebj, ecologies, econometrics, economies, education, ejihpe, electricity, electrochem, electronicmat, electronics, encyclopedia, endocrines, energies, eng, engproc, entropy, environments, environsciproc, epidemiologia, epigenomes, fermentation, fibers, fire, fishes, fluids, foods, forecasting, forensicsci, forests, fractalfract, fuels, futureinternet, futuretransp, futurepharmacol, futurephys, galaxies, games, gases, gastroent, gastrointestdisord, gels, genealogy, genes, geographies, geohazards, geomatics, geosciences, geotechnics, geriatrics, hazardousmatters, healthcare, hearts, hemato, heritage, highthroughput, histories, horticulturae, humanities, hydrogen, hydrology, hygiene, idr, ijerph, ijfs, ijgi, ijms, ijns, ijtm, ijtpp, immuno, informatics, information, infrastructures, inorganics, insects, instruments, inventions, iot, j, jcdd, jcm, jcp, jcs, jdb, jfb, jfmk, jimaging, jintelligence, jlpea, jmmp, jmp, jmse, jne, jnt, jof, joitmc, jor, journalmedia, jox, jpm, jrfm, jsan, jtaer, jzbg, kidney, land, languages, laws, life, liquids, literature, livers, logistics, lubricants, machines, macromol, magnetism, magnetochemistry, make, marinedrugs, materials, materproc, mathematics, mca, measurements, medicina, medicines, medsci, membranes, metabolites, metals, metrology, micro, microarrays, microbiolres, micromachines, microorganisms, minerals, mining, modelling, molbank, molecules, mps, mti, nanoenergyadv, nanomanufacturing, nanomaterials, ncrna, network, neuroglia, neurolint, neurosci, nitrogen, notspecified, nri, nursrep, nutrients, obesities, oceans, ohbm, onco, oncopathology, optics, oral, organics, osteology, oxygen, parasites, parasitologia, particles, pathogens, pathophysiology, pediatrrep, pharmaceuticals, pharmaceutics, pharmacy, philosophies, photochem, photonics, physchem, physics, physiolsci, plants, plasma, pollutants, polymers, polysaccharides, proceedings, processes, prosthesis, proteomes, psych, psychiatryint, publications, quantumrep, quaternary, qubs, radiation, reactions, recycling, regeneration, religions, remotesensing, reports, reprodmed, resources, risks, robotics, safety, sci, scipharm, sensors, separations, sexes, signals, sinusitis, smartcities, sna, societies, socsci, soilsystems, solids, sports, standards, stats, stresses, surfaces, surgeries, suschem, sustainability, symmetry, systems, taxonomy, technologies, telecom, textiles, thermo, tourismhosp, toxics, toxins, transplantology, traumas, tropicalmed, universe, urbansci, uro, vaccines, vehicles, vetsci, vibration, viruses, vision, water, wevj, women, world
 
@@
 % The default type of manuscript is "article", but can be replaced by:
-% abstract, addendum, article, benchmark, book, bookreview, briefreport, casereport, changes, comment, commentary, communication, conceptpaper, conferenceproceedings, correction, conferencereport, expressionofconcern, extendedabstract, meetingreport, creative, datadescriptor, discussion, editorial, essay, erratum, hypothesis, interestingimages, letter, meetingreport, newbookreceived, obituary, opinion, projectreport, reply, retraction, review, perspective, protocol, shortnote, supfile, technicalnote, viewpoint
+% abstract, addendum, article, book, bookreview, briefreport, casereport, comment, commentary, communication, conferenceproceedings, correction, conferencereport, entry, expressionofconcern, extendedabstract, datadescriptor, editorial, essay, erratum, hypothesis, interestingimage, obituary, opinion, projectreport, reply, retraction, review, perspective, protocol, shortnote, studyprotocol, systematicreview, supfile, technicalnote, viewpoint, guidelines, registeredreport, tutorial
 % supfile = supplementary materials
@@
 %=================================================================
+% MDPI internal commands
 \firstpage{1}
@@
 \makeatother
-\pubvolume{xx}
+\pubvolume{1}
 \issuenum{1}
-\articlenumber{5}
-\pubyear{2019}
-\copyrightyear{2019}
-%\externaleditor{Academic Editor: name}
-\history{Received: date; Accepted: date; Published: date}
-\updates{yes} % If there is an update available, un-comment this line
-
-%% MDPI internal command: uncomment if new journal that already uses continuous page numbers
-%\continuouspages{yes}
-
+\articlenumber{0}
+\pubyear{2021}
+\copyrightyear{2020}
+%\externaleditor{Academic Editor: Firstname Lastname} % For journal Automation, please change Academic Editor to "Communicated by"
+\datereceived{}
+\dateaccepted{}
+\datepublished{}
+\hreflink{https://doi.org/} % If needed use \linebreak
 %------------------------------------------------------------------
@@
 
+%%%% If original paper need add "Retraction", please release the following command!!%%%%%%
+%\retractiondate{Date Month Year} % For example,  13 October 2020
+%\retractionnoticeyear{Year}
+%\retractionnoticevolume{0}
+%\retractionnoticeidnumber{0000}
+%\retractionnoticedoi{10.3390/xxx}
+
 %=================================================================
 % Add packages and commands here.
-The following packages are loaded in our class file: fontenc, calc, indentfirst, fancyhdr, graphicx, lastpage, ifthen, lineno, float, amsmath, setspace, enumitem, mathpazo, booktabs, titlesec, etoolbox, amsthm, hyphenat, natbib, hyperref, footmisc, geometry, caption, url, mdframed, tabto, soul, multirow, microtype, tikz
+The following packages are loaded in our class file: fontenc, inputenc, calc, indentfirst, fancyhdr, graphicx, epstopdf, lastpage, ifthen, lineno, float, amsmath, setspace, enumitem, mathpazo, booktabs, titlesec, etoolbox, tabto, xcolor, soul, multirow, microtype, tikz, totcount, changepage, paracol, attrib, upgreek, cleveref, amsthm, hyphenat, natbib, hyperref, footmisc, url, geometry, newfloat, caption
 
 %=================================================================
-%% Please use the following mathematics environments: Theorem, Lemma, Corollary, Proposition, Characterization, Property, Problem, Example, ExamplesandDefinitions, Hypothesis, Remark, Definition
+%% Please use the following mathematics environments: Theorem, Lemma, Corollary, Proposition, Characterization, Property, Problem, Example, ExamplesandDefinitions, Hypothesis, Remark, Definition, Notation, Assumption
 %% For proofs, please use the proof environment (the amsthm package is loaded by the MDPI class).
@@
 %=================================================================
+
 % Full title of the paper (Capitalized)
@@
 
+% MDPI internal command: Title for citation in the left column
+\TitleCitation{Lessons learned and paths forward for rabies dog
+vaccination in Madagascar: a case study of pilot vaccination campaigns
+in Moramanga District}
+
 % Authors, for the paper (add full first names)
-\Author{Caitlynn Filla$^{1, \dagger}$, Malavika Rajeev$^{2,
-\dagger}$\href{https://orcid.org/0000-0001-6030-4995}{\orcidicon}, Zoavina
+\Author{Caitlynn Filla$^{1, \dagger}$, Malavika Rajeev$^{2, \dagger,
+*}$\href{https://orcid.org/0000-0001-6030-4995}{\orcidicon}, Zoavina
 Randriana$^{3}$, Chantal Hanitriniana$^{4}$, Radoniaina R.
-Rafaliarison$^{3}$, Glenn Edosoa$^{5}$, Mamitiana
+Rafaliarison$^{3}$, Glenn Torrencelli Edosoa$^{5}$, Mamitiana
 Andriamananjara$^{6}$, Nivohanitra P. Razafindraibe$^{6}$, José
@@
 \AuthorNames{Caitlynn Filla, Malavika Rajeev, Zoavina Randriana, Chantal
-Hanitriniana, Radoniaina R. Rafaliarison, Glenn Edosoa, Mamitiana
-Andriamananjara, Nivohanitra P. Razafindraibe, José Nely, Angelique
-Ferreira, Annie Yang, Fenomanana Daniel, Tara A. Clarke, Zachary
-Farris, Terry Stone, Jochem Lastdrager, Tsiky Rajaonarivelo, Katie
-Hampson, C. Jessica E. Metcalf, Kim Valenta}
+Hanitriniana, Radoniaina R. Rafaliarison, Glenn Torrencelli
+Edosoa, Mamitiana Andriamananjara, Nivohanitra P. Razafindraibe, José
+Nely, Angelique Ferreira, Annie Yang, Fenomanana Daniel, Tara A.
+Clarke, Zachary Farris, Terry Stone, Jochem Lastdrager, Tsiky
+Rajaonarivelo, Katie Hampson, C. Jessica E. Metcalf, Kim Valenta}
 
@@
 $^{1}$ \quad University of Florida, Department of Anthropology
-Gainesville, Florida, USA; \\
+Gainesville, Florida, 32611 USA; \\
 $^{2}$ \quad Department of Ecology and Evolutionary Biology, Princeton
-University Princeton, NJ, United States; \\
+University Princeton, NJ, 08544 United States; \\
 $^{3}$ \quad The Mad Dog Initiative Akanin'ny Veterinera, Akaikiniarivo,
@@
 % The commands \thirdnote{} till \eighthnote{} are available for further notes
-
 % Simple summary
@@
 
+%\conference{} % An extended version of a conference paper
+
 % The fields PACS, MSC, and JEL may be left empty or commented out if not applicable
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+% Only for the journal Encyclopedia
+%\encyclopediadef{Instead of the abstract}
+%\entrylink{The Link to this entry published on the encyclopedia platform.}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 %\setcounter{secnumdepth}{4}
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-% Pandoc citation processing
-
-\usepackage{amsmath}
-\usepackage{booktabs}
-\usepackage{caption}
-\usepackage{longtable}
 
@@
 for controlling rabies in domestic dog populations
-\citep{worldhealthorganization2013}. However, achieving this coverage
-target in low and middle income countries where the burden of human
-rabies is highest can be challenging due to economic, ecological,
-sociocultural, and political barriers \citep{fahrion2017}.
+\citep{world2018expert}. However, achieving this coverage target in low
+and middle income countries where the burden of human rabies is highest
+can be challenging due to economic, ecological, sociocultural, and
+political barriers \citep{fahrion2017}.
 
@@
 \citep{reynes2011}. Currently, there are only 31 clinics where these
-human vaccines are available, and there is minimal dog vaccination due
-to high costs to owners and lack of vaccine availability
-\citep{rajeev2018}. Recent studies have estimated a high burden of human
-deaths, masked by weak surveillance across the country
-\citep{rajeev2020, rajeev2018}.
+human vaccines are available, with one clinic serving on average greater
+than 700,000 persons and over 20,000 bite patients treated annually
+\citep{rajeev2020}. There is minimal dog vaccination due to high costs
+to owners and lack of vaccine availability \citep{rajeev2018}. Recent
+studies have estimated a high burden of human rabies deaths
+(approximately 1000 deaths annually), masked by weak surveillance across
+the country \citep{rajeev2020, rajeev2018}.
 
-The veterinary sector is largely private, with 204 veterinarians
-employed in hybrid private/public employment as designated district
-veterinarians. While dog vaccination is rare, livestock officers and
-veterinarians work together to implement cattle vaccination campaigns
-for anthrax on an annual basis as mandated by the government (owners are
-charged a fee per animal for these vaccines which vary by location)
-\citep{mondiale2017evaluation}. No routine mass dog vaccinations have
-been conducted on the island, although a few pilot programs have begun
-in recent years, largely implemented by NGO-government partnerships.
+The veterinary sector is largely private and practices are largely
+limited to urban areas, but approximately 204 veterinarians are employed
+in hybrid private/public employment as designated district veterinarians
+by the national government. While dog vaccination is rare, livestock
+officers and veterinarians work together to implement cattle vaccination
+campaigns for anthrax on an annual basis as mandated by the government
+(owners are charged a fee per animal for these vaccines which vary by
+location) \citep{mondiale2017evaluation}. No routine mass dog
+vaccinations have been conducted on the island, although a few pilot
+programs have begun in recent years, largely implemented by
+NGO-government partnerships.
 
@@
 volunteer-led pilot vaccination campaign in two communes (sub-district
-level) using a static point strategy. In 2019/2020, we provided
-vaccines, all necessary supplies for vaccine administration, and a
-per-vaccine fee to the District Veterinarian to vaccinate animals as
-part of routine services delivered continuously. We compare time, human
-resources, costs, and coverage estimates between these campaigns, and
-using a demographic and vaccination model we further explore different
+level) using a static point strategy where owners brought their animals
+to a fixed location for vaccination. In 2019/2020, we provided vaccines,
+all necessary supplies for vaccine administration, and a per-vaccine fee
+to the District Veterinarian to vaccinate animals as part of continuous
+vaccination strategy, where vaccines were delivered alongside routine
+services provided by the veterinarian. We compare time, human resources,
+costs, and coverage estimates between these campaigns, and using a
+demographic and vaccination model we further explore different
 vaccination strategies based on what we learned during implementation.
@@
 Previous work in the district has established a high burden of rabies
-exposures and deaths despite the availability of post-exposure
-prophylaxis at the district hospital \citep{rajeev2018}. While Moramanga
-is relatively close to the capital city of Antananarivo
-(\textasciitilde{} 3 hours by bus), within the district, travel times
-between locations are highly variable, with much of the population
-living in more rural areas with limited access to roads and
-transportation \citep{rajeev2020}. Before 2018, there were limited
-animal rabies vaccination services, with most animal vaccines available
-in the urban commune of Moramanga Ville where owners were often charged
-\textgreater{} 15,000 Ariary (\textasciitilde{} 4.28 USD) per vaccine
-administered.
+exposures (42 - 110 per 100,000 persons) and deaths (1 - 3 deaths per
+100,000 persons) despite the availability of post-exposure prophylaxis
+at the district hospital \citep{rajeev2018}. While Moramanga is
+relatively close to the capital city of Antananarivo (\textasciitilde{}
+3 hours by bus), within the district, travel times between locations are
+highly variable, with much of the population living in more rural areas
+with limited access to roads and transportation \citep{rajeev2020}.
+Before 2018, there were limited animal rabies vaccination services, with
+most animal vaccines available in the urban commune of Moramanga Ville
+where owners were often charged \textgreater{} 15,000 Ariary
+(\textasciitilde{} 4.28 USD) per vaccine administered.
 
@@
 with the Department of Veterinary Services, and the Ministry of Public
-Health in the District of Moramanga. This campaign focused on two
-communes in Moramanga, Moramanga Ville (the district center) and
+Health in the District of Moramanga (fig 2). This campaign focused on
+two communes in Moramanga, Moramanga Ville (the district center) and
 Andasibe (a rural commune surrounding Andasibe National Park), where
@@
 before the campaign dates, the vaccination team informed the chief of
-the fokontany about the campaign and provided posters advertising the
-date of the vaccine and that it would be available at no cost to owners.
-During the campaign, we used Rabisin (10 mL vials with 1 mL per dose,
-Boehringer Ingelheim) to vaccinate both dogs and cats presented that
-were at least 1 month old. As part of the campaign, owners were surveyed
-by vaccinators about how many dogs and cats they owned in total (split
-by \textgreater{} 1 year vs.~\textless{} 1 year in order to avoid
-language ambiguities that might result in excluding puppies and
-kittens), as well as if their dogs were free roaming (no restrictions on
-movement by the owner/`mirenyreny', tied/`mifatora', or
+the fokontany about the campaign and provided fliers advertising the
+date of the vaccine and that it would be available at no cost to owners
+(fig 2). During the campaign, we used Rabisin (10 mL vials with 1 mL per
+dose, Boehringer Ingelheim) to vaccinate both dogs and cats presented
+that were at least 1 month old based on current WHO recommendations for
+endemic settings \citep{morters2015, world2018expert}. Rabisin has a
+manufacturor stated duration of protection of 1 year given one dose, and
+of an additional three years if an additional dose is given
+approximately one year after the first dose. As part of the campaign,
+owners were surveyed by vaccinators about how many dogs and cats they
+owned in total (split by \textgreater{} 1 year vs.~\textless{} 1 year in
+order to avoid language ambiguities that might result in excluding
+puppies and kittens), as well as if their dogs were free roaming (no
+restrictions on movement by the owner/`mirenyreny', tied/`mifatora', or
 fenced/`mifefy'). Vaccinations were delivered at no cost to owners, but
@@
 Madagascar, owners were asked how much in Ariary they would be willing
-to pay to have one animal vaccinated against rabies. For each animal
+to pay to have one animal vaccinated against rabies (after being
+informed that the current vaccination was being free). For each animal
 vaccinated, we recorded the species (cat or dog), sex, approximate age
-in years, and whether the animal had been previously vaccinated.
+in years as reported by the owner, and whether the animal had been
+previously vaccinated.
 
@@
 concurrently marked with a colored, non-toxic, livestock crayon along
-the top of the head or back. Between 4 - 6 PM when dogs are most active
-\citep{sambo2017} and on the same day as the campaign, in each
-vaccination location two transect surveys were done. Each team (two
-people per team) walked a transect for 1 hour starting in opposite
-directions and accompanied by a local guide to ensure that walking paths
-did not overlap. Teams recorded any marked and unmarked dogs observed,
-their roaming status (whether roaming, inside a fence, or tied), and
-their approximate age (greater or less than 1 year of age).
+the top of the head or back. At the end of each campaign day between
+1600-1800 hours, when dogs are most active \citep{sambo2017}, two
+transect surveys were conducted on vaccination campaign days in each
+vaccination location by two teams (consisting of two volunteers and one
+local guide) for one hour on separate paths and in opposite directions.
+Marked and unmarked dogs were recorded as well as their roaming status
+(i.e., roaming, inside a fence, or tied), and their approximate age
+(\textgreater1 year or \textless{} 1 year of age).
 
-\includegraphics[width=0.8\linewidth]{/Users/mrajeev/Documents/Projects/mora_vax/figs/fig1}
-
-\hypertarget{figure-1.-photos-from-the-2018-campaign.-topleft-advertisement-for-the-campaign-posted-on-the-door-of-the-fokontany-office-as-the-campaign-starts-topright-a-dog-post-vaccination-marked-with-a-crayon-bottomleft-a-basket-of-puppies-brought-for-vaccination-bottomright-a-line-of-owners-and-dogs-waiting-for-vaccination.}{%
-\paragraph{\texorpdfstring{\textbf{Figure 1. Photos from the 2018
-campaign}. \emph{Topleft: advertisement for the campaign posted on the
-door of the fokontany office as the campaign starts; Topright: a dog
-post-vaccination marked with a crayon; Bottomleft: a basket of puppies
-brought for vaccination; Bottomright: a line of owners and dogs waiting
-for
-vaccination.}}{Figure 1.
-Photos from the 2018 campaign.
-Topleft: advertisement for the campaign posted on the door of the fokontany office as the campaign starts; Topright: a dog post-vaccination marked with a crayon; Bottomleft: a basket of puppies brought for vaccination; Bottomright: a line of owners and dogs waiting for vaccination.}}\label{figure-1.-photos-from-the-2018-campaign.-topleft-advertisement-for-the-campaign-posted-on-the-door-of-the-fokontany-office-as-the-campaign-starts-topright-a-dog-post-vaccination-marked-with-a-crayon-bottomleft-a-basket-of-puppies-brought-for-vaccination-bottomright-a-line-of-owners-and-dogs-waiting-for-vaccination.}}
+\begin{figure}
+\includegraphics[width=0.8\linewidth]{/Users/mrajeev/Documents/Projects/mora_vax/figs/fig1} \caption{Photos from the 2018 campaign.
+Topleft: advertisement for the campaign posted on the door of the fokontany office as the campaign starts; Topright: a dog post-vaccination marked with a crayon; Bottomleft: a basket of puppies brought for vaccination; Bottomright: a line of owners and dogs waiting for vaccination.
+Photo credit: Jochem Lastdrager, Traveling Animal Doctors.}\label{fig:fig1}
+\end{figure}
 
@@
 (HDRs) and human population estimates
-\citep{cleaveland2003, athingo2020}. We used a range of 8 - 25, based on
-previous data from Madagascar \citep{ratsitorahina2009} and based on
-recent estimates from household surveys in the Moramanga District
+\citep{cleaveland2003, athingo2020}. We used a ratio range of 8 - 25
+humans-to-dogs, based on previous data from Madagascar
+\citep{ratsitorahina2009} and based on recent estimates from household
+surveys in the Moramanga District
 \citep{leblancclaireRabiesMadagascarThreepronged2019}. We set the point
@@
 four age classes: puppies under the age of 1 year, juveniles aged 1 - 2
-yrs, adults aged 2 - 6 years, and older dogs aged 6+ years based on .
-With the assumption that these estimates represent the population at a
-stable age distribution, we use a Leslie matrix model to estimate annual
-adult survival probability and fertility using maximum likelihood
-estimation \citep{Fujiwara2017}. Specifically, we assume that the number
-of individuals in each age class follows a Poisson distribution with the
+years, adults aged 2 - 6 years, and older dogs aged 6+ years based on
+broad patterns of survival in comparable dog populations (i.e.~low
+survival in the first year of life, followed by plateauing survival
+probabilites, \citep{czupryna2016, davlin2012}). With the assumption
+that these estimates represent the population at a stable age
+distribution, we use a Leslie matrix model to estimate annual adult
+survival probability and fertility using maximum likelihood estimation
+\citep{Fujiwara2017}. Specifically, we assume that the number of
+individuals in each age class follows a Poisson distribution with the
 mean predicted by the stable age distribution from the model (the
@@
 \hypertarget{data-and-ethics-statement}{%
-\paragraph{Data and ethics statement}\label{data-and-ethics-statement}}
+\subsection{Data and ethics statement}\label{data-and-ethics-statement}}
 
@@
 \#130-19/MEDD/SG/DGEF/DGRNE) for its research and volunteer programs.
+Prior to vaccination, verbal informed consent was obtained from animal
+owners, and owners could opt out of answering any questions or services
+provided. No personally identifiable information was collected at any
+point during the campaigns.
 
@@
 2019 campaign, between September 2019 - June 2020, the District
-Veterinarian vaccinated a total of 2385animals (1898 dogs and 486 cats)
+Veterinarian vaccinated a total of 2385 animals (1898 dogs and 486 cats)
 over 48 days in seven communes in the Moramanga District. While more
@@
 
-\hypertarget{table-1.-summary-of-2018-and-2019-campaigns.-breakdown-of-of-animals-vaccinated-prior-vaccination-history-dog-demography-dog-ownership-daily-and-per-vaccination-rates-by-year-and-location-for-2018.-photo-credict-jochem-lastdrager-traveling-animal-doctors.}{%
-\paragraph{\texorpdfstring{\textbf{Table 1. Summary of 2018 and 2019
-campaigns}. \emph{Breakdown of of animals vaccinated, prior vaccination
-history, dog demography, dog ownership, daily and per vaccination rates
-by year and location (for 2018). Photo credict: Jochem Lastdrager,
-Traveling Animal
-Doctors.}}{Table 1.
-Summary of 2018 and 2019 campaigns.
-Breakdown of of animals vaccinated, prior vaccination history, dog demography, dog ownership, daily and per vaccination rates by year and location (for 2018).
-Photo credict: Jochem Lastdrager, Traveling Animal Doctors.}}\label{table-1.-summary-of-2018-and-2019-campaigns.-breakdown-of-of-animals-vaccinated-prior-vaccination-history-dog-demography-dog-ownership-daily-and-per-vaccination-rates-by-year-and-location-for-2018.-photo-credict-jochem-lastdrager-traveling-animal-doctors.}}
-
-\captionsetup[table]{labelformat=empty,skip=1pt}
-\begin{longtable}{lllll}
+\begin{specialtable}[H] 
+\caption{Summary of 2018 and 2019 campaigns.
+Breakdown of of animals vaccinated, prior vaccination history, dog demography, dog ownership, daily and per vaccination rates by year and location (for 2018).\label{tab1}}
+\begin{longtable}[]{@{}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.3}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.12}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}
+  >{\raggedright\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}
+  >{\centering\arraybackslash}p{(\columnwidth - 8\tabcolsep) * \real{0.15}}@{}}
 \toprule
-& \multicolumn{3}{c}{2018} & \\ 
- \cmidrule(lr){2-4}
- & Andasibe & Moramanga Ville & All Communes & All Communes \\ 
+& & 2018 & & 2019 \\ \addlinespace
 \midrule
-Total animals 
- vaccinated & 528 & 2609 & 3137 & 2385 \\ 
-Total dogs 
- vaccinated & 254 & 1,803 & 2,057 & 1,898 \\ 
-Dogs with history of vaccination & 5\% & 16\% & 15\% & 13\% \\ 
-Dogs vaccinated within last year & 5\% & 7\% & 7\% & 13\% \\ 
-Percent male dogs & 55\% & 56\% & 56\% & 65\% \\ 
-Average dogs per owner & 0.8 & 1.1 & 1.0 & NA \\ 
-Percent of owners with 
- free-roaming dogs & 67\% & 19\% & 28\% & NA \\ 
-Animals vaccinated per day (total days) & 88 (6) & 372.7 (7) & 241.3 (13) & 49.7 (48) \\ 
-Animals vaccinated per 
- vaccination point (total points) & 75.4 (7) & 186.4 (14) & 149.4 (21) & 37.3 (64) \\ 
-Animals vaccinated per 
- person day (total person days) & 11.7 (45) & 21.6 (121) & 18.9 (166) & 49.7 (48) \\ 
+\endhead
+& Andasibe & Moramanga Ville & All Communes & All
+Communes \\ \addlinespace
+Total animals vaccinated & 528 & 2609 & 3137 & 2385 \\ \addlinespace
+Total dogs vaccinated & 254 & 1,803 & 2,057 & 1,898 \\ \addlinespace
+Dogs with history of vaccination & 5\% & 16\% & 15\% &
+13\% \\ \addlinespace
+Dogs vaccinated within last year & 5\% & 7\% & 7\% &
+13\% \\ \addlinespace
+Percent male dogs & 55\% & 56\% & 56\% & 65\% \\ \addlinespace
+Average dogs per owner & 0.8 & 1.1 & 1.0 & -- \\ \addlinespace
+Percent of owners with free-roaming dogs & 67\% & 19\% & 28\% &
+-- \\ \addlinespace
+Animals vaccinated per day (total days) & 88 (6) & 372.7 (7) & 241.3
+(13) & 49.7 (48) \\ \addlinespace
+Animals vaccinated per vaccination point (total points) & 75.4 (7) &
+186.4 (14) & 149.4 (21) & 37.3 (64) \\ \addlinespace
+Animals vaccinated per person day (total person days) & 11.7 (45) & 21.6
+(121) & 18.9 (166) & 49.7 (48) \\ \addlinespace
 \bottomrule
 \end{longtable}
+\end{specialtable}
 
@@
 costs are incurred on a per animal basis, compared to the 2018 campaigns
-where the costs are largely due to setting up the static point
-vaccination stations (fig 4A). More importantly, in all cases, even a
-nominal fee would significantly reduce numbers of dogs vaccinated and
+where the costs were largely due to setting up the static point
+vaccination stations (fig 2A \& B). More importantly, in all cases, even
+a nominal fee would significantly reduce numbers of dogs vaccinated and
 thus vaccination coverage, particularly in the rural commune of Andasibe
@@
 
-\includegraphics[width=0.8\linewidth]{/Users/mrajeev/Documents/Projects/mora_vax/figs/fig2}
-
-\hypertarget{figure-2.-comparing-campaign-costs-and-willingness-to-pay.-a-vaccine-costs-broken-down-by-category-and-by-year-colors.-b-overall-cost-per-animal-vaccinated-for-the-two-campaign-years-split-by-direct-costs-of-vaccination-per-animal-i.e.-vaccine-vaccination-card-syringes-and-baseline-implementation-costs-i.e.-personnel-supplies-subsistence-costs-for-vaccinators-during-the-campaign.-c-estimated-cost-per-animal-vaccinated-under-a-willingness-to-pay-model-for-two-campaigns-examining-increasing-costs-charged-to-the-owner-with-estimated-costs-declining-due-to-cost-recovery-through-owner-payments-and-then-peaking-once-owners-report-to-be-no-longer-willing-to-pay-for-the-vaccine.-d-the-percent-reduction-in-number-of-animals-vaccinated-given-owners-willingness-to-pay.-the-curves-in-c-and-d-are-shown-based-on-the-overall-responses-on-willingness-to-pay-solid-line-from-both-moramanga-and-andasibe-and-the-responses-split-by-commune-dashed-and-dotted-lines.}{%
-\paragraph{\texorpdfstring{\textbf{Figure 2. Comparing campaign costs
-and willingness to pay}. \emph{A) Vaccine costs broken down by category
-and by year (colors). B) Overall cost per animal vaccinated for the two
-campaign years, split by direct costs of vaccination per animal
-(i.e.~vaccine, vaccination card, syringes) and baseline implementation
-costs (i.e.~personnel, supplies, subsistence costs for vaccinators
-during the campaign). C) Estimated cost per animal vaccinated under a
-willingness to pay model for two campaigns examining increasing costs
-charged to the owner, with estimated costs declining due to cost
-recovery through owner payments and then peaking once owners report to
-be no longer willing to pay for the vaccine. D) The percent reduction in
-number of animals vaccinated given owners willingness to pay. The curves
-in C and D are shown based on the overall responses on willingness to
-pay (solid line) from both Moramanga and Andasibe, and the responses
-split by Commune (dashed and dotted
-lines).}}{Figure 2.
-Comparing campaign costs and willingness to pay.
+\begin{figure}
+\includegraphics[width=0.8\linewidth]{/Users/mrajeev/Documents/Projects/mora_vax/figs/fig2} \caption{Comparing campaign costs and willingness to pay.
 A) Vaccine costs broken down by category and by year (colors).
-B) Overall cost per animal vaccinated for the two campaign years, split by direct costs of vaccination per animal (i.e.~vaccine, vaccination card, syringes) and baseline implementation costs (i.e.~personnel, supplies, subsistence costs for vaccinators during the campaign).
+B) Overall cost per animal vaccinated for the two campaign years, split by direct costs of vaccination per animal (i.e.
+vaccine, vaccination card, syringes) and baseline implementation costs (i.e.
+personnel, supplies, subsistence costs for vaccinators during the campaign).
 C) Estimated cost per animal vaccinated under a willingness to pay model for two campaigns examining increasing costs charged to the owner, with estimated costs declining due to cost recovery through owner payments and then peaking once owners report to be no longer willing to pay for the vaccine.
 D) The percent reduction in number of animals vaccinated given owners willingness to pay.
-The curves in C and D are shown based on the overall responses on willingness to pay (solid line) from both Moramanga and Andasibe, and the responses split by Commune (dashed and dotted lines).}}\label{figure-2.-comparing-campaign-costs-and-willingness-to-pay.-a-vaccine-costs-broken-down-by-category-and-by-year-colors.-b-overall-cost-per-animal-vaccinated-for-the-two-campaign-years-split-by-direct-costs-of-vaccination-per-animal-i.e.-vaccine-vaccination-card-syringes-and-baseline-implementation-costs-i.e.-personnel-supplies-subsistence-costs-for-vaccinators-during-the-campaign.-c-estimated-cost-per-animal-vaccinated-under-a-willingness-to-pay-model-for-two-campaigns-examining-increasing-costs-charged-to-the-owner-with-estimated-costs-declining-due-to-cost-recovery-through-owner-payments-and-then-peaking-once-owners-report-to-be-no-longer-willing-to-pay-for-the-vaccine.-d-the-percent-reduction-in-number-of-animals-vaccinated-given-owners-willingness-to-pay.-the-curves-in-c-and-d-are-shown-based-on-the-overall-responses-on-willingness-to-pay-solid-line-from-both-moramanga-and-andasibe-and-the-responses-split-by-commune-dashed-and-dotted-lines.}}
+The curves in C and D are shown based on the overall responses on willingness to pay (solid line) from both Moramanga and Andasibe, and the responses split by Commune (dashed and dotted lines).}\label{fig:fig2}
+\end{figure}
 
@@
 
-\includegraphics[width=0.8\linewidth]{/Users/mrajeev/Documents/Projects/mora_vax/figs/fig3}
-
-\hypertarget{figure-3.-estimates-of-coverage-achieved-by-the-2018-and-2019-campaigns.-a-the-commune-level-numbers-of-dogs-vaccinated-size-of-points-and-the-associated-coverage-estimates-color-of-points-for-the-year-2018-estimated-using-post-vaccination-transects-and-2019-estimated-using-a-human-to-dog-ratio-hdr-of-19.5-based-on-a-recent-household-survey-in-the-moramanga-district.-the-inset-shows-the-location-of-the-moramanga-district-in-madagascar.-b-a-comparison-of-coverage-estimates-by-location-and-by-method-of-estimation-post-vaccination-transects-vs.-hdr-based-estimates-for-the-transect-based-estimates-the-line-range-shows-the-95-exact-binomial-confidence-interval-while-for-the-hdr-based-estimates-the-line-range-shows-the-range-of-coverage-estimates-assuming-an-hdr-range-of-8---25-according-to-estimates-from-the-literature.}{%
-\paragraph{\texorpdfstring{\textbf{Figure 3. Estimates of coverage
-achieved by the 2018 and 2019 campaigns}. \emph{A) The commune level
-numbers of dogs vaccinated (size of points) and the associated coverage
-estimates (color of points) for the year 2018 (estimated using
-post-vaccination transects) and 2019 (estimated using a human-to-dog
-ratio (HDR) of 19.5, based on a recent household survey in the Moramanga
-District). The inset shows the location of the Moramanga District in
-Madagascar. B) A comparison of coverage estimates by location and by
-method of estimation (post-vaccination transects vs.~HDR based
-estimates); for the transect based estimates, the line range shows the
-95\% exact binomial confidence interval while for the HDR based
-estimates, the line range shows the range of coverage estimates assuming
-an HDR range of 8 - 25 according to estimates from the
-literature.}}{Figure 3.
-Estimates of coverage achieved by the 2018 and 2019 campaigns.
-A) The commune level numbers of dogs vaccinated (size of points) and the associated coverage estimates (color of points) for the year 2018 (estimated using post-vaccination transects) and 2019 (estimated using a human-to-dog ratio (HDR) of 19.5, based on a recent household survey in the Moramanga District).
+\begin{figure}
+\includegraphics[width=0.8\linewidth]{/Users/mrajeev/Documents/Projects/mora_vax/figs/fig3} \caption{Estimates of coverage achieved by the 2018 and 2019 campaigns.
+A) The commune level numbers of dogs vaccinated (size of points) and the associated coverage estimates (color of points) for the year 2018 (squares, estimated using post-vaccination transects) and 2019 (circles, estimated using a human-to-dog ratio (HDR) of 19.5, based on a recent household survey in the Moramanga District).
 The inset shows the location of the Moramanga District in Madagascar.
-B) A comparison of coverage estimates by location and by method of estimation (post-vaccination transects vs.~HDR based estimates); for the transect based estimates, the line range shows the 95\% exact binomial confidence interval while for the HDR based estimates, the line range shows the range of coverage estimates assuming an HDR range of 8 - 25 according to estimates from the literature.}}\label{figure-3.-estimates-of-coverage-achieved-by-the-2018-and-2019-campaigns.-a-the-commune-level-numbers-of-dogs-vaccinated-size-of-points-and-the-associated-coverage-estimates-color-of-points-for-the-year-2018-estimated-using-post-vaccination-transects-and-2019-estimated-using-a-human-to-dog-ratio-hdr-of-19.5-based-on-a-recent-household-survey-in-the-moramanga-district.-the-inset-shows-the-location-of-the-moramanga-district-in-madagascar.-b-a-comparison-of-coverage-estimates-by-location-and-by-method-of-estimation-post-vaccination-transects-vs.-hdr-based-estimates-for-the-transect-based-estimates-the-line-range-shows-the-95-exact-binomial-confidence-interval-while-for-the-hdr-based-estimates-the-line-range-shows-the-range-of-coverage-estimates-assuming-an-hdr-range-of-8---25-according-to-estimates-from-the-literature.}}
+B) A comparison of coverage estimates by location and by method of estimation (shape of points correspond to post-vaccination transects vs.
+HDR based estimates); for the transect based estimates, the line range shows the 95\% exact binomial confidence interval while for the HDR based estimates, the line range shows the range of coverage estimates assuming an HDR range of 8 - 25 according to estimates from the literature.}\label{fig:fig3}
+\end{figure}
 
@@
 
-Demographic data from vaccinated dogs shows a population pyramid with a
+Demographic data from vaccinated dogs showed a population pyramid with a
 large base, indicative of a fast-growing population, and with a male
@@
 which resulted in stable age distributions consistent with the data (Fig
-4B). We filter to parameter estimates that are consistent with a growing
-population resulting in mean adult annual survival probability of 0.77
-(95\% quantile: 0.59 - 0.97). We use estimates of fertility (on average
-1.09, 95\% quantile: 0.82 - 1.41) to back-calculate pup survival, which
-ranges from 0.34 - 0.68. We find that given these demographic
-parameters, annual campaigns that target dogs of all ages result in
-rapid decline in vaccination coverage between campaigns, largely due to
-rapid turnover of the dog population (compared to the impact of waning
-immunity assuming a discrete 3 year period of vaccine immunity, evident
-in the additional dip at year 3, fig 4D). Continuously targeting 70\% of
-the puppy population for vaccination, while unable to achieve the peak
-coverage, consistently reaches coverage of about 50\% of the dog
-population. A combined strategy maintains the highest and most
-temporally stable levels of coverage close to the target of 70\%.
+4B). We filtered to parameter estimates that are consistent with a
+growing population resulting in mean adult annual survival probability
+of 0.77 (95\% quantile: 0.59 - 0.97). We used estimates of fertility (on
+average 1.09, 95\% quantile: 0.82 - 1.41) to back-calculate pup
+survival, which ranged from 0.34 - 0.68. We found that given these
+demographic parameters, annual campaigns that target dogs of all ages
+result in rapid decline in vaccination coverage between campaigns,
+largely due to rapid turnover of the dog population (compared to the
+impact of waning immunity assuming a discrete 3 year period of vaccine
+immunity, evident in the additional dip at year 3, fig 4D). Continuously
+targeting 70\% of the puppy population for vaccination, while unable to
+achieve the peak coverage, consistently reached coverage of about 50\%
+of the dog population. A combined strategy maintains the highest and
+most temporally stable levels of coverage close to the target of 70\%.
 
-\includegraphics[width=0.8\linewidth]{/Users/mrajeev/Documents/Projects/mora_vax/figs/fig4}
-
-\hypertarget{figure-4.-dog-demography-and-implications-for-campaign-strategies.-a-the-age-pyramid-for-vaccinated-dogs-by-sex.-b-bootstrapped-estimates-of-the-proportion-of-the-population-in-each-age-class-from-the-age-data-dark-blue-compared-to-estimates-from-the-demographic-models-fit-to-these-data-light-blue.-c-parameter-estimates-for-annual-fertility-rates-and-adult-survival-probability-with-estimates-highlighted-in-orange-showing-the-parameter-estimates-which-result-in-positive-population-growth.-b-simulated-vaccination-coverage-n-1000-using-the-demographic-parameters-from-c-in-a-hypothetical-commune-with-1000-dogs-for-three-different-campaign-strategies-1-annual-vaccination-campaigns-targeting-dogs-of-all-ages-purple-2-routine-vaccination-of-puppies-at-3-months-of-age-and-3-a-combined-strategy-with-campaigns-annually-and-continuous-puppy-vaccination-in-between-campaigns.}{%
-\paragraph{\texorpdfstring{\textbf{Figure 4. Dog demography and
-implications for campaign strategies}. \emph{A) The age pyramid for
-vaccinated dogs by sex. B) Bootstrapped estimates of the proportion of
-the population in each age class from the age data (dark blue) compared
-to estimates from the demographic models fit to these data (light blue).
-C) Parameter estimates for annual fertility rates and adult survival
-probability, with estimates highlighted in orange showing the parameter
-estimates which result in positive population growth. B) Simulated
-vaccination coverage (N = 1000) using the demographic parameters from
-(C) in a hypothetical commune with 1000 dogs for three different
-campaign strategies: 1) annual vaccination campaigns targeting dogs of
-all ages (purple), 2) routine vaccination of puppies at 3 months of age,
-and 3) a combined strategy with campaigns annually and continuous puppy
-vaccination in between
-campaigns.}}{Figure 4.
-Dog demography and implications for campaign strategies.
+\begin{figure}
+\includegraphics[width=0.8\linewidth]{/Users/mrajeev/Documents/Projects/mora_vax/figs/fig4} \caption{Dog demography and implications for campaign strategies.
 A) The age pyramid for vaccinated dogs by sex.
@@
 C) Parameter estimates for annual fertility rates and adult survival probability, with estimates highlighted in orange showing the parameter estimates which result in positive population growth.
-B) Simulated vaccination coverage (N = 1000) using the demographic parameters from (C) in a hypothetical commune with 1000 dogs for three different campaign strategies: 1) annual vaccination campaigns targeting dogs of all ages (purple), 2) routine vaccination of puppies at 3 months of age, and 3) a combined strategy with campaigns annually and continuous puppy vaccination in between campaigns.}}\label{figure-4.-dog-demography-and-implications-for-campaign-strategies.-a-the-age-pyramid-for-vaccinated-dogs-by-sex.-b-bootstrapped-estimates-of-the-proportion-of-the-population-in-each-age-class-from-the-age-data-dark-blue-compared-to-estimates-from-the-demographic-models-fit-to-these-data-light-blue.-c-parameter-estimates-for-annual-fertility-rates-and-adult-survival-probability-with-estimates-highlighted-in-orange-showing-the-parameter-estimates-which-result-in-positive-population-growth.-b-simulated-vaccination-coverage-n-1000-using-the-demographic-parameters-from-c-in-a-hypothetical-commune-with-1000-dogs-for-three-different-campaign-strategies-1-annual-vaccination-campaigns-targeting-dogs-of-all-ages-purple-2-routine-vaccination-of-puppies-at-3-months-of-age-and-3-a-combined-strategy-with-campaigns-annually-and-continuous-puppy-vaccination-in-between-campaigns.}}
+D) Simulated vaccination coverage (N = 1000) using the demographic parameters from (C) in a hypothetical commune with 1000 dogs for three different campaign strategies: 1) annual vaccination campaigns targeting dogs of all ages (purple), 2) routine vaccination of puppies at 3 months of age, and 3) a combined strategy with campaigns annually and continuous puppy vaccination in between campaigns.}\label{fig:fig4}
+\end{figure}
 
@@
 campaigns, in particular the observation that puppies were relatively
-easy to handle for vaccinators and owners, we find that continuous
-vaccination targeting puppies in particular may be an effective way to
-maintain vaccination coverage levels given high turnover in dog
-populations.
+easy to handle for vaccinators and owners (fig 2), we find that
+continuous vaccination targeting puppies in particular may be an
+effective way to maintain vaccination coverage levels given high
+turnover in dog populations.
 
@@
 willingness to pay, age of animals, and distance travelled to the
-vaccination point likely all suffer from recall bias and uncertainty,
-whilst willingness to pay studies are often overestimates of observed
-practice \citep{birhane2016}. In addition, the age of animals brought to
+vaccination point likely all suffer from recall bias and uncertainty.
+Owner based estimates of age are very coarse, but given the broad age
+classes we use, likely are of sufficient precision to capture broad
+patterns in age structure. However,, the age of animals brought to
 vaccination points may not be representative of the age structure of the
-underlying population, and other work has shown that in general puppies
+underlying population. Previous work has shown that in general puppies
 (individuals \textless{} 1 year) tend to have lower vaccination coverage
@@
 part of the age class of 0-1 years was not captured by the data).
-However, our analyses based on these data are consistent with previous
-findings from SSA that demonstrate male-biased populations, skewed
-towards puppies, and with high mortality in the first year of life
-\citep{czupryna2016, taylor2017}. Studies have also shown that charging
-fees for vaccines can greatly reduce the efficacy of campaigns and are a
-significant barrier to vaccination for owners
-\citep{schildecker2016, yoak2021, barbosacosta2020}. Similarly, distance
-to campaign points has been identified as a barrier to vaccination, and
-in most cases owners report traveling less than 1km to reach a
-vaccination point \citep{mazeri2018barriers, mazeri2019}. We used HDRs,
-which can be sensitive to underlying estimates of the human population
-and the spatial scale of estimation
+Despite these issues, our analyses based on these data are consistent
+with previous findings from sub-Saharan Africa that demonstrate
+male-biased populations, skewed towards puppies, and with high mortality
+in the first year of life \citep{czupryna2016, taylor2017}.
+
+Willingness to pay studies have been done previously for rabies
+vaccination, and have consistently shown that cost recovery is minimal
+given the price dog owners are willing to pay for vaccination
+\citep{schildecker2016, durr2008, sambo2014, barbosacosta2020, yoak2021}.
+In fact, owners generally overstate the amounts they are willing to pay
+when compared to observed practice \citep{birhane2016}, and thus our
+analysis could underestimate the impact of charging owners on coverage
+reductions. Similarly, distance to campaign points has been identified
+as a barrier to vaccination, and in most cases owners report traveling
+less than 1km to reach a vaccination point
+\citep{mazeri2018barriers, mazeri2019}. In the context of Madagascar,
+these findings are of particular relevance, as animal vaccinations can
+be mandated by the government, but at cost to animal owners (for
+example, for the anthrax vaccine in cattle). Importantly, dog owners in
+the district do believe that vaccination can prevent rabies transmission
+\citep{leblancclaireRabiesMadagascarThreepronged2019}, and given the
+observed demand during the campaigns, are amenable to vaccination of
+their animals. Our results confirm that implementing dog rabies
+vaccinations as a public health measure and removing as many barriers as
+possible to dog vaccination will be important to the success of control
+programs.
+
+We used HDRs, which can be sensitive to underlying estimates of the
+human population and the spatial scale of estimation
 \citep{sambo2018estimating, sambo2017}. However, we used HDR estimates
@@
 transects. Finally, in our vaccination model, we make several
-simplifying assumptions: we assume that immunity wanes after three years
-(simulating vaccination with a long-lasting vaccine such as Nobivac),
-but we also assume that vaccinated dogs that survive to the following
-year are revaccinated in subsequent campaigns (effectively assuming
-boosting per manufacturer recommendations for Rabisin). Overall waning
-vaccinal immunity plays a lesser role in declining vaccination coverage
-given the high population turnover in this context. In our
-age-structured models, we also do not account for population carrying
-capacity, which likely overestimates growth of the dog population.
-
-Our estimates of costs per animal vaccinated are in line with recent
-estimates from other countries \citep{undurraga2020}, although we likely
-underestimate costs given the donation of time and resources by the
-organizations and individuals involved (i.e.~costs associated with
-international volunteers including airfare and visa costs as well as
-costs of transportation donated to the campaign). We also did not
-include costs of pre-exposure prophylaxis as all of our volunteers and
-vaccinators had been vaccinated prior to the campaigns. Both pre- and
-post-exposure vaccines should however always be readily available for
-vaccinators and should be included in vaccination program budgets. While
-the volunteer-led effort resulted in significant financial and personnel
-resources being devoted to the campaign, costs were higher overall and
-per animal vaccinated. In addition, NGO and volunteer based campaigns
-may be difficult to sustain given unpredictable funding, time
-commitments, and turnover in staff \citep{ferguson2020}. Dog vaccination
-delivered through the District Veterinarian was less costly, with the
-majority of the costs directly related to vaccination. Although we
-included costs of implementing transect based coverage estimates, these
-were negligible (less than 0.10 USD per dog vaccinated), in line with
-recent studies which have shown this strategy is a cheap and effective
-way to estimate coverage \citet{sambo2017}; \citet{gibson2015}{]}.
+simplifying assumptions: we assume that the protective effect of
+vaccination is lost after three years (simulating vaccination with a
+long-lasting vaccine such as Nobivac), but we also assume that
+vaccinated dogs that survive to the following year are revaccinated in
+subsequent campaigns (effectively assuming boosting per manufacturer
+recommendations for Rabisin). Overall loss of vaccine induced immunity
+plays a lesser role in declining vaccination coverage given the high
+population turnover in this context. In our age-structured models, we
+also do not account for population carrying capacity, likely resulting
+in overestimates of growth of the dog population.
 
@@
 
+Our estimates of costs per animal vaccinated are in line with recent
+estimates from other countries \citep{undurraga2020}, although these are
+likely underestimates given the donation of time and resources by the
+organizations and individuals involved (i.e.~costs associated with
+international volunteers including airfare and visa costs as well as
+costs of transportation donated to the campaign). We also did not
+include costs of pre-exposure prophylaxis as all of our volunteers and
+vaccinators had been vaccinated prior to the campaigns. Both pre- and
+post-exposure vaccines should however always be readily available for
+vaccinators and should be included in vaccination program budgets. While
+the volunteer-led effort resulted in significant financial and personnel
+resources being devoted to the campaign, costs were higher overall and
+per animal vaccinated. In addition, NGO and volunteer based campaigns
+may be difficult to sustain given unpredictable funding, time
+commitments, and turnover in staff \citep{ferguson2020}. For volunteer
+based efforts, focusing on local volunteers (i.e.~veterinary students)
+may be a more cost-effective strategy. However similar to international
+volunteers, volunteers require subsistence during vaccination campaigns
+when not based in the communities where they study or live. Although we
+included costs of implementing transect based coverage estimates, these
+were negligible (less than 0.10 USD per dog vaccinated), in line with
+recent studies which have shown this strategy is a cheap and effective
+way to estimate coverage \citet{sambo2017}; \citet{gibson2015}{]}.
+
+Dog vaccination delivered through the District Veterinarian was less
+costly, with the majority of the costs directly related to vaccination.
 In settings with high dog ownership, moving towards community-based
@@
 and consistent coverage, particularly in hard-to-reach communities.
-During our campaign, we found that puppies at approximately 3 months old
-were easy to handle for both vaccinators and owners. Puppy vaccination
-could be carried out by local officials embedded in communities (along
-the lines of community health workers), especially given recent findings
-on thermotolerance of rabies vaccines and locally manufactured methods
-for maintaining temperatures required for sustained vaccine storage (up
-to 3 months) \citep{lugelo2020}. Incentivizing vaccinators appropriately
-will be a key challenge, as currently providing no-cost rabies
-vaccination is not seen as part of routine duties for district
-veterinarians or for livestock officers. Implementing dog vaccination
-alongside government mandated livestock vaccination campaigns may also
-be a strategy to scale up vaccination efforts at relatively low-cost.
-Expanding veterinary services across the country and relieving financial
-pressures on veterinarians and animal health workers through appropriate
-compensation could greatly improve veterinary services across Madagascar
-\citep{mondiale2017evaluation}. For volunteer based efforts, focusing on
-local volunteers (i.e.~veterinary students) may be a more cost-effective
-strategy. However similar to international volunteers, volunteers
-require subsistence during vaccination campaigns when not based in the
-communities where they study or live.
+During our campaign, we found that puppies (aged approximately 1 - 6
+months) were easier to handle compared to adult dogs for both
+vaccinators and owners. Puppy vaccination could be carried out by local
+officials embedded in communities (along the lines of community health
+workers who may not have full veterinary qualifications), especially
+given recent findings on thermotolerance of rabies vaccines and locally
+manufactured methods for maintaining temperatures required for sustained
+vaccine storage (up to 3 months) \citep{lugelo2020}. Incentivizing
+vaccinators appropriately will be a key challenge, as currently
+providing no-cost rabies vaccination is not seen as part of routine
+duties for district veterinarians or for livestock officers.
+Implementing dog vaccination alongside government mandated livestock
+vaccination campaigns may also be a strategy to scale up vaccination
+efforts at relatively low-cost. Expanding veterinary services across the
+country and relieving financial pressures on veterinarians and animal
+health workers through appropriate compensation could greatly improve
+veterinary services across Madagascar \citep{mondiale2017evaluation}.
 
-Overall, our results suggest that dog vaccination is a feasible and
-effective strategy for controlling canine rabies in Madagascar. However,
-rabies vaccination must be recognized as a public good. Strategies to
-remove barriers for dog owners and to incentivize veterinarians and
-other animal health workers to implement vaccination will be key to
-long-term campaign success. Borrowing strategies from human vaccination
-efforts, i.e.~deploying community health workers, could be a way to
-deliver vaccinations and reduce costs in the hardest-to-reach
-communities. In addition, refining vaccination strategies to local
-contexts and using improved tools and systems, such as mobile phone
-based data collection, could improve efficacy and lower costs of
-campaigns \citep{mazeri2021, gibson2016a}. To monitor the success of
-these campaigns, it will be critical to develop efficient and effective
-methods to estimate vaccination coverage and to measure their impact on
-reducing rabies incidence through robust surveillance
-\citep{lushasi2020, ferguson2020}. With limited chances for
-reintroduction from outside the island, Madagascar could be a model for
-rabies elimination, and strategies that tackle the context specific
-challenges could be a path forward for the country to reach ZeroBy30.
+Overall, our results suggest that dog vaccination is a feasible strategy
+for controlling canine rabies in Madagascar. However, rabies vaccination
+must be recognized as a public good. Removing barriers for dog owners
+and incentivizing veterinarians and other animal health workers to
+implement vaccination will be key to long-term campaign success.
+Borrowing strategies from human vaccination efforts, i.e.~deploying
+community health workers, could be a way to deliver vaccinations and
+reduce costs in the hardest-to-reach communities. In addition, refining
+vaccination strategies to local contexts and using improved tools and
+systems, such as mobile phone based data collection, could improve
+efficacy and coverage levels reached \citep{mazeri2021, gibson2016a}. To
+monitor the success of these campaigns, it will be critical to develop
+efficient and effective methods to estimate vaccination coverage and to
+measure their impact on reducing rabies incidence through robust
+surveillance \citep{lushasi2020, ferguson2020}. With limited chances for
+reintroduction from outside the island, implementing community based
+mass dog vacciantion campaigns could be a path forward for Madagascar to
+reach ZeroBy30.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\authorcontributions{Conceptualization, MR, GTE, MA, JN, ZF, TS, JL and
+KV; Data curation, CF, MR, ZR, CH, RR, AF, AY, FD, JL and TR; Formal
+analysis, CF, MR, AY and CJEM; Funding acquisition, MR, ZF, TS, JL, CJEM
+and KV; Investigation, MR, ZR, CH, RR, AF, TAC and TR; Methodology, MR,
+ZR, CH, GTE, MA, AF, AY, ZF, JL, TR, KH, CJEM and KV; Project
+administration, ZR, CH, RR, GTE, MA, NPR, JN, AF, FD, TAC, ZF, TS, JL,
+TR and KV; Supervision, GTE, MA, NPR, JN, ZF, KH, CJEM and KV;
+Visualization, MR; Writing---original draft, CF and KV; Writing---review
+\& editing, CF, MR, ZR, CH, GTE, TAC, ZF, KH, CJEM and KV.}
 
@@
 %% optional
+
+\end{paracol}
+\reftitle{References}
 

### 2021-04-03 R/figure3.R

@@ -122,7 +122,7 @@ fig3A <-
   theme_map() +
   theme(strip.text = element_text(face = "bold", hjust = 0.45)) +
   scale_fill_distiller(name = "Cov.\n estimate", direction = 1) +
-  scale_shape_manual(values = c(21, 23), guide = "none") +
+  scale_shape_manual(values = c(21, 22), guide = "none") +
   scale_size(name = "No. dogs\n vaccinated", 
              breaks = c(50, 150, 400, 1000)) +
   labs(tag = "A") +
@@ -157,8 +157,9 @@ fig3B <-
                       ymin = scales::squish(cov_est_lower, c(0, 1)), 
                       ymax = scales::squish(cov_est_upper, c(0, 1)), 
                       shape = type, color = factor(year)), 
-                  position = position_dodge(width = 0.2)) +
-  scale_shape_manual(values = c(16, 18), name = "Estimate type", 
+                  position = position_dodge(width = 0.25),
+                  fatten = 5) +
+  scale_shape_manual(values = c(16, 15), name = "Estimate type", 
                      labels = c("HDR", "Transect")) +
   scale_color_brewer(palette = "Dark2", name = "Year") +
   scale_size_identity() +

### 2021-04-05 paper/mdpi_paper.Rmd
post-acceptance files

@@
     affil: 3, 8
-  - name: Annie Yang
+  - name: Annie L. Yang
     affil: 2, \ddagger
@@
    Graduate Fellowship. KH is supported by the Wellcome Trust (207569/Z/17/Z).
-   We want to particularly acknowledge the contributions to this work by Annie Yang, 
+   We want to particularly acknowledge the contributions to this work by Annie Li Yang, 
    an undergraduate researcher and volunteer with Princeton University and the 
@@
 authorcontributions: |
-  Conceptualization, MR, GTE, MA, JN, ZF, TS, JL and KV; Data curation, CF, MR, ZR, CH, RR, AF, AY, FD, JL and TR; Formal      analysis, CF, MR, AY and CJEM; Funding acquisition, MR, ZF, TS, JL, CJEM and KV; Investigation, MR, ZR, CH, RR, AF, TAC and   TR; Methodology, MR, ZR, CH, GTE, MA, AF, AY, ZF, JL, TR, KH, CJEM and KV; Project administration, ZR, CH, RR, GTE, MA,      NPR, JN, AF, FD, TAC, ZF, TS, JL, TR and KV; Supervision, GTE, MA, NPR, JN, ZF, KH, CJEM and KV; Visualization, MR;          Writing---original draft, CF and KV; Writing---review & editing, CF, MR, ZR, CH, GTE, TAC, ZF, KH, CJEM and KV.
+  Conceptualization, MR, GTE, MA, JN, ZF, TS, JL and KV; Data curation, CF, MR, ZR, CH, RR, AF, ALY, FD, JL and TR; Formal      analysis, CF, MR, ALY and CJEM; Funding acquisition, MR, ZF, TS, JL, CJEM and KV; Investigation, MR, ZR, CH, RR, AF, TAC and   TR; Methodology, MR, ZR, CH, GTE, MA, AF, ALY, ZF, JL, TR, KH, CJEM and KV; Project administration, ZR, CH, RR, GTE, MA,      NPR, JN, AF, FD, TAC, ZF, TS, JL, TR and KV; Supervision, GTE, MA, NPR, JN, ZF, KH, CJEM and KV; Visualization, MR; Writing---original draft, CF and KV; Writing---review & editing, CF, MR, ZR, CH, GTE, TAC, ZF, KH, CJEM and KV.
 conflictsofinterest: |
@@
 
-In 2018, two NGOs (the Mad Dog Initiative (MDI) and Traveling Animal Doctors (TAD)) organized a pilot vaccination campaign in collaboration with the Department of Veterinary Services, and the Ministry of Public Health in the District of Moramanga (fig 2).
+In 2018, two NGOs (the Mad Dog Initiative (MDI) and Traveling Animal Doctors (TAD)) organized a pilot vaccination campaign in collaboration with the Department of Veterinary Services, and the Ministry of Public Health in the District of Moramanga (fig 1).
 This campaign focused on two communes in Moramanga, Moramanga Ville (the district center) and Andasibe (a rural commune surrounding Andasibe National Park), where previously high incidence of probable rabies exposures (Moramanga Ville) and a high burden of deaths (Andasibe) had been recorded [@rajeev2018].
@@
 The campaign was planned as a series of static point vaccination stations covering 1 - 3 fokontany (i.e., sub-communes) per day.
-A week before the campaign dates, the vaccination team informed the chief of the fokontany about the campaign and provided fliers advertising the date of the vaccine and that it would be available at no cost to owners (fig 2).
+A week before the campaign dates, the vaccination team informed the chief of the fokontany about the campaign and provided fliers advertising the date of the vaccine and that it would be available at no cost to owners (fig 1).
 During the campaign, we used Rabisin (10 mL vials with 1 mL per dose, Boehringer Ingelheim) to vaccinate both dogs and cats presented that were at least 1 month old based on current WHO recommendations for endemic settings [@morters2015; @world2018expert].
@@
 In addition, in the rural setting of Andasibe, the static point campaign strategy achieved lower coverage, reflecting less accessible, hard-to-reach human communities in this location.
-Based on the lessons learned through these campaigns, in particular the observation that puppies were relatively easy to handle for vaccinators and owners (fig 2), we find that continuous vaccination targeting puppies in particular may be an effective way to maintain vaccination coverage levels given high turnover in dog populations.
+Based on the lessons learned through these campaigns, in particular the observation that puppies were relatively easy to handle for vaccinators and owners (see puppies picutred in basket in fig 1), we find that continuous vaccination targeting puppies in particular may be an effective way to maintain vaccination coverage levels given high turnover in dog populations.
 

### 2021-04-05 paper/mdpi_paper.tex
post-acceptance files

@@
 with the Department of Veterinary Services, and the Ministry of Public
-Health in the District of Moramanga (fig 2). This campaign focused on
+Health in the District of Moramanga (fig 1). This campaign focused on
 two communes in Moramanga, Moramanga Ville (the district center) and
@@
 date of the vaccine and that it would be available at no cost to owners
-(fig 2). During the campaign, we used Rabisin (10 mL vials with 1 mL per
+(fig 1). During the campaign, we used Rabisin (10 mL vials with 1 mL per
 dose, Boehringer Ingelheim) to vaccinate both dogs and cats presented
@@
 campaigns, in particular the observation that puppies were relatively
-easy to handle for vaccinators and owners (fig 2), we find that
-continuous vaccination targeting puppies in particular may be an
-effective way to maintain vaccination coverage levels given high
-turnover in dog populations.
+easy to handle for vaccinators and owners (see puppies picutred in
+basket in fig 1), we find that continuous vaccination targeting puppies
+in particular may be an effective way to maintain vaccination coverage
+levels given high turnover in dog populations.
 
